### 2013-10-29 paper.tex
adding all files for initial commit

\documentclass[preprint,9pt]{sigplanconf}
\usepackage{url}
\usepackage{listings}
\usepackage{subfigure}
\usepackage{times}
\usepackage{multirow}
\usepackage{color}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{color}
\usepackage{rotating}

\newcommand{\addtodo}[1]{\textcolor{red}{[To do: #1]}}
\newcommand{\ctrap}{CTrap\xspace}
\newcommand{\ctraps}{CTraps\xspace}
\newcommand{\ctrapsfull}{CTraps-Full\xspace}
\newcommand{\ctrapsmm}{CTraps-NRR\xspace}
\newcommand{\lws}{CTraps-LWS\xspace}
\newcommand{\lwt}{LWT\xspace}

\newcommand{\Caption}[1]{\begin{minipage}{1.0\columnwidth} \caption{#1} \end{minipage}\vspace{-2ex}}
\newcommand{\CaptionWide}[1]{\caption{#1}\vspace{-2ex}}
%\doublespacing
\authorinfo{Anonymous for Submission}{}

\title{\Large Last Writer Slices and Communication Traps}

%\authorinfo{Brandon Lucia \and Luis Ceze}{University of Washington, Department of Computer Science and Engineering}{\{blucia0a,luisceze\}@cs.washington.edu\\http://sampa.cs.washington.edu\\{\bf In Submission --- Please Do Not Distribute}}

%\author{Anonymous for Submission}
\begin{document}
\maketitle

\begin{abstract}
%Monitoring inter-thread memory dependences in production is useful.  Doing so
%improves debugging information and provides information to dynamic analyses and
%runtime systems.  Prior work that continuously monitors cross-thread
%dependences typically has overheads too high for production.

We design efficient system support for collecting {\em last writer slices} for
shared-memory concurrent programs.  Last writer slices dynamically track memory
updates.  Last writer slices encode the provenance of values in memory, helping
with debugging.   We build on last writer slices, developing {\em communication
traps} (\ctraps).  \ctraps uses system support to efficiently monitor and
interpose on inter-thread memory dependences.  \ctraps provides an extensible
framework, exposing inter-thread dependences to {\em \ctraps applications}.
These can implement arbitrary dependence analyses.

We implement last writer slicing and \ctraps.  We show \ctraps is useful by
implementing two applications from prior work. We show that last writer slices
help with debugging using case studies of real-world bugs.  We evaluate our
designs on a set of server programs and standard benchmarks.  Our results show
\ctraps imposes overheads low enough for use in production systems (0--15\%)
running important services, like memcached, LevelDB, MySQL, and Apache.

\end{abstract}

\section{Introduction}
Multi-threaded programming is challenging.  In contrast to the simplicity of
sequential reasoning, multi-threaded programs require complex reasoning about
many threads of control and their interactions.  In shared memory programs,
threads interact by reading and writing memory locations.  Threads' reads and
writes interleave arbitrarily and nondeterministically, with each interleaving
potentially yielding different execution behavior.  The complexity of reasoning
about thread interactions is a central challenge to building multi-threaded
software.

When programmers reason incorrectly, they permit some interleavings that should
be forbidden.  When such interleavings occur, program executions may fail,
experience performance degradation, or suffer other undesirable behavior.
Preventing such undesirable behavior is a fundamental challenge faced by
software engineers.  One important way to understand program behavior is to
monitor which read and write operations {\em depend} on one another.  Memory
operations are dependent if their order determines the values they read or
write.  Dependences are especially useful for understanding how threads
interact.

%\cite{defuse,conseq,recon,bugaboo,raceslicing,fasttrack,falcon}
%\cite{oshajava,osha} 
%\cite{avio,dmtracker,cci,daikon}
%\cite{aviso,cfix} 
%\cite{mtperf,criticalitypredictors,schedpredictionmodel}  
%\cite{defuse}
%\cite{avio,cci,defuse,recon,bugaboo,falcon,dmtracker,aviso,cfix,criticalitypredictors,schedpredictionmodel}

Prior work has used dependence information to reduce the burden of making
reliable, efficient concurrent software.  Bug
detection~\cite{avio,fasttrack,raceslicing,dmtracker} and
debugging~\cite{tipslicingsurvey,bugaboo,recon,cci,defuse,conseq,falcon} tools
analyze inter-thread dependences to find buggy behavior and help programmers
understand their errors.
Profilers~\cite{threadclustering,schedpredictionmodel} analyze inter-thread
dependences to find optimization opportunities.  Dependences are also essential
to some program understanding systems~\cite{oshatr} and dynamic checkers for
thread interaction properties~\cite{oshajava,velodrome}.
Record-Replay~\cite{chimera,doubleplay,fdr,rtr} systems track and record the
order of dependent events to help reproduce failures.

%Deterministic execution
%systems~\cite{coredet,grace} monitor dependences to determine how and when to
%serialize portions of threads' executions to enforce determinism.  Software
%trasactional memory (STM) systems~\cite{stm} monitor dependences to identify
%conflicts between concurrently executing atomic code regions.

We note that monitoring dependences is not only a fundamental part of the value
of such systems, but also a fundamental part of the {\em cost}.  Tracking
dependences requires executing additional monitoring code while the program
runs and many of the above systems have reported overheads too high for use in
production (10-100x
slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.
Unfortunately, many of these systems are most effective when they run in
production.

We argue in this work that dependence-tracking analysis should -- and can --
have low enough overhead for use in deployed systems.  The key to reducing the
overhead of dependence tracking is to carefully select and monitor only a
{\em subset} of dependences.  We start from two simple observations: (1) Reads are
frequent so we must minimize tracking overheads on reads, and (2) When
addressing concurrency, {\em inter-thread} dependences are most important.
Based on these observations, we focus on dependences that reflect {\em
communication} between threads.  Communication occurs when a thread reads or
overwrites a memory location that was most recently written by a different
thread; {\em i.e.}, inter-thread read-after-write (RAW) and write-after-write
(WAW).  Tracking only communication dependences preserves enough information to
implement useful applications, but has low enough overhead for always-on use in
production.  Framed by this idea, there are two parts to our work.

First, we describe {\em last writer slices}
(Section~\ref{sec:lastwriterslices}).  Last writer slices record the thread and
program point that last wrote each memory location.  Last writer slices are
useful because they expose the {\em provenance} of a memory location's contents
during debugging (Section~\ref{sec:debugging}).  When a failure occurs, last
writer slices are saved with a core dump.  Last writer slices are also
available during an execution in the debugger.  Tracking last writer slices has
low overhead because it only imposes overhead on write operations:  reads are
overhead-free.  In Section~\ref{sec:eval} we experimentally show
that collecting last writer slices has low overhead.

Second, on top of last writer slices we build {\em Communication Traps} or {\em
\ctraps}  (Section~\ref{sec:ctraps}).  \ctraps monitors inter-thread
communication with overhead low enough for production use in many important
cases.  When threads communicate, \ctraps delivers a {\em trap} that can be
handled by a trap handler.  Trap handlers implement {\em \ctraps applications}
through an extensible interface that can analyze or interpose on the execution
(Section~\ref{sec:apps}).  By default, \ctraps has very low overhead.  \ctraps
applications can perform arbitrary dependence analyses, but at their own
discretion and cost.  As we show in our evaluation, the overhead of \ctraps
applications scales with the complexity of their analysis.

\paragraph{Contributions of This Work}
To summarize, this work makes several contributions:
\begin{itemize}

\item{We discuss last-writer slices, an abstraction that dynamically tracks
memory updates, capturing inter-thread dependence information.  We
design and implement system support for collecting last writer slices.  We
present two cases studies that show that last writer slices are useful for
diagnosing real-world bugs.}

\item{Building on last-writer slices, we describe and implement \ctraps, which
exposes inter-thread communication through an extensible trap interface.
We also look at an optimization to \ctraps that explores trading off
precision for performance.}

\item{We demonstrate that \ctraps provides a simple platform for building
dependence analyses by implementing two analyses from prior
work~\cite{cci,defuse,recon}. }

\item{We implement last writer slicing and \ctraps and evaluate their performance using a
variety of important applications ({\em e.g.}, MySQL, memcached and the PARSEC
benchmarks~\cite{parsec}).  We show that in many cases our designs have overheads
low enough for use in deployment (0--15\%).}

\end{itemize}

%In this work, we begin by showing that the cost of monitoring communication has
%been overstated in prior work.  We do so by designing support to monitor
%communication that, for a variety of important applications ({\em
%e.g.},databases, key-value stores, web servers) the overheads are extremely low
%-- so low that communication tracking is viable even for today's deployed
%systems.  We illustrate our claim with system support for  {\em Communication
%Traps} or \ctraps.  Our support for \ctraps tracks communication between
%threads and exposes communication events to trap handlers that can implement
%applications, such as the above.  We show the feasibility of our approach by
%implementing a selection of the above applications, as well as several novel
%applications.  We study the performance characteristics of \ctraps and provide
%a comparison to prior work, showing that our system has lower performance
%overheads -- in some cases, by one to two orders of magnitude lower.  We then
%show that even in cases that, for \ctraps, incur higher performance overheads,
%judicious use of sampling reduces overheads to an acceptable level, and retains
%many of the benefits of non-sampling \ctraps.

\section{Background}

Many systems that support programming and execution models, debugging tools,
and other software engineering tools track dynamic memory dependences during a
program's execution.  In this section we provide context for \ctraps by
discussing some of these systems, focusing on those that address concurrency.
We then briefly discuss debugging with data dependence information, in order to
motivate the role of last writer slices at debugging time.

\subsection{Tracking Dependences and Communication}
\label{sec:background:comm}
%\ctraps tracks dependences that correspond to {\em inter-thread communication}.

A good first example of a system that monitors inter-thread dependences is
software transactional memory (STM)~\cite{stm}.  In STMs, dependences
correspond to interference between accesses in different threads that must be
detected to rollback conflicting transactions.  Deterministic concurrent
runtime systems~\cite{coredet,grace} are similar -- they monitor
inter-thread dependences, to determine when to apply deterministic ordering
constraints on threads' accesses.  STMs and deterministic systems must track
all dependences and they must track them precisely, to provide strong
guarantees as programming and execution models.  Unfortunately, precision can
have a high runtime overhead -- {\em e.g.}, CoreDet incurs a mean overhead of
around 5x~\cite{coredet}.  This overhead is too high for most production
systems.

%The overhead comes from the need to precisely track read-after-write (RAW),
%write-after-read (WAR), and write-after-write (WAW) dependences, taking action
%({\em e.g.}, buffering) on most cross-thread dependences.  These systems track
%dependences with additional code that runs around memory accesses.  To ensure
%correctness, accesses and tracking code usually must be protected by
%synchronization.  The high cost of synchronization exacerbates the overheads.

In contrast to {\em imprecision intolerant} tasks like STM and deterministic
execution, other tasks are more {\em imprecision tolerant}.  Imprecision
tolerant systems help perform tasks like
debugging~\cite{defuse,conseq,recon,bugaboo,raceslicing,fasttrack,falcon},
software engineering~\cite{oshajava,oshatr}, anomaly
detection~\cite{avio,dmtracker,cci,daikon}, bug and failure
avoidance~\cite{aviso,cfix} and concurrent performance
profiling~\cite{threadcriticality,schedpredictionmodel}.  These
systems also work by monitoring how data flows between threads.  They can
tolerate two important sources of imprecision.

First, many of these techniques track only an {\em ad hoc} subset of
dependences.  For example, DefUse~\cite{defuse} tracks inter-thread WAR
dependences, but not WAW dependences.  Recon tracks some inter-thread WAW and
RAW dependences and approximates WAR dependences.  These techniques are useful
despite an incomplete set of dependences.  Second, many of these systems rely
on statistical models of software
behavior~\cite{avio,cci,defuse,recon,bugaboo,falcon,dmtracker,aviso,threadcriticality,schedpredictionmodel}.
Models often tolerate noise, so many of these techniques can relax the
correctness constraints of their implementation and remain useful.  For
example, some systems have chosen not to enforce atomicity of each memory
access and its dependence tracking code, compromising precision for a boost in
performance.

These tools and techniques work in a variety of environments.  A common theme
is the advantage reaped from use in a deployment environment.  Debugging tools,
profilers, and anomaly detectors benefit from seeing diverse, real-world
behavior.  Dynamic failure avoidance techniques must work in deployed systems
to be effective.  Deployment environments demand low overheads.

In this work, we select a subset of inter-thread dependences in a systematic
way, rather than choosing an {\em ad hoc} subset.  \ctraps tracks dependences
that correspond to inter-thread communication.  \ctraps provides enough
precision to build useful analyses and runtime systems with overhead low enough
for use in production.

\subsection{Debugging with Data Dependence Information}

Data dependence information is directly useful during program debugging.  A
wide variety of dynamic program slicing
techniques~\cite{thinslicing,tipslicingsurvey} have been proposed.  These
techniques record a subset (slice) of dynamic program instructions selected
using data and control dependence information.  Dynamic dependence slices
expose the provenance of control- and data-flows in the execution.  Work on
debugging~\cite{whylinechi,whylineicse} in the HCI and Software Engineering
community shows that when programmers debug programs, they benefit from the
ability to directly ask such provenance (``why'') questions of their debugger.

Dynamic dependence slices are especially useful for debugging if they are
available after a failure has occurred in a deployed system. However, slicing
in deployed systems is uncommon, because deployed systems demand low overheads
and slicing can have high overhead.  Our system provides provenance information
that is useful at debugging time by collecting {\em last writer slices} that
help understand inter-thread communication.  We track last writer slices
with overhead low enough for use in production code.  Deployable efficiency
makes it possible to collect and report last writer slices the same way today's
systems collect and report core dumps.

\section{Last Writer Slices}
\label{sec:lastwriterslices}

A last writer slice is a record associated with a memory location that stores
the thread and program point that last wrote data to that memory location.  A
last writer slice is a dynamic property of a memory location at a point in a
program's execution.   We formalize last writer slices in
Section~\ref{sec:lwssoundness}.

We first discuss last writer slices in more depth by describing the design of
our support for collecting them.  The system support consists of a runtime
library and a compiler analysis.  In Section~\ref{sec:debugging} we highlight
the role of last writer slices in debugging and show they help diagnose real
bugs, using case studies.

\subsection{Runtime Support: Maintaining the Last Writer Table}
We gather last writer slices at runtime by maintaining a {\em last writer
table} (\lwt).  The \lwt is a {\em shadow memory} that holds one entry for each
memory address accessed by the program.  Each \lwt entry records the thread and
program point that last wrote to the entry's corresponding memory location.
When a thread performs a write to a memory location, it updates the \lwt by
recording its thread identifier and the program point of its write operation to
the memory location's \lwt entry.  A program point may be an
instruction address, or a call stack.  We use instructions by default, but our
system supports 2-address bounded call-stacks as well.

Figure~\ref{fig:basicLWT} shows the basic operation of our runtime support.
The right side of the figure shows a program execution with three threads.  On
the left is the \lwt at the point of each access.  There are two key things to
notice.  First, write operations update the \lwt entry of the memory location
they access.  Second, there is no action necessary for read operations.  Making
read operations cheap is key to making the overhead of last writer slicing low
enough for deployment use.

\begin{figure}[h]
\centering
\includegraphics[scale=.6]{figs/BasicLWT.pdf}
\caption{\label{fig:basicLWT}Basic operation of the last writer table. }
\end{figure}

\subsection{Compiler Support: Analysis and Instrumentation}
Collecting last writer slices requires instrumentation on each write operation.
We built an instrumenting compiler pass that inserts a call into our runtime
system just before each write to a memory location that may be accessed by code
executing in another thread (escaping memory accesses).  Write instrumentation
calls are passed three pieces of information: the address of the memory word
involved in the access, the thread identifier of the accessing thread, and the
accessing program point.

\subsubsection{Escape Analysis}
We use a local function escape analysis to identify which accesses are to
memory locations that can be statically guaranteed to be inaccessible to code
running in other threads.  When a thread accesses a location that is
inaccessible to code in other threads, there is no way that access can be a
communicating access.  Such accesses are not instrumented by our compiler.

We rely on {\tt GCC}'s built-in escape analysis to determine if a memory
operation might lead to communication.  This analysis is a function escape
analysis.  Locations determined not to escape function scope also cannot escape
the thread executing the function.  However, using an inter-procedural thread
escape analysis may eliminate more instrumentation sites.  Investigating better
instrumentation strategies is orthogonal, but an interesting subject for future
work.

\subsection{Formalism and Correctness}
\label{sec:lwssoundness}
To discuss the correctness of our \lwt-based last writer slice support, we
first develop a formal description of last writer slices.  We consider an
N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_i = \{p^{i}_{1},
p^{i}_{2}, \ldots, p^{i}_{k_{i}}\}$ is the sequence of $k_{i}$ dynamic
instructions executed by thread $i$.  We assume multi-threaded executions are
sequentially consistent (SC).  Without loss of generality, we assume all
instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
$Read$ or $Write$. We also assume $m$ is a uniquely addressed memory location.
Under SC, a point in an execution of $P$ after $e$ instructions have
executed is defined by $Q = (q_{0}, \ldots, q_{e})$ where each $q \in Q$ is a
$p^{i}_{j}$ from one of the $k_i$ instructions of thread $i$. We introduce
$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
in $Q$ to write $m$ and $t$ is the thread that executed it.  For a
never-written location $n$, $LWS[n] = (\emptyset,\emptyset)$ indicating it is
unwritten.

For our design to be correct, $LWT[m]$, the entry for a memory location $m$ in
our \lwt implementation, must be the same as $LWS[m]$ at all points in an
execution.  Correctness in the \lwt requires {\em soundness} and {\em
completeness}.  A {\em sound} design only updates $LWT[m]$ when $m$ is
written.  A {\em complete} design always updates $LWT[m]$ when $m$ is written.
Our compiler inserts \lwt updates before every write
operation (for completeness) and only before write operations (for soundness).
Formally, given a program, $P$, our compiler pass transforms it to a new
program, $P'$.  Each thread in $P'$ executes a modified sequence of
instructions: $P'_{i} \in P' = \{ I(p^{i}_{1}), p^{i}_{1}, \ldots,
I(p^{i}_{k_{i}}), p^{i}_{k_{i}} \}$.  $I(p^{t}_{j})$ sets $LWT[m] =
(t,p^{t}_{j})$ if and only if $p^{t}_{j} = Write(m)$ and is a no-op otherwise.

Thus, our \lwt is correct under the two main assumptions of our formalism: (1)
there are no data races in the execution; and (2) each memory location
referenced by a write refers to a unique region of memory that does not overlap
with another referenced region.

\paragraph{Sequential Consistency and Data-races}

We assume data-race freedom and a memory model that provides SC under data-race
freedom (implying an SC execution).  We must make these assumptions because,
for performance reasons, we do not enforce the atomicity of \lwt updates and
their corresponding program accesses (a design choice made in prior
work~\cite{fasttrack,recon}).  This atomicity is also required for correctness.

We use the relation $p \prec q$ to denote that $p$ happens before $q$.  Under
data-race freedom, any write by thread $t$ to $m$, $p^{t}_{j} = Write(m)$,  is
ordered by program synchronization with $p^{t'}_{j'} = Op(m)$, a later access to
$m$ by any other thread $t'$ -- {\em i.e.}, $p^{t}_{j} \prec p^{t'}_{j'}$.  Our
compiler inserts an \lwt update $I(p^{t}_{j})$ into $P'$ before its write
operation, $p^{t}_{j}$, so $I(p^{t}_{j}) \prec p^{t}_{j}$.  Happens-before is
transitive, so $I(p^{t}_{j}) \prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec
p^{t'}_{j'}$, meaning $I(p^{t}_{j})$ happens before all accesses that
$p^{t}_{j}$ happens before {\em and before their respective \lwt updates},
$I(p^{t'}_{j'})$.  Symmetrically, for any access $p^{t''}_{j''}$ in another
thread $t''$ that precedes $p^{t}_{j}$, $p^{t''}_{j''} \prec p^{t}_{j}$.  Our
compiler inserts each $I(p^{t}_{j})$ immediately before each $p^{t}_{j}$, so
$I(p^{t''}_{j''}) \prec p^{t''}_{j''} \prec I(p^{t}_{j}) \prec p^{t}_{j}$,
happens-before ordering $I(p^{t}_{j})$ after all preceding accesses to $m$ in
other threads, {\em and after their repective \lwt updates}.  Thus, at all
points $Q$ in a data-race free execution, given $q_r = I(p^{t}_{j})$, $q_p =
p^{t}_{j}$, and $q_r \prec q_p$, $\forall_{ \{q_s | q_r \prec q_s \prec q_p
\}}: q_s \ne Op(m) \wedge q_s \ne I(Op(m)) $.  In other words, \lwt updates and
their accesses are atomic.

With data-races, \lwt updates and their accesses may not be atomic ({\em i.e.},
such a $q_s$ can exist).  Instrumentation calls for racy accesses are not
ordered by happens-before, so they may interleave arbitrarily.  With races, an
execution could reach a state $Q_{bad} = (\ldots, I(p^{t}_{j}), I(p^{t'}_{j'}),
p^{t'}_{j'}, p^{t}_{j}, \ldots)$.  In such a state, $LWS[m] = (t,p^{t}_{j})$,
but $LWT[m] = (t',p^{t'}_{j'})$.  The \lwt was effectively not updated when
$p^{t}_{j}$ executed (violating completeness), or was effectively updated to
reflect that $p^{t'}_{j'}$ was $m$'s last writer, when $p^{t}_{j}$ was
(violating soundness).  This edge case is incorrect. However, we expect it is
also uncommon in practice, so last writer slices are usually correct, even in
racy programs.

\paragraph{Non-overlapping Memory References} 
We assume each written memory location corresponds to a non-overlapping memory
region because our \lwt design is {\em granularity oblivious}.  We discuss this
issue only informally due to space constraints.  Each \lwt entry corresponds to
a memory location referenced by a program instruction.  Two instructions might
reference different locations that actually overlap because the referenced
locations have different granularities.  When that happens, both accesses
update different \lwt entries, because they reference different locations.
However, the overlapping region then has two different, inconsistent entries
reflecting its last {\em two} updates, rather than just its last write.

For example, one instruction might write a word and another might later write
the second byte of that word.  The word access updates the word's \lwt entry.
The byte access updates a second \lwt entry for the byte address.  However, the
byte is part of the previously written word, meaning the \lwt entry for the
word no longer reflects reality -- the word was last written by the instruction
and thread that wrote the byte, not the one that wrote the word.  This
situation leads to incompleteness because the entry for the word should be
updated but it is not.

Assuming non-overlapping regions ensures \lwt are correct.  Another
approach would be to assume a fixed \lwt granularity.  We chose our solution
for two reasons.  First, fixing the \lwt at a large granularity ({\em i.e.},
cache-line) precludes sound slicing for fine-grained accesses.  Second, fixing
the \lwt at a small granularity ({\em i.e.}, byte) requires multiple \lwt
updates for larger granularity accesses, increasing overheads.

\subsection{Debugging with Last Writer Slices}
\label{sec:debugging}
Prior work has demonstrated that it is useful during debugging to allow
programmers to ask ``why?'' questions~\cite{whylineicse,whylinechi} about the
provenance of program state.  To answer such questions, prior work has mostly
used dependence slices that include intra-thread
dependences~\cite{whylineicse,conseq,tipslicingsurvey}.  The overhead of
slicing dependence chains and tracking temporal sequences is often high,
relegating such techniques to working on traces~\cite{whylinechi} or debugging
executions.

The \lwt dynamically records information about write operations -- what thread
and code point last wrote each memory location.   These last writer slices do
not explicitly encode anything about read operations.  However, by looking at
the \lwt at a point in the execution, the provenance of the data being read or
over-written can be determined.  For example, if an assertion failure occurs
due to an erroneous inter-thread data flow, the provenance of the data in the
assertion's condition helps explain why the failure occurred.  Last writer
slices therefore answer an important class of ``why?'' questions related to
concurrency errors.

Computing last writer slices is fast enough to deploy and use
continuously (see Section~\ref{sec:eval}) and the \lwt can be saved with a core
dump.  Thus last writer slices can (and should) be made part of the standard
debugging information provided when a program fails.

\subsubsection{Case Studies}

We illustrate the debugging benefits of last writer slices using a set of case
studies.  In each case, we debug a buggy program using last writer slices.  We
show that for some concurrency errors, last writer slices are a good match
because they provide answers to important {\em ``why?''}
questions~\cite{whylineicse}.

\paragraph{Case Study: PBZip2}
We conducted a case study using a buggy version of PBZip2-0.9.1.    The bug is
a use-after-delete bug that leads to an assertion failure.  A worker thread
tries to acquire a mutex after it has been deleted by the main thread during
shutdown, causing the assertion to fail.

Last writer slices provide the key piece of information to debug this failure: the last
write to the mutex before the worker tried acquiring it was when the main
thread deleted it.  A developer is likely to begin debugging by checking the
validity of the mutex.  Upon finding it is invalid ({\em
i.e.}, it has been deleted), the developer is likely to probe to determine who 
deleted it.  Without the information in the \lwt, answering that question is difficult
because the main thread may have executed arbitrarily far from where it deleted
the mutex.  With the \lwt, answering that question only requires the developer
to check the mutex's entry.

\paragraph{Case Study: MySQL}
Last writer slices were also useful for debugging a logging system failure in
MySQL-4.0.12.  The bug is an atomicity violation involving two threads.  One
thread rotates the database's log file.  Log rotation involves closing the log
file, marking the log closed, opening a new log file, and marking the log open.
Log rotation should be atomic, but lack of synchronization allows another
thread to run queries during rotation.  The query code checks if the log is
marked closed; if it is not closed, the queries are logged, but if it is
closed, the queries are not logged and the program simply continues.  We added
an assertion to the non-logging case because this behavior is a security
vulnerability.  The bug report describes the bug as ``critical'', so it is
reasonable to assume a programmer that expected all queries to be logged would
write such an assertion after the bug was reported.

We ran the program and concurrently issued a log rotation command, and an
insert query.  The bug manifested ({\em i.e.}, the queries went unlogged) and
our assertion failed, dumping core.  Examining the log's last writer slice in
GDB allowed examining the last thread to mark the log closed.  We discerned it
was the log rotation thread that last marked the log closed during log
rotation.  Knowing how the log ended up marked closed enables reasoning about
what code to change ({\em e.g.}, added synchronization) to prevent the
situation in the future.

While a developer is likely to understand the symptom of the bug (queries go
unlogged) without the information in the \lwt, understanding {\em why} it is marked closed would be
difficult.  This example illustrates how the \lwt fills in the crucial inter-thread
dependence that helps understand why the program failed.

\section{Communication Traps}
\label{sec:ctraps}

\ctraps is a platform for analyzing and interposing on inter-thread
communication events during a concurrent program's execution.  We build our
support for \ctraps on top of the support for last writer slicing that we
described in Section~\ref{sec:lastwriterslices}.  Using last writer slice
information, \ctraps dynamically tracks dependences representing inter-thread
communication events. \ctraps exposes communication events as {\em traps}
during an execution that {\em trap handlers} can handle to perform analysis
and interposition.  We formalize trap delivery and handling in
Section~\ref{sec:ctsoundness}.

\subsection{\ctraps Design}

We now discuss \ctraps in more detail by describing the design of our \ctraps
system support.  Figure~\ref{fig:systemdiagram} shows an overview of our system
support.  The programmer writes their program as usual.  They compile it using
the \ctraps compiler. The \ctraps compiler includes the compiler support for
collecting last writer slices.  In addition, the \ctraps compiler also inserts
calls to the \ctraps runtime at points where communication may occur.  The
resulting compiled executable links to the \ctraps runtime that collects last
writer slices and monitors communication.  The runtime loads and manages \ctraps
handlers when the execution starts via a plugin interface.  \ctraps handlers,
written independently of the original program implement \ctraps applications.
Traps are delivered to handlers by the runtime when operations communicate.
Together, the \ctraps executable and the \ctraps handlers are deployed.

\begin{figure}[htb]
\centering
\includegraphics[width=.90\columnwidth]{figs/SystemDiagram.pdf}
\caption{\label{fig:systemdiagram}CTraps system support.}
\end{figure}

%Our system support for \ctraps incorporates the support for collecting last
%writer slices described in Section~\ref{sec:lastwriterslices}.  In addition,
%our \ctraps system support also monitors all read and write operations that
%might communicate and traps on all those that communicate.  
In addition to our base \ctraps design, we also investigate \ctrapsmm, a
variant of \ctraps that uses a flow-sensitive optimization to eliminate some
sources of overhead at a cost in analysis precision. \ctrapsmm tracks all write
operations that might communicate, but leaves some redundant read operations
untracked (\ctraps, with {\bf N}o {\bf R}edundant {\bf R}eads).

\subsubsection{\ctraps System Support}

Our base \ctraps design delivers traps on all communicating read and write
operations using compiler and runtime support.

\paragraph{Compiler Support}
Our compiler instruments all read and write operations that access potentially
escaping memory locations with a call to a {\em trap hook} function.  Trap hooks
are passed the address of the memory word involved in the access, the thread
identifier of the accessing thread, the type of operation being performed, and
the program point of the access.

\paragraph{Runtime Support}
The \ctraps runtime implements the trap hook function.  Trap hooks have two
purposes: first, to detect communication; second, to deliver traps to \ctraps
applications.  Trap hooks detect communication by looking up the \lwt entry for
the memory location being accessed.  If the thread identifier in the \lwt entry is
different from the accessing thread, then the access is a {\em communicating
memory access}.  At each communicating memory access, the runtime delivers a
trap to the accessing thread just before the memory access is allowed to
proceed.  Traps are delivered by trap hooks to {\em trap handler routines} that
implement \ctraps applications.  Trap handler routines are like signal
handlers.  They are registered with the \ctraps runtime when the program
starts.  Trap handlers may contain arbitrary application-specific code.
Figure~\ref{fig:hookapi} shows the \ctraps trap handling API.  Trap
handlers are passed the four pieces of information passed to the trap hook, as
well as the \lwt's record of the program point and thread identifier of the
last write to the involved memory location.

\begin{figure}[htb]
\centering
\begin{verbatim}
void trap(AccessType type, 
          ThdId currThd, ProgPt currPt,
          ThdId lastThd, ProgPt lastPt);
\end{verbatim}
\caption{\label{fig:hookapi}The \ctraps trap handling interface.}
\end{figure}

\subsubsection{\ctrapsmm: Eliminating Redundant Instrumentation}
\label{sec:dpo}

The runtime support for \ctrapsmm is mostly the same as the runtime support for
our base \ctraps design, but \ctrapsmm adds trap hooks on only a subset of
reads.  \ctrapsmm's compiler support uses the program's control-flow structure
to identify reads that are likely to be {\em redundant}, and eliminates trap
hooks on those accesses.

We use dominator-based analysis to eliminate instrumentation that is likely to
be redundant.  Our analysis is function-local and works on code in SSA form.
The compiler looks at each pair of reads that are instrumented with trap hooks.
If a pair of instrumented reads both refer to a memory address stored in the
same SSA variable and the instrumentation points are {\em control equivalent},
then they are redundant.  Two reads, $A$ and $B$ are control equivalent if $A$
dominates $B$ and $B$ post-dominates $A$ -- every path to $B$ passes through
$A$, and every path passing through $A$ reaches $B$.  The latter read ("$B$")
is not instrumented.  Using this technique to remove instrumentation sites
preserves instrumentation along conditional control flow paths.  For example,
if the second of a pair of reads is conditionally executed, and the first is
not, then the pair are not control equivalent and both remain.

Figure~\ref{fig:insthooks} illustrates how a pair of control equivalent reads
are instrumented by the compiler for \ctrapsmm.  Pictured on the left is the
original program.  The shapes represent memory accesses with writes marked 'W'
and reads marked 'R'.  Notice that the the pentagon and the diamond are control
equivalent, despite the control flow structure separating them.  Pictured on
the right is the program after instrumentation.  There are two important things
to notice.  First, only the first of the control equivalent accesses are
instrumented, eliminating instrumentation overhead.  Second, the conditionally
executed accesses on either side of the branch are both instrumented.
Preserving these accesses is important for enabling flow-sensitive applications
using \ctrapsmm.

\begin{figure}[htb]
\centering
\includegraphics[width=.9\columnwidth]{figs/CTMMHooks.pdf}
\caption{\label{fig:insthooks}Placement of trap hooks for \ctrapsmm.}
\end{figure}

\subsection{Formalism and Correctness}
\label{sec:ctsoundness}
To discuss the correctness of \ctraps, we first formalize trap delivery by
building on our last writer slicing formalism from
Section~\ref{sec:lastwriterslices}.

Given a program $P'$ already instrumented with \lwt updates, the \ctraps
compiler produces $P''$. A thread $i$ in $P''$ executes a modified instruction
sequence, $P''_{i} = \{ T(p^{i}_{j}), I(p^{i}_{1}), p^{i}_{1},$ $\ldots,
T(p^{i}_{j}), I(p^{i}_{k_{i}}), p^{i}_{k_{i}} \}$, where $T(p^{i}_{j})$ is a
trap hook call.  We consider a point $Q = \{q_1, \ldots, q_{e-2}\}$ after $e-2$
instructions of a sequentially consistent execution have executed and $q_{e} =
p^{i}_{j}$ is about to execute after its trap hook $T(p^{i}_{j})$ and \lwt
update $I(p^{i}_{j})$ execute.  The trap hook $T(p^{i}_{j})$ {\em delivers} a
trap before if $LWT[m] = (t,p^{t}_{\ell w})$ and $i \ne t$.  Informally, the
trap hook delivers a trap before the instruction if the thread that last wrote
$m$ is different from the one executing the instruction.  When a trap hook,
$T(p^{t}_{j})$, delivers a trap, thread $t$ executes an arbitrary trap handler
$H = \{h^{t}_{1}, \ldots, h^{t}_{h_{n}}\}$.  Thus, after the trap, the \lwt
update (if $p^{i}_{j} = Write(m)$), and the instruction execute, the execution
reaches the new point $Q = (q_{1}, \ldots, T(q_{e}), h^{t}_{1}, \ldots,
h^{t}_{h_{n}}, I(q_{e}), q_{e})$, and continues.

We define what makes a \ctraps design sound and complete, and then discuss the
correctness of our design.  A design is {\em sound} if it only delivers a trap
when threads have communicated.  A design is {\em complete} if it always
delivers a trap when threads have communicated.  The \ctraps compiler puts a
trap hook on all read and write operations, and only on those operations.
Assuming a correct \lwt, \ctraps is correct.  The correctness of \ctraps also
relies on data-race freedom and non-overlapping memory references.  The same
arguments for \lwt correctness apply to the correctness of trap delivery and
handling.  Most importantly, data-race freedom implies that trap hooks, \lwt
updates, and accesses are all atomic.

%\paragraph{Escape Analysis}
%All our designs eliminate instrumentation on accesses to memory locations
%statically guaranteed to be inaccessible to other threads.  Doing so has no
%effect on soundness or completeness because accesses to such locations are
%guaranteed never to be communicating.

\paragraph{\ctrapsmm}

\ctrapsmm has the same soundness guarantees as our base \ctraps design, but
compromises on completeness.  The concession in completeness is the result of
removing instrumentation on redundant reads.

\ctraps and \ctrapsmm rely on the same \lwt implementation, so all traps
delivered in \ctrapsmm are also delivered in the base \ctraps design.  Hence,
\ctrapsmm has the same soundness guarantee as basic \ctraps.  \ctrapsmm removes
trap hooks on some reads that have trap hooks in our base \ctraps design.
Without trap hooks, \ctrapsmm cannot deliver a trap before these reads, so it
is incomplete.  We designed our redundancy analysis to sacrifice completeness,
but preserve enough trap hooks to remain useful.  In \ctrapsmm, at least one of
each pair of control-equivalent accesses is instrumented with a trap hook.
Preserving at least one instrumenation site per control-equivalent pair
guarantees some monitoring still occurs (although the set of trapping accesses
may decrease substantially compared to our base \ctraps design).  Furthermore,
as Section~\ref{sec:dpo} discusses, instrumentation sites on interesting
control flow paths are preserved under our redundancy analysis.  For many
debugging applications, preserving this control flow information is important.

\subsection{\ctraps Applications}
\label{sec:apps}
\ctraps applications are implemented as shared object plugins ({\em i.e.}, {\tt
.so} files on Linux) that are loaded by our runtime.  They must implement the
\ctraps handler API, which requires a trap handler function and permits a
constructor, destructor,  thread constructor,  thread destructor, which allow
initialization and disposal of global and thread-local state.  A wide variety
of \ctraps applications are possible.  Several such applications were outlined
in Section~\ref{sec:background:comm}.    We implemented two applications, a
version of a debugging analysis from prior work~\cite{cci} and communication
graph collection, similar to that in DefUse~\cite{defuse}, Recon~\cite{recon},
and DMTracker~\cite{dmtracker}.

\paragraph{CCI-Prev Implementation}
CCI-Prev is a technique from CCI~\cite{cci} that computes a set of code points
that access a memory location when the previous access to the same location was
by a different thread.  We implemented a variant of CCI-Prev using \ctraps.  On
communcating read operations, the runtime updates the LWT like it normally does
on writes.  Under this policy, any operation that accesses a location that the
LWT indicates was last accessed by a different thread is recorded.  The set of
recorded accesses is the output of the analysis.  Our \ctraps implementation of
CCI-Prev took about 10 lines of code (on top of our base system).

\paragraph{Communication Graph Implementation}
Communication graphs are fundamental to several prior debugging
techniques~\cite{recon, bugaboo, defuse}.  We implement simple communication
graph collection using \ctraps.  When a trap is delivered, our implementation records
communication graph edges composed of the code point of the last writer to the
location being accessed and the code point of the trapping access.  The set of
recorded graph edges is the communication graph that we output.  Our \ctraps
implementation of communication graphs took about 50 lines of code (including
output formatting code).

\paragraph{Other Applications}
In addition to those applications, we envision that a variety of applications
will be enabled by our simple interface.  Anomaly detection and ``likely
invariant'' techniques similar to AVIO~\cite{avio}, Daikon~\cite{daikon}, and
others are a natural fit for \ctraps.  \ctraps provides the support necessary
for implementing specification checkers for specifications related to
inter-thread communication~\cite{velodrome,oshajava}.  Guided testing
tools~\cite{cuzz,chess} and bug detection tools~\cite{ctrigger} focused on
concurrency could also easily implemented as \ctraps applications.  \ctraps
also provides an interesting platform for developing new approaches to
record/replay~\cite{chimera,fdr}.  We expect these and other
applications will become more prevalent with the adoption of \ctraps'
low-overhead, deployable support for communication tracking.

Additionally, \ctraps affords the ability for {\em users} of systems to add
functionality to or extend systems.  This facet seems especially promising for
monitoring or enforcing installation-specific security or performance policies
without the need to modify applications.  While interesting, we relegate such
user-built applications to future work.

%Our redundancy analysis applies only to instrumentation on loads,
%not stores.  Eliminating instrumentation on stores may lead to stale
%information in the \lwt, which could lead to spurious traps.  Eliminating
%reads, on the other hand, can only lead to overlooked communication events,
%never false traps.

%\subsection{Design 3: \lws}
%
%Our final design, \lws, varies the compiler support used by \ctrapsfull and
%\ctrapsmm to simply eliminate instrumentation on all read operations.  With no
%instrumentation on read operations, \lws has the lowest runtime overhead, but
%the most imprecise.  However, despite not collecting information about reads,
%\lws precisely collects last writer slices: all writes are instrumented, so the
%\lwt is a sound, complete reflection of program state.

%\section{Design Justification}

%We designed \ctraps to satisfy two design goals.
%
%First, our design must be efficient enough to run in useful deployment
%environments without hardware support or sampling.  Second, our design must
%collect enough information at run time to be useful for debugging and to
%facilitate building applications.  
%
%The constraints conflict because deployment use requires minimizing
%instrumentation overheads, but collecting information requires potentially
%high-overhead instrumentation.  Our design was guided by prior work on slicing
%and concurrency debugging, and our empirical assessment of our system
%(Section~\ref{sec:eval}).  The main design challenge is deciding which memory
%dependences to track during the program's execution.  
%
%\paragraph{Tracking Most Dependences with High Overhead} 
%
%At one extreme is a design that captures all RAW, WAW, and WAR dependence
%edges.    
%%Another example is the {\em CCI-Prev} feature 
%%used by CCI~\cite{cci} tracks inter-thread RAW, WAR, WAW, and RAR dependences (and some
%%intra-thread ones) with their {\em CCI-Prev} feature. 
%Such a design collects all dependence information, which is useful for
%debugging and building analysis applications.  However, such a design falls
%short on our efficiency goal. Tracking all dependences requires work on every
%memory operation.  The additional work must update a data structure storing
%dependences and access a thread-shared structure like the \lwt that indicates
%which threads and code points are dependent.  Prior
%work~\cite{stm,coredet,raceslicing} illustrates that the overhead of this
%design point is frequently too high for use in deployment. 
%
%%Tallam, {\em et al} see overheads nearing 1000\% for server
%%programs ({\em e.g.} Apache, MySQL) with largely independent tasks.  Without
%%sampling, CCI's overheads as high as 57,000\% for some applications. In most
%%other cases, overheads are lower, but still too high for deployed use without
%%sampling.  Note that with sampling, CCI has deployable overheads, but we are
%%looking for a deployable design that does not require sampling (Constraint 1).
%
%\paragraph{Tracking Many Dependences with Moderate Overhead}
%Short of tracking all dependences, a system may make the decision to track an
%{\em ad hoc} subset of dependences,  as we discussed in
%Section~\ref{sec:background:comm}.  Such systems work well at their specific task
%({\em e.g.}, debugging a certain class of errors).  They collect enough
%dependence information to be useful for their particular application.  However,
%their usefulness is limited because tracking dependences specific to one
%application does not generalize to new applications.  
%
%One reason for tracking a subset of dependences is to reduce overheads.  Systems
%that track {\em ad hoc} dependences tend to have lower overheads than those
%that track all dependences.
%%DefUse tracks intra- and inter-thread RAR and RAW dependences; Recon tracks
%%inter-thread WAW and WAR dependences, and approximates some other dependences.
%However, these systems mostly fall short of our efficiency goal as well, often having
%overheads greater than 1000\%~\cite{defuse,recon}. Such overheads are too high for deployed
%systems.  Some systems avoid such high overheads, but require hardware
%support~\cite{avio, bugaboo, fdr} or sampling~\cite{cbi,cci} to do so, also
%failing to satisfy our design goal.
%
%\paragraph{CTraps: Tracking Select Dependences with Low Overhead}
%We make the case that our \ctraps designs satisfies our design goals.
%
%The tracking support for \ctraps is built atop last writer slices.  Last writer
%slices require tracking only inter-thread WAW dependences.  Tracking only these
%dependences does not require read operations to be instrumented.   Read
%instrumentation has high overhead because reads are more frequent than writes.
%\lws is efficient because tracking last writer slices does not need to track
%any reads.  The overhead of \ctrapsfull and \ctrapsmm is low by default because
%communication hooks (on reads) are no-ops by default.  \ctraps applications may
%perform more analysis on reads and their performance overhead scales with the
%complexity of their analysis.  We argue that such a ``pay-as-you-go'' model
%that is efficient by default is the right approach, satisfying our efficiency
%design goal.  We carefully evalute the overheads of our systems in
%Section~\ref{sec:eval}.
%
%
%Our \ctraps designs also satisfy our usefulness design goal.  \ctraps is useful for
%implementing analyses, especially those focused on concurrent program behavior.
%Inter-thread communication dependences show how values modified in one thread
%flow to operations in another thread.  This systematically selected subset of
%dependences is general enough for \ctrapsfull and \ctrapsmm to implement
%dynamic analyses related to inter-thread data flow.  Furthermore, with
%overheads low enough for deployement, \ctraps applications are useful not just
%during development, but also at runtime and in production.  \lws is also useful
%for providing important debugging information absent from current systems.  We
%describe the debugging benefits of \lws in in Section~\ref{sec:debugging}.  
%
%%\ctrapsfull and \ctrapsmm provide the ability to track reads, but they remain efficient.  The key to
%%their efficiency is the flexibility of read hooks. By default, the hooks are
%%no-ops.  CTraps {\em applications} can carefully select which dependences to
%%track, potentially increasing overhead with increases in functionality.  The
%%baseline performance of both designs is efficient, satisfying Constraint 1.
%%The flexibility provided by read and write hooks enables a variety of useful
%%applications (two of which we implement and evaluate), satisfying Constraint 2.
%

%\section{Formalism and Correctness}
%\label{sec:soundness}

%We formalize last writer slices and \ctraps.  We then discuss the soundness and
%completeness properties of our last writer slices and \ctraps designs.
%We also discuss the impact on soundness and completeness of eliminating
%redundant read instrumentation in \ctrapsmm.

%\paragraph{Preserving Write Instrumentation}
%Our system does not eliminate any write instrumentation.  Removing
%instrumentation on writes may lead to {\em false traps}.  A false trap occurs
%when a thread executes a memory access, refers to the \lwt, and the \lwt has an
%out-of-date version of the last writer information.  False traps reflect unsoundness.
%
%
%Figure~\ref{fig:unsoundness} illustrates how removing write instrumentation
%leads to unsound trap delivery.  The figure shows memory operations in two
%threads, all of which are assumed instrumented except the triangle marked
%as uninstrumented.  The actual last thread and instruction to write is shown at
%the left alongside the state of the LWT.  The figure assumes the behavior of
%\ctrapsfull, except for the single uninstrumented write operation (all of our
%designs instrument all writes).  The LWT becomes corrupted when Thread 2
%performs its uninstrumented write -- the LWT indicates the last writer was
%Thread 1, but Thread 2's uninstrumented write is the actual last writer.  When
%Thread 2 executes its next read instruction, it refers to the corrupted LWT,
%which indicates the read is communicating.  The system delivers a false trap on
%the read.  This example illustrates the risk of eliminating write
%instrumentation.  Our designs preserve instrumentation on all writes to avoid
%this problematic situation.
%
%\begin{figure}
%\centering
%\includegraphics[width=\columnwidth]{figs/unsoundness.pdf}
%\caption{\label{fig:unsoundness}Removing write instrumentation leads to unsoundness.  CTraps does not eliminate instrumentation on writes because it is unsound.}
%\end{figure}
%

%\section{Extensions to \ctraps}
%
%\subsection{Sampling}
%
%\subsection{Context-Sensitivity}

\section{System Implementation and Applications}

We implemented our last writer slicing design, our base \ctraps designs, and
our \ctrapsmm design.  We built our compiler support as a plugin for GCC ({\tt
gcc-4.7.0 20111112}).  We used GCC's points-to and escape analysis support to
prune instrumentation to non-escaping memory locations.  We used GCC's
dominator and post-dominator computations to implement our redundancy analysis.
Our compiler instruments calls to {\tt free} and {\tt delete} as writes to the
pointer being deallocated.  Our analysis handles all code compilable using this
version of GCC except for a few cases: we do not handle accesses that compile
to GCC {\tt BITFIELD\_REFERENCE} IR types because these are undocumented, we do
not handle inlined assembly instructions, and we do not handle C++ exceptions.
Note that these limitations of our research prototype are not fundamental and
the limited cases are uncommon.

We built our runtime system from scratch in about 1000 lines of C and C++ code.
We implemented the LWT as a fixed size 2GB array of 64-bit words.  Each entry
packs a program point and a thread identifier into the 64-bit word.  By
default, program points are instruction addresses, but we also support bounded
(2-address) context-sensitivity by packing the current instruction's address
and the nearest return address into an LWT entry. When an access to a memory
location occurs, we index into the LWT with the lower bits of the location's
address.  Our prototype implementation uses a lossy resolution policy for hash
collisions.  We use this policy in our prototype because it is faster and
simpler than a chaining concurrent hash table implementation.  Collisions are
extremely rare, so this simplification is unlikely to be a problem even in a
more industrial strength implementation.

\section{Evaluation}
\label{sec:eval}
We evaluated our design along several axes. Recall that in
Section~\ref{sec:debugging} we illustrated the debugging benefits of last
writer slices using case studies.   We evaluate our performance overheads to
demonstrate that last writer slices and \ctraps are feasible for use in
deployed systems.  We discuss the sources of \ctraps' overhead.   We show that
\ctraps is useful by implementing two existing applications and analyzing their
performance.  We characterize our instrumentation and the precision and
performance impact of our redundancy analysis.

\subsection{Setup and Benchmarks}

We conducted our evaluation on a machine running Linux 2.6.27-7, with a 2.27GHz
8-core Xeon E5520 processor with 2-way SMT ({\em i.e.}, 16 execution contexts)
and 10GB of memory.  We evaluated \ctraps using PARSEC, a set of architectural
benchmark programs~\cite{parsec} and a collection of real server and desktop
applications.

\subsubsection{PARSEC}

We ran PARSEC programs with their {\tt native} input set, and with each
program's 8 thread configuration. We have omitted three of the PARSEC
benchmarks because our compiler pass does not handle C++ exceptions.

\subsubsection{Servers}

\paragraph{MySQL-5.1.65} MySQL is an industrial-strength database
server. To benchmark MySQL, we used {\tt SysBench OLTP} running with its
default 8 threaded configuration, measuring MySQL's performance to be the
throughput reported by {\tt SysBench}.

\paragraph{Apache-2.4.3} 
Apache-httpd is a popular web server.  To benchmark Apache, we used {\tt
ApacheBench} to request a static {\tt html} page 1,000,000 times using 8 threads,
measuring Apache's performance to be the throughput reported by {\tt
ApacheBench}.

\paragraph{LevelDB-1.5.0}
LevelDB is a carefully tuned high-performance key-value store written by Google
developers. To benchmark LevelDB, we used the included {\tt db\_bench} utility,
running its ``Read while Writing'' test with 8 threads.  We measure LevelDB's
performance to be the operation throughput reported by {\tt db\_bench}.

\paragraph{Memcached-1.4.4}
Memcached is an in-memory key-value store frequently used as a cache for web
services.  We were unable to find a standard benchmark for Memcached.  Instead,
we wrote a C program that uses libmemcached-0.49 to issue a
mixture of 10\% load requests and 90\% store requests for a single key 
simultaneously from 8 threads.  We measured Memcached's performance to be the
total time to complete 10,000 requests in each thread.

%\begin{table}
%\begin{tabular}{ l | l | c | l }
%& & & \\ \hline
%\multirow{9}{*}{\begin{sideways}PARSEC\end{sideways}} & blackscholes & & \\
%&dedup & & \\
%&canneal & & \\
%&x264& & \\
%&vips& & \\
%&ferret& & \\
%&fluidanimate& & \\
%&swaptions & & \\
%&streamcluster & & \\ \hline
%\multirow{4}{*}{\begin{sideways}Servers\end{sideways}}&MySQL & & \\
%&Apache & & \\
%&LevelDB & & \\
%&Memcached & & \\ \hline
%\multirow{3}{*}{\begin{sideways}App\end{sideways}}&PBZip2 & & \\
%&AGet & & \\
%&Transmission & & \\
%
%\end{tabular}
%\caption{\label{tab:benchmarks}\addtodo{build this table, show LOC, \#threads, sharing pattern(?), etc}}
%\end{table}

\subsection{Performance}
\label{sec:eval:perf}

The main performance result of our work is that our designs impose performance
overheads that are low enough for use in deployed systems.
Figure~\ref{fig:perfall} shows the slowdown suffered by our benchmarks due to last writer slicing
and \ctraps running with ``no-op'' trap hooks ({\em i.e.}, no
handlers), relative to the natively executing baseline.  These data reflect
our main performance findings.  We defer considering \ctrapsmm to
Section~\ref{sec:appperf} because no-op trap hooks have negligible overhead, so
redundancy analysis provides negligible improvement.  In applications that
perform analysis in read instrumentation, the benefit of redundancy analysis is
non-negligible.

\paragraph{Last Writer Slices}
The data show that last writer slicing has very low overheads with a geometric mean of less
than 10\% across our server programs and less than 50\% across PARSEC.  Such
low overheads are very likely to be tolerable in deployed systems.  In many
cases (Apache, MySQL, {\tt dedup}, {\tt canneal}), overheads are negligible.
In all but two cases ({\tt vips} \& {\tt swaptions}), the overhead of collecting last writer slices is
less than 100\%.

\paragraph{\ctraps}
\ctraps imposes a geometric mean overhead of 14\% for server applications
and 110\% for PARSEC applications.  In 7 out of 13 of our tests, the overhead
of \ctraps is less than 50\%.  These seven low overhead benchmarks include
all of our server programs, as well as {\tt blackscholes}, {\tt dedup} and {\tt
canneal} from PARSEC.  The overhead for these applications is likely to be
tolerable in production.  {\tt vips} and {\tt swaptions} saw the highest
overheads -- around 400\%.  We discuss sources of high overheads in
Sections~\ref{sec:eval:conservative} and~\ref{sec:eval:parsecserver}.

By comparing \ctraps and last writer slicing, we see four applications ({\tt
dedup}, {\tt streamcluster}, {\tt fluidanimate}, {\tt ferret}) that have
\ctraps overhead that is probably too high for always-on use (averaging 153\%)
drop to a last writer slicing overhead that is probably low enough for always-on use
(averaging 38\%).  The large decrease indicates these programs perform
relatively more reads than other applications, and thus benefit more from
eliminating read instrumentation.
%\paragraph{\ctrapsmm}
%\ctrapsmm has very similar overheads to \ctrapsfull in most cases.   The only case
%with a major difference was {\tt vips}, indicating it executes more redundant
%reads than the other applications.  The similarity in the two designs'
%overheads is expected because our default ``no-op'' read instrumentation does
%not contribute much to overheads in most cases.  Hence, unless \ctrapsmm culls
%calls with high dynamic frequency, it is unlikely to reduce overheads much.  We
%show in Section~\ref{sec:appperf} that \ctrapsmm provides more performance
%benefit when used to implement applications that make read instrumentation sites
%more costly.
\\
\\
These high-level results support our main claim that the runtime overhead of
\ctraps is low enough for deployed systems for many 
applications, especially for servers.  We now further analyze our observed overheads, illuminating
differences between the lower-overhead server programs and the higher-overhead
PARSEC programs.

\begin{figure}
\centering
\includegraphics[width=.9\columnwidth]{plots/perf.pdf}
\caption{\label{fig:perfall}Runtime overhead of \ctraps and Last Writer Slices (LWS).}
\end{figure}

\subsubsection{Overheads Due to Conservative Escape Analysis}
\label{sec:eval:conservative}
{\tt swaptions} had the highest overheads in our tests.  We investigated their
cause and determined they are mostly artificial and due to shortcomings of our
escape analysis.  We located the inner-loop function {\tt
HJM\_SimPath\_Forward\_Blocking}.  The function uses two local matrices that
are allocated through an external call.  Our escape analysis cannot determine
that the call only does allocation and that the matrices are local, so it
conservatively assumes the matrices escape.  We manually inlined the allocation
(and a matching deallocation), and escape analysis eliminated their
instrumentation.  The program's overhead dropped from about 420\% overhead for
\ctraps to about 300\% overhead; the change for last writer slicing was from 300\% to about
130\% overhead.  We also looked at an output matrix passed into the inner-loop
function.  The matrix is allocated and deallocated in the inner-loop's caller,
but is never shared.  Eliminating instrumentation on accesses to this matrix
yielded an overhead of about 46\% for \ctraps, and to just 30\% for
last writer slicing.

Note that both these changes amount to inlining and preserve the semantics and
sharing pattern of the program.  We expect that more sophisticated escape
analysis would find these local accesses and achieve these performance gains
automatically.

\subsubsection{PARSEC vs. Servers}
\label{sec:eval:parsecserver}
In our tests, server programs were better able to tolerate the overhead of
last writer slicing and \ctraps.   We discern three reasons for this difference.

\paragraph{Independent Parallelism}
In most cases, servers have more independent parallel work.  Server programs
set up a pool of threads to handle requests.  Each thread incurs some
instrumentation latency, but that latency is hidden by other threads making
progress on independent requests.  This analysis also applies to the PARSEC
applications that had performance similar to the servers.  For instance, {\tt
blackscholes} uses largely independent threads to compute on different regions
of a matrix.  In addition, threads performing independent computation ({\em
i.e.}, non-communicating computation) are likely to operate mostly on
non-escaping, local data.  Local data are not instrumented by our compiler,
sparing the overhead.  We characterize the amount of local data in each
application in detail in Section~\ref{sec:char}.  We show that the server
programs have a larger proportion of provably local accesses than PARSEC.  The
difference supports the fact that the servers have more independent work ({\em
i.e.}, more local accesses) than PARSEC.

\paragraph{Latency Hiding I/O}
The servers perform more I/O.  Similarly to the effect of abundant parallelism,
I/O helps hide instrumentation latency.  While I/O is naturally part of many
applications, we aimed to minimize the effect of I/O in our experiments.  For
MySQL and Apache, we used local disks for logs and resources, and connected via
a local socket, rather than using the network.  In spite of these controls, I/O
remains, and hides latency in our server applications.  LevelDB's benchmark is
designed to avoid all I/O, creating and manipulating a database entirely in a
single process.  The absence of I/O may account for some of the difference in
overhead between LevelDB and the other servers.  Note that the effects of I/O
are not merely experimental noise -- in real deployed software the latency
hiding effects of I/O are likely to be even {\em more} pronounced.

\paragraph{Interference with Optimization}
The PARSEC programs are more heavily hand-optimized. In several cases
resorting to hand-coded assembly and cache-aware memory access patterns.
Running instrumentation that accesses the \lwt alongside these computations may
disrupt their cache performance, or other delicately optimized behavior,
leading to higher overheads.

Such carefully optimized code is likely to have been written by expert
programmers.  As with other code written by experts ({\em e.g.}, lock-free
data-structures), this code may not require as much {\em in situ} analysis
support as code written by distributed teams of open source developers or
novice programmers.  Depending on the maturity and level of verification of a
piece of carefully hand-optimized code, it might make sense to elide all
communication tracking instrumentation to preserve performance.

%\subsubsection{Breaking Down the Design Space}
%Figure~\ref{fig:perfall} also shows the impact of context-sensitivity, lossy
%optimization, and sampling on our overheads.  
%
%\paragraph{Context Sensitivity}
%Context-sensitivity imposes a substantial overhead in some cases.  The
%additional cost of context-sensitivity is about 10\% of the total overhead for
%PARSEC.  For server applications, the cost is less substantial, but still a
%non-negligible 4\% over the baseline.  The cost of context-sensitivity comes
%from the need to compute and store multiple addresses, rather than a single
%program counter.  
%
%In the context insensitive case, updating the \lwt requires a single register
%read ({\em rip}) a bitwise {\tt or} to combine the thread identifier and the
%instruction pointer, and the actual write to update the \lwt entry.  In the
%context sensitive case, updating the \lwt requires looking back over the call
%stack, and a bitwise {\tt or} operation at each return address.  The cost is at
%least a multiple of the cost of the constext insensitive case.  However, to
%restrict overhead to write instrumentation calls, we walk stack frames, rather
%than iteratively computing call stacks at each call and return.  As a result,
%the cost is amplified by the need to walk the stack to find return addresses.
%Future work could reduce this cost using Breadcrumbs~\cite{breadcrumbs} or
%Probabilistic Calling Contexts~\cite{probcallcont}. \addtodo{Use the Aviso
%iterative context tracker -- maybe it's actually faster?}

%\paragraph{Redundant Read Optimization}
%
%Our static analysis to remove read instrumentation reduces performance
%overheads for PARSEC programs, but does not provide much benefit in server
%programs.  In particular, removing redundant read instrumentation benefits {\tt
%x264} and {\tt swaptions}.  In the case of {\tt swaptions}, the gains are due
%to a \addtodo{Read Swaptions Code -- where's the beef?}.  In the case of {\tt
%x264}, the improvement is due to \addtodo{What is it?}.
%
%Note that the experiments depticted in Figure~\ref{fig:perfall} reflect our
%runtime in its base configuration -- writes update the \lwt, and reads only
%call a {\em no-op} hook.  Our base configuration targets manual debugging,
%which does not require actions on reads.  In more sophisticated applications of
%\ctraps, such as communication graph collection, action is required on reads.
%In Section~\ref{sec:graphperf}, we show that this optimization more
%substantially improves performance when read instrumentation sites do more
%work.
%
%\paragraph{Sampling}
%Sampling reduces overheads for both server and PARSEC applications.
%\addtodo{Explain why sampling is awesome sometimes and less awesome other
%times}

%\paragraph{Annotating Away Instrumentation}
%\addtodo{Discuss the /*Sequential*/ region in streamcluster and how not instrumenting those loads and stores might increase perf a lot}

%blackscholes 2 4
%canneal 124 226
%dedup 1 171
%ferret 40 792
%fluidanimate 3 149
%streamcluster 22 47
%swaptions 2 19
%vips 1919 9116
%x264 2987 5960

%blackscholes 3.77358490566038 7.54716981132075
%canneal 15.7360406091371 28.6802030456853
%dedup 0.128865979381443 22.0360824742268
%ferret 1.0584810796507 20.9314633500926
%fluidanimate 0.398406374501992 19.6547144754316
%streamcluster 5.44554455445545 11.6336633663366
%swaptions 0.87719298245614 8.33333333333333

\subsection{Context-Sensitivity}
We briefly evaluated the cost of context-sensitivity for \ctraps.  With a
bounded 2-address context, the average overhead for server programs was 18\%
and the average overhead for PARSEC was 145\%.  These data show that a
restricted form of context-sensitive analysis is possible in \ctraps with
overheads low enough for production.  This possibility broadens the scope of
applicability for \ctraps.  We omit a full analysis of these results here due
to space constraints.

\subsection{Evaluating CTraps Applications}
\label{sec:appperf}
We evaluated our CCI-Prev and communication graph collection CTraps
applications.  Our results show that these CTraps applications have modest
overheads that scale with their analysis complexity.  In several cases -- most
notably, in tests with our server benchmarks -- their overhead is low enough
for production use.  We also show that our redundancy analysis saves some
performance overhead in some cases.

\subsubsection{Performance with \ctraps}
Figure~\ref{fig:perfapps} shows the performance overheads of our implementation
as the slowdown with respect to the natively executing baseline.  The overheads
vary substantially across applications.  On average, the overheads are higher
than for our baseline implementation (Section~\ref{sec:eval:perf}).

In six cases (all servers, {\tt blackscholes}, and {\tt dedup}), the overheads
for CCI-Prev implementations are tolerable for deployment use (around 50\%).
This result supports our claim that \ctraps helps build useful analyses that
have overheads low enough for deployment for an important class of
applications.

Overheads for communication graph collection are generally higher and probably
too high for production use in most instances.  The higher overheads are not a
negative result; instead, they illustrate the ``complexity proportional''
nature of \ctraps' overheads.  Our communication graph collection
implementation does more costly data structure manipulations on each trap, so
its overheads are higher.  \ctraps provides a low-overhead foundation and
applications scale their overhead as necessary.

%Of these cases, eliminating redundant read instrumentation
%(\ctrapsmm) improved performance in 5 cases.  For {\tt dedup} the eliminating
%redundant reads halved the performance overhead.  This result illustrates that
%for these applications, redundancy elimination provides additional performance
%benefit.

The PARSEC benchmarks have higher overhead than the servers for CCI-Prev and
graph collection.  Graph collection overheads for {\tt vips} and {\tt
streamcluster} are the worst cases we saw (23x and 47x, respectively).  We
described some reasons for the higher overheads in these tests in
Section~\ref{sec:eval:parsecserver}.  The overheads discussed there are further
exacerbated by the increased cost of reads in these applications.  These
overheads are too high for use in deployment;  however, we note that PARSEC
applications are batch focused and compute-intensive; we expect the
long-running, event-driven nature of server applications makes them a more
important target than PARSEC for these types of analyses in production.

%In two of these cases, {\tt
%streamcluster} and {\tt swaptions}, we saw a substantial improvement in
%performance due to eliminating redundant read instrumentation ($~$4x for {\tt
%swaptions}).

%?The overheads we report are considerably lower than the overheads reported for
%?the compute-intensive benchmarks shown in the original CCI paper~\cite{cci}({\em
%?e.g.}LU Factorization \addtodo{BE SURE IT IS LU}).  While not efficient enough
%?to be deployed, these overheads are adequate for use in debugging.  Combining
%?\ctraps with sampling, as in CCI, is a promising route to deployable
%?efficiency.

\begin{figure}
\centering
\includegraphics[width=.9\columnwidth]{plots/appperf.pdf}
\caption{\label{fig:perfapps}Performance overheads of CCI-Prev and Graph Collection implemented with \ctraps.}
\end{figure}

\subsubsection{Precision and Performance with \ctrapsmm}
We also evaluated the tradeoff of precision for performance of \ctrapsmm by
applying our redundancy analysis to these applications.  We compute the loss in
precision for CCI-Prev as the symmetric difference of the sets of points
reported by \ctraps and \ctrapsmm.  We compute the loss in precision for
communication graph collection as the symmetric difference in the edge lists
reported by \ctraps and \ctrapsmm.  Table~\ref{tab:precisionapps} shows our
precision results.

\begin{table}
\small
\centering
\begin{tabular}{c | r | r  }
\multirow{2}{*}{\bf Application} & \multicolumn{2}{c}{\bf \% Symmetric Difference} \\
& {\bf \em CCI-Prev} & {\bf \em Comm. Graph} \\ \hline
Apache& 0.2\% & 0.6\% \\
MySQL & 12.9\%  & 37.9\% \\
memcached  & 6.2\% & 4.3\% \\
LevelDB & 2.2\% & 9.9\% \\ \hline
blackscholes & 6.7\% & 0.0\% \\
dedup & 1.5\% & 3.5\% \\
canneal & 37.0\%  & 3.9\% \\
x264 & 0.0\% & 0.0\% \\
vips & 47.2\% & 11.3\% \\
ferret & 25.7\% & 11.2\%\\
fluidanimate & 1.4\% & 1.6\%\\
swaptions & 0.0\% & 0.5\%\\
streamcluster & 2.2\% & 3.8\%\\
\end{tabular}
\caption{\label{tab:precisionapps}The difference in precision for \ctraps applications using our base \ctraps design versus \ctrapsmm}
\end{table}

The results show that in many cases \ctrapsmm's redundancy analysis preserves
analysis information.  For CCI-Prev, {\tt swaptions} and {\tt x264} had
identical outputs under \ctrapsmm and our base \ctraps design.  The difference for other
benchmarks was often tolerable -- $0.2\%$ for Apache, $1.5$\% for {\tt dedup},
$1.4\%$ for {\tt fluidanimate}, and 2.2\% for LevelDB.  Results were similar
for communication graph collection.   The maximum difference in precision we saw (37.9\%)
was less the maximum for CCI-Prev (47.2\%).  Several applications saw
miniscule differences in their output ({\tt swaptions}, {\tt Apache}, {\tt
fluidanimate}) or had identical outputs under both schemes ({\tt blackscholes},
{\tt x264}).

We also quantified the performance improvement of \ctrapsmm.
Figure~\ref{fig:perfcci} shows the speedup of each application with \ctrapsmm
versus our base \ctraps design, as a percentage.  The speedups were low on average, with a
geometric mean speedup of around 4\% for communication graph collection.  The
key result here is that some applications that reaped larger performance
benefits ({\em e.g.} $>$10\% for {\tt x264}) had no precision loss (cf.
Table~\ref{tab:precisionapps}).  Others, like LevelDB had a modest performance
boost (about 7\%) at a modest precision cost (about 2\% for CCI-Prev, 10\% for graph collection).

These data show that if performance is paramount and imprecision is tolerable,
\ctrapsmm can provide marginally better overheads with little cost.  We note
that for applications with very high overhead, a 5-10\% improvement in
overheads is likely negligible.  However, for applications with very low
overhead to begin with ({\em e.g.} servers), a 5-10\%
performance improvement is more likely to matter.

\begin{figure}
\centering
\includegraphics[width=.9\columnwidth]{plots/appmm.pdf}
\caption{\label{fig:perfcci}Performance overheads of CCI-Prev and Graph Collection implemented with \ctrapsmm.}
\end{figure}

\subsection{Characterizing \ctraps}
\label{sec:char}
We characterized the instrumentation pruned by escape and redundancy analysis.
We report our results in Table~\ref{tab:char}.  Each column shows the percent
of instrumentation sites pruned and the absolute number of sites in
parenthesis.  The first column shows the amount of static instrumentation
pruned by escape analysis.  The second column shows the amount of static
instrumentation pruned by our redundancy analysis.  The third column shows how
many dynamic instances of instrumentation were eliminated by redundancy analysis.  To
collect the data in the third column, we implemented a special instrumentation
pass that marked pruned functions statically so they could be counted
separately from non-pruned functions at runtime.

\begin{table}
\renewcommand{\tabcolsep}{2pt}
\centering
\small
\begin{tabular}{l | r r | r r | r r}
\multirow{3}{*}{\bf Application} & \multicolumn{6}{c}{\bf Pruned Instrumentation} \\
               & \multicolumn{2}{c|}{\bf \em Static Local}          & \multicolumn{4}{c}{\bf \em  Redundant Reads}                 \\ 
               & \multicolumn{2}{c|}{\bf \em Operations}          & \multicolumn{2}{c|}{ \em  Static} & \multicolumn{2}{c}{ \em  Dynamic}                \\ \hline
Apache         &  21.4\%&(12K)                &  2.6\%& (1.4K)                        & 3.0\%   & (15.2M)                   \\
MySQL          &  21.3\%&(53K)                &  6.8\%& (17K)                         & 0.1\%   & (230M)                    \\
memcached      &  65.8\%&(7.5K)               & 1.4\% & (160)                        & $<$0.1\% & (10.2M)                   \\
LevelDB        &  44.0\%&(4.5K)               & 5.8\% & (598)                         & 2.2\%   & (6.8M)                    \\ \hline
blackscholes   &   7.5\%&(4)                  & 3.8\% & (2)                           & $<$0.1\%& (2)                       \\
dedup          &   22.0\%&(171)               & 0.1\% & (1)                           &  0.1\%  & (899K)                    \\
canneal        &   28.6\%&(226)               & 15.7\%& (124)                         &  15.2\% & (977M)                    \\
x264           &   29.8\%&(5960)              & 14.9\%& (2987)                        &  23.3\% & (869M)                    \\
vips           &   24.6\%&(9116)              & 5.2\% & (1919)                        &  16.0\% & (3.3B)                    \\
ferret         &   20.9\%&(792)               & 1.1\% & (40)                          &   1.6\% & (923M)                    \\
fluidanimate   &   19.6\%&(149)               & 0.4\% & (3)                           &   0.1\% & (138M)                    \\
streamcluster  &   11.6\%&(47)                & 5.4\% & (22)                          & $<$0.1\%& (5.8M)                    \\
swaptions      &   8.3\% &(19)                & 1.0\% & (2)                           & $<$0.1\%& (2)                       \\
\end{tabular}
\caption{\label{tab:char}Instrumentation pruned by static analysis.}
\end{table}

The data show that escape analysis can prove a substantial fraction of
operations are local and do not require instrumentation.  Across PARSEC,
between 8\% and 30\% of accesses were proved local.  In the server programs, a
larger proportion of accesses were able to be proved local (21\% -- 66\%).  The
difference is likely due to the fact that server threads perform more
independent work (as discussed in Section~\ref{sec:eval:parsecserver}).  Eliminating
instrumentation using escape analysis is key to high performance.

The data also show that fewer accesses were eliminated by redundancy analysis
than by escape analysis.  In some cases ({\em e.g.}, {\tt swaptions}), the
fraction of static and dynamic instrumentation sites eliminated is small.  In
these cases, the few pruned calls were infrequently executed and as
Figure~\ref{fig:perfapps} shows, there was little reduction in overhead.  In
other cases, like {\tt x264}, the fraction of calls pruned was substantial 
and yielded the largest performance boost.

\section{Related Work}

There are several categories of prior work related to last writer slices and \ctraps.

\paragraph{Program Slicing}
Slicing techniques~\cite{tipslicingsurvey} are techniques for
identifying useful subsequences of a program's instructions.   Slicing can work
by analyzing program code statically or analyzing execution traces dynamically.
Many slicing techniques discussed in ~\cite{tipslicingsurvey} monitor control and
data dependences to identify which parts of a program might be relevant to a
task ({\em e.g., debugging}).

Thin Slicing~\cite{thinslicing} is related in its approach. Thin
Slicing monitors and reports a select subset of dependent operations from an
execution.  The goal is to eliminate some operations that would be included in a traditional
slice to make the slice simpler and less costly to collect.  While this work and
Thin Slicing differ in what they collect and how, our motivation to monitor a
select subset of dependences was similar to theirs.    Prior work
on slicing is unlike our work in that little has focused on monitoring
inter-thread dependences in production code without sampling or hardware support.

\paragraph{Communication Tracking in Concurrent Programs}
Several techniques have been proposed do some form of dependence and
communication tracking to analyze concurrent programs.

CCI~\cite{cci}, Bugaboo~\cite{bugaboo}, Recon~\cite{recon}, and
DefUse~\cite{defuse} all explicitly track inter-thread memory dependences at
runtime.  Unlike our work, these techniques were not designed to run in
production systems~\cite{recon,defuse} without sampling~\cite{cci} or hardware
support~\cite{bugaboo}.

Some other techniques~\cite{threadclustering,threadcriticality} have relied on
hardware debugging registers to monitor dependences efficiently enough for
production use.  These techniques are similar to our work in that they track how
threads interact ({\em i.e.}, share).  These techniques are different in that
their approach to dependence tracking is specific to the problem they address;
our system provides general support.

\paragraph{Extensible Program Instrumentation}
There are many systems for building analyses using extensible program
instrumentation.  Binary instrumentation
systems~\cite{pin,dynamorio,valgrind,roadrunner} provide completely general
support to instrument arbitrary code -- instrumenting a program to track
communciation is possible in such systems.  These systems are similar to
\ctraps in that they aim to enable dynamic analyses and runtime systems.  These
are different from \ctraps because they rely on dynamic rewriting, yielding
overheads too high for deployment.  \ctraps provides extensible support for a
restricted set of instrumentation tasks ({\em i.e.}, tracking communication)
with overheads low enough for production.  Such general systems also increase
the burden on implementors: to monitor communication, they must manually track
dependences, crafting delicate, performance sensitive code.  \ctraps tracks
only those dependences corresponding to communcation.  Furthermore, 
instrumentation systems lack the debugging benefits afforded by last writer slices.

\section{Conclusions}
This work stems from the observation that monitoring inter-thread communication
is often fast enough for use in deployed systems.  Based on this premise, we
develop system support for efficiently collecting last writer slices.  Using
that support, we developed \ctraps, a system for extensibly tracking
communicating inter-thread dependences.  \ctraps works efficiently at runtime,
in production, and without sampling or hardware support.  We showed that
\ctraps exposes a sound and complete (assuming race freedom) communication
tracking interface, enabling trap handlers to implement applications in just a
few lines of code.  We showed that a useful class of analyses can be
implemented using \ctraps that are efficient enough for deployment use in an
important set of programs ({\em e.g.}, servers, key-value stores, databases,
{\em etc.}).  We showed that last writer slices provide information that is
helpful for understanding concurrency errors with nearly no overhead.  Our
future work includes studying the generality of possible \ctraps applications
and the role of minimal hardware support for \ctraps.  We will also further study the
impact of provenance information on debugging in both sequential and concurrent
applications.

\bibliography{bib}{}
\bibliographystyle{abbrv}

\end{document}

### 2013-10-29 paper.tex
comitting the paper with some initial changes to the formalism.
Lots more to do!

@@
 
-\subsection{Formalism and Correctness}
+\subsection{\lwt Formalism and Correctness}
 \label{sec:lwssoundness}
@@
 first develop a formal description of last writer slices.  We consider an
-N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_i = \{p^{i}_{1},
-p^{i}_{2}, \ldots, p^{i}_{k_{i}}\}$ is the sequence of $k_{i}$ dynamic
-instructions executed by thread $i$.  We assume multi-threaded executions are
-sequentially consistent (SC).  Without loss of generality, we assume all
+N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_i = (p^{i}_{1},
+p^{i}_{2}, \ldots, p^{i}_{k_{i}})$ is the sequence of $k_{i}$ dynamic
+instructions executed by thread $i$.  Without loss of generality, we assume all
 instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
-$Read$ or $Write$. We also assume $m$ is a uniquely addressed memory location.
-Under SC, a point in an execution of $P$ after $e$ instructions have
+$Read$ or $Write$ and $m$ is a uniquely addressed memory location.
+A point in an execution of $P$ after $e$ instructions have
 executed is defined by $Q = (q_{0}, \ldots, q_{e})$ where each $q \in Q$ is a
@@
 
-\paragraph{Sequential Consistency and Data-races}
-
-We assume data-race freedom and a memory model that provides SC under data-race
-freedom (implying an SC execution).  We must make these assumptions because,
-for performance reasons, we do not enforce the atomicity of \lwt updates and
-their corresponding program accesses (a design choice made in prior
-work~\cite{fasttrack,recon}).  This atomicity is also required for correctness.
-
-We use the relation $p \prec q$ to denote that $p$ happens before $q$.  Under
-data-race freedom, any write by thread $t$ to $m$, $p^{t}_{j} = Write(m)$,  is
-ordered by program synchronization with $p^{t'}_{j'} = Op(m)$, a later access to
-$m$ by any other thread $t'$ -- {\em i.e.}, $p^{t}_{j} \prec p^{t'}_{j'}$.  Our
-compiler inserts an \lwt update $I(p^{t}_{j})$ into $P'$ before its write
-operation, $p^{t}_{j}$, so $I(p^{t}_{j}) \prec p^{t}_{j}$.  Happens-before is
-transitive, so $I(p^{t}_{j}) \prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec
+\paragraph{Performance, Atomicity, and Sequential Consistency}
+
+We do not enforce the atomicity of \lwt updates and their corresponding program
+accesses.  This design choice, made in prior work~\cite{fasttrack,recon},
+allows us to omit the synchronization operations that enforce the atomicity of
+these operations.  The benefit of omitting these synchronization operations is
+that overhead of our analysis is reduced. The downside is that our analysis is
+vulnerable to violations of the atomicity of \lwt updates and their
+corresponding accesses. Importantly, however, these violations can only occur
+in programs with data-races and all programs with data-races do not necessarily
+corrupt the \lwt.
+
+We use our \lwt formalism from above to characterize how data-races can
+interfere with \lwt updates.  The relation $p \prec q$ denotes that $p$ happens
+before $q$.  In the absence of data-races, any write by thread $t$ to $m$,
+$p^{t}_{j} = Write(m)$, is ordered by program synchronization with $p^{t'}_{j'}
+= Op(m)$, a later access to $m$ by any other thread $t'$ -- {\em i.e.},
+$p^{t}_{j} \prec p^{t'}_{j'}$.  Our compiler inserts an \lwt update
+$I(p^{t}_{j})$ into $P'$ before its write operation, $p^{t}_{j}$, so
+$I(p^{t}_{j}) \prec p^{t}_{j}$.  Happens-before is transitive, so without
+data-races, $I(p^{t}_{j}) \prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec
 p^{t'}_{j'}$, meaning $I(p^{t}_{j})$ happens before all accesses that
@@
 relegating such techniques to working on traces~\cite{whylinechi} or debugging
-executions.
+executions run ``in the lab''.
 
@@
 
-Computing last writer slices is fast enough to deploy and use
-continuously (see Section~\ref{sec:eval}) and the \lwt can be saved with a core
-dump.  Thus last writer slices can (and should) be made part of the standard
-debugging information provided when a program fails.
+Computing last writer slices is fast enough to deploy and use continuously (see
+Section~\ref{sec:eval}) and the \lwt can be saved with a core dump.  Thus, last
+writer slices can be made part of the standard debugging information provided
+when a program fails.
 

### 2013-10-31 paper.tex
paper changes

@@
 Preventing such undesirable behavior is a fundamental challenge faced by
-software engineers.  One important way to understand program behavior is to
-monitor which read and write operations {\em depend} on one another.  Memory
+software engineers.  An important way to understand program behavior is to
+monitor which read and write operations depend on one another.  Memory
 operations are dependent if their order determines the values they read or
-write.  Dependences are especially useful for understanding how threads
-interact.
+write.  Inter-thread dependence analysis is especially useful for understanding
+how threads interact.
 
@@
 
-We note that monitoring dependences is not only a fundamental part of the value
-of such systems, but also a fundamental part of the {\em cost}.  Tracking
-dependences requires executing additional monitoring code while the program
-runs and many of the above systems have reported overheads too high for use in
-production (10-100x
+Many of the systems mentioned above are especially valuable when they run in
+production systems on deployed code.  Unfortunately, tracking dependences is
+not only a fundamental part of the value of these systems, but also a
+fundamental part of their {\em cost}.  Tracking dependences requires executing
+additional monitoring code while the program runs and many of the above systems
+have reported overheads too high for use in production (10-100x
 slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.
-Unfortunately, many of these systems are most effective when they run in
-production.
-
-We argue in this work that dependence-tracking analysis should -- and can --
-have low enough overhead for use in deployed systems.  The key to reducing the
-overhead of dependence tracking is to carefully select and monitor only a
-{\em subset} of dependences.  We start from two simple observations: (1) Reads are
-frequent so we must minimize tracking overheads on reads, and (2) When
-addressing concurrency, {\em inter-thread} dependences are most important.
-Based on these observations, we focus on dependences that reflect {\em
-communication} between threads.  Communication occurs when a thread reads or
-overwrites a memory location that was most recently written by a different
+
+The key to reducing the overhead of dependence tracking is to carefully select
+and monitor only a {\em subset} of dependences.  We start from two simple
+observations: (1) Reads are frequent so we must minimize tracking overheads on
+reads, and (2) In concurrency analyses, {\em inter-thread} dependences are most
+important.  Based on these observations, we focus on dependences that reflect
+{\em communication} between threads.  Communication occurs when a thread reads
+or overwrites a memory location that was most recently written by a different
 thread; {\em i.e.}, inter-thread read-after-write (RAW) and write-after-write
 (WAW).  Tracking only communication dependences preserves enough information to
-implement useful applications, but has low enough overhead for always-on use in
-production.  Framed by this idea, there are two parts to our work.
+implement useful applications and analyses, but has low enough overhead for
+always-on use in production systems.
+
+Framed by these observations, we demonstrate in this work that an important,
+versatile form of inter-thread dependence-tracking analysis has inherent
+overheads that are low enough for use in deployed systems.  There are two parts
+to our work:
 
@@
 writer slices are saved with a core dump.  Last writer slices are also
-available during an execution in the debugger.  Tracking last writer slices has
-low overhead because it only imposes overhead on write operations:  reads are
-overhead-free.  In Section~\ref{sec:eval} we experimentally show
-that collecting last writer slices has low overhead.
+available during an execution in the debugger.  Note that while prior work
+({\em e.g.}, ~\cite{recon,fasttrack,conmem,conseq,cci}) has relied on {\em ad
+hoc} last-writer analyses, in this work, we isolate the essential last-writer
+information in the form of last writer slices and provide an implementation
+with overheads appropriate for production systems.  Tracking last writer slices
+has low overhead because it imposes overhead on only write operations:  reads
+are overhead-free.  In Section~\ref{sec:eval} we experimentally show that
+collecting last writer slices indeed has overheads suitable for production
+systems for a variety of applications.
 
@@
 communication with overhead low enough for production use in many important
-cases.  When threads communicate, \ctraps delivers a {\em trap} that can be
+cases.  When threads communicate, \ctraps delivers a software {\em trap} that can be
 handled by a trap handler.  Trap handlers implement {\em \ctraps applications}
@@
 %might communicate and traps on all those that communicate.
-In addition to our base \ctraps design, we also investigate \ctrapsmm, a
-variant of \ctraps that uses a flow-sensitive optimization to eliminate some
-sources of overhead at a cost in analysis precision. \ctrapsmm tracks all write
-operations that might communicate, but leaves some redundant read operations
-untracked (\ctraps, with {\bf N}o {\bf R}edundant {\bf R}eads).
 
@@
 
-\subsubsection{\ctrapsmm: Eliminating Redundant Instrumentation}
-\label{sec:dpo}
-
-The runtime support for \ctrapsmm is mostly the same as the runtime support for
-our base \ctraps design, but \ctrapsmm adds trap hooks on only a subset of
-reads.  \ctrapsmm's compiler support uses the program's control-flow structure
-to identify reads that are likely to be {\em redundant}, and eliminates trap
-hooks on those accesses.
-
-We use dominator-based analysis to eliminate instrumentation that is likely to
-be redundant.  Our analysis is function-local and works on code in SSA form.
-The compiler looks at each pair of reads that are instrumented with trap hooks.
-If a pair of instrumented reads both refer to a memory address stored in the
-same SSA variable and the instrumentation points are {\em control equivalent},
-then they are redundant.  Two reads, $A$ and $B$ are control equivalent if $A$
-dominates $B$ and $B$ post-dominates $A$ -- every path to $B$ passes through
-$A$, and every path passing through $A$ reaches $B$.  The latter read ("$B$")
-is not instrumented.  Using this technique to remove instrumentation sites
-preserves instrumentation along conditional control flow paths.  For example,
-if the second of a pair of reads is conditionally executed, and the first is
-not, then the pair are not control equivalent and both remain.
-
-Figure~\ref{fig:insthooks} illustrates how a pair of control equivalent reads
-are instrumented by the compiler for \ctrapsmm.  Pictured on the left is the
-original program.  The shapes represent memory accesses with writes marked 'W'
-and reads marked 'R'.  Notice that the the pentagon and the diamond are control
-equivalent, despite the control flow structure separating them.  Pictured on
-the right is the program after instrumentation.  There are two important things
-to notice.  First, only the first of the control equivalent accesses are
-instrumented, eliminating instrumentation overhead.  Second, the conditionally
-executed accesses on either side of the branch are both instrumented.
-Preserving these accesses is important for enabling flow-sensitive applications
-using \ctrapsmm.
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=.9\columnwidth]{figs/CTMMHooks.pdf}
-\caption{\label{fig:insthooks}Placement of trap hooks for \ctrapsmm.}
-\end{figure}
-
 \subsection{Formalism and Correctness}
@@
 
-\paragraph{\ctrapsmm}
-
-\ctrapsmm has the same soundness guarantees as our base \ctraps design, but
-compromises on completeness.  The concession in completeness is the result of
-removing instrumentation on redundant reads.
-
-\ctraps and \ctrapsmm rely on the same \lwt implementation, so all traps
-delivered in \ctrapsmm are also delivered in the base \ctraps design.  Hence,
-\ctrapsmm has the same soundness guarantee as basic \ctraps.  \ctrapsmm removes
-trap hooks on some reads that have trap hooks in our base \ctraps design.
-Without trap hooks, \ctrapsmm cannot deliver a trap before these reads, so it
-is incomplete.  We designed our redundancy analysis to sacrifice completeness,
-but preserve enough trap hooks to remain useful.  In \ctrapsmm, at least one of
-each pair of control-equivalent accesses is instrumented with a trap hook.
-Preserving at least one instrumenation site per control-equivalent pair
-guarantees some monitoring still occurs (although the set of trapping accesses
-may decrease substantially compared to our base \ctraps design).  Furthermore,
-as Section~\ref{sec:dpo} discusses, instrumentation sites on interesting
-control flow paths are preserved under our redundancy analysis.  For many
-debugging applications, preserving this control flow information is important.
-
 \subsection{\ctraps Applications}
@@
 
-We implemented our last writer slicing design, our base \ctraps designs, and
-our \ctrapsmm design.  We built our compiler support as a plugin for GCC ({\tt
-gcc-4.7.0 20111112}).  We used GCC's points-to and escape analysis support to
-prune instrumentation to non-escaping memory locations.  We used GCC's
-dominator and post-dominator computations to implement our redundancy analysis.
-Our compiler instruments calls to {\tt free} and {\tt delete} as writes to the
-pointer being deallocated.  Our analysis handles all code compilable using this
-version of GCC except for a few cases: we do not handle accesses that compile
-to GCC {\tt BITFIELD\_REFERENCE} IR types because these are undocumented, we do
-not handle inlined assembly instructions, and we do not handle C++ exceptions.
-Note that these limitations of our research prototype are not fundamental and
-the limited cases are uncommon.
+We implemented last writer slicing and \ctraps.  We built our compiler support
+as a plugin for GCC ({\tt gcc-4.7.0 20111112}).  We used GCC's points-to and
+escape analysis support to prune instrumentation to non-escaping memory
+locations.  We used GCC's dominator and post-dominator computations to
+implement our redundancy analysis.  Our compiler instruments calls to {\tt
+free} and {\tt delete} as writes to the pointer being deallocated.  Our
+analysis handles all code compilable using this version of GCC except for a few
+cases: we do not handle accesses that compile to GCC {\tt BITFIELD\_REFERENCE}
+IR types because these are undocumented, we do not handle inlined assembly
+instructions, and we do not handle C++ exceptions.  Note that these limitations
+of our research prototype are not fundamental and the limited cases are
+uncommon.
 
@@
 handlers), relative to the natively executing baseline.  These data reflect
-our main performance findings.  We defer considering \ctrapsmm to
-Section~\ref{sec:appperf} because no-op trap hooks have negligible overhead, so
-redundancy analysis provides negligible improvement.  In applications that
-perform analysis in read instrumentation, the benefit of redundancy analysis is
-non-negligible.
+our main performance findings.
 
@@
 
-\subsubsection{Precision and Performance with \ctrapsmm}
-We also evaluated the tradeoff of precision for performance of \ctrapsmm by
-applying our redundancy analysis to these applications.  We compute the loss in
-precision for CCI-Prev as the symmetric difference of the sets of points
-reported by \ctraps and \ctrapsmm.  We compute the loss in precision for
-communication graph collection as the symmetric difference in the edge lists
-reported by \ctraps and \ctrapsmm.  Table~\ref{tab:precisionapps} shows our
-precision results.
-
-\begin{table}
-\small
-\centering
-\begin{tabular}{c | r | r  }
-\multirow{2}{*}{\bf Application} & \multicolumn{2}{c}{\bf \% Symmetric Difference} \\
-& {\bf \em CCI-Prev} & {\bf \em Comm. Graph} \\ \hline
-Apache& 0.2\% & 0.6\% \\
-MySQL & 12.9\%  & 37.9\% \\
-memcached  & 6.2\% & 4.3\% \\
-LevelDB & 2.2\% & 9.9\% \\ \hline
-blackscholes & 6.7\% & 0.0\% \\
-dedup & 1.5\% & 3.5\% \\
-canneal & 37.0\%  & 3.9\% \\
-x264 & 0.0\% & 0.0\% \\
-vips & 47.2\% & 11.3\% \\
-ferret & 25.7\% & 11.2\%\\
-fluidanimate & 1.4\% & 1.6\%\\
-swaptions & 0.0\% & 0.5\%\\
-streamcluster & 2.2\% & 3.8\%\\
-\end{tabular}
-\caption{\label{tab:precisionapps}The difference in precision for \ctraps applications using our base \ctraps design versus \ctrapsmm}
-\end{table}
-
-The results show that in many cases \ctrapsmm's redundancy analysis preserves
-analysis information.  For CCI-Prev, {\tt swaptions} and {\tt x264} had
-identical outputs under \ctrapsmm and our base \ctraps design.  The difference for other
-benchmarks was often tolerable -- $0.2\%$ for Apache, $1.5$\% for {\tt dedup},
-$1.4\%$ for {\tt fluidanimate}, and 2.2\% for LevelDB.  Results were similar
-for communication graph collection.   The maximum difference in precision we saw (37.9\%)
-was less the maximum for CCI-Prev (47.2\%).  Several applications saw
-miniscule differences in their output ({\tt swaptions}, {\tt Apache}, {\tt
-fluidanimate}) or had identical outputs under both schemes ({\tt blackscholes},
-{\tt x264}).
-
-We also quantified the performance improvement of \ctrapsmm.
-Figure~\ref{fig:perfcci} shows the speedup of each application with \ctrapsmm
-versus our base \ctraps design, as a percentage.  The speedups were low on average, with a
-geometric mean speedup of around 4\% for communication graph collection.  The
-key result here is that some applications that reaped larger performance
-benefits ({\em e.g.} $>$10\% for {\tt x264}) had no precision loss (cf.
-Table~\ref{tab:precisionapps}).  Others, like LevelDB had a modest performance
-boost (about 7\%) at a modest precision cost (about 2\% for CCI-Prev, 10\% for graph collection).
-
-These data show that if performance is paramount and imprecision is tolerable,
-\ctrapsmm can provide marginally better overheads with little cost.  We note
-that for applications with very high overhead, a 5-10\% improvement in
-overheads is likely negligible.  However, for applications with very low
-overhead to begin with ({\em e.g.} servers), a 5-10\%
-performance improvement is more likely to matter.
-
-\begin{figure}
-\centering
-\includegraphics[width=.9\columnwidth]{plots/appmm.pdf}
-\caption{\label{fig:perfcci}Performance overheads of CCI-Prev and Graph Collection implemented with \ctrapsmm.}
-\end{figure}
-
 \subsection{Characterizing \ctraps}
@@
 \begin{table}
-\renewcommand{\tabcolsep}{2pt}
+%\renewcommand{\tabcolsep}{2pt}
 \centering
 \small
-\begin{tabular}{l | r r | r r | r r}
-\multirow{3}{*}{\bf Application} & \multicolumn{6}{c}{\bf Pruned Instrumentation} \\
-               & \multicolumn{2}{c|}{\bf \em Static Local}          & \multicolumn{4}{c}{\bf \em  Redundant Reads}                 \\ 
-               & \multicolumn{2}{c|}{\bf \em Operations}          & \multicolumn{2}{c|}{ \em  Static} & \multicolumn{2}{c}{ \em  Dynamic}                \\ \hline
-Apache         &  21.4\%&(12K)                &  2.6\%& (1.4K)                        & 3.0\%   & (15.2M)                   \\
-MySQL          &  21.3\%&(53K)                &  6.8\%& (17K)                         & 0.1\%   & (230M)                    \\
-memcached      &  65.8\%&(7.5K)               & 1.4\% & (160)                        & $<$0.1\% & (10.2M)                   \\
-LevelDB        &  44.0\%&(4.5K)               & 5.8\% & (598)                         & 2.2\%   & (6.8M)                    \\ \hline
-blackscholes   &   7.5\%&(4)                  & 3.8\% & (2)                           & $<$0.1\%& (2)                       \\
-dedup          &   22.0\%&(171)               & 0.1\% & (1)                           &  0.1\%  & (899K)                    \\
-canneal        &   28.6\%&(226)               & 15.7\%& (124)                         &  15.2\% & (977M)                    \\
-x264           &   29.8\%&(5960)              & 14.9\%& (2987)                        &  23.3\% & (869M)                    \\
-vips           &   24.6\%&(9116)              & 5.2\% & (1919)                        &  16.0\% & (3.3B)                    \\
-ferret         &   20.9\%&(792)               & 1.1\% & (40)                          &   1.6\% & (923M)                    \\
-fluidanimate   &   19.6\%&(149)               & 0.4\% & (3)                           &   0.1\% & (138M)                    \\
-streamcluster  &   11.6\%&(47)                & 5.4\% & (22)                          & $<$0.1\%& (5.8M)                    \\
-swaptions      &   8.3\% &(19)                & 1.0\% & (2)                           & $<$0.1\%& (2)                       \\
+\begin{tabular}{l | r r }
+\multirow{3}{*}{\bf Application} & \multicolumn{2}{c}{\bf \em Pruned } \\
+               & \multicolumn{2}{c}{\bf \em Static Local}  \\ 
+               & \multicolumn{2}{c}{\bf \em Operations}    \\ \hline
+
+Apache         &  21.4\%&(12K)                    \\
+MySQL          &  21.3\%&(53K)                    \\
+memcached      &  65.8\%&(7.5K)                   \\
+LevelDB        &  44.0\%&(4.5K)                   \\ \hline
+blackscholes   &   7.5\%&(4)                      \\
+dedup          &   22.0\%&(171)                   \\
+canneal        &   28.6\%&(226)                   \\
+x264           &   29.8\%&(5960)                  \\
+vips           &   24.6\%&(9116)                  \\
+ferret         &   20.9\%&(792)                   \\
+fluidanimate   &   19.6\%&(149)                   \\
+streamcluster  &   11.6\%&(47)                    \\
+swaptions      &   8.3\% &(19)                    \\
 \end{tabular}

### 2013-11-01 paper.tex
committing a version of the paper with a background that serves the main message.

@@
 operations are dependent if their order determines the values they read or
-write.  Inter-thread dependence analysis is especially useful for understanding
-how threads interact.
+write.  The value of dependence information is made especially clear by its
+prevalent use during manual program debugging and in support of program
+analysis tools.
+
+Debugging a concurrent program requires understanding how threads interact and
+dependence information furnishes that understanding.  Programmers debug
+programs more effectively when they can ask questions about the provenance of
+values during a program's execution~\cite{slicingsurvey,whylineicse}.
+Dependence information helps programmers with the difficult task of identifying
+control- and data-flows that influenced a particular data value at a particular
+point.  In concurrent programs, data provenance information is vital because
+influential control- and data-flows may be in any thread, making the task of
+understanding the origin of a value much more complex.
+
+Compounding this complexity, many concurrency bugs manifest as failures rarely.
+When a deployed system fails and a programmer receives a bug report,
+reproducing the reported failure to collect dependence information may be
+difficult or impossible.  It is essential to collect as much information as
+possible in deployed systems to reduce the need for time-consuming forensic bug
+reproduction.  Unfortunately, prior approaches to tracking dependences in
+concurrent code~\cite{raceslicing,tipslicingsurvey} impose high run time
+overheads.  These overheads are prohibitively high for use in production.
 
@@
 
-Prior work has used dependence information to reduce the burden of making
-reliable, efficient concurrent software.  Bug
+Dependence information is also an important part of many program analysis
+tools.  Prior work has used dependence information to reduce the burden of
+making reliable, efficient concurrent software.  Bug
 detection~\cite{avio,fasttrack,raceslicing,dmtracker} and
@@
 
-Many of the systems mentioned above are especially valuable when they run in
-production systems on deployed code.  Unfortunately, tracking dependences is
-not only a fundamental part of the value of these systems, but also a
-fundamental part of their {\em cost}.  Tracking dependences requires executing
-additional monitoring code while the program runs and many of the above systems
-have reported overheads too high for use in production (10-100x
+These systems would be especially valuable, if run in production systems on
+deployed code.  However, as with dependence tracking on its own, most of these
+systems have overheads too high for use in production (5-100x
 slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.
@@
 
-Framed by these observations, we demonstrate in this work that an important,
+Framed by these observations, we demonstrate in this work that a 
 versatile form of inter-thread dependence-tracking analysis has inherent
@@
 
-Many systems that support programming and execution models, debugging tools,
-and other software engineering tools track dynamic memory dependences during a
-program's execution.  In this section we provide context for \ctraps by
-discussing some of these systems, focusing on those that address concurrency.
-We then briefly discuss debugging with data dependence information, in order to
-motivate the role of last writer slices at debugging time.
+In this section we discuss debugging with data dependence information, using an
+example, in order to motivate collecting last writer slices in deployed
+systems.  We then provide context for \ctraps by discussing some more
+sophisticated concurrency analyses that rely on the kind of dependence
+information we are focusing on in this work.
+
+%Many systems that support programming and execution models, debugging tools,
+%and other software engineering tools track dynamic memory dependences during a
+%program's execution.
+
+\subsection{Debugging and Data Provenance}
+
+Data dependence information is directly useful during program debugging.  Prior
+work has proposed a wide variety of dynamic program slicing
+techniques~\cite{thinslicing,tipslicingsurvey}.  These techniques record a
+subset (slice) of dynamic program instructions selected using data and control
+dependence information.  Dynamic dependence slices expose the provenance of
+data in the execution.  Work on debugging~\cite{whylinechi,whylineicse} in the
+HCI and Software Engineering community shows that when programmers debug
+programs, they benefit from the ability to directly ask such provenance
+(``why'') questions of their debugger.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\columnwidth]{figs/BackgroundCoreDumpFail.pdf}
+\caption{\label{fig:coreDumpFail}{\bf Ordering violation bug in HTTrack.} Simply examining memory does not provide provenance information that helps understand this bug.}
+\end{figure}
+
+Figure~\ref{fig:coreDumpFail} illustrates the role that provenance information
+plays in debugging, using an example of a real bug from HTTrack, a web caching
+tool.  The figure depicts two threads manipulating a data structure that
+contains a mutex\_lock.  The program has an ordering violation bug: the program
+crashes when Thread 2 tries to acquire the lock before Thread 1 has initialized
+it.  Assuming the run time environment supports it, a core dump delivered to
+the programmer after such a failure reveals very little about how to fix the
+bug.  The lock is uninitizalized, so it may hold {\tt NULL} or it may hold
+random bits.  Both scenarios provide no insight into the root cause of the
+failure, which is that the code permits a use of the lock before its
+initialization.   Comparing the core dump from the failing run to the memory
+image from a correct execution -- a likely debugging strategy -- reveals only
+that in a correct execution, the lock contains valid mutex state.  This
+revalation is of limited use as it does not rule out arbitrary data corruption
+as the bug's root cause.
+
+The vital, missing piece to solving this bug is provenance information.
+Debugging this bug would be easier if the core dump included information about
+the thread and code point that last wrote each variable.  The programmer would
+see that no thread had initialized the lock in the failing execution, but that
+Thread 1 initialized the lock in non-failing executions.  This information
+rules out data corruption as a root cause and guides the programmer to add
+synchronization ensuring the correct ordering relationship.
+
+Another facet to debugging this bug is that it manifests as a failure
+infrequently and under only real-world conditions.  If the programmer runs an
+instrumented or tooled version of the program, the failure may not manifest in
+a reasonable amount of time and running it may require unavailable real-world
+inputs.  Thus, collecting provenance or other information offline is
+time-consuming or impossible.  A better strategy is to continuously collect
+that information in deployed systems and package it with core dumps sent to
+programmers with bug reports.  The last writer slicing design we propose in
+this work espouses this {\em in situ} data collection strategy and does so with
+overheads appropriate for production systems.
+
+%Dynamic dependence slices are especially useful for debugging if they are
+%available after a failure has occurred in a deployed system. However, slicing
+%in deployed systems is uncommon, because deployed systems demand low overheads
+%and slicing can have high overhead.  Our system provides provenance information
+%that is useful at debugging time by collecting {\em last writer slices} that
+%help understand inter-thread communication.  We track last writer slices with
+%overhead low enough for use in production code.  Deployable efficiency makes it
+%possible to collect and report last writer slices the same way today's systems
+%collect and report core dumps.
 
@@
 for use in production.
-
-\subsection{Debugging with Data Dependence Information}
-
-Data dependence information is directly useful during program debugging.  A
-wide variety of dynamic program slicing
-techniques~\cite{thinslicing,tipslicingsurvey} have been proposed.  These
-techniques record a subset (slice) of dynamic program instructions selected
-using data and control dependence information.  Dynamic dependence slices
-expose the provenance of control- and data-flows in the execution.  Work on
-debugging~\cite{whylinechi,whylineicse} in the HCI and Software Engineering
-community shows that when programmers debug programs, they benefit from the
-ability to directly ask such provenance (``why'') questions of their debugger.
-
-Dynamic dependence slices are especially useful for debugging if they are
-available after a failure has occurred in a deployed system. However, slicing
-in deployed systems is uncommon, because deployed systems demand low overheads
-and slicing can have high overhead.  Our system provides provenance information
-that is useful at debugging time by collecting {\em last writer slices} that
-help understand inter-thread communication.  We track last writer slices
-with overhead low enough for use in production code.  Deployable efficiency
-makes it possible to collect and report last writer slices the same way today's
-systems collect and report core dumps.
 
@@
 \label{sec:debugging}
-Prior work has demonstrated that it is useful during debugging to allow
-programmers to ask ``why?'' questions~\cite{whylineicse,whylinechi} about the
-provenance of program state.  To answer such questions, prior work has mostly
-used dependence slices that include intra-thread
-dependences~\cite{whylineicse,conseq,tipslicingsurvey}.  The overhead of
-slicing dependence chains and tracking temporal sequences is often high,
-relegating such techniques to working on traces~\cite{whylinechi} or debugging
-executions run ``in the lab''.
+The WhyLine debugger~\cite{whylineicse} demonstrated that it is useful to
+enable programmers to make data provenance queries during debugging.  To answer
+such questions, prior work has mostly used dependence slices that include
+intra-thread dependences~\cite{whylineicse,conseq,tipslicingsurvey}.  The
+overhead of slicing dependence chains and tracking temporal sequences is often
+high, relegating such techniques to working on traces~\cite{whylinechi} or
+debugging executions run ``in the lab''.
 
@@
 dependence that helps understand why the program failed.
+
+\paragraph{Case Study: HTTrack}
 

### 2013-11-02 paper.tex
committing a draft with more figure edits and some other fun changes

@@
 
-Data dependence information is directly useful during program debugging.  Prior
-work has proposed a wide variety of dynamic program slicing
-techniques~\cite{thinslicing,tipslicingsurvey}.  These techniques record a
-subset (slice) of dynamic program instructions selected using data and control
-dependence information.  Dynamic dependence slices expose the provenance of
-data in the execution.  Work on debugging~\cite{whylinechi,whylineicse} in the
-HCI and Software Engineering community shows that when programmers debug
-programs, they benefit from the ability to directly ask such provenance
-(``why'') questions of their debugger.
+Data provenance information is directly useful during program debugging.
+Work on debugging~\cite{whylinechi,whylineicse} in the HCI and Software
+Engineering community shows that when programmers debug programs, they benefit
+from the ability to directly ask such provenance (``why'') questions of their
+debugger.
 
@@
 tool.  The figure depicts two threads manipulating a data structure that
-contains a mutex\_lock.  The program has an ordering violation bug: the program
-crashes when Thread 2 tries to acquire the lock before Thread 1 has initialized
-it.  Assuming the run time environment supports it, a core dump delivered to
-the programmer after such a failure reveals very little about how to fix the
-bug.  The lock is uninitizalized, so it may hold {\tt NULL} or it may hold
-random bits.  Both scenarios provide no insight into the root cause of the
-failure, which is that the code permits a use of the lock before its
-initialization.   Comparing the core dump from the failing run to the memory
-image from a correct execution -- a likely debugging strategy -- reveals only
-that in a correct execution, the lock contains valid mutex state.  This
-revalation is of limited use as it does not rule out arbitrary data corruption
-as the bug's root cause.
+contains a mutex\_lock.  The program has an ordering violation bug: the
+program crashes when Thread 2 tries to acquire the lock before Thread 1 has
+initialized it.  Assuming the run time environment supports it, a core dump
+delivered to the programmer after such a failure reveals very little about how
+to fix the bug.  The lock is uninitizalized, so the core dump may show that it
+holds {\tt NULL} or it may hold random bits.  Both scenarios provide no
+insight into the root cause of the failure, which is that the code permits a
+use of the lock before its initialization.   Comparing the core dump from the
+failing run to the memory image from a correct execution -- a helpful debugging
+strategy -- reveals only that in a correct execution, the lock contains valid
+mutex state.  This revalation is of limited use as it does not rule out
+arbitrary data corruption as the bug's root cause.
 
 The vital, missing piece to solving this bug is provenance information.
-Debugging this bug would be easier if the core dump included information about
-the thread and code point that last wrote each variable.  The programmer would
-see that no thread had initialized the lock in the failing execution, but that
-Thread 1 initialized the lock in non-failing executions.  This information
-rules out data corruption as a root cause and guides the programmer to add
-synchronization ensuring the correct ordering relationship.
-
-Another facet to debugging this bug is that it manifests as a failure
-infrequently and under only real-world conditions.  If the programmer runs an
-instrumented or tooled version of the program, the failure may not manifest in
-a reasonable amount of time and running it may require unavailable real-world
-inputs.  Thus, collecting provenance or other information offline is
-time-consuming or impossible.  A better strategy is to continuously collect
-that information in deployed systems and package it with core dumps sent to
-programmers with bug reports.  The last writer slicing design we propose in
-this work espouses this {\em in situ} data collection strategy and does so with
-overheads appropriate for production systems.
+Debugging would be easier if the core dump included information about the
+thread and code point that last wrote the lock.  The programmer would see that
+no thread had initialized the lock in the failing execution, but that Thread 1
+initialized the lock in non-failing executions.  This information guides the
+programmer to fix the bug by adding synchronization that ensures the correct
+operation ordering.
+
+Another challenge in debugging this bug is that it manifests as a failure
+infrequently and under only real-world conditions.  If the programmer runs the
+program in a debugger, the failure may not manifest in a reasonable amount of
+time and running it may require unavailable real-world inputs.  Thus,
+collecting provenance or other information offline is time-consuming or
+impossible.  A better strategy is to continuously collect that information in
+deployed systems and package it with core dumps sent to programmers with bug
+reports.  In this work, we propose a last writer slicing design that espouses
+this {\em in situ} data collection strategy and does so with overheads
+appropriate for production systems.
 
@@
 
-A good first example of a system that monitors inter-thread dependences is
-software transactional memory (STM)~\cite{stm}.  In STMs, dependences
-correspond to interference between accesses in different threads that must be
-detected to rollback conflicting transactions.  Deterministic concurrent
-runtime systems~\cite{coredet,grace} are similar -- they monitor
-inter-thread dependences, to determine when to apply deterministic ordering
-constraints on threads' accesses.  STMs and deterministic systems must track
-all dependences and they must track them precisely, to provide strong
-guarantees as programming and execution models.  Unfortunately, precision can
-have a high runtime overhead -- {\em e.g.}, CoreDet incurs a mean overhead of
-around 5x~\cite{coredet}.  This overhead is too high for most production
-systems.
+Monitoring inter-thread dependences is the foundation of many useful
+concurrency analyses.  For example, in a software transactional memory (STM)
+system~\cite{stm} dependences correspond to interference between accesses in
+different threads that must be detected to rollback conflicting transactions.
+Deterministic concurrent runtime systems~\cite{coredet,grace} are similar --
+they monitor inter-thread dependences, to determine when to apply
+deterministic ordering constraints on threads' accesses.  STMs and
+deterministic systems must track all dependences and they must track them
+precisely, to provide strong guarantees as programming and execution models.
+Unfortunately, precision has a high runtime overhead -- {\em e.g.},
+CoreDet incurs a mean overhead of around 5x~\cite{coredet}.  This overhead is
+too high for use in production.
 
@@
 
-In this work, we select a subset of inter-thread dependences in a systematic
-way, rather than choosing an {\em ad hoc} subset.  \ctraps tracks dependences
-that correspond to inter-thread communication.  \ctraps provides enough
-precision to build useful analyses and runtime systems with overhead low enough
-for use in production.
+The design for \ctraps that we develop in this work monitors a subset of
+inter-thread dependences that correspond to inter-thread communication.
+\ctraps provides enough precision to build useful analyses and runtime systems
+with overhead low enough for use in production.
 
@@
 \label{sec:lastwriterslices}
-
-A last writer slice is a record associated with a memory location that stores
-the thread and program point that last wrote data to that memory location.  A
-last writer slice is a dynamic property of a memory location at a point in a
-program's execution.   We formalize last writer slices in
-Section~\ref{sec:lwssoundness}.
-
-We first discuss last writer slices in more depth by describing the design of
-our support for collecting them.  The system support consists of a runtime
-library and a compiler analysis.  In Section~\ref{sec:debugging} we highlight
-the role of last writer slices in debugging and show they help diagnose real
-bugs, using case studies.
-
-\subsection{Runtime Support: Maintaining the Last Writer Table}
-We gather last writer slices at runtime by maintaining a {\em last writer
-table} (\lwt).  The \lwt is a {\em shadow memory} that holds one entry for each
-memory address accessed by the program.  Each \lwt entry records the thread and
-program point that last wrote to the entry's corresponding memory location.
-When a thread performs a write to a memory location, it updates the \lwt by
-recording its thread identifier and the program point of its write operation to
-the memory location's \lwt entry.  A program point may be an
-instruction address, or a call stack.  We use instructions by default, but our
-system supports 2-address bounded call-stacks as well.
-
-Figure~\ref{fig:basicLWT} shows the basic operation of our runtime support.
-The right side of the figure shows a program execution with three threads.  On
-the left is the \lwt at the point of each access.  There are two key things to
+A last writer slice is a dynamic property of a memory location at a point in a
+program's execution.   A memory location's last writer slice is a record
+containing the program point and the identifier of the thread that last wrote
+data to that memory location.  We gather last writer slices at runtime by
+maintaining a {\em last writer table} (\lwt) that maps from a memory
+location's address to its last writer slice.  Just before a thread writes to a
+memory location,  the thread first finds the memory location's entry in the
+\lwt and updates that entry with its thread identifier and its current point in the
+program.  After updating the \lwt, the thread performs the original memory
+access.
+
+Figure~\ref{fig:basicLWT} shows the basic operation of the \lwt.  The right
+side of the figure shows a program execution with three threads.  On the left
+is the \lwt at the point of each memory access.  There are two key things to
 notice.  First, write operations update the \lwt entry of the memory location
-they access.  Second, there is no action necessary for read operations.  Making
-read operations cheap is key to making the overhead of last writer slicing low
-enough for deployment use.
+they access.  Second, there is no action necessary for read operations.
+Making read operations cheap is key to making the overhead of last writer
+slicing low enough for deployment use.
 
@@
 
-\subsection{Compiler Support: Analysis and Instrumentation}
+\subsection{Debugging with Last Writer Slices}
+\label{sec:debugging}
+%To answer data provenance questions during debugging, prior work has mostly
+%used dependence slices that include intra-thread
+%dependences~\cite{whylineicse,conseq,tipslicingsurvey}.  The overhead of
+%slicing dependence chains and tracking temporal sequences is often high,
+%relegating such techniques to working on traces~\cite{whylinechi} or debugging
+%executions run ``in the lab''.
+
+At any point in an execution, the information stored in a memory location's
+\lwt entry represents the provenance of the data stored in that memory
+location.   Last writer slices do not explicitly record any information about
+read operations and they do not need to to be useful.  Interrupting the
+execution while debugging and examining an entry in the \lwt reveals the
+provenance of a value just when it is needed.  For example, if an assertion
+failure occurs due to an erroneous inter-thread data flow, the provenance of
+the data in the assertion's condition helps explain why the failure occurred.
+Computing last writer slices is fast enough to deploy and use continuously
+(see Section~\ref{sec:eval}) and the \lwt can be saved with a core dump.
+Thus, last writer slices can be made part of the standard debugging
+information provided when a program fails.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\columnwidth]{figs/LWSHTTDebug.pdf}
+\caption{\label{fig:httlws}{\bf Using last writer slices to debug an
+ordering violation in HTTrack.} The provenance information stored in the last
+writer slice for the mutex variable helps understand the bug's root cause.}
+\end{figure}
+
+%The core dump does not provide any
+%insight into why the mutex is invalid at the time of the call or what its
+%value should have been instead.
+
+Figure~\ref{fig:httlws} shows how last writer slices help understand the root
+cause of the concurrency bug from Figure~\ref{fig:coreDumpFail}.  Recall
+that the the program crashes at point {\bf B} when it calls {\tt
+hts\_mutex\_lock()} with an invalid mutex.    The root cause of this bug is
+that Thread 1 fails to initialize the mutex before Thread 2 uses that mutex.
+The last writer slices in the figure make this root cause clear: in the
+failing execution, the last writer slice indicates the mutex is uninitialized; in
+the non-failing execution, the last writer slice indicates the mutex was 
+initialized by Thread 1 at point {\bf A}.  This information tells the
+programmer that there is an ordering relationship between {\bf A} and {\bf B}
+that must be enforced to fix the bug.  Note that without last writer slices,
+the program crashes at {\bf B} and the core dump provides no information about
+the connection between point {\bf A} and point {\bf B} -- the programmer would be left to guess that the bug was an initialization problem.
+
+Another important trait of last writer slices illustrated by
+Figure~\ref{fig:httlws} is that last writer slices are reported back to the
+programmer {\em with the core dump on a failure}.  Capturing good debugging
+information is especially important for concurrency bugs that manifest very
+rarely or in the presence of only peculiar real-world inputs.  A primary goal
+of this work is to show that we can collect last writer slices with overheads
+that are low enough for production use so they can be included pervasively in
+bug reports.
+
+%We discuss last writer slices in more depth by first describing our run time
+%support for collecting them.  The system support consists of a runtime library
+%and a compiler analysis.  Section~\ref{sec:debugging} shows how last writer
+%slices help with debugging using our running example from
+%Figure~\ref{fig:bgcoredumpfail}.  We formalize last writer slices in
+%Section~\ref{sec:lwssoundness}.
+
+\subsection{Runtime \& Compiler Support}
+ The \lwt is implemented as a {\em shadow memory} that holds one entry for
+each memory address accessed by the program.  Each \lwt entry records the
+thread and program point that last wrote to the entry's corresponding memory
+location.  When a thread performs a write to a memory location, it updates the
+\lwt by recording its thread identifier and the program point of its write
+operation to the memory location's \lwt entry.  A program point may be an
+instruction address, or a call stack.  We use instructions by default, but we experimented with 2-address bounded call-stacks as well.
+
 Collecting last writer slices requires instrumentation on each write operation.
@@
 
-\subsection{Debugging with Last Writer Slices}
-\label{sec:debugging}
-The WhyLine debugger~\cite{whylineicse} demonstrated that it is useful to
-enable programmers to make data provenance queries during debugging.  To answer
-such questions, prior work has mostly used dependence slices that include
-intra-thread dependences~\cite{whylineicse,conseq,tipslicingsurvey}.  The
-overhead of slicing dependence chains and tracking temporal sequences is often
-high, relegating such techniques to working on traces~\cite{whylinechi} or
-debugging executions run ``in the lab''.
-
-The \lwt dynamically records information about write operations -- what thread
-and code point last wrote each memory location.   These last writer slices do
-not explicitly encode anything about read operations.  However, by looking at
-the \lwt at a point in the execution, the provenance of the data being read or
-over-written can be determined.  For example, if an assertion failure occurs
-due to an erroneous inter-thread data flow, the provenance of the data in the
-assertion's condition helps explain why the failure occurred.  Last writer
-slices therefore answer an important class of ``why?'' questions related to
-concurrency errors.
-
-Computing last writer slices is fast enough to deploy and use continuously (see
-Section~\ref{sec:eval}) and the \lwt can be saved with a core dump.  Thus, last
-writer slices can be made part of the standard debugging information provided
-when a program fails.
-
-\subsubsection{Case Studies}
-
-We illustrate the debugging benefits of last writer slices using a set of case
-studies.  In each case, we debug a buggy program using last writer slices.  We
-show that for some concurrency errors, last writer slices are a good match
-because they provide answers to important {\em ``why?''}
-questions~\cite{whylineicse}.
-
-\paragraph{Case Study: PBZip2}
-We conducted a case study using a buggy version of PBZip2-0.9.1.    The bug is
-a use-after-delete bug that leads to an assertion failure.  A worker thread
-tries to acquire a mutex after it has been deleted by the main thread during
-shutdown, causing the assertion to fail.
-
-Last writer slices provide the key piece of information to debug this failure: the last
-write to the mutex before the worker tried acquiring it was when the main
-thread deleted it.  A developer is likely to begin debugging by checking the
-validity of the mutex.  Upon finding it is invalid ({\em
-i.e.}, it has been deleted), the developer is likely to probe to determine who 
-deleted it.  Without the information in the \lwt, answering that question is difficult
-because the main thread may have executed arbitrarily far from where it deleted
-the mutex.  With the \lwt, answering that question only requires the developer
-to check the mutex's entry.
-
-\paragraph{Case Study: MySQL}
-Last writer slices were also useful for debugging a logging system failure in
-MySQL-4.0.12.  The bug is an atomicity violation involving two threads.  One
-thread rotates the database's log file.  Log rotation involves closing the log
-file, marking the log closed, opening a new log file, and marking the log open.
-Log rotation should be atomic, but lack of synchronization allows another
-thread to run queries during rotation.  The query code checks if the log is
-marked closed; if it is not closed, the queries are logged, but if it is
-closed, the queries are not logged and the program simply continues.  We added
-an assertion to the non-logging case because this behavior is a security
-vulnerability.  The bug report describes the bug as ``critical'', so it is
-reasonable to assume a programmer that expected all queries to be logged would
-write such an assertion after the bug was reported.
-
-We ran the program and concurrently issued a log rotation command, and an
-insert query.  The bug manifested ({\em i.e.}, the queries went unlogged) and
-our assertion failed, dumping core.  Examining the log's last writer slice in
-GDB allowed examining the last thread to mark the log closed.  We discerned it
-was the log rotation thread that last marked the log closed during log
-rotation.  Knowing how the log ended up marked closed enables reasoning about
-what code to change ({\em e.g.}, added synchronization) to prevent the
-situation in the future.
-
-While a developer is likely to understand the symptom of the bug (queries go
-unlogged) without the information in the \lwt, understanding {\em why} it is marked closed would be
-difficult.  This example illustrates how the \lwt fills in the crucial inter-thread
-dependence that helps understand why the program failed.
-
-\paragraph{Case Study: HTTrack}
-
 \section{Communication Traps}
@@
 \label{sec:eval}
-We evaluated our design along several axes. Recall that in
-Section~\ref{sec:debugging} we illustrated the debugging benefits of last
-writer slices using case studies.   We evaluate our performance overheads to
-demonstrate that last writer slices and \ctraps are feasible for use in
-deployed systems.  We discuss the sources of \ctraps' overhead.   We show that
-\ctraps is useful by implementing two existing applications and analyzing their
-performance.  We characterize our instrumentation and the precision and
-performance impact of our redundancy analysis.
+We evaluated our design along several axes.  We illustrate the debugging
+benefits of last writer slices in Section~\ref{sec:eval:debugging} using case
+studies conducted using real, buggy programs on real.   We evaluate our
+performance overheads to demonstrate that last writer slices and \ctraps are
+feasible for use in deployed systems.  We discuss the sources of \ctraps'
+overhead.   We show that \ctraps is useful by implementing two existing
+applications and analyzing their performance.  We characterize our
+instrumentation and the precision and performance impact of our redundancy
+analysis.
+
+\subsection{Debugging with Last Writer Slices}
+\label{sec:eval:debugging}
+We illustrate the debugging benefits of last writer slices using three case
+studies.
+
+\paragraph{Case Study: HTTrack}
+The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
+and the text in Section~\ref{sec:httdebug} describes the benefit provided by
+last writer slices in debugging an ordering violation in HTTrack-3.43.9.
+
+\paragraph{Case Study: PBZip2}
+We conducted a case study using a buggy version of PBZip2-0.9.1.    The bug is
+a use-after-delete bug that leads to an assertion failure.  A worker thread
+tries to acquire a mutex after it has been deleted by the main thread during
+shutdown, causing the assertion to fail.
+
+Last writer slices provide the key piece of information to debug this failure:
+the last write to the mutex before the worker tried acquiring it was when the
+main thread deleted it.  A developer is likely to begin debugging by checking
+the validity of the mutex.  Upon finding it is invalid ({\em i.e.}, it has
+been deleted), the developer is likely to probe to determine who deleted it.
+Without the information in the \lwt, answering that question is difficult
+because the main thread may have executed arbitrarily far from where it
+deleted the mutex.  With the \lwt, answering that question only requires the
+developer to check the mutex's entry.
+
+\paragraph{Case Study: MySQL}
+Last writer slices were also useful for debugging a logging system failure in
+MySQL-4.0.12.  The bug is an atomicity violation involving two threads.  One
+thread rotates the database's log file.  Log rotation involves closing the log
+file, marking the log closed, opening a new log file, and marking the log open.
+Log rotation should be atomic, but lack of synchronization allows another
+thread to run queries during rotation.  The query code checks if the log is
+marked closed; if it is not closed, the queries are logged, but if it is
+closed, the queries are not logged and the program simply continues.  We added
+an assertion to the non-logging case because this behavior is a security
+vulnerability.  The bug report describes the bug as ``critical'', so it is
+reasonable to assume a programmer that expected all queries to be logged would
+write such an assertion after the bug was reported.
+
+We ran the program and concurrently issued a log rotation command, and an
+insert query.  The bug manifested ({\em i.e.}, the queries went unlogged) and
+our assertion failed, dumping core.  Examining the log's last writer slice in
+GDB allowed examining the last thread to mark the log closed.  We discerned it
+was the log rotation thread that last marked the log closed during log
+rotation.  Knowing how the log ended up marked closed enables reasoning about
+what code to change ({\em e.g.}, added synchronization) to prevent the
+situation in the future.
+
+While a developer is likely to understand the symptom of the bug (queries go
+unlogged) without the information in the \lwt, understanding {\em why} it is marked closed would be
+difficult.  This example illustrates how the \lwt fills in the crucial inter-thread
+dependence that helps understand why the program failed.
 

### 2013-11-03 paper.tex
committing new text in intro, background, and formalism

@@
 threads interact by reading and writing memory locations.  Threads' reads and
-writes interleave arbitrarily and nondeterministically, with each interleaving
-potentially yielding different execution behavior.  The complexity of reasoning
-about thread interactions is a central challenge to building multi-threaded
-software.
-
-When programmers reason incorrectly, they permit some interleavings that should
-be forbidden.  When such interleavings occur, program executions may fail,
-experience performance degradation, or suffer other undesirable behavior.
-Preventing such undesirable behavior is a fundamental challenge faced by
-software engineers.  An important way to understand program behavior is to
-monitor which read and write operations depend on one another.  Memory
-operations are dependent if their order determines the values they read or
-write.  The value of dependence information is made especially clear by its
-prevalent use during manual program debugging and in support of program
-analysis tools.
+writes interleave arbitrarily and nondeterministically and each interleaving
+can yield different behavior.  Reasoning about thread interactions is a central
+challenge to building multi-threaded software because some thread interactions
+have undesirable consequences, like causing a crash or degrading performance.
+
+A common approach to this challenge is to reason about dependent memory
+operations.  Memory operations are dependent if their order determines the
+values they read or write.  The utility of dependence information is made
+especially clear by its prevalent use during manual program debugging and in
+support of program analysis tools.
 
 Debugging a concurrent program requires understanding how threads interact and
-dependence information furnishes that understanding.  Programmers debug
-programs more effectively when they can ask questions about the provenance of
-values during a program's execution~\cite{slicingsurvey,whylineicse}.
-Dependence information helps programmers with the difficult task of identifying
-control- and data-flows that influenced a particular data value at a particular
-point.  In concurrent programs, data provenance information is vital because
-influential control- and data-flows may be in any thread, making the task of
-understanding the origin of a value much more complex.
+dependence information furnishes that understanding.  Dependence information
+helps programmers with the difficult task of identifying control- and
+data-flows that influenced a particular data value at a particular point.
+Programmers debug programs more effectively when they can ask questions about
+the provenance of values during a program's
+execution~\cite{slicingsurvey,whylineicse}.  In concurrent programs, data
+provenance information is particularly valuable because data may originate from
+any of a potentially large number of threads, making reasoning complex.
 
@@
 When a deployed system fails and a programmer receives a bug report,
-reproducing the reported failure to collect dependence information may be
+reproducing the reported failure to collect provenance information may be
 difficult or impossible.  It is essential to collect as much information as
-possible in deployed systems to reduce the need for time-consuming forensic bug
-reproduction.  Unfortunately, prior approaches to tracking dependences in
-concurrent code~\cite{raceslicing,tipslicingsurvey} impose high run time
-overheads.  These overheads are prohibitively high for use in production.
+performance margins permit from deployed systems, to reduce the need for
+time-consuming failure reproduction.  Unfortunately, prior approaches to
+tracking dependences in concurrent code~\cite{raceslicing,tipslicingsurvey}
+impose high run time overheads.  These overheads are prohibitively high for use
+in production.
 
@@
 Dependence information is also an important part of many program analysis
-tools.  Prior work has used dependence information to reduce the burden of
-making reliable, efficient concurrent software.  Bug
-detection~\cite{avio,fasttrack,raceslicing,dmtracker} and
+tools.    Bug detection~\cite{avio,fasttrack,raceslicing,dmtracker} and
 debugging~\cite{tipslicingsurvey,bugaboo,recon,cci,defuse,conseq,falcon} tools
@@
 Record-Replay~\cite{chimera,doubleplay,fdr,rtr} systems track and record the
-order of dependent events to help reproduce failures.
+order of dependent events to help reproduce failures.  These systems would be
+even more valuable, if their underlying dependence analysis was run in
+production.  However, the heavy-weight analyses proposed by many of these
+systems impose overheads too high for production use (5-100x
+slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.
 
@@
 
-These systems would be especially valuable, if run in production systems on
-deployed code.  However, as with dependence tracking on its own, most of these
-systems have overheads too high for use in production (5-100x
-slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.
-
-The key to reducing the overhead of dependence tracking is to carefully select
-and monitor only a {\em subset} of dependences.  We start from two simple
-observations: (1) Reads are frequent so we must minimize tracking overheads on
-reads, and (2) In concurrency analyses, {\em inter-thread} dependences are most
-important.  Based on these observations, we focus on dependences that reflect
-{\em communication} between threads.  Communication occurs when a thread reads
-or overwrites a memory location that was most recently written by a different
-thread; {\em i.e.}, inter-thread read-after-write (RAW) and write-after-write
-(WAW).  Tracking only communication dependences preserves enough information to
-implement useful applications and analyses, but has low enough overhead for
-always-on use in production systems.
-
-Framed by these observations, we demonstrate in this work that a 
-versatile form of inter-thread dependence-tracking analysis has inherent
-overheads that are low enough for use in deployed systems.  There are two parts
-to our work:
+Our goal in this work is to collect, with deployable efficiency, useful
+dependence information that captures data provenance for use in debugging and
+serves as a foundation for more sophisticated program analyses.  We start from
+two simple observations: (1) Reads are frequent so we must minimize tracking
+overheads on reads, and (2) In concurrency analyses, {\em inter-thread}
+dependences are most important.  Based on these observations, we focus first on
+tracking write-after-write dependences (WAWs), which reflect data provenance.
+Building on that, we describe support for tracking {\em communication
+dependences}.  Communication occurs when a thread reads or overwrites a memory
+location that was most recently written by a different thread; {\em i.e.},
+inter-thread read-after-write dependences (RAWs) as well as (WAWs).  In this
+work, we demonstrate that this form of light-weight dependence tracking is
+useful and versatile, and has inherent overheads that are low enough for use in
+deployed systems.  There are two parts to our work:
 
@@
 communication with overhead low enough for production use in many important
-cases.  When threads communicate, \ctraps delivers a software {\em trap} that can be
-handled by a trap handler.  Trap handlers implement {\em \ctraps applications}
-through an extensible interface that can analyze or interpose on the execution
-(Section~\ref{sec:apps}).  By default, \ctraps has very low overhead.  \ctraps
-applications can perform arbitrary dependence analyses, but at their own
-discretion and cost.  As we show in our evaluation, the overhead of \ctraps
-applications scales with the complexity of their analysis.
+cases.  When threads communicate, \ctraps delivers a software {\em trap} that
+can be handled by a trap handler.  Trap handlers implement {\em \ctraps
+applications} through an extensible interface that can analyze or interpose on
+the execution (Section~\ref{sec:apps}).  By default, \ctraps has very low
+overhead.  \ctraps applications can perform arbitrary dependence analyses, but
+at their own discretion and cost.  As we show in our evaluation, the overhead
+of \ctraps applications scales with the complexity of their analysis.
 
@@
 
-\item{We discuss last-writer slices, an abstraction that dynamically tracks
-memory updates, capturing inter-thread dependence information.  We
-design and implement system support for collecting last writer slices.  We
-present two cases studies that show that last writer slices are useful for
-diagnosing real-world bugs.}
-
-\item{Building on last-writer slices, we describe and implement \ctraps, which
-exposes inter-thread communication through an extensible trap interface.
-We also look at an optimization to \ctraps that explores trading off
-precision for performance.}
-
-\item{We demonstrate that \ctraps provides a simple platform for building
+\item{We describe last writer slicing, a dynamic analysis that tracks
+memory updates, capturing data provenance information.  We
+design and implement system support for collecting last writer slices.  }
+
+\item{Building on last writer slicing, we describe and implement \ctraps, which
+exposes inter-thread communication through an extensible trap interface.  }
+
+\item{We evaluate the utility of last writer slices and \ctraps. We first
+present several case studies that show that last writer slices are useful for
+debugging real-world bugs.  We then show how \ctraps is useful for building
 dependence analyses by implementing two analyses from prior
@@
 
-\item{We implement last writer slicing and \ctraps and evaluate their performance using a
-variety of important applications ({\em e.g.}, MySQL, memcached and the PARSEC
-benchmarks~\cite{parsec}).  We show that in many cases our designs have overheads
-low enough for use in deployment (0--15\%).}
+\item{We evaluate our system's performance using a variety of important
+applications ({\em e.g.}, MySQL, memcached and the PARSEC
+benchmarks~\cite{parsec}).  We show that in many cases our designs have
+overheads low enough for use in deployment (0--15\%).}
 
@@
 \includegraphics[scale=.6]{figs/BasicLWT.pdf}
-\caption{\label{fig:basicLWT}Basic operation of the last writer table. }
+\caption{\label{fig:basicLWT}Basic operation of the last writer table. Different shapes signify different distinct memory operations. }
 \end{figure}
@@
 communicating access.  Such accesses are not instrumented by our compiler.  
-
-We rely on {\tt GCC}'s built-in escape analysis to determine if a memory
-operation might lead to communication.  This analysis is a function escape
-analysis.  Locations determined not to escape function scope also cannot escape
-the thread executing the function.  However, using an inter-procedural thread
-escape analysis may eliminate more instrumentation sites.  Investigating better
-instrumentation strategies is orthogonal, but an interesting subject for future
-work.
+We rely on {\tt GCC}'s built-in function escape analysis to determine if a
+memory operation might lead to communication.  Locations determined not to
+escape function scope also cannot escape the thread executing the function.
+%Using an inter-procedural thread escape analysis may eliminate more
+%instrumentation sites.
 
@@
 first develop a formal description of last writer slices.  We consider an
-N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_i = (p^{i}_{1},
-p^{i}_{2}, \ldots, p^{i}_{k_{i}})$ is the sequence of $k_{i}$ dynamic
-instructions executed by thread $i$.  Without loss of generality, we assume all
+N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_t = (p^{t}_{1},
+p^{t}_{2}, \ldots, p^{t}_{k_{t}})$ is the sequence of $k_{t}$ dynamic
+instructions executed by thread $t$.  Without loss of generality, we assume all
 instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
@@
 A point in an execution of $P$ after $e$ instructions have
-executed is defined by $Q = (q_{0}, \ldots, q_{e})$ where each $q \in Q$ is a
-$p^{i}_{j}$ from one of the $k_i$ instructions of thread $i$.  We introduce
+executed is defined by $Q = \{q_{0}, \ldots, q_{e}\}$ where each $q \in Q$ is a
+$p^{t}_{j}$ from one of the $k_t$ instructions of thread $t$.  We introduce
 $LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
@@
 execution.  Correctness in the \lwt requires {\em soundness} and {\em
-completeness}.  A {\em sound} design only updates $LWT[m]$ when $m$ is
-written.  A {\em complete} design always updates $LWT[m]$ when $m$ is written.
-Our compiler inserts \lwt updates before every write
-operation (for completeness) and only before write operations (for soundness).
-Formally, given a program, $P$, our compiler pass transforms it to a new
-program, $P'$.  Each thread in $P'$ executes a modified sequence of
-instructions: $P'_{i} \in P' = \{ I(p^{i}_{1}), p^{i}_{1}, \ldots,
-I(p^{i}_{k_{i}}), p^{i}_{k_{i}} \}$.  $I(p^{t}_{j})$ sets $LWT[m] =
-(t,p^{t}_{j})$ if and only if $p^{t}_{j} = Write(m)$ and is a no-op otherwise.
-
-Thus, our \lwt is correct under the two main assumptions of our formalism: (1)
-there are no data races in the execution; and (2) each memory location
-referenced by a write refers to a unique region of memory that does not overlap
-with another referenced region.
-
-\paragraph{Performance, Atomicity, and Sequential Consistency}
-
-We do not enforce the atomicity of \lwt updates and their corresponding program
-accesses.  This design choice, made in prior work~\cite{fasttrack,recon},
-allows us to omit the synchronization operations that enforce the atomicity of
-these operations.  The benefit of omitting these synchronization operations is
-that overhead of our analysis is reduced. The downside is that our analysis is
-vulnerable to violations of the atomicity of \lwt updates and their
-corresponding accesses. Importantly, however, these violations can only occur
-in programs with data-races and all programs with data-races do not necessarily
-corrupt the \lwt.
-
-We use our \lwt formalism from above to characterize how data-races can
-interfere with \lwt updates.  The relation $p \prec q$ denotes that $p$ happens
-before $q$.  In the absence of data-races, any write by thread $t$ to $m$,
-$p^{t}_{j} = Write(m)$, is ordered by program synchronization with $p^{t'}_{j'}
-= Op(m)$, a later access to $m$ by any other thread $t'$ -- {\em i.e.},
-$p^{t}_{j} \prec p^{t'}_{j'}$.  Our compiler inserts an \lwt update
-$I(p^{t}_{j})$ into $P'$ before its write operation, $p^{t}_{j}$, so
-$I(p^{t}_{j}) \prec p^{t}_{j}$.  Happens-before is transitive, so without
-data-races, $I(p^{t}_{j}) \prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec
-p^{t'}_{j'}$, meaning $I(p^{t}_{j})$ happens before all accesses that
-$p^{t}_{j}$ happens before {\em and before their respective \lwt updates},
-$I(p^{t'}_{j'})$.  Symmetrically, for any access $p^{t''}_{j''}$ in another
-thread $t''$ that precedes $p^{t}_{j}$, $p^{t''}_{j''} \prec p^{t}_{j}$.  Our
-compiler inserts each $I(p^{t}_{j})$ immediately before each $p^{t}_{j}$, so
-$I(p^{t''}_{j''}) \prec p^{t''}_{j''} \prec I(p^{t}_{j}) \prec p^{t}_{j}$,
-happens-before ordering $I(p^{t}_{j})$ after all preceding accesses to $m$ in
-other threads, {\em and after their repective \lwt updates}.  Thus, at all
-points $Q$ in a data-race free execution, given $q_r = I(p^{t}_{j})$, $q_p =
-p^{t}_{j}$, and $q_r \prec q_p$, $\forall_{ \{q_s | q_r \prec q_s \prec q_p
-\}}: q_s \ne Op(m) \wedge q_s \ne I(Op(m)) $.  In other words, \lwt updates and
-their accesses are atomic.
-
-With data-races, \lwt updates and their accesses may not be atomic ({\em i.e.},
-such a $q_s$ can exist).  Instrumentation calls for racy accesses are not
-ordered by happens-before, so they may interleave arbitrarily.  With races, an
-execution could reach a state $Q_{bad} = (\ldots, I(p^{t}_{j}), I(p^{t'}_{j'}),
-p^{t'}_{j'}, p^{t}_{j}, \ldots)$.  In such a state, $LWS[m] = (t,p^{t}_{j})$,
-but $LWT[m] = (t',p^{t'}_{j'})$.  The \lwt was effectively not updated when
-$p^{t}_{j}$ executed (violating completeness), or was effectively updated to
-reflect that $p^{t'}_{j'}$ was $m$'s last writer, when $p^{t}_{j}$ was
-(violating soundness).  This edge case is incorrect. However, we expect it is
-also uncommon in practice, so last writer slices are usually correct, even in
-racy programs.
+completeness}.  A {\em sound} design only updates $LWT[m]$ when $m$ is written.
+A {\em complete} design always updates $LWT[m]$ when $m$ is written.  Our
+compiler inserts \lwt updates before every write operation (for completeness)
+and only before write operations (for soundness).  Formally, given a program,
+$P$, our compiler pass transforms it to a new program, $P'$.  Each thread in
+$P'$ executes a modified sequence of instructions: $P'_{t} \in P' = \{
+I(p^{t}_{1}), p^{t}_{1}, \ldots, I(p^{t}_{k_{t}}), p^{t}_{k_{t}} \}$.
+$I(p^{t}_{j})$ sets $LWT[m] = (t,p^{t}_{j})$ if and only if $p^{t}_{j} =
+Write(m)$ and is a no-op otherwise.  This reasoning shows our \lwt is both
+sound, complete, and thus correct.
+
+Next we discuss two nuances of our design: (1) the interaction of our \lwt
+implementation with code containing data-races; and (2) the correctness of the
+\lwt when a program addresses data that overlap in memory.
+
+\paragraph{Consistency and Atomicity}
+
+Our implementation of the \lwt does not add any synchronization to the program.
+As a result our analysis does not incur the overheads or possible thread
+schedule perturbation synchronization can cause.  However, there are two
+potential issues that stem from this design choice. First, we do not enforce
+memory consistency of meta-data reads and writes in different threads.  Second,
+our implementation does not enforce the atomicity of \lwt updates and their
+corresponding program accesses.  In the absence of data-races our \lwt
+implementation is {\em always correct} and neither of these issues exist.  Both
+atomicity and ordering is enforced by the program synchronization that ensures
+consistency for the data being written by the instructions our system
+instruments.
+
+We use our \lwt formalism to show that without data-races, \lwt updates and
+their corresponding memory accesses are atomic and well-ordered.  The relation
+$p \prec q$ denotes that $p$ happens before $q$.  In correctly synchronized
+code, any write by thread $t$ to $m$, $p^{t}_{j} = Write(m)$, is ordered by
+program synchronization with $p^{t'}_{j'} = Op(m)$, a later access to $m$ by
+any other thread $t'$ -- {\em i.e.}, $p^{t}_{j} \prec p^{t'}_{j'}$.  Our
+compiler inserts an \lwt update $I(p^{t}_{j})$ into $P'$ before its write
+operation, $p^{t}_{j}$. Thus, $I(p^{t}_{j}) \prec p^{t}_{j}$ due to program
+order.  Happens-before is transitive, so without data-races, $I(p^{t}_{j})
+\prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec p^{t'}_{j'}$, meaning $I(p^{t}_{j})$
+happens before all accesses that $p^{t}_{j}$ happens before {\em and before
+their respective \lwt updates}, $I(p^{t'}_{j'})$.  Symmetrically, for any
+access $p^{t''}_{j''}$ in another thread $t''$ that precedes $p^{t}_{j}$,
+$p^{t''}_{j''} \prec p^{t}_{j}$.  Our compiler inserts each $I(p^{t}_{j})$
+immediately before each $p^{t}_{j}$, so $I(p^{t''}_{j''}) \prec p^{t''}_{j''}
+\prec I(p^{t}_{j}) \prec p^{t}_{j}$, happens-before ordering $I(p^{t}_{j})$
+after all preceding accesses to $m$ in other threads, {\em and after their
+repective \lwt updates}.  Thus, at all points $Q$ in a data-race free
+execution, given $q_r = I(p^{t}_{j})$, $q_p = p^{t}_{j}$, and $q_r \prec q_p$,
+$\forall_{ \{q_s | q_r \prec q_s \prec q_p \}}: q_s \ne Op(m) \wedge q_s \ne
+I(Op(m)) $.  In other words, \lwt updates and their accesses are atomic.
+
+In a program with a data-race, there are some accesses, $p^{t}_{j}$ and
+$p^{t'}_{j'}$, that are concurrent.  The \lwt updates for these accesses are
+also concurrent because they are not transitively ordered by the accesses' ordering.  The unordered accesses and \lwt updates can interleave arbitrarily
+because they are concurrent.  Some racy thread schedules may violate the
+atomicity of accesses and their \lwt updates.  Such an execution could reach a
+state $Q_{bad} = (\ldots, I(p^{t}_{j}), I(p^{t'}_{j'}), p^{t'}_{j'}, p^{t}_{j},
+\ldots)$ and in that state, $LWS[m] = (t,p^{t}_{j})$, but $LWT[m] =
+(t',p^{t'}_{j'})$.  The \lwt was effectively not updated when $p^{t}_{j}$
+executed (violating completeness), or was effectively updated to reflect that
+$p^{t'}_{j'}$ was $m$'s last writer, when $p^{t}_{j}$ was (violating
+soundness).
+
+According to the C++11 language specification, in the presence of a data-race,
+a program's behavior is undefined and arbitrary state corruption can occur.  In
+the presence of data-races, {\em any} analysis, including our \lwt analysis,
+running on such a program could experience arbitrary corruption.  In practice
+({\em e.g.}, on x86) these atomicity violations may cause limited \lwt
+corruption, but even this is unlikely.  These atomicity violations only occur
+only when accesses to the same data occur within a very short time interval,
+which is uncommon, {\em and} an unserializable interleaving of accesses and
+\lwt updates also occurs.
+
+{\bf TODO: Figure?}
 

### 2013-11-05 paper.tex
committing some changes to the debugging evaluation section, including two new figures.  also added some new changes to the introduction and the section on last writer slices.

@@
 shared-memory concurrent programs.  Last writer slices dynamically track memory
-updates.  Last writer slices encode the provenance of values in memory, helping
-with debugging.   We build on last writer slices, developing {\em communication
-traps} (\ctraps).  \ctraps uses system support to efficiently monitor and
-interpose on inter-thread memory dependences.  \ctraps provides an extensible
-framework, exposing inter-thread dependences to {\em \ctraps applications}.
-These can implement arbitrary dependence analyses.
-
-We implement last writer slicing and \ctraps.  We show \ctraps is useful by
-implementing two applications from prior work. We show that last writer slices
-help with debugging using case studies of real-world bugs.  We evaluate our
-designs on a set of server programs and standard benchmarks.  Our results show
-\ctraps imposes overheads low enough for use in production systems (0--15\%)
-running important services, like memcached, LevelDB, MySQL, and Apache.
+updates.  Last writer slices reflect the provenance of values in memory,
+information that is useful during debugging.   Using last writer slices,
+we build {\em communication traps} (\ctraps).  \ctraps is a mechanism that
+efficiently monitors and can interpose on memory operations that share data
+between threads.  \ctraps is an extensible framework, exposing
+inter-thread dependences to {\em \ctraps applications} that can implement
+arbitrary dependence analyses.
+
+We implement last writer slicing and \ctraps.  We show that last writer slices
+provide information that is useful for debugging several real-world bugs.  We
+show \ctraps is useful by implementing two concurrency analyses from prior
+work.  We evaluate our designs on a set of server programs and standard
+benchmarks.  Our results show that collecting last writer slices has low run
+time overheads (0--15\%) for many important applications, like memcached,
+LevelDB, MySQL, and Apache.  We show that \ctraps imposes low run time
+overheads by default and that the overheads scale with the complexity of the
+analysis being implemented.
 
@@
 dependence information furnishes that understanding.  Dependence information
-helps programmers with the difficult task of identifying control- and
-data-flows that influenced a particular data value at a particular point.
-Programmers debug programs more effectively when they can ask questions about
-the provenance of values during a program's
-execution~\cite{slicingsurvey,whylineicse}.  In concurrent programs, data
+helps programmers with the difficult task of reasoning about the {\em
+provenance} of a memory location's value -- control- and data-flows that
+influenced that value.  Programmers debug programs more effectively when they
+have access to a value's provenance at points during a program's
+execution~\cite{tipslicingsurvey,whylineicse}.  In concurrent programs, data
 provenance information is particularly valuable because data may originate from
-any of a potentially large number of threads, making reasoning complex.
+any of a potentially large number of threads, making reasoning without explicit
+provenance information complex and difficult.
 
@@
 dependences are most important.  Based on these observations, we focus first on
-tracking write-after-write dependences (WAWs), which reflect data provenance.
-Building on that, we describe support for tracking {\em communication
-dependences}.  Communication occurs when a thread reads or overwrites a memory
-location that was most recently written by a different thread; {\em i.e.},
-inter-thread read-after-write dependences (RAWs) as well as (WAWs).  In this
-work, we demonstrate that this form of light-weight dependence tracking is
-useful and versatile, and has inherent overheads that are low enough for use in
-deployed systems.  There are two parts to our work:
+tracking write-after-write dependences (WAWs), which provide a useful form of
+data provenance information.  Building on that, we describe support for
+tracking {\em communication dependences}.  Communication occurs when a thread
+reads or overwrites a memory location that was most recently written by a
+different thread; {\em i.e.}, inter-thread read-after-write dependences (RAWs)
+as well as WAWs.  In this work, we demonstrate that this form of light-weight
+dependence tracking is useful and versatile, and has inherent overheads that
+are low enough for use in deployed systems.  There are two parts to our work:
 
@@
 program point that last wrote each memory location.     Last writer slices are
-useful because they expose the {\em provenance} of a memory location's contents
+useful because they expose the provenance of a memory location's contents
 during debugging (Section~\ref{sec:debugging}).  When a failure occurs, last
 writer slices are saved with a core dump.  Last writer slices are also
-available during an execution in the debugger.  Note that while prior work
-({\em e.g.}, ~\cite{recon,fasttrack,conmem,conseq,cci}) has relied on {\em ad
-hoc} last-writer analyses, in this work, we isolate the essential last-writer
+available during an execution in the debugger ({\em e.g.}, at breakpoints).
+Note that while prior work ({\em e.g.},
+~\cite{recon,fasttrack,conmem,conseq,cci}) has relied on {\em ad hoc}
+last-writer analyses, in this work, we isolate the essential last-writer
 information in the form of last writer slices and provide an implementation
@@
 region because our \lwt design is {\em granularity oblivious}.  We discuss this
-issue only informally due to space constraints.  Each \lwt entry corresponds to
-a memory location referenced by a program instruction.  Two instructions might
-reference different locations that actually overlap because the referenced
-locations have different granularities.  When that happens, both accesses
-update different \lwt entries, because they reference different locations.
-However, the overlapping region then has two different, inconsistent entries
-reflecting its last {\em two} updates, rather than just its last write.
-
-For example, one instruction might write a word and another might later write
-the second byte of that word.  The word access updates the word's \lwt entry.
-The byte access updates a second \lwt entry for the byte address.  However, the
-byte is part of the previously written word, meaning the \lwt entry for the
-word no longer reflects reality -- the word was last written by the instruction
-and thread that wrote the byte, not the one that wrote the word.  This
-situation leads to incompleteness because the entry for the word should be
-updated but it is not.
-
-Assuming non-overlapping regions ensures \lwt are correct.  Another
-approach would be to assume a fixed \lwt granularity.  We chose our solution
-for two reasons.  First, fixing the \lwt at a large granularity ({\em i.e.},
-cache-line) precludes sound slicing for fine-grained accesses.  Second, fixing
-the \lwt at a small granularity ({\em i.e.}, byte) requires multiple \lwt
-updates for larger granularity accesses, increasing overheads.
+issue informally due to space constraints.  Each \lwt entry corresponds to a
+memory location referenced by a program instruction.  Two instructions might
+reference different locations that overlap because the referenced locations
+have different granularities ({\em e.g.}, accessing a byte that is part of a
+previously accessed word).  When that happens, both accesses update different
+\lwt entries, because they reference different locations.  However, the
+overlapping region then has two different entries reflecting the last {\em two}
+updates to its parts rather than just its last write.
+
+The \lwt remains correct, assuming the program only accesses non-overlapping
+regions.  In programs that access overlapping regions, there is always some
+entry in the \lwt that correctly reflects each memory location's last writer.
+However, the entry for a coarse-grained region may be obsolete because of an
+access to an overlapping sub-region.
+
+One approach to mitigating this issue would be to allocate \lwt entries for
+regions of fixed granularity.  We chose not to do so for two reasons.  First,
+associating \lwt entries with regions of  large granularity only ({\em i.e.},
+cache-line) precludes sound analysis for fine-grained accesses.  Second,
+associating \lwt entries with regions of small granularity ({\em i.e.}, byte)
+requires multiple \lwt updates for larger granularity accesses, increasing
+their overheads proportionately to the granularity of the access.
 
@@
 
-\paragraph{Case Study: HTTrack}
+\paragraph{HTTrack}
 The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
@@
 
-\paragraph{Case Study: PBZip2}
+\paragraph{PBZip2}
 We conducted a case study using a buggy version of PBZip2-0.9.1.    The bug is
@@
 
-\paragraph{Case Study: MySQL}
-Last writer slices were also useful for debugging a logging system failure in
-MySQL-4.0.12.  The bug is an atomicity violation involving two threads.  One
-thread rotates the database's log file.  Log rotation involves closing the log
-file, marking the log closed, opening a new log file, and marking the log open.
-Log rotation should be atomic, but lack of synchronization allows another
-thread to run queries during rotation.  The query code checks if the log is
-marked closed; if it is not closed, the queries are logged, but if it is
-closed, the queries are not logged and the program simply continues.  We added
-an assertion to the non-logging case because this behavior is a security
-vulnerability.  The bug report describes the bug as ``critical'', so it is
-reasonable to assume a programmer that expected all queries to be logged would
-write such an assertion after the bug was reported.
-
-We ran the program and concurrently issued a log rotation command, and an
-insert query.  The bug manifested ({\em i.e.}, the queries went unlogged) and
-our assertion failed, dumping core.  Examining the log's last writer slice in
-GDB allowed examining the last thread to mark the log closed.  We discerned it
-was the log rotation thread that last marked the log closed during log
-rotation.  Knowing how the log ended up marked closed enables reasoning about
-what code to change ({\em e.g.}, added synchronization) to prevent the
-situation in the future.
-
-While a developer is likely to understand the symptom of the bug (queries go
-unlogged) without the information in the \lwt, understanding {\em why} it is marked closed would be
-difficult.  This example illustrates how the \lwt fills in the crucial inter-thread
-dependence that helps understand why the program failed.
-
-\subsection{Setup and Benchmarks}
+\paragraph{MySQL}
+
+We used last writer slices to debug an atomicity violation in MySQL-4.0.12.
+Figure~\ref{fig:mysqllws} shows involved code.  Thread 1 rotates the database's
+log file, marking it closed, opening a new log file, then marking it open
+again.  Log rotation should be atomic, but the lack of synchronization allows
+Thread 2 to run its insert during log rotation.  The insert code checks if the
+log is marked closed. If it is insert is not logged, manifesting the failure.
+We added an {\tt abort} statement to the non-logging case because this behavior
+is a security vulnerability that the bug report describes as ``critical''.
+Adding this {\tt abort} is reasonable, as the bug report describes the
+condition in the {\tt else} block as being an error.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\columnwidth]{figs/MySQLDebug.pdf}
+\caption{\label{fig:mysqllws}{\bf Using last writer slices to debug an
+atomicity violation in MySQL 4.0.12.} The last writer slice reveals the code that closes the log before a database insertion that goes unlogged, which is a security problem.}
+\end{figure}
+
+We triggered the bug by running an insert query during a log rotation.  The bug
+manifested, stopping execution at point {\bf C}.  As described in the bug
+report, the query went unlogged and examining the core dump revealed the log
+was marked closed at point {\bf B}.  However, the core dump did not reveal {\em
+why} the log was marked closed there.  Looking at the last writer slice for the
+{\tt log\_type} variable shows that it was written last Thread 1 at {\bf A}.
+Seeing here that {\tt log\_type} is accessed by multiple threads
+suggests adding synchronization to prevent this atomicity violation.
+
+\paragraph{JDK-StringBuffer}
+
+Last writer slices help elucidate a subtle atomicity violation in the
+StringBuffer library in version 1.4 of the Java JDK.  We use the C++ port of
+the buggy JDK and a driver program code developed in prior concurrency debugging
+work~\cite{concurrencybugs}.
+
+Figure~\ref{fig:jdklws} shows how the bug can manifest as a failure. The string
+object pointed to by {\tt sb} in Thread 1's code contains a string buffer and a
+variable called {\tt count} that represents its length.  The {\tt append()}
+code should atomically read {\tt sb->count} and update the string's buffer.
+The failure occurs when the {\tt append()} code in Thread 1 is interleaved with
+the {\tt erase()} code in Thread 2 as indicated in the figure.  The {\tt
+erase()} code updates the buffer's length, making the version Thread 1 read
+stale.  Thread 1 uses its stale version in the function called at point {\bf
+B}, causing a crash.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\columnwidth]{figs/JDKStringBufferDebug.pdf}
+\caption{\label{fig:jdklws}{\bf Using last writer slices to debug an
+ordering violation in the JDK 1.4 StringBuffer library.} The last writer slice reveals the interaction between the erase function and the append function that leads to a failure.}
+\end{figure}
+
+When the atomicity violation manifests, the appending thread is updating the
+string's underlying buffer.  The erasing thread, however, may have completed
+its erase operation.  The execution context of the erasing thread at the point
+of the failure provides no insight as to why the program failed.  Examining the memory image, the programmer can only conclude that the length and buffer are 
+inconsistent, but not why they are inconsistent.
+
+This situation illustrates the value of the provenenace information furnished
+by last writer slices.  At the point of the failure the last writer slice for
+the length variable points to a decrement of that variable in the erase
+function.  The last writer slice illustrates the need for synchronization
+between the append function and the erase function.  In the absence of
+provenance information, the programmer would not be aware of that relationship.
+
+\subsection{Performance Evaluation: Setup and Benchmarks}
 
@@
 
-\subsection{Performance}
+\subsection{Performance Evaluation}
 \label{sec:eval:perf}

### 2013-11-05 paper.tex
committing before pushing to dbox and printing.

@@
 The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
-and the text in Section~\ref{sec:httdebug} describes the benefit provided by
+and the text in Section~\ref{sec:debugging} describes the benefit provided by
 last writer slices in debugging an ordering violation in HTTrack-3.43.9.
@@
 {\tt log\_type} variable shows that it was written last Thread 1 at {\bf A}.
-Seeing here that {\tt log\_type} is accessed by multiple threads
-suggests adding synchronization to prevent this atomicity violation.
+Seeing that {\tt log\_type} is accessed concurrently in inserts and
+log rotations shows where synchronization needs to be added to fix the bug.
 

### 2013-11-06 paper.tex
lots of edits.  up through the eval.  soon to be a full work-over since the "rebirth"

@@
 
-\ctraps is a platform for analyzing and interposing on inter-thread
-communication events during a concurrent program's execution.  We build our
-support for \ctraps on top of the support for last writer slicing that we
-described in Section~\ref{sec:lastwriterslices}.  Using last writer slice
-information, \ctraps dynamically tracks dependences representing inter-thread
-communication events. \ctraps exposes communication events as {\em traps}
-during an execution that {\em trap handlers} can handle to perform analysis
-and interposition.  We formalize trap delivery and handling in
-Section~\ref{sec:ctsoundness}.
+\ctraps is a framework for analyzing and interposing on inter-thread
+communication events during a program's execution.  We build \ctraps on top of
+last writer slicing that we described in Section~\ref{sec:lastwriterslices}.
+Using last writer slice information, \ctraps dynamically tracks 
+inter-thread communication events. \ctraps exposes communication
+events as {\em traps} during an execution that {\em trap handlers} can handle
+to perform analysis and interposition.  We formalize trap delivery and handling
+in Section~\ref{sec:ctsoundness}.
 
@@
 
-\subsubsection{\ctraps System Support}
-
-Our base \ctraps design delivers traps on all communicating read and write
-operations using compiler and runtime support.
+\subsubsection{Compiler and Runtime Support}
+
+\ctraps delivers software traps before all read and write operations executed
+by one thread that access a memory location that was last written by another
+thread, {\em i.e.}, communicating memory operations.  Traps are simply function
+calls made before such operations execution.  \ctraps implements traps using a
+combination of compiler and runtime library support.
 
 \paragraph{Compiler Support}
-Our compiler instruments all read and write operations that access potentially
-escaping memory locations with a call to a {\em trap hook} function.  Trap hooks
-are passed the address of the memory word involved in the access, the thread
-identifier of the accessing thread, the type of operation being performed, and
-the program point of the access.
+Our compiler inserts a call to the \ctraps {\em trap hook} function immediately
+before all read and write operations that access potentially escaping memory
+locations.  Trap hooks are passed the address of the memory location being
+accessed, the thread identifier of the accessing thread, the type of operation
+being performed, and the program point of the access.
 
 \paragraph{Runtime Support}
-The \ctraps runtime implements the trap hook function.  Trap hooks have two
-purposes: first, to detect communication; second, to deliver traps to \ctraps
-applications.  Trap hooks detect communication by looking up the \lwt entry for
-the memory location being accessed.  If the thread identifier in the \lwt entry is
-different from the accessing thread, then the access is a {\em communicating
-memory access}.  At each communicating memory access, the runtime delivers a
-trap to the accessing thread just before the memory access is allowed to
-proceed.  Traps are delivered by trap hooks to {\em trap handler routines} that
-implement \ctraps applications.  Trap handler routines are like signal
-handlers.  They are registered with the \ctraps runtime when the program
-starts.  Trap handlers may contain arbitrary application-specific code.
-Figure~\ref{fig:hookapi} shows the \ctraps trap handling API.  Trap
-handlers are passed the four pieces of information passed to the trap hook, as
-well as the \lwt's record of the program point and thread identifier of the
-last write to the involved memory location.
+The \ctraps runtime implements the trap hook function,  which has two purposes.
+The trap hook function detects communication and delivers traps to \ctraps
+applications when communication occurs.  The trap hook function detects
+communication by looking up the \lwt entry for the memory location being
+accessed.  If the thread identifier in the \lwt entry is different from the
+accessing thread, then the trap hook function delivers a trap to the accessing
+thread, just before the memory access is allowed to proceed.  To deliver a
+trap, the trap hook function calls {\em trap handler routines} that are defined
+in \ctraps applications.  Trap handler routines are like signal handlers.  They
+are registered with the \ctraps runtime when the program starts.  Trap handlers
+may contain arbitrary application-specific code.  Figure~\ref{fig:hookapi}
+shows the \ctraps trap handling API.  Trap handlers are passed the four pieces
+of information passed to the trap hook, as well as the \lwt's record of the
+program point and thread identifier of the last write to the involved memory
+location.
 
@@
 Given a program $P'$ already instrumented with \lwt updates, the \ctraps
-compiler produces $P''$. A thread $i$ in $P''$ executes a modified instruction
-sequence, $P''_{i} = \{ T(p^{i}_{j}), I(p^{i}_{1}), p^{i}_{1},$ $\ldots,
-T(p^{i}_{j}), I(p^{i}_{k_{i}}), p^{i}_{k_{i}} \}$, where $T(p^{i}_{j})$ is a
+compiler produces $P''$. A thread $t$ in $P''$ executes a modified instruction
+sequence, $P''_{t} = \{ T(p^{t}_{j}), I(p^{t}_{1}), p^{t}_{1},$ $\ldots,
+T(p^{t}_{j}), I(p^{t}_{k_{t}}), p^{t}_{k_{t}} \}$, where $T(p^{t}_{j})$ is a
 trap hook call.  We consider a point $Q = \{q_1, \ldots, q_{e-2}\}$ after $e-2$
-instructions of a sequentially consistent execution have executed and $q_{e} =
-p^{i}_{j}$ is about to execute after its trap hook $T(p^{i}_{j})$ and \lwt
-update $I(p^{i}_{j})$ execute.  The trap hook $T(p^{i}_{j})$ {\em delivers} a
-trap before if $LWT[m] = (t,p^{t}_{\ell w})$ and $i \ne t$.  Informally, the
-trap hook delivers a trap before the instruction if the thread that last wrote
-$m$ is different from the one executing the instruction.  When a trap hook,
-$T(p^{t}_{j})$, delivers a trap, thread $t$ executes an arbitrary trap handler
-$H = \{h^{t}_{1}, \ldots, h^{t}_{h_{n}}\}$.  Thus, after the trap, the \lwt
-update (if $p^{i}_{j} = Write(m)$), and the instruction execute, the execution
-reaches the new point $Q = (q_{1}, \ldots, T(q_{e}), h^{t}_{1}, \ldots,
-h^{t}_{h_{n}}, I(q_{e}), q_{e})$, and continues.
-
-We define what makes a \ctraps design sound and complete, and then discuss the
-correctness of our design.  A design is {\em sound} if it only delivers a trap
-when threads have communicated.  A design is {\em complete} if it always
-delivers a trap when threads have communicated.  The \ctraps compiler puts a
-trap hook on all read and write operations, and only on those operations.
-Assuming a correct \lwt, \ctraps is correct.  The correctness of \ctraps also
-relies on data-race freedom and non-overlapping memory references.  The same
-arguments for \lwt correctness apply to the correctness of trap delivery and
-handling.  Most importantly, data-race freedom implies that trap hooks, \lwt
-updates, and accesses are all atomic.
+instructions of an execution have executed and $q_{e} = p^{t}_{j}$ is about to
+execute after its trap hook $T(p^{t}_{j})$ and \lwt update $I(p^{t}_{j})$
+execute.  The trap hook $T(p^{t}_{j})$ {\em delivers} a trap before if $LWT[m]
+= (t',p^{t'}_{\ell w})$ and $t \ne t'$.  Informally, the trap hook delivers a
+trap before the instruction if the thread that last wrote $m$ is different from
+the one executing the instruction.  When a trap hook, $T(p^{t}_{j})$, delivers
+a trap, thread $t$ executes an arbitrary trap handler $H = \{h^{t}_{1}, \ldots,
+h^{t}_{h_{n}}\}$.  Thus, after the trap, the \lwt update, and the instruction
+execute, the execution reaches the new point $Q = (q_{1}, \ldots, T(q_{e}),
+h^{t}_{1}, \ldots, h^{t}_{h_{n}}, I(q_{e}), q_{e})$, and continues.
+
+A \ctraps implementation must be sound and complete to be correct.  A design is
+{\em sound} if it only delivers a trap when threads have communicated.  A
+design is {\em complete} if it always delivers a trap when threads have
+communicated.  The \ctraps compiler puts a trap hook on all read and write
+operations, and only on those operations.  Assuming a correct \lwt, \ctraps is
+correct.  Our \ctraps implementation also does not insert any synchronization
+into the program, so it comes with the same caveats related to data-race
+freedom as the \lwt.  The same arguments for \lwt correctness apply to the
+correctness of trap delivery and handling.  Most importantly, data-race freedom
+implies that trap hooks, \lwt updates, and accesses are all atomic.
 
@@
 \label{sec:apps}
-\ctraps applications are implemented as shared object plugins ({\em i.e.}, {\tt
-.so} files on Linux) that are loaded by our runtime.  They must implement the
-\ctraps handler API, which requires a trap handler function and permits a
-constructor, destructor,  thread constructor,  thread destructor, which allow
-initialization and disposal of global and thread-local state.  A wide variety
-of \ctraps applications are possible.  Several such applications were outlined
-in Section~\ref{sec:background:comm}.    We implemented two applications, a
-version of a debugging analysis from prior work~\cite{cci} and communication
-graph collection, similar to that in DefUse~\cite{defuse}, Recon~\cite{recon},
-and DMTracker~\cite{dmtracker}.
+\ctraps applications are implemented as shared library plugins that are loaded
+by our runtime.  They must implement the \ctraps trap handler API, which
+requires a trap handler function and permits a constructor, destructor, thread
+constructor, and thread destructor, which allow initialization and disposal of
+global and thread-local state.  Many useful \ctraps applications are possible,
+several of which were outlined in Section~\ref{sec:background:comm}.    To
+demonstrate that \ctraps can implement real analyses with little work, we
+implemented two applications. The first application is a version of a debugging
+analysis from CCI~\cite{cci}.  The second application is a communication graph
+collector, similar to DefUse~\cite{defuse}, Recon~\cite{recon}, and
+DMTracker~\cite{dmtracker}.
 
 \paragraph{CCI-Prev Implementation}
-CCI-Prev is a technique from CCI~\cite{cci} that computes a set of code points
+CCI-Prev is a technique from CCI~\cite{cci} that records a set of code points
 that access a memory location when the previous access to the same location was
-by a different thread.  We implemented a variant of CCI-Prev using \ctraps.  On
-communcating read operations, the runtime updates the LWT like it normally does
-on writes.  Under this policy, any operation that accesses a location that the
-LWT indicates was last accessed by a different thread is recorded.  The set of
-recorded accesses is the output of the analysis.  Our \ctraps implementation of
-CCI-Prev took about 10 lines of code (on top of our base system).
+by a different thread.  We implemented a variant of CCI-Prev using \ctraps that
+records such a set of code points.  In our implementation, communcating read
+operations update the LWT like writes normally do.  Under this policy, any
+operation that accesses a location that was last accessed by a different thread
+is recorded.  The set of recorded accesses is the output of the analysis.  Our
+\ctraps implementation of CCI-Prev took about 10 lines of code (on top of our
+base system).
 
 \paragraph{Communication Graph Implementation}
-Communication graphs are fundamental to several prior debugging
-techniques~\cite{recon, bugaboo, defuse}.  We implement simple communication
-graph collection using \ctraps.  When a trap is delivered, our implementation records
+Communication graphs are the basis of several prior debugging
+techniques~\cite{recon, bugaboo, defuse}.  We implement communication graph
+collection using \ctraps.  When a trap is delivered, our implementation records
 communication graph edges composed of the code point of the last writer to the
 location being accessed and the code point of the trapping access.  The set of
-recorded graph edges is the communication graph that we output.  Our \ctraps
-implementation of communication graphs took about 50 lines of code (including
-output formatting code).
+recorded edges is the tool's output.  Our \ctraps implementation of
+communication graphs took about 50 lines of code (including output formatting
+code).
 
@@
 tools~\cite{cuzz,chess} and bug detection tools~\cite{ctrigger} focused on
-concurrency could also easily implemented as \ctraps applications.  \ctraps
-also provides an interesting platform for developing new approaches to
-record/replay~\cite{chimera,fdr}.  We expect these and other
-applications will become more prevalent with the adoption of \ctraps'
-low-overhead, deployable support for communication tracking.
-
-Additionally, \ctraps affords the ability for {\em users} of systems to add
-functionality to or extend systems.  This facet seems especially promising for
-monitoring or enforcing installation-specific security or performance policies
-without the need to modify applications.  While interesting, we relegate such
-user-built applications to future work.
-
-%Our redundancy analysis applies only to instrumentation on loads,
-%not stores.  Eliminating instrumentation on stores may lead to stale
-%information in the \lwt, which could lead to spurious traps.  Eliminating
-%reads, on the other hand, can only lead to overlooked communication events,
-%never false traps.
-
-%\subsection{Design 3: \lws}
-%
-%Our final design, \lws, varies the compiler support used by \ctrapsfull and
-%\ctrapsmm to simply eliminate instrumentation on all read operations.  With no
-%instrumentation on read operations, \lws has the lowest runtime overhead, but
-%the most imprecise.  However, despite not collecting information about reads,
-%\lws precisely collects last writer slices: all writes are instrumented, so the
-%\lwt is a sound, complete reflection of program state.
-
-%\section{Design Justification}
-
-%We designed \ctraps to satisfy two design goals.
-%
-%First, our design must be efficient enough to run in useful deployment
-%environments without hardware support or sampling.  Second, our design must
-%collect enough information at run time to be useful for debugging and to
-%facilitate building applications.  
-%
-%The constraints conflict because deployment use requires minimizing
-%instrumentation overheads, but collecting information requires potentially
-%high-overhead instrumentation.  Our design was guided by prior work on slicing
-%and concurrency debugging, and our empirical assessment of our system
-%(Section~\ref{sec:eval}).  The main design challenge is deciding which memory
-%dependences to track during the program's execution.  
-%
-%\paragraph{Tracking Most Dependences with High Overhead} 
-%
-%At one extreme is a design that captures all RAW, WAW, and WAR dependence
-%edges.    
-%%Another example is the {\em CCI-Prev} feature 
-%%used by CCI~\cite{cci} tracks inter-thread RAW, WAR, WAW, and RAR dependences (and some
-%%intra-thread ones) with their {\em CCI-Prev} feature. 
-%Such a design collects all dependence information, which is useful for
-%debugging and building analysis applications.  However, such a design falls
-%short on our efficiency goal. Tracking all dependences requires work on every
-%memory operation.  The additional work must update a data structure storing
-%dependences and access a thread-shared structure like the \lwt that indicates
-%which threads and code points are dependent.  Prior
-%work~\cite{stm,coredet,raceslicing} illustrates that the overhead of this
-%design point is frequently too high for use in deployment. 
-%
-%%Tallam, {\em et al} see overheads nearing 1000\% for server
-%%programs ({\em e.g.} Apache, MySQL) with largely independent tasks.  Without
-%%sampling, CCI's overheads as high as 57,000\% for some applications. In most
-%%other cases, overheads are lower, but still too high for deployed use without
-%%sampling.  Note that with sampling, CCI has deployable overheads, but we are
-%%looking for a deployable design that does not require sampling (Constraint 1).
-%
-%\paragraph{Tracking Many Dependences with Moderate Overhead}
-%Short of tracking all dependences, a system may make the decision to track an
-%{\em ad hoc} subset of dependences,  as we discussed in
-%Section~\ref{sec:background:comm}.  Such systems work well at their specific task
-%({\em e.g.}, debugging a certain class of errors).  They collect enough
-%dependence information to be useful for their particular application.  However,
-%their usefulness is limited because tracking dependences specific to one
-%application does not generalize to new applications.  
-%
-%One reason for tracking a subset of dependences is to reduce overheads.  Systems
-%that track {\em ad hoc} dependences tend to have lower overheads than those
-%that track all dependences.
-%%DefUse tracks intra- and inter-thread RAR and RAW dependences; Recon tracks
-%%inter-thread WAW and WAR dependences, and approximates some other dependences.
-%However, these systems mostly fall short of our efficiency goal as well, often having
-%overheads greater than 1000\%~\cite{defuse,recon}. Such overheads are too high for deployed
-%systems.  Some systems avoid such high overheads, but require hardware
-%support~\cite{avio, bugaboo, fdr} or sampling~\cite{cbi,cci} to do so, also
-%failing to satisfy our design goal.
-%
-%\paragraph{CTraps: Tracking Select Dependences with Low Overhead}
-%We make the case that our \ctraps designs satisfies our design goals.
-%
-%The tracking support for \ctraps is built atop last writer slices.  Last writer
-%slices require tracking only inter-thread WAW dependences.  Tracking only these
-%dependences does not require read operations to be instrumented.   Read
-%instrumentation has high overhead because reads are more frequent than writes.
-%\lws is efficient because tracking last writer slices does not need to track
-%any reads.  The overhead of \ctrapsfull and \ctrapsmm is low by default because
-%communication hooks (on reads) are no-ops by default.  \ctraps applications may
-%perform more analysis on reads and their performance overhead scales with the
-%complexity of their analysis.  We argue that such a ``pay-as-you-go'' model
-%that is efficient by default is the right approach, satisfying our efficiency
-%design goal.  We carefully evalute the overheads of our systems in
-%Section~\ref{sec:eval}.
-%
-%
-%Our \ctraps designs also satisfy our usefulness design goal.  \ctraps is useful for
-%implementing analyses, especially those focused on concurrent program behavior.
-%Inter-thread communication dependences show how values modified in one thread
-%flow to operations in another thread.  This systematically selected subset of
-%dependences is general enough for \ctrapsfull and \ctrapsmm to implement
-%dynamic analyses related to inter-thread data flow.  Furthermore, with
-%overheads low enough for deployement, \ctraps applications are useful not just
-%during development, but also at runtime and in production.  \lws is also useful
-%for providing important debugging information absent from current systems.  We
-%describe the debugging benefits of \lws in in Section~\ref{sec:debugging}.  
-%
-%%\ctrapsfull and \ctrapsmm provide the ability to track reads, but they remain efficient.  The key to
-%%their efficiency is the flexibility of read hooks. By default, the hooks are
-%%no-ops.  CTraps {\em applications} can carefully select which dependences to
-%%track, potentially increasing overhead with increases in functionality.  The
-%%baseline performance of both designs is efficient, satisfying Constraint 1.
-%%The flexibility provided by read and write hooks enables a variety of useful
-%%applications (two of which we implement and evaluate), satisfying Constraint 2.
-%
-
-%\section{Formalism and Correctness}
-%\label{sec:soundness}
-
-%We formalize last writer slices and \ctraps.  We then discuss the soundness and
-%completeness properties of our last writer slices and \ctraps designs.
-%We also discuss the impact on soundness and completeness of eliminating
-%redundant read instrumentation in \ctrapsmm.
-
-%\paragraph{Preserving Write Instrumentation}
-%Our system does not eliminate any write instrumentation.  Removing
-%instrumentation on writes may lead to {\em false traps}.  A false trap occurs
-%when a thread executes a memory access, refers to the \lwt, and the \lwt has an
-%out-of-date version of the last writer information.  False traps reflect unsoundness.
-%
-%
-%Figure~\ref{fig:unsoundness} illustrates how removing write instrumentation
-%leads to unsound trap delivery.  The figure shows memory operations in two
-%threads, all of which are assumed instrumented except the triangle marked
-%as uninstrumented.  The actual last thread and instruction to write is shown at
-%the left alongside the state of the LWT.  The figure assumes the behavior of
-%\ctrapsfull, except for the single uninstrumented write operation (all of our
-%designs instrument all writes).  The LWT becomes corrupted when Thread 2
-%performs its uninstrumented write -- the LWT indicates the last writer was
-%Thread 1, but Thread 2's uninstrumented write is the actual last writer.  When
-%Thread 2 executes its next read instruction, it refers to the corrupted LWT,
-%which indicates the read is communicating.  The system delivers a false trap on
-%the read.  This example illustrates the risk of eliminating write
-%instrumentation.  Our designs preserve instrumentation on all writes to avoid
-%this problematic situation.
-%
-%\begin{figure}
-%\centering
-%\includegraphics[width=\columnwidth]{figs/unsoundness.pdf}
-%\caption{\label{fig:unsoundness}Removing write instrumentation leads to unsoundness.  CTraps does not eliminate instrumentation on writes because it is unsound.}
-%\end{figure}
-%
-
-%\section{Extensions to \ctraps}
-%
-%\subsection{Sampling}
-%
-%\subsection{Context-Sensitivity}
+concurrency could also be implemented as \ctraps applications.  \ctraps also
+provides an interesting platform for developing new approaches to
+record/replay~\cite{chimera,fdr}.  \ctraps is an enabling platform for
+developing these or other analyses with low baseline overheads and without the
+need to build an entire concurrency analysis from scratch.
 
@@
 escape analysis support to prune instrumentation to non-escaping memory
-locations.  We used GCC's dominator and post-dominator computations to
-implement our redundancy analysis.  Our compiler instruments calls to {\tt
-free} and {\tt delete} as writes to the pointer being deallocated.  Our
-analysis handles all code compilable using this version of GCC except for a few
-cases: we do not handle accesses that compile to GCC {\tt BITFIELD\_REFERENCE}
-IR types because these are undocumented, we do not handle inlined assembly
-instructions, and we do not handle C++ exceptions.  Note that these limitations
-of our research prototype are not fundamental and the limited cases are
-uncommon.
+locations.  Our compiler instruments calls to {\tt free} and {\tt delete} as
+writes to the pointer being deallocated.  Our analysis handles all code
+compilable using this version of GCC except for a few cases: we do not handle
+accesses that compile to GCC {\tt BITFIELD\_REFERENCE} IR types because these
+are undocumented, we do not handle inlined assembly instructions, and we do not
+handle C++ exceptions.  Note that these limitations of our research prototype
+are not fundamental and the limited cases are uncommon.
 
@@
 packs a program point and a thread identifier into the 64-bit word.  By
-default, program points are instruction addresses, but we also support bounded
-(2-address) context-sensitivity by packing the current instruction's address
-and the nearest return address into an LWT entry. When an access to a memory
-location occurs, we index into the LWT with the lower bits of the location's
-address.  Our prototype implementation uses a lossy resolution policy for hash
-collisions.  We use this policy in our prototype because it is faster and
-simpler than a chaining concurrent hash table implementation.  Collisions are
-extremely rare, so this simplification is unlikely to be a problem even in a
-more industrial strength implementation.
+default, program points are single instruction addresses, but we also support
+limited (2-address) context-sensitivity by packing the current instruction's
+address and the nearest return address into a 64-bit LWT entry. When an access
+to a memory location occurs, we index into the LWT with the lower bits of the
+location's address.  Our prototype implementation uses a lossy resolution
+policy for hash collisions.  We use this policy in our prototype because it is
+faster and simpler than a chaining concurrent hash table implementation.
+Collisions are extremely rare, so this simplification is unlikely to be a
+problem even in a more industrial strength implementation.
+
+We released our implementation, free and open-sourced~\cite{ctrapsonweb}, on
+the web and have had contributions to our codebase from members of the GCC
+development community.
 
@@
 \label{sec:eval}
-We evaluated our design along several axes.  We illustrate the debugging
-benefits of last writer slices in Section~\ref{sec:eval:debugging} using case
-studies conducted using real, buggy programs on real.   We evaluate our
-performance overheads to demonstrate that last writer slices and \ctraps are
-feasible for use in deployed systems.  We discuss the sources of \ctraps'
-overhead.   We show that \ctraps is useful by implementing two existing
-applications and analyzing their performance.  We characterize our
-instrumentation and the precision and performance impact of our redundancy
-analysis.
+We evaluated last writer slices and \ctraps along several axes.  We show that
+last writer slices provide information that is useful during debugging in
+Section~\ref{sec:eval:debugging} using case studies on several buggy programs.
+We evaluate our performance overheads and show that last writer slices and
+\ctraps are feasible for use in deployed systems for a majority of workloads we
+studied.  We then characterize our instrumentation and discuss the sources of
+our system's run time overhead.   We show that \ctraps is useful by
+implementing two existing applications and analyzing their performance.
 
@@
 \label{sec:eval:debugging}
-We illustrate the debugging benefits of last writer slices using three case
-studies.
+We illustrate the debugging benefits of last writer slices by using our system
+to debug several real-world programs studied previously in the debugging
+literature.
 
@@
 
-Last writer slices provide the key piece of information to debug this failure:
-the last write to the mutex before the worker tried acquiring it was when the
-main thread deleted it.  A developer is likely to begin debugging by checking
-the validity of the mutex.  Upon finding it is invalid ({\em i.e.}, it has
-been deleted), the developer is likely to probe to determine who deleted it.
-Without the information in the \lwt, answering that question is difficult
-because the main thread may have executed arbitrarily far from where it
-deleted the mutex.  With the \lwt, answering that question only requires the
-developer to check the mutex's entry.
+Last writer slices provide the key piece of information to debug this failure.
+When the program fails, the core dump shows that the mutex is invalid because
+it has been deleted.  The mutex's last writer slice shows that the last write
+to the mutex before the worker tried to acquire it was when the main thread
+deleted it.  The last writer slice reveals the connection between the the two
+pieces of code involved in the bug: the code that acquires the mutex and the
+code that deletes the mutex.  Note that this connection is not obvious when the
+program crashes because the main thread may have executed arbitrarily far from
+where it deleted the mutex before the worker thread crashes.
 
@@
 
-We used last writer slices to debug an atomicity violation in MySQL-4.0.12.
-Figure~\ref{fig:mysqllws} shows involved code.  Thread 1 rotates the database's
-log file, marking it closed, opening a new log file, then marking it open
-again.  Log rotation should be atomic, but the lack of synchronization allows
-Thread 2 to run its insert during log rotation.  The insert code checks if the
-log is marked closed. If it is insert is not logged, manifesting the failure.
+We used last writer slices to debug an atomicity violation in MySQL-4.0.12 that
+is a critical security vulnerability.  Figure~\ref{fig:mysqllws} shows the
+buggy code.  Thread 1 rotates the database's log file, marking it closed,
+opening a new log file, then marking it open again.  Log rotation should be
+atomic, but the lack of synchronization allows Thread 2 to run its insert
+during log rotation.  The insert code checks if the log is marked closed. If
+the log is marked closed, the insert is not logged, manifesting the failure.
 We added an {\tt abort} statement to the non-logging case because this behavior
@@
 We triggered the bug by running an insert query during a log rotation.  The bug
-manifested, stopping execution at point {\bf C}.  As described in the bug
-report, the query went unlogged and examining the core dump revealed the log
-was marked closed at point {\bf B}.  However, the core dump did not reveal {\em
-why} the log was marked closed there.  Looking at the last writer slice for the
-{\tt log\_type} variable shows that it was written last Thread 1 at {\bf A}.
-Seeing that {\tt log\_type} is accessed concurrently in inserts and
-log rotations shows where synchronization needs to be added to fix the bug.
+manifested, stopping execution at {\bf C}.  As described in the bug report, the
+query went unlogged and examining the core dump revealed the log was marked
+closed at point {\bf B} ({\em e.g.}, {\tt log\_type} was {\tt LOG\_CLOSED}).
+However, the core dump did not reveal {\em why} the log was marked closed
+there.  The last writer slice at {\bf C} for the {\tt log\_type} variable shows
+that it was written last by Thread 1 at {\bf A}.  Without the last writer
+slice, the relationship between the log rotation code and the insert code is
+not apparent. The last writer slice reveals that the log rotation code
+interacts with the insert code incorrectly via the {\tt log\_type} variable.
 
@@
 object pointed to by {\tt sb} in Thread 1's code contains a string buffer and a
-variable called {\tt count} that represents its length.  The {\tt append()}
+variable called {\tt count} that stores its length.  The {\tt append()}
 code should atomically read {\tt sb->count} and update the string's buffer.
@@
 function.  The last writer slice illustrates the need for synchronization
-between the append function and the erase function.  In the absence of
-provenance information, the programmer would not be aware of that relationship.
+between the append function and the erase function.
 
@@
 8-core Xeon E5520 processor with 2-way SMT ({\em i.e.}, 16 execution contexts)
-and 10GB of memory.  We evaluated \ctraps using PARSEC, a set of architectural
-benchmark programs~\cite{parsec} and a collection of real server and desktop
-applications.
+and 10GB of memory.  We evaluated our system using PARSEC, a set of
+architectural benchmark programs~\cite{parsec} and a collection of real server
+and desktop applications.
 
@@
 Apache-httpd is a popular web server.  To benchmark Apache, we used {\tt
-ApacheBench} to request a static {\tt html} page 1,000,000 times using 8 threads,
-measuring Apache's performance to be the throughput reported by {\tt
+ApacheBench} to request a static {\tt html} page 1,000,000 times using 8
+threads, measuring Apache's performance to be the throughput reported by {\tt
 ApacheBench}.
@@
 The main performance result of our work is that our designs impose performance
-overheads that are low enough for use in deployed systems.
-Figure~\ref{fig:perfall} shows the slowdown suffered by our benchmarks due to last writer slicing
-and \ctraps running with ``no-op'' trap hooks ({\em i.e.}, no
-handlers), relative to the natively executing baseline.  These data reflect
-our main performance findings.
+overheads that are low enough for use in deployed systems for many of our
+experimental workloads.  Figure~\ref{fig:perfall} shows the slowdown suffered
+by our benchmarks due to last writer slicing and \ctraps running with ``no-op''
+trap hooks ({\em i.e.}, no handlers), relative to the natively executing
+baseline.  These data reflect our main performance findings.
 
 \paragraph{Last Writer Slices}
-The data show that last writer slicing has very low overheads with a geometric mean of less
-than 10\% across our server programs and less than 50\% across PARSEC.  Such
-low overheads are very likely to be tolerable in deployed systems.  In many
-cases (Apache, MySQL, {\tt dedup}, {\tt canneal}), overheads are negligible.
-In all but two cases ({\tt vips} \& {\tt swaptions}), the overhead of collecting last writer slices is
-less than 100\%.
+The data show that last writer slicing has very low overheads with a geometric
+mean of less than 10\% across our server programs and less than 50\% across
+PARSEC.  Such low overheads are very likely to be tolerable in deployed
+systems.  In many cases (Apache, MySQL, {\tt dedup}, {\tt canneal}), overheads
+are negligible.  In all but two cases ({\tt vips} \& {\tt swaptions}), the
+overhead of collecting last writer slices is less than 100\%.
 

### 2013-11-06 paper.tex
committing more changes in eval.  got rid of redundancy stuff

@@
 \paragraph{\ctraps}
-\ctraps imposes a geometric mean overhead of 14\% for server applications
-and 110\% for PARSEC applications.  In 7 out of 13 of our tests, the overhead
-of \ctraps is less than 50\%.  These seven low overhead benchmarks include
-all of our server programs, as well as {\tt blackscholes}, {\tt dedup} and {\tt
+\ctraps imposes a geometric mean overhead of 14\% for server applications and
+110\% for PARSEC applications.  In 7 out of 13 of our tests, the overhead of
+\ctraps is less than 50\%.  These seven low overhead benchmarks include all of
+our server programs, as well as {\tt blackscholes}, {\tt dedup} and {\tt
 canneal} from PARSEC.  The overhead for these applications is likely to be
@@
 \ctraps overhead that is probably too high for always-on use (averaging 153\%)
-drop to a last writer slicing overhead that is probably low enough for always-on use
-(averaging 38\%).  The large decrease indicates these programs perform
-relatively more reads than other applications, and thus benefit more from
-eliminating read instrumentation.
-%\paragraph{\ctrapsmm}
-%\ctrapsmm has very similar overheads to \ctrapsfull in most cases.   The only case
-%with a major difference was {\tt vips}, indicating it executes more redundant
-%reads than the other applications.  The similarity in the two designs'
-%overheads is expected because our default ``no-op'' read instrumentation does
-%not contribute much to overheads in most cases.  Hence, unless \ctrapsmm culls
-%calls with high dynamic frequency, it is unlikely to reduce overheads much.  We
-%show in Section~\ref{sec:appperf} that \ctrapsmm provides more performance
-%benefit when used to implement applications that make read instrumentation sites
-%more costly.
-\\
-\\
+drop to a last writer slicing overhead that is probably low enough for
+always-on use (averaging 38\%).  The large decrease indicates these programs
+perform relatively more reads than other applications, and thus benefit more
+from eliminating read instrumentation.
+
 These high-level results support our main claim that the runtime overhead of
-\ctraps is low enough for deployed systems for many 
-applications, especially for servers.  We now further analyze our observed overheads, illuminating
+\ctraps is low enough for deployed systems for many applications, especially
+for servers.  We now further analyze our observed overheads, illuminating
 differences between the lower-overhead server programs and the higher-overhead
@@
 \label{sec:eval:conservative}
-{\tt swaptions} had the highest overheads in our tests.  We investigated their
-cause and determined they are mostly artificial and due to shortcomings of our
+{\tt swaptions} had the highest run time overhead in our tests.  We
+investigated the cause and determined it is largely due to shortcomings in our
 escape analysis.  We located the inner-loop function {\tt
@@
 instrumentation.  The program's overhead dropped from about 420\% overhead for
-\ctraps to about 300\% overhead; the change for last writer slicing was from 300\% to about
-130\% overhead.  We also looked at an output matrix passed into the inner-loop
-function.  The matrix is allocated and deallocated in the inner-loop's caller,
-but is never shared.  Eliminating instrumentation on accesses to this matrix
-yielded an overhead of about 46\% for \ctraps, and to just 30\% for
-last writer slicing.
+\ctraps to about 300\% overhead; the change for last writer slicing was from
+300\% to about 130\% overhead.  We also looked at an output matrix passed into
+the inner-loop function.  The matrix is allocated and deallocated in the
+inner-loop's caller, but is never shared.  Eliminating instrumentation on
+accesses to this matrix yielded an overhead of about 46\% for \ctraps, and to
+just 30\% for last writer slicing.
 
 Note that both these changes amount to inlining and preserve the semantics and
-sharing pattern of the program.  We expect that more sophisticated escape
+sharing pattern of the program.  We expect a more sophisticated escape
 analysis would find these local accesses and achieve these performance gains
@@
 In our tests, server programs were better able to tolerate the overhead of
-last writer slicing and \ctraps.   We discern three reasons for this difference.
+last writer slices and \ctraps.   We discern three reasons for this difference.
 
 \paragraph{Independent Parallelism}
-In most cases, servers have more independent parallel work.  Server programs
-set up a pool of threads to handle requests.  Each thread incurs some
-instrumentation latency, but that latency is hidden by other threads making
-progress on independent requests.  This analysis also applies to the PARSEC
-applications that had performance similar to the servers.  For instance, {\tt
-blackscholes} uses largely independent threads to compute on different regions
-of a matrix.  In addition, threads performing independent computation ({\em
-i.e.}, non-communicating computation) are likely to operate mostly on
-non-escaping, local data.  Local data are not instrumented by our compiler,
-sparing the overhead.  We characterize the amount of local data in each
-application in detail in Section~\ref{sec:char}.  We show that the server
+In most cases, the servers we evaluated have abundant, completely independent
+parallel work.  Server programs use a pool of threads to handle requests.
+Each thread incurs some instrumentation latency, but that latency is hidden by
+other threads making progress on independent requests.  This analysis also
+applies to the PARSEC applications that had performance similar to the servers.
+For instance, {\tt blackscholes} uses largely independent threads to compute on
+different regions of a matrix.  In addition, threads performing independent
+computation ({\em i.e.}, non-communicating computation) are likely to operate
+mostly on non-escaping, local data.  Local data are not instrumented by our
+compiler, sparing the overhead.  We characterize the amount of local data in
+each application in detail in Section~\ref{sec:char}.  We show that the server
 programs have a larger proportion of provably local accesses than PARSEC.  The
@@
 a local socket, rather than using the network.  In spite of these controls, I/O
-remains, and hides latency in our server applications.  LevelDB's benchmark is
-designed to avoid all I/O, creating and manipulating a database entirely in a
-single process.  The absence of I/O may account for some of the difference in
-overhead between LevelDB and the other servers.  Note that the effects of I/O
-are not merely experimental noise -- in real deployed software the latency
-hiding effects of I/O are likely to be even {\em more} pronounced.
+remains, and hides added instrumentation latency in our server applications.
+LevelDB's benchmark is designed to avoid all I/O, creating and manipulating an
+in-memory database in a single process.  The absence of I/O may account for some
+of the difference in overhead between LevelDB and the other servers.  Note that
+the effects of I/O are not merely experimental noise -- in real deployed
+software the latency hiding effects of I/O are likely to be even {\em more}
+pronounced, favoring the use of our system.
 
 \paragraph{Interference with Optimization}
-The PARSEC programs are more heavily hand-optimized. In several cases
+The PARSEC programs are heavily hand-optimized. In several cases
 resorting to hand-coded assembly and cache-aware memory access patterns.
@@
 novice programmers.  Depending on the maturity and level of verification of a
-piece of carefully hand-optimized code, it might make sense to elide all
-communication tracking instrumentation to preserve performance.
-
-%\subsubsection{Breaking Down the Design Space}
-%Figure~\ref{fig:perfall} also shows the impact of context-sensitivity, lossy
-%optimization, and sampling on our overheads.  
-%
-%\paragraph{Context Sensitivity}
-%Context-sensitivity imposes a substantial overhead in some cases.  The
-%additional cost of context-sensitivity is about 10\% of the total overhead for
-%PARSEC.  For server applications, the cost is less substantial, but still a
-%non-negligible 4\% over the baseline.  The cost of context-sensitivity comes
-%from the need to compute and store multiple addresses, rather than a single
-%program counter.  
-%
-%In the context insensitive case, updating the \lwt requires a single register
-%read ({\em rip}) a bitwise {\tt or} to combine the thread identifier and the
-%instruction pointer, and the actual write to update the \lwt entry.  In the
-%context sensitive case, updating the \lwt requires looking back over the call
-%stack, and a bitwise {\tt or} operation at each return address.  The cost is at
-%least a multiple of the cost of the constext insensitive case.  However, to
-%restrict overhead to write instrumentation calls, we walk stack frames, rather
-%than iteratively computing call stacks at each call and return.  As a result,
-%the cost is amplified by the need to walk the stack to find return addresses.
-%Future work could reduce this cost using Breadcrumbs~\cite{breadcrumbs} or
-%Probabilistic Calling Contexts~\cite{probcallcont}. \addtodo{Use the Aviso
-%iterative context tracker -- maybe it's actually faster?}
-
-%\paragraph{Redundant Read Optimization}
-%
-%Our static analysis to remove read instrumentation reduces performance
-%overheads for PARSEC programs, but does not provide much benefit in server
-%programs.  In particular, removing redundant read instrumentation benefits {\tt
-%x264} and {\tt swaptions}.  In the case of {\tt swaptions}, the gains are due
-%to a \addtodo{Read Swaptions Code -- where's the beef?}.  In the case of {\tt
-%x264}, the improvement is due to \addtodo{What is it?}.
-%
-%Note that the experiments depticted in Figure~\ref{fig:perfall} reflect our
-%runtime in its base configuration -- writes update the \lwt, and reads only
-%call a {\em no-op} hook.  Our base configuration targets manual debugging,
-%which does not require actions on reads.  In more sophisticated applications of
-%\ctraps, such as communication graph collection, action is required on reads.
-%In Section~\ref{sec:graphperf}, we show that this optimization more
-%substantially improves performance when read instrumentation sites do more
-%work.
-%
-%\paragraph{Sampling}
-%Sampling reduces overheads for both server and PARSEC applications.
-%\addtodo{Explain why sampling is awesome sometimes and less awesome other
-%times}
-
-%\paragraph{Annotating Away Instrumentation}
-%\addtodo{Discuss the /*Sequential*/ region in streamcluster and how not instrumenting those loads and stores might increase perf a lot}
-
-%blackscholes 2 4
-%canneal 124 226
-%dedup 1 171
-%ferret 40 792
-%fluidanimate 3 149
-%streamcluster 22 47
-%swaptions 2 19
-%vips 1919 9116
-%x264 2987 5960
-
-%blackscholes 3.77358490566038 7.54716981132075
-%canneal 15.7360406091371 28.6802030456853
-%dedup 0.128865979381443 22.0360824742268
-%ferret 1.0584810796507 20.9314633500926
-%fluidanimate 0.398406374501992 19.6547144754316
-%streamcluster 5.44554455445545 11.6336633663366
-%swaptions 0.87719298245614 8.33333333333333
+piece of carefully hand-optimized code, it might make sense to elide 
+last writer slicing and \ctraps code to preserve performance.
 
 \subsection{Context-Sensitivity}
+{\bf TODO: Was this LWS or CTRAPS-NOOP or what?}
 We briefly evaluated the cost of context-sensitivity for \ctraps.  With a
@@
 restricted form of context-sensitive analysis is possible in \ctraps with
-overheads low enough for production.  This possibility broadens the scope of
-applicability for \ctraps.  We omit a full analysis of these results here due
-to space constraints.
+overheads low enough for production.
 
@@
 notably, in tests with our server benchmarks -- their overhead is low enough
-for production use.  We also show that our redundancy analysis saves some
-performance overhead in some cases.
+for production use.
 
@@
 \label{sec:char}
-We characterized the instrumentation pruned by escape and redundancy analysis.
-We report our results in Table~\ref{tab:char}.  Each column shows the percent
-of instrumentation sites pruned and the absolute number of sites in
-parenthesis.  The first column shows the amount of static instrumentation
-pruned by escape analysis.  The second column shows the amount of static
-instrumentation pruned by our redundancy analysis.  The third column shows how
-many dynamic instances of instrumentation were eliminated by redundancy analysis.  To
-collect the data in the third column, we implemented a special instrumentation
-pass that marked pruned functions statically so they could be counted
-separately from non-pruned functions at runtime.
+We characterized the instrumentation pruned by escape analysis.  We report our
+results in Table~\ref{tab:char}.  The table shows the percent of
+instrumentation sites pruned and the absolute number of sites in parenthesis.
 
@@
 instrumentation using escape analysis is key to high performance.
-
-The data also show that fewer accesses were eliminated by redundancy analysis
-than by escape analysis.  In some cases ({\em e.g.}, {\tt swaptions}), the
-fraction of static and dynamic instrumentation sites eliminated is small.  In
-these cases, the few pruned calls were infrequently executed and as
-Figure~\ref{fig:perfapps} shows, there was little reduction in overhead.  In
-other cases, like {\tt x264}, the fraction of calls pruned was substantial 
-and yielded the largest performance boost.
 

### 2013-11-06 paper.tex
committing this little daddy after a onceover and a new conclusion.

@@
 
-\section{System Implementation and Applications}
+\section{System Implementation}
 
@@
 
-We released our implementation, free and open-sourced~\cite{ctrapsonweb}, on
-the web and have had contributions to our codebase from members of the GCC
-development community.
+\subsection{Release}
+We released our implementation, free and open-sourced, on the
+web~\cite{withheldforanonymity} and published it to the GCC plugin repository.
+Since its release, we have seen contributions to our codebase from members of
+the GCC development community and downloads from interested developers.
 
@@
 
-%\begin{table}
-%\begin{tabular}{ l | l | c | l }
-%& & & \\ \hline
-%\multirow{9}{*}{\begin{sideways}PARSEC\end{sideways}} & blackscholes & & \\
-%&dedup & & \\
-%&canneal & & \\
-%&x264& & \\
-%&vips& & \\
-%&ferret& & \\
-%&fluidanimate& & \\
-%&swaptions & & \\
-%&streamcluster & & \\ \hline
-%\multirow{4}{*}{\begin{sideways}Servers\end{sideways}}&MySQL & & \\
-%&Apache & & \\
-%&LevelDB & & \\
-%&Memcached & & \\ \hline
-%\multirow{3}{*}{\begin{sideways}App\end{sideways}}&PBZip2 & & \\
-%&AGet & & \\
-%&Transmission & & \\
-%
-%\end{tabular}
-%\caption{\label{tab:benchmarks}\addtodo{build this table, show LOC, \#threads, sharing pattern(?), etc}}
-%\end{table}
-
 \subsection{Performance Evaluation}
@@
 trap hooks ({\em i.e.}, no handlers), relative to the natively executing
-baseline.  These data reflect our main performance findings.
+baseline.  In these experiments, we consider a run time overhead of 10\% or
+less to be ideal for production use and a run time overhead of 50\% as
+a reasonable upper bound on acceptable overhead for production use.
 
@@
 mean of less than 10\% across our server programs and less than 50\% across
-PARSEC.  Such low overheads are very likely to be tolerable in deployed
-systems.  In many cases (Apache, MySQL, {\tt dedup}, {\tt canneal}), overheads
-are negligible.  In all but two cases ({\tt vips} \& {\tt swaptions}), the
-overhead of collecting last writer slices is less than 100\%.
+PARSEC.  Such low overheads are likely to be acceptable in production.  In many
+cases (Apache, MySQL, {\tt dedup}, {\tt canneal}), overheads are negligible.
+In all but two cases ({\tt vips} \& {\tt swaptions}), the overhead of
+collecting last writer slices is less than 100\%.
 
@@
 dedup}, {\tt streamcluster}, {\tt fluidanimate}, {\tt ferret}) that have
-\ctraps overhead that is probably too high for always-on use (averaging 153\%)
-drop to a last writer slicing overhead that is probably low enough for
-always-on use (averaging 38\%).  The large decrease indicates these programs
-perform relatively more reads than other applications, and thus benefit more
-from eliminating read instrumentation.
+\ctraps overhead that is probably too high for production use (averaging
+153\%).  By comparison, those applications have a last writer slicing overhead
+that is likely acceptable for production use (averaging 38\%).  The difference
+indicates that these programs perform relatively more reads than other
+applications, experiencing more overhead due to \ctraps's read instrumentation.
 
@@
 are allocated through an external call.  Our escape analysis cannot determine
-that the call only does allocation and that the matrices are local, so it
-conservatively assumes the matrices escape.  We manually inlined the allocation
-(and a matching deallocation), and escape analysis eliminated their
+that the matrices are only used locally in that function, so it conservatively
+assumes the matrices escape.  We manually inlined the matrix allocation calls
+(and a matching deallocation calls), and escape analysis eliminated their
 instrumentation.  The program's overhead dropped from about 420\% overhead for
@@
 
-Note that both these changes amount to inlining and preserve the semantics and
-sharing pattern of the program.  We expect a more sophisticated escape
-analysis would find these local accesses and achieve these performance gains
-automatically.
+These manual improvements to the escape analysis reduced the run time overheads
+for these workloads from being unacceptable for any production environment to
+being reasonable for use in production systems.  We expect a more sophisticated
+escape analysis would find these local accesses and achieve these performance
+gains automatically.
 
@@
 \subsubsection{Performance with \ctraps}
-Figure~\ref{fig:perfapps} shows the performance overheads of our implementation
-as the slowdown with respect to the natively executing baseline.  The overheads
-vary substantially across applications.  On average, the overheads are higher
-than for our baseline implementation (Section~\ref{sec:eval:perf}).
-
-In six cases (all servers, {\tt blackscholes}, and {\tt dedup}), the overheads
+
+ Figure~\ref{fig:perfapps} shows the performance overheads of
+our implementation as the slowdown with respect to the natively executing
+baseline.  The overheads vary substantially across applications and the average
+run time overheads are higher than for our baseline implementation
+(Section~\ref{sec:eval:perf}).
+
+In six cases (the four servers, {\tt blackscholes}, and {\tt dedup}), the overheads
 for CCI-Prev implementations are tolerable for deployment use (around 50\%).
@@
 
+{\bf TODO: Compare with these analyses implemented using binary instrumentation}
+
 %In two of these cases, {\tt
@@
 \begin{table}
-%\renewcommand{\tabcolsep}{2pt}
 \centering
@@
 difference is likely due to the fact that server threads perform more
-independent work (as discussed in Section~\ref{sec:eval:parsecserver}).  Eliminating
-instrumentation using escape analysis is key to high performance.
+independent work (as discussed in Section~\ref{sec:eval:parsecserver}).
+Eliminating instrumentation using escape analysis is key to high performance.
+
+{\bf TODO: Run without escape analysis?}
 
@@
 
-There are several categories of prior work related to last writer slices and \ctraps.
+There are several categories of prior work related to last writer slices and
+\ctraps.
 
 \paragraph{Program Slicing}
-Slicing techniques~\cite{tipslicingsurvey} are techniques for
-identifying useful subsequences of a program's instructions.   Slicing can work
-by analyzing program code statically or analyzing execution traces dynamically.
-Many slicing techniques discussed in ~\cite{tipslicingsurvey} monitor control and
-data dependences to identify which parts of a program might be relevant to a
-task ({\em e.g., debugging}).
-
-Thin Slicing~\cite{thinslicing} is related in its approach. Thin
+Slicing techniques~\cite{tipslicingsurvey} are techniques for identifying
+useful subsequences of a program's instructions.   Slicing can work by
+analyzing program code statically or analyzing execution traces dynamically.
+Many slicing techniques discussed in ~\cite{tipslicingsurvey} monitor control
+and data dependences to identify which parts of a program might be relevant to
+a task ({\em e.g., debugging}).
+
+Thin Slicing~\cite{thinslicing} is especially related to this work. Thin
 Slicing monitors and reports a select subset of dependent operations from an
-execution.  The goal is to eliminate some operations that would be included in a traditional
-slice to make the slice simpler and less costly to collect.  While this work and
-Thin Slicing differ in what they collect and how, our motivation to monitor a
-select subset of dependences was similar to theirs.    Prior work
-on slicing is unlike our work in that little has focused on monitoring
-inter-thread dependences in production code without sampling or hardware support.
+execution.  The goal is to eliminate some operations that would be included in
+a traditional slice to make the slice simpler and less costly to collect.
+While this work and Thin Slicing differ in what they collect and how, our
+motivation to monitor a select subset of dependences was similar to theirs.
+
+%is unlike our work in that our work focused on
+%monitoring inter-thread dependences in production code without sampling or
+%hardware support.
+
+\paragraph{Debugging with Data Provenance Information}
+
+Prior work has looked at using data provenance information in
+debugging~\cite{whylineicse, whylinechi}.  The WhyLine debugger allowed
+programmers to ask questions about why particular variables held particular
+values during debugging.  WhyLine answered those questions by presenting data-
+and control-flows that influenced those variables.  WhyLine showed that
+provenance information is intuitive to programmers and makes debugging easier,
+justifying our focus on provenance information in our last writer slicing
+design.
+
+There are several characteristics of this work that distinguish it from prior
+work on debugging with data provenance information.  Unlike prior work, our
+work focuses on isolating a limited subset of dependence information that we
+can collect with run time overheads low enough for production use.
+Furthermore, we focus on {\em concurrent} program debugging.  Prior work on
+provenance debugging that focused mainly on single threaded programs and had
+dependence slicing overheads are probably too high for production.
 
@@
 DefUse~\cite{defuse} all explicitly track inter-thread memory dependences at
-runtime.  Unlike our work, these techniques were not designed to run in
-production systems~\cite{recon,defuse} without sampling~\cite{cci} or hardware
-support~\cite{bugaboo}.
-
-Some other techniques~\cite{threadclustering,threadcriticality} have relied on
-hardware debugging registers to monitor dependences efficiently enough for
-production use.  These techniques are similar to our work in that they track how
-threads interact ({\em i.e.}, share).  These techniques are different in that
-their approach to dependence tracking is specific to the problem they address;
-our system provides general support.
+runtime.  Unlike our work, several of these techniques were not designed to run
+in production systems~\cite{recon,defuse}.  A distinguishing characteristic of
+our work is that it is effective and fast enough for production without the
+need to aggregate information from multiple different sampled executions, like
+CCI~\cite{cci}.  Additionally, our system does not require invasive hardware
+changes like Bugaboo~\cite{bugaboo} or LBA~\cite{paralog}.
+
+Other techniques~\cite{threadclustering,threadcriticality} use  hardware debug
+registers to monitor dependences efficiently enough for production use.  These
+techniques are similar to our work in that they track how threads interact.
+These techniques are different from our work in that they track information
+specific to the performance and our system provides general support.
 
@@
 support to instrument arbitrary code -- instrumenting a program to track
-communciation is possible in such systems.  These systems are similar to
-\ctraps in that they aim to enable dynamic analyses and runtime systems.  These
-are different from \ctraps because they rely on dynamic rewriting, yielding
-overheads too high for deployment.  \ctraps provides extensible support for a
-restricted set of instrumentation tasks ({\em i.e.}, tracking communication)
-with overheads low enough for production.  Such general systems also increase
-the burden on implementors: to monitor communication, they must manually track
-dependences, crafting delicate, performance sensitive code.  \ctraps tracks
-only those dependences corresponding to communcation.  Furthermore, 
-instrumentation systems lack the debugging benefits afforded by last writer slices.
+communciation is possible in such systems.
+
+These systems are similar to our work in that they aim to enable dynamic
+analyses and runtime systems.  They are different from our work because they
+rely on dynamic rewriting, yielding overheads often too high for deployment.
+Instead, we trade generality for performance. Last writer slices and \ctraps
+provide extensible support in the restricted domain of tracking communication
+with overheads low enough for production.  Another difference is that our work
+handles some of the delicate, performance sensitive code required to track
+communication.  In general instrumentation systems, programmers write that code
+from scratch at their own risk.
 
 \section{Conclusions}
-This work stems from the observation that monitoring inter-thread communication
-is often fast enough for use in deployed systems.  Based on this premise, we
-develop system support for efficiently collecting last writer slices.  Using
-that support, we developed \ctraps, a system for extensibly tracking
-communicating inter-thread dependences.  \ctraps works efficiently at runtime,
-in production, and without sampling or hardware support.  We showed that
-\ctraps exposes a sound and complete (assuming race freedom) communication
-tracking interface, enabling trap handlers to implement applications in just a
-few lines of code.  We showed that a useful class of analyses can be
-implemented using \ctraps that are efficient enough for deployment use in an
-important set of programs ({\em e.g.}, servers, key-value stores, databases,
-{\em etc.}).  We showed that last writer slices provide information that is
-helpful for understanding concurrency errors with nearly no overhead.  Our
-future work includes studying the generality of possible \ctraps applications
-and the role of minimal hardware support for \ctraps.  We will also further study the
-impact of provenance information on debugging in both sequential and concurrent
+This work stems observes that we can collect a useful form of data provenance
+information with overheads low enough for use in deployed systems.
+
+We developed system support that collects last writer slices with overheads
+well within an acceptable range for production.  We showed that, during
+debugging, the provenance information provided by last writer slices reveals
+crucial connections between different threads' otherwise disparate regions of
+code.  Understanding how code in different threads interacts is essential to
+concurrency debugging and this work helps with that understanding.
+
+Using last writer slices, we then built \ctraps, an extensible framework for
+implementing concurrent program analyses that interpose on inter-thread
+communication.  Trap handlers implement such analyses in just a few lines of
+code.  We showed that it is possible to build useful analyses using \ctraps
+that are efficient enough for deployment use in an important set of programs
+({\em e.g.}, servers, key-value stores, databases, {\em etc.}).
+
+An interesting direction for future work is to evaluate the role of minimal
+hardware support for \ctraps.  We will also further study the impact of
+provenance information on debugging in both sequential and concurrent
 applications.

### 2013-11-07 paper.tex
committing?
committing before changing networks

@@
 
-Data provenance information is directly useful during program debugging.
-Work on debugging~\cite{whylinechi,whylineicse} in the HCI and Software
-Engineering community shows that when programmers debug programs, they benefit
-from the ability to directly ask such provenance (``why'') questions of their
-debugger.
+Data provenance information is directly useful during program debugging.  Work
+on debugging~\cite{whylinechi,whylineicse} in the HCI and Software Engineering
+community shows that when programmers debug programs, they benefit from the
+ability to directly ask such provenance (``why'') questions of their debugger.
+Note that that like prior work on debugging with provenance, in this work we
+focus on {\em debugging} -- understanding the root cause of a failure -- rather than {\em bug detection} -- finding new, unknown bugs.
 
@@
 dependences.  For example, DefUse~\cite{defuse} tracks inter-thread WAR
-dependences, but not WAW dependences.  Recon tracks some inter-thread WAW and
+dependences, but not WAW dependences.  Recon~\cite{recon} tracks some inter-thread WAW and
 RAW dependences and approximates WAR dependences.  These techniques are useful
@@
 
-{\bf TODO: Figure?}
+%{\bf TODO: Figure?}
 
@@
 calls made before such operations execution.  \ctraps implements traps using a
-combination of compiler and runtime library support.
+combination of compiler and runtime library support.  Note that we use the name
+``trap'' for the interposition points in \ctraps, but we implement traps using 
+function calls only, as opposed to signals or some other mechanism.
 
@@
 
-We built our runtime system from scratch in about 1000 lines of C and C++ code.
-We implemented the LWT as a fixed size 2GB array of 64-bit words.  Each entry
-packs a program point and a thread identifier into the 64-bit word.  By
-default, program points are single instruction addresses, but we also support
-limited (2-address) context-sensitivity by packing the current instruction's
-address and the nearest return address into a 64-bit LWT entry. When an access
-to a memory location occurs, we index into the LWT with the lower bits of the
-location's address.  Our prototype implementation uses a lossy resolution
-policy for hash collisions.  We use this policy in our prototype because it is
-faster and simpler than a chaining concurrent hash table implementation.
-Collisions are extremely rare, so this simplification is unlikely to be a
-problem even in a more industrial strength implementation.
+We built our runtime system from scratch.  We implemented the LWT as a fixed
+size 2GB array of 64-bit words, fixing our memory overheads.  Each entry packs a program point and a thread
+identifier into the 64-bit word.  By default, program points are single
+instruction addresses, but we also support limited (2-address)
+context-sensitivity by packing the current instruction's address and the
+nearest return address into a 64-bit LWT entry. When an access to a memory
+location occurs, we index into the LWT with the lower bits of the location's
+address.  Our prototype implementation uses a lossy resolution policy for hash
+collisions.  We use this policy in our prototype because it is faster and
+simpler than a chaining concurrent hash table implementation.  Collisions are
+extremely rare, so this simplification is unlikely to be a problem even in a
+more industrial strength implementation.
 
@@
 \subsection{Context-Sensitivity}
-{\bf TODO: Was this LWS or CTRAPS-NOOP or what?}
-We briefly evaluated the cost of context-sensitivity for \ctraps.  With a
-bounded 2-address context, the average overhead for server programs was 18\%
-and the average overhead for PARSEC was 145\%.  These data show that a
-restricted form of context-sensitive analysis is possible in \ctraps with
-overheads low enough for production.
+%{\bf TODO: Was this LWS or CTRAPS-NOOP or what?}
+We briefly evaluated the cost of context-sensitivity using \ctraps with no-op
+trap handlers.  With a bounded 2-address context, the average overhead for
+server programs was 18\% and the average overhead for PARSEC was 145\%.  These
+data show that a restricted form of context-sensitive analysis is possible in
+\ctraps with overheads low enough for production.
 
@@
 
+Bad Apple Tracking~\cite{bondapple} is also a form of debugging with data
+provenance information. This work records the {\em origin} of an unusable value
+and can report it to a programmer when it is used in a way that causes a
+failure.
+
 \paragraph{Communication Tracking in Concurrent Programs}
@@
 \section{Conclusions}
-This work stems observes that we can collect a useful form of data provenance
+This work observes that we can collect a useful form of data provenance
 information with overheads low enough for use in deployed systems.

### 2013-11-07 paper.tex
committing an update that includes the debugging of apache-2.0.48 memcache bug (21285)

@@
 
-\subsection{Formalism and Correctness}
-\label{sec:ctsoundness}
-To discuss the correctness of \ctraps, we first formalize trap delivery by
-building on our last writer slicing formalism from
-Section~\ref{sec:lastwriterslices}.
-
-Given a program $P'$ already instrumented with \lwt updates, the \ctraps
-compiler produces $P''$. A thread $t$ in $P''$ executes a modified instruction
-sequence, $P''_{t} = \{ T(p^{t}_{j}), I(p^{t}_{1}), p^{t}_{1},$ $\ldots,
-T(p^{t}_{j}), I(p^{t}_{k_{t}}), p^{t}_{k_{t}} \}$, where $T(p^{t}_{j})$ is a
-trap hook call.  We consider a point $Q = \{q_1, \ldots, q_{e-2}\}$ after $e-2$
-instructions of an execution have executed and $q_{e} = p^{t}_{j}$ is about to
-execute after its trap hook $T(p^{t}_{j})$ and \lwt update $I(p^{t}_{j})$
-execute.  The trap hook $T(p^{t}_{j})$ {\em delivers} a trap before if $LWT[m]
-= (t',p^{t'}_{\ell w})$ and $t \ne t'$.  Informally, the trap hook delivers a
-trap before the instruction if the thread that last wrote $m$ is different from
-the one executing the instruction.  When a trap hook, $T(p^{t}_{j})$, delivers
-a trap, thread $t$ executes an arbitrary trap handler $H = \{h^{t}_{1}, \ldots,
-h^{t}_{h_{n}}\}$.  Thus, after the trap, the \lwt update, and the instruction
-execute, the execution reaches the new point $Q = (q_{1}, \ldots, T(q_{e}),
-h^{t}_{1}, \ldots, h^{t}_{h_{n}}, I(q_{e}), q_{e})$, and continues.
-
-A \ctraps implementation must be sound and complete to be correct.  A design is
-{\em sound} if it only delivers a trap when threads have communicated.  A
-design is {\em complete} if it always delivers a trap when threads have
-communicated.  The \ctraps compiler puts a trap hook on all read and write
-operations, and only on those operations.  Assuming a correct \lwt, \ctraps is
-correct.  Our \ctraps implementation also does not insert any synchronization
-into the program, so it comes with the same caveats related to data-race
-freedom as the \lwt.  The same arguments for \lwt correctness apply to the
-correctness of trap delivery and handling.  Most importantly, data-race freedom
-implies that trap hooks, \lwt updates, and accesses are all atomic.
+%MAYBE PUT THIS BACK IN
+%\subsection{Formalism and Correctness}
+%\label{sec:ctsoundness}
+%To discuss the correctness of \ctraps, we first formalize trap delivery by
+%building on our last writer slicing formalism from
+%Section~\ref{sec:lastwriterslices}.  
+%
+%Given a program $P'$ already instrumented with \lwt updates, the \ctraps
+%compiler produces $P''$. A thread $t$ in $P''$ executes a modified instruction
+%sequence, $P''_{t} = \{ T(p^{t}_{j}), I(p^{t}_{1}), p^{t}_{1},$ $\ldots,
+%T(p^{t}_{j}), I(p^{t}_{k_{t}}), p^{t}_{k_{t}} \}$, where $T(p^{t}_{j})$ is a
+%trap hook call.  We consider a point $Q = \{q_1, \ldots, q_{e-2}\}$ after $e-2$
+%instructions of an execution have executed and $q_{e} = p^{t}_{j}$ is about to
+%execute after its trap hook $T(p^{t}_{j})$ and \lwt update $I(p^{t}_{j})$
+%execute.  The trap hook $T(p^{t}_{j})$ {\em delivers} a trap before if $LWT[m]
+%= (t',p^{t'}_{\ell w})$ and $t \ne t'$.  Informally, the trap hook delivers a
+%trap before the instruction if the thread that last wrote $m$ is different from
+%the one executing the instruction.  When a trap hook, $T(p^{t}_{j})$, delivers
+%a trap, thread $t$ executes an arbitrary trap handler $H = \{h^{t}_{1}, \ldots,
+%h^{t}_{h_{n}}\}$.  Thus, after the trap, the \lwt update, and the instruction
+%execute, the execution reaches the new point $Q = (q_{1}, \ldots, T(q_{e}),
+%h^{t}_{1}, \ldots, h^{t}_{h_{n}}, I(q_{e}), q_{e})$, and continues.
+%
+%A \ctraps implementation must be sound and complete to be correct.  A design is
+%{\em sound} if it only delivers a trap when threads have communicated.  A
+%design is {\em complete} if it always delivers a trap when threads have
+%communicated.  The \ctraps compiler puts a trap hook on all read and write
+%operations, and only on those operations.  Assuming a correct \lwt, \ctraps is
+%correct.  Our \ctraps implementation also does not insert any synchronization
+%into the program, so it comes with the same caveats related to data-race
+%freedom as the \lwt.  The same arguments for \lwt correctness apply to the
+%correctness of trap delivery and handling.  Most importantly, data-race freedom
+%implies that trap hooks, \lwt updates, and accesses are all atomic.
 
@@
 literature.
+
+\paragraph{Apache}
+%http://issues.apache.org/bugzilla/show_bug.cgi?id=21287
+%concurrency-bugs
+We conducted a case study using Apache httpd-2.0.48, which has a bug in its
+script cache. The cache uses reference counting to determine when to deallocate
+cached objects.  When a thread sees a reference count hit zero, it frees the
+object.  The bug is that the check that the count is zero and the deallocation
+are not atomic but it should be.  Multiple threads may see the zero reference
+count and free the object more than once, leading to a crash.
+
+We debugged this issue by examining the core dump from a crashing execution.
+The core dump showed that the memory allocator aborted due to one thread
+freeing already freed memory at {\tt mod\_mem\_cache.c}, line 287.  We examined
+the last writer table entry for the object being freed at that point.  That
+entry revealed that a different thread last wrote that object at the same point
+in the code.  The information in the last writer table illustrates why the
+double free bug is happening -- two threads end up in the object deallocation
+routine for the same object, which should not happen.
 

### 2013-11-08 paper.tex
committing for sw update

@@
 
-\title{\Large Last Writer Slices and Communication Traps}
+\title{\Large Last Writer Slicing: Efficient Data Provenance for Concurrent Program Debugging and Analysis}
 
@@
 %dependences typically has overheads too high for production.  
-
+{\bf TODO Re-write}
 We design efficient system support for collecting {\em last writer slices} for
-shared-memory concurrent programs.  Last writer slices dynamically track memory
-updates.  Last writer slices reflect the provenance of values in memory,
-information that is useful during debugging.   Using last writer slices,
-we build {\em communication traps} (\ctraps).  \ctraps is a mechanism that
-efficiently monitors and can interpose on memory operations that share data
-between threads.  \ctraps is an extensible framework, exposing
-inter-thread dependences to {\em \ctraps applications} that can implement
-arbitrary dependence analyses.
+shared-memory concurrent programs.  Last writer slices dynamically record the
+code responsible for shared variable updates.  Last writer slices reflect the
+provenance of values in memory, information that is useful during debugging.
+Using last writer slices, we build {\em communication traps} (\ctraps).
+\ctraps is a mechanism that efficiently monitors and can interpose on memory
+operations that share data between threads.  \ctraps is an extensible
+framework, exposing inter-thread dependences to {\em \ctraps applications} that
+can implement arbitrary dependence analyses.
 
@@
 
-A common approach to this challenge is to reason about dependent memory
-operations.  Memory operations are dependent if their order determines the
-values they read or write.  The utility of dependence information is made
-especially clear by its prevalent use during manual program debugging and in
-support of program analysis tools.
+%A common approach to this challenge is to reason about dependent memory
+%operations.  Memory operations are dependent if their order determines the
+%values they read or write.  The utility of dependence information is made
+%especially clear by its prevalent use during manual program debugging and in
+%support of program analysis tools.
 
 Debugging a concurrent program requires understanding how threads interact and
-dependence information furnishes that understanding.  Dependence information
-helps programmers with the difficult task of reasoning about the {\em
-provenance} of a memory location's value -- control- and data-flows that
-influenced that value.  Programmers debug programs more effectively when they
-have access to a value's provenance at points during a program's
-execution~\cite{tipslicingsurvey,whylineicse}.  In concurrent programs, data
-provenance information is particularly valuable because data may originate from
-any of a potentially large number of threads, making reasoning without explicit
-provenance information complex and difficult.
-
-Compounding this complexity, many concurrency bugs manifest as failures rarely.
-When a deployed system fails and a programmer receives a bug report,
-reproducing the reported failure to collect provenance information may be
-difficult or impossible.  It is essential to collect as much information as
-performance margins permit from deployed systems, to reduce the need for
-time-consuming failure reproduction.  Unfortunately, prior approaches to
-tracking dependences in concurrent code~\cite{raceslicing,tipslicingsurvey}
-impose high run time overheads.  These overheads are prohibitively high for use
-in production.
+{\em data provenance information} furnishes that understanding. A value's
+provenance is a description of the reason it is held by a variable.  Provenance
+can take on many forms, for example, a value's provenance could be the
+instruction that wrote it or the function that returned it.  Provenance
+information helps answer the question: "why does this variable contain this
+value?".  During debugging a programmer can use that information to connect 
+disparate parts of a program that interact, leading to a failure.
+
+In concurrent programs, collecting debugging information, like data provenance,
+from production systems is particularly important.  Concurrency bugs manifest
+as failures rarely.  When a deployed system fails and a programmer receives a
+bug report, reproducing the reported failure to collect provenance information
+may be difficult or impossible.  It is essential to collect as much information
+as performance margins permit from deployed systems, to reduce the need for
+time-consuming failure reproduction.
+
+Prior work has shown that provenence helps with debugging~\cite{badapples,
+whylineicse, tipslicingsurvey}.  One effort tracked code that wrote unusable
+values into variables~\cite{badapples}.  That work showed that revealing the
+connection between a bad value's origin and a failure it causes is helpful
+during debugging.  Other work showed that programmers debug programs more
+effectively when they can explicitly ask "why" questions during
+debugging~\cite{tipslicingsurvey,whylineicse}.
+
+These efforts have shown the promise of provenance, but are not a panacea.
+First, prior work~\cite{badapples} has focused on providing provenance
+information for only some unusable {\em values}, like {\tt NULL}.  Limiting the
+applicability of their technique to only some failures, like {\tt Null Pointer
+Exceptions}.  Second, while prior work has shown that limited provenance
+tracking for Java programs is possible with overheads low enough for
+production~\cite{badapples}, a more general form of provenance tracking for C
+and C++ programs has been elusive.  Prior approaches that use
+slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight memory
+checkers~\cite{badapples} impose overheads too high for production for C and
+C++ programs. Third, no prior technique that we know of has collected
+provenance information that is tailored to debugging concurrent programs with
+overheads that are low enough for use in production systems.  Focusing on
+concurrency is important because a value could have provenance in any of a
+large number of threads, making reasoning about why a variable has a value
+complex and difficult.
+
+In that context, we design a novel form of in-production provenance tracking
+for concurrent programs with three goals.   (1) To collect a general form of
+provenance information, not limited to certain values.  (2) To provide
+provenance information that helps understand how threads interact, which is
+especially useful for concurrent program debugging.  (3) To work with overheads
+low enough for use in production (ideally, less than 10\%).
+
+We develop a novel provenance tracking mechanism called {\em last writer
+slicing} that satisfies these goals.   As the program executes in production,
+last writer slicing tracks the thread and program point that last wrote any
+value to each potentially shared memory location --- its provenance.  Last
+writer slices are saved with a core dump when the program fails and are
+available in the debugger ({\em e.g.}, for breakpoint debugging).
+
+Last writer slices are primarily useful for concurrency debugging, but we also
+observe that provenance information in the form of last writer slices is a
+useful foundation for implementing other concurrent program analyses.  Several
+tools from prior work has used {\em ad hoc} last-writer information as part of
+more heavy-weight, sophisticated in domains like bug detection ({\em e.g.},
+~\cite{dmtracker,avio}), automatic debugging ({\em e.g.},
+~\cite{recon,conmem,conseq,cci}), profiling ({\em
+e.g.},~\cite{threadclustering, schedpredictionmodel}), and program
+understanding~\cite{oshatr}.
+
+Motivated by these prior {\em ad hoc} efforts, we build {\em Communication
+Traps} or {\em \ctraps}  on top of last writer slices
+(Section~\ref{sec:ctraps}).  \ctraps checks the last writer slice of each potentially
+shared memory location when threads access those locations.  If the accessing thread
+is different from the thread that last wrote, the access signfies {\em communication}
+between the threads.  Communication is a common subject of
+
+monitors inter-thread communication with
+overhead low enough for production use in many important cases.  When threads
+communicate, \ctraps delivers a software {\em trap} that can be handled by a
+trap handler.  Trap handlers implement {\em \ctraps applications} through an
+extensible interface that can analyze or interpose on the execution
+(Section~\ref{sec:apps}).  By default, \ctraps has very low overhead.  \ctraps
+applications can perform arbitrary dependence analyses, but at their own
+discretion and cost.  As we show in our evaluation, the overhead of \ctraps
+applications scales with the complexity of their analysis.
+
+%Provenance is also useful in implementing concurrent program analyses.
+%Dependence information is also an important part of many program analysis
+%tools.    Bug detection~\cite{avio,fasttrack,raceslicing,dmtracker} and
+%debugging~\cite{tipslicingsurvey,bugaboo,recon,cci,defuse,conseq,falcon} tools
+%analyze inter-thread dependences to find buggy behavior and help programmers
+%understand their errors.
+%Profilers~\cite{threadclustering,schedpredictionmodel} analyze inter-thread
+%dependences to find optimization opportunities.  Dependences are also essential
+%to some program understanding systems~\cite{oshatr} and dynamic checkers for
+%thread interaction properties~\cite{oshajava,velodrome}.
+%Record-Replay~\cite{chimera,doubleplay,fdr,rtr} systems track and record the
+%order of dependent events to help reproduce failures.  These systems would be
+%even more valuable, if their underlying dependence analysis was run in
+%production.  However, the heavy-weight analyses proposed by many of these
+%systems impose overheads too high for production use (5-100x
+%slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.
+%
+
+%First, we describe {\em last writer slices}
+%(Section~\ref{sec:lastwriterslices}).       Last writer slices are
+%useful because they expose the provenance of a memory location's contents
+%during debugging (Section~\ref{sec:debugging}).  When a failure occurs, last
+%writer slices are saved with a core dump.  Last writer slices are also
+%available during an execution in the debugger ({\em e.g.}, at breakpoints).
+%Note that while prior work ({\em e.g.},
+%~\cite{recon,fasttrack,conmem,conseq,cci}) has relied on {\em ad hoc}
+%last-writer analyses, in this work, we isolate the essential last-writer
+%information in the form of last writer slices and provide an implementation
+%with overheads appropriate for production systems.  Tracking last writer slices
+%has low overhead because it imposes overhead on only write operations:  reads
+%are overhead-free.  In Section~\ref{sec:eval} we experimentally show that
+%collecting last writer slices indeed has overheads suitable for production
+%systems for a variety of applications.
 
@@
 
-Dependence information is also an important part of many program analysis
-tools.    Bug detection~\cite{avio,fasttrack,raceslicing,dmtracker} and
-debugging~\cite{tipslicingsurvey,bugaboo,recon,cci,defuse,conseq,falcon} tools
-analyze inter-thread dependences to find buggy behavior and help programmers
-understand their errors.
-Profilers~\cite{threadclustering,schedpredictionmodel} analyze inter-thread
-dependences to find optimization opportunities.  Dependences are also essential
-to some program understanding systems~\cite{oshatr} and dynamic checkers for
-thread interaction properties~\cite{oshajava,velodrome}.
-Record-Replay~\cite{chimera,doubleplay,fdr,rtr} systems track and record the
-order of dependent events to help reproduce failures.  These systems would be
-even more valuable, if their underlying dependence analysis was run in
-production.  However, the heavy-weight analyses proposed by many of these
-systems impose overheads too high for production use (5-100x
-slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.
-
 %Deterministic execution
@@
 
-Our goal in this work is to collect, with deployable efficiency, useful
-dependence information that captures data provenance for use in debugging and
-serves as a foundation for more sophisticated program analyses.  We start from
-two simple observations: (1) Reads are frequent so we must minimize tracking
-overheads on reads, and (2) In concurrency analyses, {\em inter-thread}
-dependences are most important.  Based on these observations, we focus first on
-tracking write-after-write dependences (WAWs), which provide a useful form of
-data provenance information.  Building on that, we describe support for
-tracking {\em communication dependences}.  Communication occurs when a thread
-reads or overwrites a memory location that was most recently written by a
-different thread; {\em i.e.}, inter-thread read-after-write dependences (RAWs)
-as well as WAWs.  In this work, we demonstrate that this form of light-weight
-dependence tracking is useful and versatile, and has inherent overheads that
-are low enough for use in deployed systems.  There are two parts to our work:
-
-First, we describe {\em last writer slices}
-(Section~\ref{sec:lastwriterslices}).  Last writer slices record the thread and
-program point that last wrote each memory location.     Last writer slices are
-useful because they expose the provenance of a memory location's contents
-during debugging (Section~\ref{sec:debugging}).  When a failure occurs, last
-writer slices are saved with a core dump.  Last writer slices are also
-available during an execution in the debugger ({\em e.g.}, at breakpoints).
-Note that while prior work ({\em e.g.},
-~\cite{recon,fasttrack,conmem,conseq,cci}) has relied on {\em ad hoc}
-last-writer analyses, in this work, we isolate the essential last-writer
-information in the form of last writer slices and provide an implementation
-with overheads appropriate for production systems.  Tracking last writer slices
-has low overhead because it imposes overhead on only write operations:  reads
-are overhead-free.  In Section~\ref{sec:eval} we experimentally show that
-collecting last writer slices indeed has overheads suitable for production
-systems for a variety of applications.
-
-Second, on top of last writer slices we build {\em Communication Traps} or {\em
-\ctraps}  (Section~\ref{sec:ctraps}).  \ctraps monitors inter-thread
-communication with overhead low enough for production use in many important
-cases.  When threads communicate, \ctraps delivers a software {\em trap} that
-can be handled by a trap handler.  Trap handlers implement {\em \ctraps
-applications} through an extensible interface that can analyze or interpose on
-the execution (Section~\ref{sec:apps}).  By default, \ctraps has very low
-overhead.  \ctraps applications can perform arbitrary dependence analyses, but
-at their own discretion and cost.  As we show in our evaluation, the overhead
-of \ctraps applications scales with the complexity of their analysis.
+%Our goal in this work is to collect, with deployable efficiency, useful
+%dependence information that captures data provenance for use in debugging and
+%serves as a foundation for more sophisticated program analyses.  We start from
+%two simple observations: (1) Reads are frequent so we must minimize tracking
+%overheads on reads, and (2) In concurrency analyses, {\em inter-thread}
+%dependences are most important.  Based on these observations, we focus first on
+%tracking write-after-write dependences (WAWs), which provide a useful form of
+%data provenance information.  Building on that, we describe support for
+%tracking {\em communication dependences}.  Communication occurs when a thread
+%reads or overwrites a memory location that was most recently written by a
+%different thread; {\em i.e.}, inter-thread read-after-write dependences (RAWs)
+%as well as WAWs.  In this work, we demonstrate that this form of light-weight
+%dependence tracking is useful and versatile, and has inherent overheads that
+%are low enough for use in deployed systems.  There are two parts to our work:
 

### 2013-11-08 paper.tex
committing some stuff before a break

@@
 {\em data provenance information} furnishes that understanding. A value's
-provenance is a description of the reason it is held by a variable.  Provenance
-can take on many forms, for example, a value's provenance could be the
-instruction that wrote it or the function that returned it.  Provenance
-information helps answer the question: "why does this variable contain this
-value?".  During debugging a programmer can use that information to connect 
-disparate parts of a program that interact, leading to a failure.
+provenance is a description of the reason that it is stored in a variable.
+Provenance can take on many forms, for example, a value's provenance could be
+the instruction that wrote it or the function that returned it.  Provenance
+information helps answer the question: "Why does this variable contain this
+value?", which is central to debugging.
+
+%During debugging provenance helps a programmer better understand not just where failures occur, but where values came from, connectng disparate
+%parts of a program that interact, leading to a failure.
 
@@
 during debugging.  Other work showed that programmers debug programs more
-effectively when they can explicitly ask "why" questions during
-debugging~\cite{tipslicingsurvey,whylineicse}.
-
-These efforts have shown the promise of provenance, but are not a panacea.
-First, prior work~\cite{badapples} has focused on providing provenance
-information for only some unusable {\em values}, like {\tt NULL}.  Limiting the
-applicability of their technique to only some failures, like {\tt Null Pointer
-Exceptions}.  Second, while prior work has shown that limited provenance
-tracking for Java programs is possible with overheads low enough for
+effectively when they can explicitly ask "why" questions about the origin of
+certain values during debugging~\cite{tipslicingsurvey,whylineicse}.
+
+These efforts have shown some of the promise of provenance, but are not a
+debugging panacea.  First, prior work~\cite{badapples} has focused on providing
+provenance information for only some unusable {\em values}, like {\tt NULL}.
+Limiting the applicability of their technique to only some failures, like {\tt
+Null Pointer Exceptions}.  Second, while prior work has shown that limited
+provenance tracking for Java programs is possible with overheads low enough for
 production~\cite{badapples}, a more general form of provenance tracking for C
 and C++ programs has been elusive.  Prior approaches that use
-slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight memory
-checkers~\cite{badapples} impose overheads too high for production for C and
-C++ programs. Third, no prior technique that we know of has collected
+slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight
+memory checkers~\cite{badapples} impose overheads too high for production for C
+and C++ programs. Third, no prior technique that we know of has collected
 provenance information that is tailored to debugging concurrent programs with
@@
 tools from prior work has used {\em ad hoc} last-writer information as part of
-more heavy-weight, sophisticated in domains like bug detection ({\em e.g.},
+more heavy-weight analyses in domains like bug detection ({\em e.g.},
 ~\cite{dmtracker,avio}), automatic debugging ({\em e.g.},
@@
 Traps} or {\em \ctraps}  on top of last writer slices
-(Section~\ref{sec:ctraps}).  \ctraps checks the last writer slice of each potentially
-shared memory location when threads access those locations.  If the accessing thread
-is different from the thread that last wrote, the access signfies {\em communication}
-between the threads.  Communication is a common subject of
-
-monitors inter-thread communication with
-overhead low enough for production use in many important cases.  When threads
-communicate, \ctraps delivers a software {\em trap} that can be handled by a
-trap handler.  Trap handlers implement {\em \ctraps applications} through an
-extensible interface that can analyze or interpose on the execution
-(Section~\ref{sec:apps}).  By default, \ctraps has very low overhead.  \ctraps
-applications can perform arbitrary dependence analyses, but at their own
-discretion and cost.  As we show in our evaluation, the overhead of \ctraps
-applications scales with the complexity of their analysis.
+(Section~\ref{sec:ctraps}).  \ctraps checks the last writer slice of each
+potentially shared memory location when a thread accesses such a location.  If
+the accessing thread is different from the thread that last wrote the location,
+the access signifies {\em communication} between the threads.  When
+communication occurs, \ctraps delivers a software {\em trap} to the accessing
+thread giving it information about how communication occurred.  The trap can be
+handled by an analysis-specific trap handler.  Trap handlers implement {\em
+\ctraps applications} through an extensible interface that can interpose on the
+execution (Section~\ref{sec:apps}).  By default, \ctraps has very low overheads
+and similarly to last writer slicing are appropriate for use in production in
+many cases.  \ctraps applications can perform arbitrary concurrency analyses,
+but at their own discretion and cost.  As we show in our evaluation, the
+overhead of \ctraps applications scales with the complexity of their analysis.
 
@@
 
-\item{We describe last writer slicing, a dynamic analysis that tracks
-memory updates, capturing data provenance information.  We
-design and implement system support for collecting last writer slices.  }
-
-\item{Building on last writer slicing, we describe and implement \ctraps, which
-exposes inter-thread communication through an extensible trap interface.  }
+\item{We describe last writer slicing, an novel data provenance tracking
+mechanism, that produces information that is useful for debugging concurrent
+programs.  We design and implement system support for collecting last writer
+slices that is efficient enough for use in production.  }
+
+\item{Building on last writer slicing, we develop \ctraps, an extensible
+framework for implementing analyses of inter-thread communication in
+multi-threaded programs.  }
 
 \item{We evaluate the utility of last writer slices and \ctraps. We first
-present several case studies that show that last writer slices are useful for
-debugging real-world bugs.  We then show how \ctraps is useful for building
-dependence analyses by implementing two analyses from prior
+present a study of several real concurrent program bugs and show that last
+writer slices are useful for debugging.  We then show how \ctraps is useful for
+building dependence analyses by implementing two analyses from prior
 work~\cite{cci,defuse,recon}. }

### 2013-11-09 paper.tex
committing many TODOs, a new fig, some new Background text and some copy/paste-blasted Eval text.  Lots to do!

@@
 
-\title{\Large Last Writer Slicing: Efficient Data Provenance for Concurrent Program Debugging and Analysis}
+\title{Data Provenance for Concurrent Program Debugging and Analysis in Production Systems}
 
@@
 %dependences typically has overheads too high for production.  
-{\bf TODO Re-write}
-We design efficient system support for collecting {\em last writer slices} for
-shared-memory concurrent programs.  Last writer slices dynamically record the
-code responsible for shared variable updates.  Last writer slices reflect the
-provenance of values in memory, information that is useful during debugging.
-Using last writer slices, we build {\em communication traps} (\ctraps).
-\ctraps is a mechanism that efficiently monitors and can interpose on memory
-operations that share data between threads.  \ctraps is an extensible
-framework, exposing inter-thread dependences to {\em \ctraps applications} that
-can implement arbitrary dependence analyses.
-
-We implement last writer slicing and \ctraps.  We show that last writer slices
-provide information that is useful for debugging several real-world bugs.  We
+We propose {\em last writer slices}, an analysis for shared-memory
+multi-threaded programs that tracks {\em data provenance information} in
+production code.  Last writer slices dynamically record the thread and
+operation that last wrote each variable, reflecting the {\em provenance} of
+values.  Provenance helps understand bugs and we provide last writer slices to
+programmers during debugging.  We build {\em communication traps}
+(\ctraps) using last writer slices.  \ctraps efficiently
+monitors and can interpose on operations that share data between
+threads.  \ctraps is extensible, exposing inter-thread dependences
+to {\em \ctraps applications} that implement arbitrary analyses.
+
+We implement last writer slicing and \ctraps.  We conduct debugging studies of
+real, buggy programs showing last writer slices are useful for debugging.  We
 show \ctraps is useful by implementing two concurrency analyses from prior
-work.  We evaluate our designs on a set of server programs and standard
-benchmarks.  Our results show that collecting last writer slices has low run
-time overheads (0--15\%) for many important applications, like memcached,
-LevelDB, MySQL, and Apache.  We show that \ctraps imposes low run time
-overheads by default and that the overheads scale with the complexity of the
-analysis being implemented.
+work.  We evaluate the performance of our designs using server programs
+and standard benchmarks.  We show that collecting last writer slices
+has low run time overheads (0--15\%) for many applications,including  
+memcached, LevelDB, MySQL, and Apache.  We show that \ctraps imposes low 
+overheads by default and that overheads scale with analysis complexity.
 
@@
 
-In concurrent programs, collecting debugging information, like data provenance,
-from production systems is particularly important.  Concurrency bugs manifest
-as failures rarely.  When a deployed system fails and a programmer receives a
+Collecting debugging information, like data provenance, from production systems
+is particularly important for concurrent software.  Concurrency bugs manifest
+as failures rarely.  When a production system fails and a programmer receives a
 bug report, reproducing the reported failure to collect provenance information
 may be difficult or impossible.  It is essential to collect as much information
-as performance margins permit from deployed systems, to reduce the need for
+as performance margins permit from production, to reduce the need for
 time-consuming failure reproduction.
 
-Prior work has shown that provenence helps with debugging~\cite{badapples,
-whylineicse, tipslicingsurvey}.  One effort tracked code that wrote unusable
-values into variables~\cite{badapples}.  That work showed that revealing the
-connection between a bad value's origin and a failure it causes is helpful
-during debugging.  Other work showed that programmers debug programs more
-effectively when they can explicitly ask "why" questions about the origin of
-certain values during debugging~\cite{tipslicingsurvey,whylineicse}.
+Prior work have previously shown that some limited forms of provenence help
+with debugging~\cite{badapples, whylineicse, tipslicingsurvey}.  One effort
+tracked code that wrote unusable values into variables~\cite{badapples}.  That
+work showed that revealing the connection between a bad value's origin and a
+failure it causes is helpful during debugging.  Other work showed that
+programmers debug programs more effectively when they can explicitly ask "why"
+questions about the origin of certain values during
+debugging~\cite{tipslicingsurvey,whylineicse}.
 
@@
 provenance information that is tailored to debugging concurrent programs with
-overheads that are low enough for use in production systems.  Focusing on
+overheads that are low enough for use in production systems.  Addressing
 concurrency is important because a value could have provenance in any of a
-large number of threads, making reasoning about why a variable has a value
-complex and difficult.
-
-In that context, we design a novel form of in-production provenance tracking
-for concurrent programs with three goals.   (1) To collect a general form of
-provenance information, not limited to certain values.  (2) To provide
-provenance information that helps understand how threads interact, which is
-especially useful for concurrent program debugging.  (3) To work with overheads
-low enough for use in production (ideally, less than 10\%).
+large number of threads, making reasoning about why a variable contains that
+value complex and difficult.
+
+In this work we design a novel form of in-production provenance tracking for
+concurrent programs with three goals.  (1) To collect a general form of
+provenance information, not limited to only certain values.  (2) To provide
+provenance information that helps understand how threads interact, which is the
+key to understanding concurrency bugs.  (3) To work with overheads low enough
+for use in production (ideally, less than 10\%).
 
@@
 
-Data provenance information is directly useful during program debugging.  Work
-on debugging~\cite{whylinechi,whylineicse} in the HCI and Software Engineering
-community shows that when programmers debug programs, they benefit from the
-ability to directly ask such provenance (``why'') questions of their debugger.
-Note that that like prior work on debugging with provenance, in this work we
-focus on {\em debugging} -- understanding the root cause of a failure -- rather than {\em bug detection} -- finding new, unknown bugs.
+Data provenance information is directly useful during program debugging.  prior
+work~\cite{whylinechi,whylineicse} community showed that when programmers debug
+programs, they benefit from the ability to directly ask such provenance
+(``why'') questions of their debugger.  Origin tracking~\cite{badapples} and
+program slicing~\cite{tipslicingsurvey} also make debugging easier by reporting
+the provenance of values.  Like these prior efforts, in this work we focus on
+using data provenance information for {\em debugging} -- understanding the root
+cause of a failure -- as opposed to {\em bug detection} -- finding new, unknown
+bugs.
 
@@
 \centering
-\includegraphics[width=\columnwidth]{figs/BackgroundCoreDumpFail.pdf}
-\caption{\label{fig:coreDumpFail}{\bf Ordering violation bug in HTTrack.} Simply examining memory does not provide provenance information that helps understand this bug.}
+\includegraphics[width=\columnwidth]{figs/JDKStringBufferFail.pdf}
+\caption{\label{fig:coreDumpFail}{\bf Atomicity violation bug in a string buffer implementation.} The core dump does not reveal the connection between {\tt append()} and {\tt erase()}, which is crucial to understand the bug.}
 \end{figure}
 
-Figure~\ref{fig:coreDumpFail} illustrates the role that provenance information
-plays in debugging, using an example of a real bug from HTTrack, a web caching
-tool.  The figure depicts two threads manipulating a data structure that
-contains a mutex\_lock.  The program has an ordering violation bug: the
-program crashes when Thread 2 tries to acquire the lock before Thread 1 has
-initialized it.  Assuming the run time environment supports it, a core dump
-delivered to the programmer after such a failure reveals very little about how
-to fix the bug.  The lock is uninitizalized, so the core dump may show that it
-holds {\tt NULL} or it may hold random bits.  Both scenarios provide no
-insight into the root cause of the failure, which is that the code permits a
-use of the lock before its initialization.   Comparing the core dump from the
-failing run to the memory image from a correct execution -- a helpful debugging
-strategy -- reveals only that in a correct execution, the lock contains valid
-mutex state.  This revalation is of limited use as it does not rule out
-arbitrary data corruption as the bug's root cause.
-
-The vital, missing piece to solving this bug is provenance information.
-Debugging would be easier if the core dump included information about the
-thread and code point that last wrote the lock.  The programmer would see that
-no thread had initialized the lock in the failing execution, but that Thread 1
-initialized the lock in non-failing executions.  This information guides the
-programmer to fix the bug by adding synchronization that ensures the correct
-operation ordering.
-
-Another challenge in debugging this bug is that it manifests as a failure
+Figure~\ref{fig:coreDumpFail} illustrates the role of provenance information in
+debugging, using buggy code from the JDK-1.4 stringbuffer library, adapted to
+C++ by other researchers and used variously in prior
+work~\cite{concurrencybugs}.  The stringbuffer object pointed to by {\tt sb} in
+Thread 1's code contains a buffer and a variable called {\tt count} that stores
+its length.  The {\tt append()} code should atomically read {\tt sb->count} and
+update the string's buffer (the full update code is not pictured).  The failure
+occurs when the {\tt append()} code in Thread 1 is interleaved with the {\tt
+erase()} code in Thread 2 as indicated in the figure.  The {\tt erase()} code
+updates the buffer's {\tt count}, making the version that Thread 1 read into
+{\tt len} stale.  Thread 1's assertion fails at {\bf B} using the stale {\tt
+len} value, causing a crash.
+
+When the failure manifests, Thread 1 is at the assertion in {\tt append()}, as
+picture.  However, Thread 2 may have completed its erase operation and moved on
+to execute some unrelated code.  The call stack for Thread 2 may show nothing
+about its involvement in the failure, despite the fact that its interleaving
+with the {\tt append()} code is the root cause.  Examining the core dump, the
+programmer can see that the length and buffer are inconsistent, but is left to
+ask {\em why} they are inconsistent.
+
+Understanding the connection between {\tt append()} and {\tt erase()} is the
+key to fixing this bug.  The provenance of the value stored in {\tt sb->count}
+at {\bf B} was the write at {\bf A} in {\tt erase()} and with that provenance
+information, the programmer would understand that key connection.  Prior
+provenance work on origin tracking for unusable values~\cite{badapples} does
+not help here.  The involved values are inconsistent, but they are all {\em
+usable} so their provenance would not be tracked.
+
+Another challenge posed by this bug is that it manifests as a failure
 infrequently and under only real-world conditions.  If the programmer runs the
@@
 time and running it may require unavailable real-world inputs.  Thus,
-collecting provenance or other information offline is time-consuming or
+collecting provenance or other information offline may be time-consuming or
 impossible.  A better strategy is to continuously collect that information in
@@
 
-%Dynamic dependence slices are especially useful for debugging if they are
-%available after a failure has occurred in a deployed system. However, slicing
-%in deployed systems is uncommon, because deployed systems demand low overheads
-%and slicing can have high overhead.  Our system provides provenance information
-%that is useful at debugging time by collecting {\em last writer slices} that
-%help understand inter-thread communication.  We track last writer slices with
-%overhead low enough for use in production code.  Deployable efficiency makes it
-%possible to collect and report last writer slices the same way today's systems
-%collect and report core dumps.
-
 \subsection{Tracking Dependences and Communication}
 \label{sec:background:comm}
-%\ctraps tracks dependences that correspond to {\em inter-thread communication}.
-
+{\bf TODO: Clean this up, downplay the "dependences" obsession and make it 
+a better match to the LWS mission, as stated in the intro -- provenance + instrumentation = analysis}
 Monitoring inter-thread dependences is the foundation of many useful
@@
 
+{\bf TODO: Fix this figure to reflect availability of LWS info if the program crashes/is stopped at *any* point in the exec (like the read X in thread 2).}
+
 \begin{figure}[h]
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=\columnwidth]{figs/LWSHTTDebug.pdf}
-\caption{\label{fig:httlws}{\bf Using last writer slices to debug an
-ordering violation in HTTrack.} The provenance information stored in the last
-writer slice for the mutex variable helps understand the bug's root cause.}
-\end{figure}
-
 %The core dump does not provide any
@@
 
-Figure~\ref{fig:httlws} shows how last writer slices help understand the root
-cause of the concurrency bug from Figure~\ref{fig:coreDumpFail}.  Recall
-that the the program crashes at point {\bf B} when it calls {\tt
-hts\_mutex\_lock()} with an invalid mutex.    The root cause of this bug is
-that Thread 1 fails to initialize the mutex before Thread 2 uses that mutex.
-The last writer slices in the figure make this root cause clear: in the
-failing execution, the last writer slice indicates the mutex is uninitialized; in
-the non-failing execution, the last writer slice indicates the mutex was 
-initialized by Thread 1 at point {\bf A}.  This information tells the
-programmer that there is an ordering relationship between {\bf A} and {\bf B}
-that must be enforced to fix the bug.  Note that without last writer slices,
-the program crashes at {\bf B} and the core dump provides no information about
-the connection between point {\bf A} and point {\bf B} -- the programmer would be left to guess that the bug was an initialization problem.
+{\bf TODO: Replace with the "fixed" version of the JDK figure (currently in eval).  Update the text to match the new figure.}
 
@@
 \subsection{\lwt Formalism and Correctness}
+{\bf TODO: Necessary?  Is it really interesting?}
 \label{sec:lwssoundness}
@@
 \label{sec:eval:debugging}
+
+{\bf TODO: Table showing the bugs we looked at, vital stats, comparison to origin tracking}
+
 We illustrate the debugging benefits of last writer slices by using our system
@@
 \paragraph{HTTrack}
-The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
-and the text in Section~\ref{sec:debugging} describes the benefit provided by
-last writer slices in debugging an ordering violation in HTTrack-3.43.9.
+{\bf TODO: Deal with this duplication from the copy/paste}
+Figure~\ref{fig:coreDumpFail} illustrates the role that provenance information
+plays in debugging, using an example of a real bug from HTTrack, a web caching
+tool.  The figure depicts two threads manipulating a data structure that
+contains a mutex\_lock.  The program has an ordering violation bug: the
+program crashes when Thread 2 tries to acquire the lock before Thread 1 has
+initialized it.  Assuming the run time environment supports it, a core dump
+delivered to the programmer after such a failure reveals very little about how
+to fix the bug.  The lock is uninitizalized, so the core dump may show that it
+holds {\tt NULL} or it may hold random bits.  Both scenarios provide no
+insight into the root cause of the failure, which is that the code permits a
+use of the lock before its initialization.   Comparing the core dump from the
+failing run to the memory image from a correct execution -- a helpful debugging
+strategy -- reveals only that in a correct execution, the lock contains valid
+mutex state.  This revalation is of limited use as it does not rule out
+arbitrary data corruption as the bug's root cause.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\columnwidth]{figs/LWSHTTDebug.pdf}
+\caption{\label{fig:httlws}{\bf Using last writer slices to debug an
+ordering violation in HTTrack.} The provenance information stored in the last
+writer slice for the mutex variable helps understand the bug's root cause.}
+\end{figure}
+
+The vital, missing piece to solving this bug is provenance information.
+Debugging would be easier if the core dump included information about the
+thread and code point that last wrote the lock.  The programmer would see that
+no thread had initialized the lock in the failing execution, but that Thread 1
+initialized the lock in non-failing executions.  This information guides the
+programmer to fix the bug by adding synchronization that ensures the correct
+operation ordering.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\columnwidth]{figs/LWSHTTDebug.pdf}
+\caption{\label{fig:httlws}{\bf Using last writer slices to debug an
+ordering violation in HTTrack.} The provenance information stored in the last
+writer slice for the mutex variable helps understand the bug's root cause.}
+\end{figure}
+
+Figure~\ref{fig:httlws} shows how last writer slices help understand the root
+cause of the concurrency bug from Figure~\ref{fig:coreDumpFail}.  Recall
+that the the program crashes at point {\bf B} when it calls {\tt
+hts\_mutex\_lock()} with an invalid mutex.    The root cause of this bug is
+that Thread 1 fails to initialize the mutex before Thread 2 uses that mutex.
+The last writer slices in the figure make this root cause clear: in the
+failing execution, the last writer slice indicates the mutex is uninitialized; in
+the non-failing execution, the last writer slice indicates the mutex was 
+initialized by Thread 1 at point {\bf A}.  This information tells the
+programmer that there is an ordering relationship between {\bf A} and {\bf B}
+that must be enforced to fix the bug.  Note that without last writer slices,
+the program crashes at {\bf B} and the core dump provides no information about
+the connection between point {\bf A} and point {\bf B} -- the programmer would be left to guess that the bug was an initialization problem.
 
@@
 
-Last writer slices help elucidate a subtle atomicity violation in the
-StringBuffer library in version 1.4 of the Java JDK.  We use the C++ port of
-the buggy JDK and a driver program code developed in prior concurrency debugging
-work~\cite{concurrencybugs}.
-
-Figure~\ref{fig:jdklws} shows how the bug can manifest as a failure. The string
-object pointed to by {\tt sb} in Thread 1's code contains a string buffer and a
-variable called {\tt count} that stores its length.  The {\tt append()}
-code should atomically read {\tt sb->count} and update the string's buffer.
-The failure occurs when the {\tt append()} code in Thread 1 is interleaved with
-the {\tt erase()} code in Thread 2 as indicated in the figure.  The {\tt
-erase()} code updates the buffer's length, making the version Thread 1 read
-stale.  Thread 1 uses its stale version in the function called at point {\bf
-B}, causing a crash.
-
+The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
+and the text in Section~\ref{sec:debugging} describes the benefit provided by
+last writer slices in debugging an ordering violation in HTTrack-3.43.9.
+
+{\bf TODO: Move this to Section 3 where it will live.}
 \begin{figure}[h]
@@
 \end{figure}
-
-When the atomicity violation manifests, the appending thread is updating the
-string's underlying buffer.  The erasing thread, however, may have completed
-its erase operation.  The execution context of the erasing thread at the point
-of the failure provides no insight as to why the program failed.  Examining the memory image, the programmer can only conclude that the length and buffer are 
-inconsistent, but not why they are inconsistent.
-
-This situation illustrates the value of the provenenace information furnished
-by last writer slices.  At the point of the failure the last writer slice for
-the length variable points to a decrement of that variable in the erase
-function.  The last writer slice illustrates the need for synchronization
-between the append function and the erase function.
 

### 2013-11-10 paper.tex
committing some changes before sending a draft out

@@
 Data provenance information is directly useful during program debugging.  prior
-work~\cite{whylinechi,whylineicse} community showed that when programmers debug
+work~\cite{whylinechi,whylineicse} showed that when programmers debug
 programs, they benefit from the ability to directly ask such provenance
@@
 When the failure manifests, Thread 1 is at the assertion in {\tt append()}, as
-picture.  However, Thread 2 may have completed its erase operation and moved on
+pictured.  However, Thread 2 may have completed its erase operation and moved on
 to execute some unrelated code.  The call stack for Thread 2 may show nothing
@@
 
-{\bf TODO: Replace with the "fixed" version of the JDK figure (currently in eval).  Update the text to match the new figure.}
+{\bf TODO: Update the text to explain how the last writer slice shown makes the conection between erase() and append(), showing the programmer their mistake.}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\columnwidth]{figs/JDKStringBufferDebug2.pdf}
+\caption{\label{fig:jdklws}{\bf Using last writer slices to debug an
+ordering violation in the JDK 1.4 StringBuffer library.} The last writer slice reveals the interaction between the erase function and the append function that leads to a failure.}
+\end{figure}
 
@@
 and the text in Section~\ref{sec:debugging} describes the benefit provided by
-last writer slices in debugging an ordering violation in HTTrack-3.43.9.
-
-{\bf TODO: Move this to Section 3 where it will live.}
-\begin{figure}[h]
-\centering
-\includegraphics[width=\columnwidth]{figs/JDKStringBufferDebug.pdf}
-\caption{\label{fig:jdklws}{\bf Using last writer slices to debug an
-ordering violation in the JDK 1.4 StringBuffer library.} The last writer slice reveals the interaction between the erase function and the append function that leads to a failure.}
-\end{figure}
+last writer slices in debugging an ordering violation in the JDK-1.4 stringbuffer library.
 

### 2013-11-11 paper.tex
committing new eval stuff

@@
 key to fixing this bug.  The provenance of the value stored in {\tt sb->count}
-at {\bf B} was the write at {\bf A} in {\tt erase()} and with that provenance
-information, the programmer would understand that key connection.  Prior
-provenance work on origin tracking for unusable values~\cite{badapples} does
-not help here.  The involved values are inconsistent, but they are all {\em
-usable} so their provenance would not be tracked.
+at {\bf B} was the write at {\bf A} in {\tt erase()}.  Provenance information
+shows the programmer that key connection.  Prior provenance work on origin
+tracking for unusable values~\cite{badapples} does not help here.  The involved
+values are inconsistent, but they are all {\em usable} so their provenance
+would not be tracked.
 
 Another challenge posed by this bug is that it manifests as a failure
-infrequently and under only real-world conditions.  If the programmer runs the
-program in a debugger, the failure may not manifest in a reasonable amount of
-time and running it may require unavailable real-world inputs.  Thus,
-collecting provenance or other information offline may be time-consuming or
-impossible.  A better strategy is to continuously collect that information in
-deployed systems and package it with core dumps sent to programmers with bug
-reports.  In this work, we propose a last writer slicing design that espouses
-this {\em in situ} data collection strategy and does so with overheads
-appropriate for production systems.
-
-\subsection{Tracking Dependences and Communication}
+infrequently and under only real-world conditions.  Prior slicing and
+provenance debugging
+techniques~\cite{tipslicingsurvey,thinslicing,whylineicse}, have overheads too
+high to use in production.  Used offline, these techniques may not observe the
+failure in a reasonable amount of time and triggering the bug may require
+unavailable real-world inputs making offline analysis impossible.   A better
+strategy is to continuously collect that information in deployed systems and
+package it with core dumps sent to programmers with bug reports.  In this work,
+we propose a last writer slicing design that espouses this {\em in situ} data
+collection strategy and does so with overheads appropriate for production
+systems.
+
+\subsection{Communication Tracking in Concurrency Analyses}
 \label{sec:background:comm}
-{\bf TODO: Clean this up, downplay the "dependences" obsession and make it 
-a better match to the LWS mission, as stated in the intro -- provenance + instrumentation = analysis}
-Monitoring inter-thread dependences is the foundation of many useful
-concurrency analyses.  For example, in a software transactional memory (STM)
-system~\cite{stm} dependences correspond to interference between accesses in
-different threads that must be detected to rollback conflicting transactions.
-Deterministic concurrent runtime systems~\cite{coredet,grace} are similar --
-they monitor inter-thread dependences, to determine when to apply
-deterministic ordering constraints on threads' accesses.  STMs and
-deterministic systems must track all dependences and they must track them
-precisely, to provide strong guarantees as programming and execution models.
-Unfortunately, precision has a high runtime overhead -- {\em e.g.},
-CoreDet incurs a mean overhead of around 5x~\cite{coredet}.  This overhead is
-too high for use in production.
-
-%The overhead comes from the need to precisely track read-after-write (RAW),
-%write-after-read (WAR), and write-after-write (WAW) dependences, taking action
-%({\em e.g.}, buffering) on most cross-thread dependences.  These systems track
-%dependences with additional code that runs around memory accesses.  To ensure
-%correctness, accesses and tracking code usually must be protected by
-%synchronization.  The high cost of synchronization exacerbates the overheads.
-
-In contrast to {\em imprecision intolerant} tasks like STM and deterministic
-execution, other tasks are more {\em imprecision tolerant}.  Imprecision
-tolerant systems help perform tasks like
+
+Many useful concurrency analyses rely on monitoring inter-thread
+communication.  The domains of these analyses include 
 debugging~\cite{defuse,conseq,recon,bugaboo,raceslicing,fasttrack,falcon},
@@
 avoidance~\cite{aviso,cfix} and concurrent performance
-profiling~\cite{threadcriticality,schedpredictionmodel}.  These
-systems also work by monitoring how data flows between threads.  They can
-tolerate two important sources of imprecision.
-
-First, many of these techniques track only an {\em ad hoc} subset of
-dependences.  For example, DefUse~\cite{defuse} tracks inter-thread WAR
-dependences, but not WAW dependences.  Recon~\cite{recon} tracks some inter-thread WAW and
-RAW dependences and approximates WAR dependences.  These techniques are useful
-despite an incomplete set of dependences.  Second, many of these systems rely
-on statistical models of software
-behavior~\cite{avio,cci,defuse,recon,bugaboo,falcon,dmtracker,aviso,threadcriticality,schedpredictionmodel}.
-Models often tolerate noise, so many of these techniques can relax the
-correctness constraints of their implementation and remain useful.  For
-example, some systems have chosen not to enforce atomicity of each memory
-access and its dependence tracking code, compromising precision for a boost in
-performance.
-
-These tools and techniques work in a variety of environments.  A common theme
-is the advantage reaped from use in a deployment environment.  Debugging tools,
-profilers, and anomaly detectors benefit from seeing diverse, real-world
-behavior.  Dynamic failure avoidance techniques must work in deployed systems
-to be effective.  Deployment environments demand low overheads.
-
-The design for \ctraps that we develop in this work monitors a subset of
-inter-thread dependences that correspond to inter-thread communication.
-\ctraps provides enough precision to build useful analyses and runtime systems
-with overhead low enough for use in production.
+profiling~\cite{threadcriticality,schedpredictionmodel}.  The breadth of
+applications illustrates the generality of communication tracking.
+
+Despite the potential of a general communication tracking mechanism, prior work
+has largely relied on {\em ad hoc} communication analyses.  For example,
+DefUse~\cite{defuse} tracks communication via read-after-write (RAW)
+dependences, but not write-after-write (WAW) dependences.  Recon~\cite{recon}
+tracks some inter-thread WAW and RAW dependences.  These analyses could both be
+built with a general communication tracking mechanism.  Unfortunately, their
+single-purpose implementations track slightly different information, making
+them incompatible and difficult to compose.
+
+These analyses vary in their purpose and environments, but a cross-cutting
+theme is the advantage reaped from their use in a deployment environment.
+Debugging tools, profilers, and anomaly detectors benefit from seeing diverse,
+real-world behavior.  Dynamic failure avoidance techniques must work in
+deployed systems to be effective.  Deployment environments demand low overheads
+and all of these analyses benefit from a general, high-performance
+communication tracking framework.
+Unfortunately, many of these techniques are implemented using heavy-weight infrastructure, like binary instrumentation~\cite{pin}, making them inappropriate for use in production.
+
+There is a need for a general communication tracking framework that has
+performance overheads that are low enough to use in production systems.  As we
+describe in Section~\ref{sec:ctraps}, we aim to satisfy that need in this work.
+
+%For example, in a software transactional memory (STM) system~\cite{stm}
+%communication corresponds to interference between accesses in different threads
+%that must be detected to rollback conflicting transactions.  Deterministic
+%concurrent runtime systems~\cite{coredet,grace} similarly monitor
+%communication, to apply deterministic ordering constraints.  STMs and
+%deterministic systems must track all dependences and they must track them
+%precisely, to provide strong guarantees as programming and execution models.
+%Unfortunately, precision has a high runtime overhead -- {\em e.g.}, CoreDet
+%incurs a mean overhead of around 5x~\cite{coredet}.  This overhead is too high
+%for use in production.
 
@@
 
-At any point in an execution, the information stored in a memory location's
-\lwt entry represents the provenance of the data stored in that memory
-location.   Last writer slices do not explicitly record any information about
-read operations and they do not need to to be useful.  Interrupting the
-execution while debugging and examining an entry in the \lwt reveals the
-provenance of a value just when it is needed.  For example, if an assertion
-failure occurs due to an erroneous inter-thread data flow, the provenance of
-the data in the assertion's condition helps explain why the failure occurred.
-Computing last writer slices is fast enough to deploy and use continuously
-(see Section~\ref{sec:eval}) and the \lwt can be saved with a core dump.
-Thus, last writer slices can be made part of the standard debugging
-information provided when a program fails.
+At each point in an execution, a memory location's \lwt entry represents the
+provenance of the data stored in that memory location.   Last writer slicing
+performs no action on read operations and to be useful, it does not need to.
+Instead, a programmer can examine an \lwt entry at any arbitrary point in an
+execution ({\em e.g.}, at a breakpoint), including after a failure, and see a
+value's provenance.  Including the \lwt with core dumps means provenance is
+available when post-mortem debugging production failures.
+
+Figure~\ref{fig:jdklws} illustrates how the information in the \lwt helps debug
+the JDK stringbuffer bug introduced in Figure~\ref{fig:coreDumpFail}.  Recall
+that the failure occurs because Thread 2's execution of {\tt erase()} violates
+the atomicity of Thread 1's execution of {\tt append()}.  The \lwt entry for
+{\tt sb->count} at {\bf B} shows the programmer that its last writer was Thread
+2 at {\bf A}. Thread 2 may have run ahead to an arbitrary point in its
+execution.  Nothing in the core dump implicates {\tt erase()} in causing
+this bug.  The last writer slice reveals the interference between {\tt append()}
+and {\tt erase()} via {\tt sb->count}, helping the programmer understand the
+bug's cause.
 
@@
 %value should have been instead.
-
-{\bf TODO: Update the text to explain how the last writer slice shown makes the conection between erase() and append(), showing the programmer their mistake.}
 
@@
 Another important trait of last writer slices illustrated by
-Figure~\ref{fig:httlws} is that last writer slices are reported back to the
-programmer {\em with the core dump on a failure}.  Capturing good debugging
-information is especially important for concurrency bugs that manifest very
-rarely or in the presence of only peculiar real-world inputs.  A primary goal
-of this work is to show that we can collect last writer slices with overheads
-that are low enough for production use so they can be included pervasively in
-bug reports.
+Figure~\ref{fig:jdklws} is that when the assertion fails at {\bf B}, stopping
+the execution, the \lwt is saved and can be sent to the programmer with the
+core dump.  Capturing good debugging information is especially important for
+concurrency bugs that manifest very rarely or in the presence of only peculiar
+real-world inputs.  A primary goal of this work is to show that we can collect
+last writer slices with overheads that are low enough for production use so
+they can be included pervasively in bug reports.
 
@@
 to debug several real-world programs studied previously in the debugging
-literature.
+literature.  Table~\ref{tab:bugs} provides an overview of the bugs we studied
+describing the program, bug type, and bug report identifier where applicable.
+We then provide a summary of last writer slicing's debugging benefits.  To do
+so, we mirror the evaluation strategy of bad value origin
+tracking~\cite{badapples}, as it is the only other low-overhead provenance
+tracking mechanism we know of.  We evaluate our system using their three
+evaluation criteria - (1) effectiveness, (2) triviality, and (3) usefulness.
+Our technique is effective if the last writer slice reports a code point that
+wrote a value that subsequently caused a failure.  A failure is trivially
+debuggable if its cause is made obvious by examining the core dump alone, {\em
+i.e.}, a last writer slice is unnecessary.  A last writer slice is useful if it
+is effective and the thread and code point in the slice actually help
+understand why the failure occurred or how to fix it.   To draw contrast to bad
+value origin tracking, we indicate whether their analysis could help debug each
+bug, based on the description of their technique.
+
+\begin{table}
+\begin{tabular}{l|l|ccc}
+{\bf App} & {\bf Type} & {\bf Eff.?} & {\bf Triv.?} & {\bf Usef.?} \\ \hline
+mysql   & Atom.  &Yes  &No & Yes    \\
+pbzip2  & Order  &Yes  &No & Yes    \\
+httrack & Order  &Yes  &No & Yes    \\
+jdk-sb  & Atom.  &Yes  &No & Yes    \\
+apache1 & Atom.  &Yes  &No & Yes    \\ 
+apache2 & Atom.  &Yes  &No & Maybe  \\
+\end{tabular}
+\caption{{\bf Evaluation of debugging with last writer slices.} We describe each bug and summarize the debugging benefit of last writer slices using the evaluation criteria from ~\cite{badapples}.}
+\end{table}
 
@@
 \paragraph{HTTrack}
-{\bf TODO: Deal with this duplication from the copy/paste}
-Figure~\ref{fig:coreDumpFail} illustrates the role that provenance information
-plays in debugging, using an example of a real bug from HTTrack, a web caching
-tool.  The figure depicts two threads manipulating a data structure that
-contains a mutex\_lock.  The program has an ordering violation bug: the
-program crashes when Thread 2 tries to acquire the lock before Thread 1 has
-initialized it.  Assuming the run time environment supports it, a core dump
-delivered to the programmer after such a failure reveals very little about how
-to fix the bug.  The lock is uninitizalized, so the core dump may show that it
-holds {\tt NULL} or it may hold random bits.  Both scenarios provide no
-insight into the root cause of the failure, which is that the code permits a
-use of the lock before its initialization.   Comparing the core dump from the
-failing run to the memory image from a correct execution -- a helpful debugging
-strategy -- reveals only that in a correct execution, the lock contains valid
-mutex state.  This revalation is of limited use as it does not rule out
-arbitrary data corruption as the bug's root cause.
+%{\bf TODO: Deal with this duplication from the copy/paste}
+%Figure~\ref{fig:coreDumpFail} illustrates the role that provenance information
+%plays in debugging, using an example of a real bug from HTTrack, a web caching
+%tool.  The figure depicts two threads manipulating a data structure that
+%contains a mutex\_lock.  The program has an ordering violation bug: the
+%program crashes when Thread 2 tries to acquire the lock before Thread 1 has
+%initialized it.  Assuming the run time environment supports it, a core dump
+%delivered to the programmer after such a failure reveals very little about how
+%to fix the bug.  The lock is uninitizalized, so the core dump may show that it
+%holds {\tt NULL} or it may hold random bits.  Both scenarios provide no
+%insight into the root cause of the failure, which is that the code permits a
+%use of the lock before its initialization.     This revalation is of limited use as it does not rule out
+%arbitrary data corruption as the bug's root cause.
+%
+%\begin{figure}[h]
+%\centering
+%\includegraphics[width=\columnwidth]{figs/LWSHTTDebug.pdf}
+%\caption{\label{fig:httlws}{\bf Using last writer slices to debug an
+%ordering violation in HTTrack.} The provenance information stored in the last
+%writer slice for the mutex variable helps understand the bug's root cause.}
+%\end{figure}
+%
+%
+%The vital, missing piece to solving this bug is provenance information.
+%Debugging would be easier if the core dump included information about the
+%thread and code point that last wrote the lock.  The programmer would see that
+%no thread had initialized the lock in the failing execution, but that Thread 1
+%initialized the lock in non-failing executions.  This information guides the
+%programmer to fix the bug by adding synchronization that ensures the correct
+%operation ordering.
+
+Figure~\ref{fig:httlws} shows how last writer slices help understand the root
+cause of the concurrency bug from Figure~\ref{fig:coreDumpFail}.  The program
+crashes at point {\bf B} when it calls {\tt hts\_mutex\_lock()} with an invalid
+mutex.    The root cause of this bug is that Thread 1 fails to initialize the
+mutex before Thread 2 uses it.   The core dump is of little debugging use, only
+showing that, indeed, the lock is uninitialized.  Even if the programmer
+compared the core dump from the failing run to the memory image from a correct
+execution -- a helpful debugging strategy -- they would find only that in a
+correct execution, the lock contains valid mutex state.
 
@@
 
-The vital, missing piece to solving this bug is provenance information.
-Debugging would be easier if the core dump included information about the
-thread and code point that last wrote the lock.  The programmer would see that
-no thread had initialized the lock in the failing execution, but that Thread 1
-initialized the lock in non-failing executions.  This information guides the
-programmer to fix the bug by adding synchronization that ensures the correct
-operation ordering.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=\columnwidth]{figs/LWSHTTDebug.pdf}
-\caption{\label{fig:httlws}{\bf Using last writer slices to debug an
-ordering violation in HTTrack.} The provenance information stored in the last
-writer slice for the mutex variable helps understand the bug's root cause.}
-\end{figure}
-
-Figure~\ref{fig:httlws} shows how last writer slices help understand the root
-cause of the concurrency bug from Figure~\ref{fig:coreDumpFail}.  Recall
-that the the program crashes at point {\bf B} when it calls {\tt
-hts\_mutex\_lock()} with an invalid mutex.    The root cause of this bug is
-that Thread 1 fails to initialize the mutex before Thread 2 uses that mutex.
-The last writer slices in the figure make this root cause clear: in the
-failing execution, the last writer slice indicates the mutex is uninitialized; in
-the non-failing execution, the last writer slice indicates the mutex was 
-initialized by Thread 1 at point {\bf A}.  This information tells the
-programmer that there is an ordering relationship between {\bf A} and {\bf B}
-that must be enforced to fix the bug.  Note that without last writer slices,
-the program crashes at {\bf B} and the core dump provides no information about
-the connection between point {\bf A} and point {\bf B} -- the programmer would be left to guess that the bug was an initialization problem.
+Looking at last writer slices make this root cause clear: in the failing
+execution, the last writer slice indicates the mutex is uninitialized; in the
+non-failing execution, the last writer slice indicates the mutex was
+initialized by Thread 1 at point {\bf A}. While the core dump provides no
+information about the connection between point {\bf A} and point {\bf B}, the
+last writer slice reveals that the programmer should ensure {\bf B} follows
+{\bf A} to prevent the failure.  Note that even augmenting the core dump with
+bad value origin tracking~\cite{badapples} does not help with this bug.  In the
+failing execution, origin tracking shows that the unusable lock is
+uninitialized, but in the non-failing exeuction, origin tracking does not
+monitor the {\em usable} lock value.  By contrast, last writer slicing tracks
+that crucial information, helping debug this complex concurrency bug.
 
@@
 
-Bad Apple Tracking~\cite{bondapple} is also a form of debugging with data
+Bad Apple Tracking~\cite{badapples} is also a form of debugging with data
 provenance information. This work records the {\em origin} of an unusable value

### 2013-11-11 paper.tex
committing changes to the eval that add a table and our new eval criteria

@@
 
-{\bf TODO: Table showing the bugs we looked at, vital stats, comparison to origin tracking}
-
 We illustrate the debugging benefits of last writer slices by using our system
@@
 \begin{table}
-\begin{tabular}{l|l|ccc}
-{\bf App} & {\bf Type} & {\bf Eff.?} & {\bf Triv.?} & {\bf Usef.?} \\ \hline
-mysql   & Atom.  &Yes  &No & Yes    \\
-pbzip2  & Order  &Yes  &No & Yes    \\
-httrack & Order  &Yes  &No & Yes    \\
-jdk-sb  & Atom.  &Yes  &No & Yes    \\
-apache1 & Atom.  &Yes  &No & Yes    \\ 
-apache2 & Atom.  &Yes  &No & Maybe  \\
+\scriptsize
+\centering
+\begin{tabular}{l|ll|l|ccc|c}
+{\bf App} & {\bf Ver. \#} & {\bf Bug \#} &{\bf Type} & {\bf Eff.?} & {\bf Triv.?} & {\bf Usef.?} & {\bf O.T.?}\\ \hline
+mysql  &  4.0.12 &  791     & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
+pbzip2 &  0.9.4  &  N/A     & Order  &Yes  &No    & Yes   & Yes \\
+httrack& 3.43.9  &  N/A     & Order  &Yes  &No    & Yes   & No  \\%Used w/o origin - no help
+jdk1.4 &  1.70   &          & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
+apache1&  2.0.48 &  21285   & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
+apache2&  2.2.9  &  45605   & Atom.  &Yes  &No    & Maybe & ??  \\
 \end{tabular}
-\caption{{\bf Evaluation of debugging with last writer slices.} We describe each bug and summarize the debugging benefit of last writer slices using the evaluation criteria from ~\cite{badapples}.}
+\caption{{\bf Debugging with last writer slices.} We describe each bug and summarize the debugging benefit of last writer slices using the evaluation criteria from ~\cite{badapples}.  }
+\normalsize
 \end{table}

### 2013-11-12 paper.tex
committing an intermediate state on the ride home...

@@
 
-Prior work have previously shown that some limited forms of provenence help
-with debugging~\cite{badapples, whylineicse, tipslicingsurvey}.  One effort
-tracked code that wrote unusable values into variables~\cite{badapples}.  That
-work showed that revealing the connection between a bad value's origin and a
+Prior work has previously shown that some limited forms of provenence help with
+debugging~\cite{badapples, whylineicse, tipslicingsurvey}.  One effort tracked
+code that wrote unusable values into variables~\cite{badapples}.  That work
+showed that revealing the connection between an unusable value's origin and a
 failure it causes is helpful during debugging.  Other work showed that
-programmers debug programs more effectively when they can explicitly ask "why"
-questions about the origin of certain values during
-debugging~\cite{tipslicingsurvey,whylineicse}.
+programmers debug more effectively when they can ask "why" questions about
+values during debugging~\cite{tipslicingsurvey,whylineicse}.
 
@@
 overheads that are low enough for use in production systems.  Addressing
-concurrency is important because a value could have provenance in any of a
+concurrency is important because a value could have originated in any of a
 large number of threads, making reasoning about why a variable contains that
@@
 ~\cite{dmtracker,avio}), automatic debugging ({\em e.g.},
-~\cite{recon,conmem,conseq,cci}), profiling ({\em
+~\cite{recon,conseq,cci}), profiling ({\em
 e.g.},~\cite{threadclustering, schedpredictionmodel}), and program
@@
 communication occurs, \ctraps delivers a software {\em trap} to the accessing
-thread giving it information about how communication occurred.  The trap can be
+thread, giving it information about how communication occurred.  The trap can be
 handled by an analysis-specific trap handler.  Trap handlers implement {\em
@@
 containing the program point and the identifier of the thread that last wrote
-data to that memory location.  We gather last writer slices at runtime by
-maintaining a {\em last writer table} (\lwt) that maps from a memory
-location's address to its last writer slice.  Just before a thread writes to a
-memory location,  the thread first finds the memory location's entry in the
-\lwt and updates that entry with its thread identifier and its current point in the
-program.  After updating the \lwt, the thread performs the original memory
-access.
+data to that memory location.
+
+We gather last writer slices at runtime by maintaining a {\em last writer
+table} (\lwt) that maps from a memory location's address to its last writer
+slice.  Just before a thread writes to a memory location, the thread first
+finds the memory location's entry in the \lwt and updates that entry with its
+thread identifier and its current point in the program.  After updating the
+\lwt, the thread performs the original memory access.
 
@@
 \subsection{Runtime \& Compiler Support}
- The \lwt is implemented as a {\em shadow memory} that holds one entry for
-each memory address accessed by the program.  Each \lwt entry records the
-thread and program point that last wrote to the entry's corresponding memory
-location.  When a thread performs a write to a memory location, it updates the
-\lwt by recording its thread identifier and the program point of its write
-operation to the memory location's \lwt entry.  A program point may be an
-instruction address, or a call stack.  We use instructions by default, but we experimented with 2-address bounded call-stacks as well.
+
+Figure~\ref{fig:lwsoverview} shows an overview of our system support for last
+writer slicing.  The programmer writes their program as usual. Our compiler
+instruments write operations in the program with calls to our runtime that
+update the \lwt.  The \lwt is maintained in the runtime and saved for use
+during debugging, like the core dump.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=.9\columnwidth]{figs/LWS_Overview.pdf}
+\caption{\label{fig:lwsoverview}{\bf Overview of system support for last writer slicing.} Our compiler and runtime library ensure the \lwt remains updated so it can be used during debugging.}
+\end{figure}
+
+The \lwt implemented in the runtime library is a {\em shadow memory} that
+holds one entry for each potentially shared memory address accessed by the
+program.  Each \lwt entry records the thread and program point that last wrote
+to the entry's corresponding memory location.  When a thread performs a write
+to a memory location, it updates the \lwt by recording its thread identifier
+and the program point of its write operation to the memory location's \lwt
+entry.  A program point may be an instruction address, or a call stack.  We use
+instructions by default, but we experimented with 2-address bounded call-stacks
+as well.
 
 Collecting last writer slices requires instrumentation on each write operation.
-We built an instrumenting compiler pass that inserts a call into our runtime
-system just before each write to a memory location that may be accessed by code
-executing in another thread (escaping memory accesses).  Write instrumentation
-calls are passed three pieces of information: the address of the memory word
-involved in the access, the thread identifier of the accessing thread, and the
-accessing program point.
-
-\subsubsection{Escape Analysis}
-We use a local function escape analysis to identify which accesses are to
-memory locations that can be statically guaranteed to be inaccessible to code
-running in other threads.  When a thread accesses a location that is
-inaccessible to code in other threads, there is no way that access can be a
-communicating access.  Such accesses are not instrumented by our compiler.  
-We rely on {\tt GCC}'s built-in function escape analysis to determine if a
-memory operation might lead to communication.  Locations determined not to
-escape function scope also cannot escape the thread executing the function.
+We built an instrumenting compiler pass that inserts a call to an \lwt update
+function in our runtime library just before each write that may access data
+last written by another thread (escaping memory accesses).  Write
+instrumentation calls are passed three pieces of information: the address of
+the memory word involved in the access, the thread identifier of the accessing
+thread, and the accessing program point.
+
+We use a {\tt GCC}'s local function escape analysis to identify accesses to
+memory locations that are provably inaccessible to code running in other
+threads.  Such operations access unshared data and they are not instrumented by
+our compiler.  
 %Using an inter-procedural thread escape analysis may eliminate more
@@
 \centering
-\includegraphics[width=.90\columnwidth]{figs/SystemDiagram.pdf}
-\caption{\label{fig:systemdiagram}CTraps system support.}
+\includegraphics[width=.90\columnwidth]{figs/CTraps_Overview.pdf}
+\caption{\label{fig:systemdiagram}{\bf CTraps system support.}}
 \end{figure}

### 2013-11-12 paper.tex
edits to intro

@@
 \section{Introduction}
-Multi-threaded programming is challenging.  In contrast to the simplicity of
+Multi-threaded programming is challenging.  In contrast to 
 sequential reasoning, multi-threaded programs require complex reasoning about
@@
 
-Debugging a concurrent program requires understanding how threads interact and
-{\em data provenance information} furnishes that understanding. A value's
-provenance is a description of the reason that it is stored in a variable.
-Provenance can take on many forms, for example, a value's provenance could be
-the instruction that wrote it or the function that returned it.  Provenance
-information helps answer the question: "Why does this variable contain this
-value?", which is central to debugging.
+A value's provenance is a description of the reason that it is stored in a
+variable. Provenance information helps answer the question: "Why does this
+variable contain this value?", which is central to debugging. Debugging a
+concurrent program requires understanding how threads interact, and since
+communication happens via reads and writes to shared variables, {\em data
+provenance information is essential}. Provenance can take on many forms, for
+example, a value's provenance could be the instruction that wrote it, the
+function that returned it, and in the context of concurrent programs, the
+thread that produced it.
 
@@
 
-Collecting debugging information, like data provenance, from production systems
-is particularly important for concurrent software.  Concurrency bugs manifest
-as failures rarely.  When a production system fails and a programmer receives a
-bug report, reproducing the reported failure to collect provenance information
-may be difficult or impossible.  It is essential to collect as much information
-as performance margins permit from production, to reduce the need for
-time-consuming failure reproduction.
-
-Prior work has previously shown that some limited forms of provenence help with
-debugging~\cite{badapples, whylineicse, tipslicingsurvey}.  One effort tracked
-code that wrote unusable values into variables~\cite{badapples}.  That work
-showed that revealing the connection between an unusable value's origin and a
-failure it causes is helpful during debugging.  Other work showed that
-programmers debug more effectively when they can ask "why" questions about
+Collecting debugging information, like data provenance, from production
+systems is particularly important for concurrent software since bugs
+manifest as failures rarely. When a production system fails and a programmer
+receives a bug report, reproducing the reported failure to collect provenance
+information may be difficult or impossible. It is essential to collect as much
+information as performance margins permit from production, to reduce the need
+for time-consuming failure reproduction.
+
+Prior work has previously shown that some limited forms of provenance help
+with debugging~\cite{badapples, whylineicse, tipslicingsurvey}. One effort
+tracked code that wrote unusable values into variables~\cite{badapples}. That
+work showed that revealing the connection between an unusable value's origin
+and a failure it causes is helpful during debugging. Other work showed that
+programmers debug more effectively when they can ask ``why'' questions about
 values during debugging~\cite{tipslicingsurvey,whylineicse}.
@@
 These efforts have shown some of the promise of provenance, but are not a
-debugging panacea.  First, prior work~\cite{badapples} has focused on providing
-provenance information for only some unusable {\em values}, like {\tt NULL}.
-Limiting the applicability of their technique to only some failures, like {\tt
-Null Pointer Exceptions}.  Second, while prior work has shown that limited
-provenance tracking for Java programs is possible with overheads low enough for
+debugging panacea and are insufficient for dealing with concurrent program
+bugs. First, prior work~\cite{badapples} has focused on providing provenance
+information for only some unusable {\em values}, like {\tt NULL}, which limits its applicability. 
+% Limiting the applicability of their technique to only some failures, like {\tt
+% Null Pointer Exceptions}.
+Second, while prior work has shown that limited provenance
+tracking for Java programs is possible with overheads low enough for
 production~\cite{badapples}, a more general form of provenance tracking for C
@@
 slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight
-memory checkers~\cite{badapples} impose overheads too high for production for C
-and C++ programs. Third, no prior technique that we know of has collected
+memory checkers~\cite{badapples}{\bf TODO: Brandon, check this reference, since you just cited it in the previous sentence.} impose overheads too high for production for
+C and C++ programs. Third, no prior technique that we know of has collected
 provenance information that is tailored to debugging concurrent programs with
@@
 value complex and difficult.
-
-In this work we design a novel form of in-production provenance tracking for
-concurrent programs with three goals.  (1) To collect a general form of
-provenance information, not limited to only certain values.  (2) To provide
-provenance information that helps understand how threads interact, which is the
-key to understanding concurrency bugs.  (3) To work with overheads low enough
-for use in production (ideally, less than 10\%).
+In summary, the challenges in using provenance to debug concurrent programs are: (1) collecting the right information to characterize the concurrent execution behavior; and (2) making it efficient enough for general programs running in production settings.
+
+In this work we address the challenges above and develop a novel form of
+in-production provenance tracking for concurrent programs. We have three
+goals. (1) To collect a general form of provenance information, not limited to
+only certain values. (2) To provide provenance information that helps
+understand how threads interact, which is the key to understanding concurrency
+bugs. (3) To work with overheads low enough for use in production (ideally,
+less than 10\%).
 
@@
 
-Motivated by these prior {\em ad hoc} efforts, we build {\em Communication
-Traps} or {\em \ctraps}  on top of last writer slices
-(Section~\ref{sec:ctraps}).  \ctraps checks the last writer slice of each
-potentially shared memory location when a thread accesses such a location.  If
-the accessing thread is different from the thread that last wrote the location,
-the access signifies {\em communication} between the threads.  When
-communication occurs, \ctraps delivers a software {\em trap} to the accessing
-thread, giving it information about how communication occurred.  The trap can be
-handled by an analysis-specific trap handler.  Trap handlers implement {\em
-\ctraps applications} through an extensible interface that can interpose on the
+Motivated by these prior {\em ad hoc} efforts, we build a general mechanism
+called {\em Communication Traps} or {\em \ctraps} on top of last writer slices
+(Section~\ref{sec:ctraps}). \ctraps can be used to implement several dynamic
+analysis tailored to concurrent programs, like the ones discussed above.
+\ctraps works by checking the last writer slice of each potentially shared
+memory location when a thread accesses such a location. If the accessing
+thread is different from the thread that last wrote the location, the access
+signifies {\em communication} between the threads. When communication occurs,
+\ctraps delivers a software {\em trap} to the accessing thread, giving it
+information about how communication occurred. The trap can be handled by an
+analysis-specific trap handler. Trap handlers implement {\em \ctraps
+applications} through an extensible interface that can interpose on the
 execution (Section~\ref{sec:apps}). By default, \ctraps has very low overheads
 and similarly to last writer slicing are appropriate for use in production in
-many cases.  \ctraps applications can perform arbitrary concurrency analyses,
-but at their own discretion and cost.  As we show in our evaluation, the
+many cases. 
+%\ctraps applications can perform arbitrary concurrency analyses,
+%but at their own discretion and cost. 
+As we show in our evaluation, the
 overhead of \ctraps applications scales with the complexity of their analysis.

### 2013-11-12 paper.tex
committing a draft with more discussion of the main table in eval

--- 2013-11-12 paper.tex
@@
 We illustrate the debugging benefits of last writer slices by using our system
-to debug several real-world programs studied previously in the debugging
-literature.  Table~\ref{tab:bugs} provides an overview of the bugs we studied
-describing the program, bug type, and bug report identifier where applicable.
-We then provide a summary of last writer slicing's debugging benefits.  To do
-so, we mirror the evaluation strategy of bad value origin
+to debug six real-world programs studied previously in the debugging
+literature.  Table~\ref{tab:bugs} provides an overview of this evaluation and
+Section~\ref{sec:eval:debugging:cases} looks in detail at several cases to 
+better illustrate how last writer slices helps with debugging.
+
+In Table~\ref{tab:bugs}, we show the bugs we studied describing the program,
+bug type, and bug report identifier where applicable.  For each bug, we
+followed the procedure for installation and bug triggering described by the
+author of~\cite{concurrencybugs}.  We study a mixture of concurrency bug types
+in this work, including both atomicity and ordering violations.
+
+The table also summarizes our evaluation of last writer slicing's debugging
+benefits.  We mirror the evaluation strategy of bad value origin
 tracking~\cite{badapples}, as it is the only other low-overhead provenance
 tracking mechanism we know of.  We evaluate our system using their three
-evaluation criteria - (1) effectiveness, (2) triviality, and (3) usefulness.
-Our technique is effective if the last writer slice reports a code point that
-wrote a value that subsequently caused a failure.  A failure is trivially
-debuggable if its cause is made obvious by examining the core dump alone, {\em
-i.e.}, a last writer slice is unnecessary.  A last writer slice is useful if it
-is effective and the thread and code point in the slice actually help
-understand why the failure occurred or how to fix it.   To draw contrast to bad
-value origin tracking, we indicate whether their analysis could help debug each
-bug, based on the description of their technique.
+evaluation criteria - (1) {\bf effectiveness}, (2) {\bf triviality}, and (3)
+{\bf usefulness}.  Our technique is {\bf effective} if the last writer slice
+reports a code point that wrote a value that subsequently caused a failure.  A
+failure is {\bf trivial} to debug if its cause is made obvious by examining the
+core dump alone, {\em i.e.}, a last writer slice is unnecessary.  A last writer
+slice is {\bf useful} if it is effective and the thread and code point in the
+slice help understand why the failure occurred or how to fix it.   We directly
+compare to bad value origin tracking.    The {\bf O.T.} column in
+Table~\ref{tab:bugs} shows  whether their analysis could help debug each bug,
+based on the description of their technique.
 
@@
 \end{tabular}
-\caption{{\bf Debugging with last writer slices.} We describe each bug and summarize the debugging benefit of last writer slices using the evaluation criteria from ~\cite{badapples}.  }
+\caption{\label{tab:bugs}{\bf Concurrency debugging with last writer slices.} We describe each bug and summarize the debugging benefit of last writer slices using the evaluation criteria from ~\cite{badapples}.
+}
 \normalsize
@@
 
-\paragraph{Apache}
-%http://issues.apache.org/bugzilla/show_bug.cgi?id=21287
-%concurrency-bugs
-We conducted a case study using Apache httpd-2.0.48, which has a bug in its
-script cache. The cache uses reference counting to determine when to deallocate
-cached objects.  When a thread sees a reference count hit zero, it frees the
-object.  The bug is that the check that the count is zero and the deallocation
-are not atomic but it should be.  Multiple threads may see the zero reference
-count and free the object more than once, leading to a crash.
-
-We debugged this issue by examining the core dump from a crashing execution.
-The core dump showed that the memory allocator aborted due to one thread
-freeing already freed memory at {\tt mod\_mem\_cache.c}, line 287.  We examined
-the last writer table entry for the object being freed at that point.  That
-entry revealed that a different thread last wrote that object at the same point
-in the code.  The information in the last writer table illustrates why the
-double free bug is happening -- two threads end up in the object deallocation
-routine for the same object, which should not happen.
+The results in Table~\ref{tab:bugs} show that for both atomicity and ordering
+violation bugs, last writer slices are effective and useful.  In all cases, we
+found that the last writer slice revealed the provenance of the data involved
+in the operation that caused the failure, so last writer slicing was effective.
+We found that none of the failures we looked at were trivially debuggable.
+This finding is consistent with the difficulty of debugging concurrency bugs --
+the core dump does not reveal the interaction between threads that led to a
+failure, only the result of that interaction that leads to the failure.
+
+Last writer slices were obviously useful in five of six cases.  In these cases,
+the root cause of the bug becomes clear when last writer slices show the
+inter-thread data flow led to a variable having a value that caused a failure.
+Using that information, we expect it would be straightforward in these cases to
+add synchronization to fix these bugs.  Note that without last writer slices,
+understanding the responsible inter-thread data flow and fixing the bug would
+be difficult.  For the remaining case, {\tt apache2}, last writer slices would
+probably help with debugging, but it is less obvious because this bug is a
+complex, multi-variable bug.  We provide more detail on {\tt apache2} in Section~\ref{sec:eval:debugging:cases}.
+
+Our evaluation shows that bad value origin tracking~\cite{badapples} is not
+likely to be helpful in five out of six cases where last writer slices is
+useful.  In one case, a use-after-delete bug in {\tt pbzip2}, bad value origin
+tracking helps.  For this bug, when one thread deletes a variable, it makes
+that variable unusable and bad value origin tracking would track that value.
+When the failure occurs, the programmer would then know why the value was
+deleted.  In the other five cases we looked at, the values involved in causing
+the failure are usable and would not be tracked by bad value origin tracking.
+
+\subsubsection{Case Studies}
+\label{sec:eval:debugging:cases}
+
+We now provide detailed cases studies showing the debugging role of last writer
+slices in several of the bugs we examined.  Space constraints prevent us from
+providing details for all cases.
+
+\paragraph{JDK-StringBuffer}
+The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
+and the text in Section~\ref{sec:debugging} describes the benefit provided by
+last writer slices in debugging an ordering violation in the JDK-1.4.
+
+%\paragraph{Apache}
+%%http://issues.apache.org/bugzilla/show_bug.cgi?id=21287
+%%concurrency-bugs
+%We conducted a case study using Apache httpd-2.0.48, which has a bug in its
+%script cache. The cache uses reference counting to determine when to deallocate
+%cached objects.  When a thread sees a reference count hit zero, it frees the
+%object.  The bug is that the check that the count is zero and the deallocation
+%are not atomic but it should be.  Multiple threads may see the zero reference
+%count and free the object more than once, leading to a crash. 
+%
+%We debugged this issue by examining the core dump from a crashing execution.
+%The core dump showed that the memory allocator aborted due to one thread
+%freeing already freed memory at {\tt mod\_mem\_cache.c}, line 287.  We examined
+%the last writer table entry for the object being freed at that point.  That
+%entry revealed that a different thread last wrote that object at the same point
+%in the code.  The information in the last writer table illustrates why the
+%double free bug is happening -- two threads end up in the object deallocation
+%routine for the same object, which should not happen.
 
@@
 
-\paragraph{PBZip2}
-We conducted a case study using a buggy version of PBZip2-0.9.1.    The bug is
-a use-after-delete bug that leads to an assertion failure.  A worker thread
-tries to acquire a mutex after it has been deleted by the main thread during
-shutdown, causing the assertion to fail.
-
-Last writer slices provide the key piece of information to debug this failure.
-When the program fails, the core dump shows that the mutex is invalid because
-it has been deleted.  The mutex's last writer slice shows that the last write
-to the mutex before the worker tried to acquire it was when the main thread
-deleted it.  The last writer slice reveals the connection between the the two
-pieces of code involved in the bug: the code that acquires the mutex and the
-code that deletes the mutex.  Note that this connection is not obvious when the
-program crashes because the main thread may have executed arbitrarily far from
-where it deleted the mutex before the worker thread crashes.
+%
+%\paragraph{PBZip2}
+%We conducted a case study using a buggy version of PBZip2-0.9.1.    The bug is
+%a use-after-delete bug that leads to an assertion failure.  A worker thread
+%tries to acquire a mutex after it has been deleted by the main thread during
+%shutdown, causing the assertion to fail.  
+%
+%Last writer slices provide the key piece of information to debug this failure.
+%When the program fails, the core dump shows that the mutex is invalid because
+%it has been deleted.  The mutex's last writer slice shows that the last write
+%to the mutex before the worker tried to acquire it was when the main thread
+%deleted it.  The last writer slice reveals the connection between the the two
+%pieces of code involved in the bug: the code that acquires the mutex and the
+%code that deletes the mutex.  Note that this connection is not obvious when the
+%program crashes because the main thread may have executed arbitrarily far from
+%where it deleted the mutex before the worker thread crashes.
 
@@
 is a critical security vulnerability.  Figure~\ref{fig:mysqllws} shows the
-buggy code.  Thread 1 rotates the database's log file, marking it closed,
-opening a new log file, then marking it open again.  Log rotation should be
+buggy code.  Thread 1 rotates the database's log file.  Log rotation should be
 atomic, but the lack of synchronization allows Thread 2 to run its insert
-during log rotation.  The insert code checks if the log is marked closed. If
-the log is marked closed, the insert is not logged, manifesting the failure.
-We added an {\tt abort} statement to the non-logging case because this behavior
-is a security vulnerability that the bug report describes as ``critical''.
-Adding this {\tt abort} is reasonable, as the bug report describes the
-condition in the {\tt else} block as being an error.
+during log rotation.  In that case, the insert query not logged, manifesting
+the failure.  We added an {\tt abort} statement to the non-logging case because
+this behavior is a security vulnerability that the bug report describes as
+``critical''.  Adding the {\tt abort} is reasonable, as the bug report
+describes the {\tt else} block as an error case.  Running in a debugger, a
+programmer might equivalently use a breakpoint, rather than adding an
+{\tt abort}.
 
@@
 
-We triggered the bug by running an insert query during a log rotation.  The bug
-manifested, stopping execution at {\bf C}.  As described in the bug report, the
-query went unlogged and examining the core dump revealed the log was marked
-closed at point {\bf B} ({\em e.g.}, {\tt log\_type} was {\tt LOG\_CLOSED}).
-However, the core dump did not reveal {\em why} the log was marked closed
-there.  The last writer slice at {\bf C} for the {\tt log\_type} variable shows
-that it was written last by Thread 1 at {\bf A}.  Without the last writer
-slice, the relationship between the log rotation code and the insert code is
-not apparent. The last writer slice reveals that the log rotation code
-interacts with the insert code incorrectly via the {\tt log\_type} variable.
-
-\paragraph{JDK-StringBuffer}
-
-The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
-and the text in Section~\ref{sec:debugging} describes the benefit provided by
-last writer slices in debugging an ordering violation in the JDK-1.4 stringbuffer library.
+We triggered the bug by running an insert query during a log rotation and
+execution stopped at {\bf C}.  The core dump showed the log was marked closed,
+which is the failure state ({\em e.g.}, {\tt log\_type} was {\tt LOG\_CLOSED}).
+However, the core dump did not reveal {\em why} the log was marked closed.  The
+last writer slice at {\bf C} for {\tt log\_type} shows that it was written last
+by Thread 1 at {\bf A}, in the log rotation code, which should be atomic.
+Thus, the last writer slice shows the programmer the need for log rotation to
+be atomic with respect to the insert code to prevent this failure.
 
@@
 
-{\bf TODO: Compare with these analyses implemented using binary instrumentation}
-
 %In two of these cases, {\tt
@@
 
-{\bf TODO: Run without escape analysis?}
-
 \section{Related Work}

### 2013-11-12 paper.tex
committing many changes from a read on 11/11.  Now, prepping to rework related & be sure the eval isn't lying or bad.

@@
 
-\title{Data Provenance for Concurrent Program Debugging and Analysis in Production Systems}
+\title{Efficient Data Provenance Analysis for Concurrent Programs}
 
@@
 % Null Pointer Exceptions}.
-Second, while prior work has shown that limited provenance
-tracking for Java programs is possible with overheads low enough for
-production~\cite{badapples}, a more general form of provenance tracking for C
-and C++ programs has been elusive. Prior approaches that use
-slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight
-memory checkers~\cite{badapples}{\bf TODO: Brandon, check this reference, since you just cited it in the previous sentence.} impose overheads too high for production for
-C and C++ programs. Third, no prior technique that we know of has collected
-provenance information that is tailored to debugging concurrent programs with
-overheads that are low enough for use in production systems. Addressing
-concurrency is important because a value could have originated in any of a
-large number of threads, making reasoning about why a variable contains that
-value complex and difficult.
-In summary, the challenges in using provenance to debug concurrent programs are: (1) collecting the right information to characterize the concurrent execution behavior; and (2) making it efficient enough for general programs running in production settings.
+Second, while prior work has shown that limited provenance tracking for Java
+programs is possible with overheads low enough for production~\cite{badapples},
+a more general form of provenance tracking for C and C++ programs has been
+elusive. Prior approaches that build on slicing~\cite{thinslicing,
+tipslicingsurvey, whylineicse} or heavy-weight memory checkers~\cite{badapples}
+impose overheads too high for production for C and C++ programs. Third, no
+prior technique that we know of has collected provenance information that is
+tailored to debugging concurrent programs with overheads that are low enough
+for use in production systems. Addressing concurrency is important because a
+value could have originated in any of a large number of threads, making
+reasoning about why a variable contains that value complex and difficult. In
+summary, the challenges in using provenance to debug concurrent programs are:
+(1) collecting the right information to characterize the concurrent execution
+behavior; and (2) making it efficient enough for general programs running in
+production settings.
 
@@
 the memory word involved in the access, the thread identifier of the accessing
-thread, and the accessing program point.
-
-We use a {\tt GCC}'s local function escape analysis to identify accesses to
-memory locations that are provably inaccessible to code running in other
-threads.  Such operations access unshared data and they are not instrumented by
-our compiler.  
+thread, and the accessing program point.   We use a {\tt GCC}'s local function
+escape analysis to identify accesses to memory locations that are provably
+inaccessible to code running in other threads.  Such operations access unshared
+data and they are not instrumented by our compiler.
+
 %Using an inter-procedural thread escape analysis may eliminate more
@@
 \subsection{\lwt Formalism and Correctness}
-{\bf TODO: Necessary?  Is it really interesting?}
 \label{sec:lwssoundness}
@@
 A point in an execution of $P$ after $e$ instructions have
-executed is defined by $Q = \{q_{0}, \ldots, q_{e}\}$ where each $q \in Q$ is a
+executed is defined by $Q = (q_{0}, \ldots, q_{e})$ where each $q_{i} \in Q$ is a
 $p^{t}_{j}$ from one of the $k_t$ instructions of thread $t$.  We introduce
@@
 atomicity and ordering is enforced by the program synchronization that ensures
-consistency for the data being written by the instructions our system
-instruments.
+consistency for program data.
 
@@
 operation, $p^{t}_{j}$. Thus, $I(p^{t}_{j}) \prec p^{t}_{j}$ due to program
-order.  Happens-before is transitive, so without data-races, $I(p^{t}_{j})
+order.
+$I(p^{t'}_{j'})$ is inserted before $p^{t'}_{j'}$, but after the synchronization that orders $p^{t'}_{j'}$ and $p^{t}_{j}$, so $p^{t}_{j} \prec I(p^{t'}_{j'})$.
+Happens-before is transitive, so without data-races, $I(p^{t}_{j})
 \prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec p^{t'}_{j'}$, meaning $I(p^{t}_{j})$
@@
 
-%{\bf TODO: Figure?}
-
 \paragraph{Non-overlapping Memory References} 
@@
 events as {\em traps} during an execution that {\em trap handlers} can handle
-to perform analysis and interposition.  We formalize trap delivery and handling
-in Section~\ref{sec:ctsoundness}.
+to perform analysis and interposition.
 
@@
 by one thread that access a memory location that was last written by another
-thread, {\em i.e.}, communicating memory operations.  Traps are simply function
-calls made before such operations execution.  \ctraps implements traps using a
-combination of compiler and runtime library support.  Note that we use the name
-``trap'' for the interposition points in \ctraps, but we implement traps using 
-function calls only, as opposed to signals or some other mechanism.
+thread, {\em i.e.}, communicating memory operations.  Traps are simply inserted
+function calls (as opposed to, {\em e.g.}, signals) made before such operations
+execution.  \ctraps implements traps using a combination of compiler and
+runtime library support.
 
 \paragraph{Compiler Support}
-Our compiler inserts a call to the \ctraps {\em trap hook} function immediately
+Our compiler inserts a call to the \ctraps~{\em trap hook} function immediately
 before all read and write operations that access potentially escaping memory
@@
 trap, the trap hook function calls {\em trap handler routines} that are defined
-in \ctraps applications.  Trap handler routines are like signal handlers.  They
-are registered with the \ctraps runtime when the program starts.  Trap handlers
-may contain arbitrary application-specific code.  Figure~\ref{fig:hookapi}
-shows the \ctraps trap handling API.  Trap handlers are passed the four pieces
-of information passed to the trap hook, as well as the \lwt's record of the
-program point and thread identifier of the last write to the involved memory
-location.
-
-\begin{figure}[htb]
-\centering
-\begin{verbatim}
-void trap(AccessType type, 
-          ThdId currThd, ProgPt currPt,
-          ThdId lastThd, ProgPt lastPt);
-\end{verbatim}
-\caption{\label{fig:hookapi}The \ctraps trap handling interface.}
-\end{figure}
+in \ctraps applications.  Trap handler routines are like signal handlers.  The
+\ctraps runtime maintains a list of trap handlers that are registered when the
+program starts via a configuration stored in the environment.  Trap handlers
+may contain arbitrary application-specific code.  \ctraps defines a trap
+handling API.  Trap handlers are passed the four pieces of information passed
+to the trap hook, as well as the \lwt's record of the program point and thread
+identifier of the last write to the involved memory location.
+
+%\begin{figure}[htb]
+%\centering
+%\begin{verbatim}
+%void trap(AccessType type, 
+%          ThdId currThd, ProgPt currPt,
+%          ThdId lastThd, ProgPt lastPt);
+%\end{verbatim}
+%\caption{\label{fig:hookapi}The \ctraps trap handling interface.}
+%\end{figure}
 
@@
 We released our implementation, free and open-sourced, on the
-web~\cite{withheldforanonymity} and published it to the GCC plugin repository.
+web~\cite{ctrapsrelease} and published it to the GCC plugin repository.
 Since its release, we have seen contributions to our codebase from members of
@@
 last writer slices provide information that is useful during debugging in
-Section~\ref{sec:eval:debugging} using case studies on several buggy programs.
-We evaluate our performance overheads and show that last writer slices and
-\ctraps are feasible for use in deployed systems for a majority of workloads we
-studied.  We then characterize our instrumentation and discuss the sources of
-our system's run time overhead.   We show that \ctraps is useful by
-implementing two existing applications and analyzing their performance.
+Section~\ref{sec:eval:debugging} looking at a set of buggy programs and
+comparing to prior, related work.  We evaluate our performance overheads and
+show that last writer slices and \ctraps are feasible for use in deployed
+systems for a majority of workloads we studied.  We then characterize our
+instrumentation and discuss the sources of our system's run time overhead.   We
+show that \ctraps is useful by implementing two existing applications and
+analyzing their performance.
 
@@
 We illustrate the debugging benefits of last writer slices by using our system
-to debug six real-world programs studied previously in the debugging
+to debug five real-world programs studied previously in the debugging
 literature.  Table~\ref{tab:bugs} provides an overview of this evaluation and
@@
 author of~\cite{concurrencybugs}.  We study a mixture of concurrency bug types
-in this work, including both atomicity and ordering violations.
+in this work, including both single- and multi-variable bugs and both atomicity
+and ordering violations.
 
@@
 jdk1.4 &  1.70   &          & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
-apache1&  2.0.48 &  21285   & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
-apache2&  2.2.9  &  45605   & Atom.  &Yes  &No    & Maybe & ??  \\
+apache&  2.0.48 &  21285   & Atom.  &Yes  &No    & Yes   & Maybe\\%All values usable
+%apache2&  2.2.9  &  45605   & Atom.  &Yes  &No    & Maybe & ??  \\
 \end{tabular}
@@
 The results in Table~\ref{tab:bugs} show that for both atomicity and ordering
-violation bugs, last writer slices are effective and useful.  In all cases, we
-found that the last writer slice revealed the provenance of the data involved
-in the operation that caused the failure, so last writer slicing was effective.
-We found that none of the failures we looked at were trivially debuggable.
-This finding is consistent with the difficulty of debugging concurrency bugs --
-the core dump does not reveal the interaction between threads that led to a
-failure, only the result of that interaction that leads to the failure.
-
-Last writer slices were obviously useful in five of six cases.  In these cases,
+violation bugs, last writer slices are effective.  In all cases, we found that
+the last writer slice revealed the provenance of the data involved in the
+operation that caused the failure, so last writer slicing was effective.  We
+found that none of the failures we looked at were trivially debuggable.  This
+finding is consistent with the difficulty of debugging concurrency bugs -- the
+core dump does not reveal the interaction between threads that led to a
+failure, only the result of the interaction, which leads to the failure.
+
+Last writer slices were obviously useful in all five cases.  In these cases,
 the root cause of the bug becomes clear when last writer slices show the
@@
 understanding the responsible inter-thread data flow and fixing the bug would
-be difficult.  For the remaining case, {\tt apache2}, last writer slices would
-probably help with debugging, but it is less obvious because this bug is a
-complex, multi-variable bug.  We provide more detail on {\tt apache2} in Section~\ref{sec:eval:debugging:cases}.
+be difficult.
+
+%For the remaining case, {\tt apache2}, last writer slices would
+%probably help with debugging, but it is less obvious because this bug is a
+%complex, multi-variable bug.  We provide more detail on {\tt apache2} in
+%Section~\ref{sec:eval:debugging:cases}.
 
 Our evaluation shows that bad value origin tracking~\cite{badapples} is not
-likely to be helpful in five out of six cases where last writer slices is
+helpful in three out of five cases where last writer slices is
 useful.  In one case, a use-after-delete bug in {\tt pbzip2}, bad value origin
@@
 When the failure occurs, the programmer would then know why the value was
-deleted.  In the other five cases we looked at, the values involved in causing
-the failure are usable and would not be tracked by bad value origin tracking.
+deleted.
+
+In the remaining case, {\tt apache}, bad value origin tracking may help.  The
+bug is an atomicity violation that leads to a double free.  Two threads
+manipulate an object's reference count and deallocate the object when the count
+hits zero. Zero-checks and deallocation should be atomic, but the code does not
+ensure they are.  If threads interleave their code, multiple threads might see
+zero reference counts, which is incorrect.  When that happens, both threads try
+to free the object and the second free causes a crash.  Bad value origin
+tracking may help in this case.  At the first free, the freed pointer becomes
+unusable and its origin is tracked.  At the crash, the bad value's reported
+origin is the code point that frees the object.  That information potentially
+helps debug the program by suggesting a double free.  However, in {\tt apache},
+there are many threads, some perform different tasks, and all may free that
+object at that code point.  Bad value origin tracking does not record the
+thread of the origin.  The programmer is left unsure of whether the problem is
+with single-threaded logic or with synchronization.  In contrast, last writer
+slices reports the freed value's origin's thread and code point, making it
+clear that the programmer is dealing with a concurrency bug, and which threads
+are involved.
 
@@
 \paragraph{JDK-StringBuffer}
-The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
+The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:jdklws},
 and the text in Section~\ref{sec:debugging} describes the benefit provided by
@@
 \paragraph{HTTrack}
-%{\bf TODO: Deal with this duplication from the copy/paste}
-%Figure~\ref{fig:coreDumpFail} illustrates the role that provenance information
-%plays in debugging, using an example of a real bug from HTTrack, a web caching
-%tool.  The figure depicts two threads manipulating a data structure that
-%contains a mutex\_lock.  The program has an ordering violation bug: the
-%program crashes when Thread 2 tries to acquire the lock before Thread 1 has
-%initialized it.  Assuming the run time environment supports it, a core dump
-%delivered to the programmer after such a failure reveals very little about how
-%to fix the bug.  The lock is uninitizalized, so the core dump may show that it
-%holds {\tt NULL} or it may hold random bits.  Both scenarios provide no
-%insight into the root cause of the failure, which is that the code permits a
-%use of the lock before its initialization.     This revalation is of limited use as it does not rule out
-%arbitrary data corruption as the bug's root cause.
-%
-%\begin{figure}[h]
-%\centering
-%\includegraphics[width=\columnwidth]{figs/LWSHTTDebug.pdf}
-%\caption{\label{fig:httlws}{\bf Using last writer slices to debug an
-%ordering violation in HTTrack.} The provenance information stored in the last
-%writer slice for the mutex variable helps understand the bug's root cause.}
-%\end{figure}
-%
-%
-%The vital, missing piece to solving this bug is provenance information.
-%Debugging would be easier if the core dump included information about the
-%thread and code point that last wrote the lock.  The programmer would see that
-%no thread had initialized the lock in the failing execution, but that Thread 1
-%initialized the lock in non-failing executions.  This information guides the
-%programmer to fix the bug by adding synchronization that ensures the correct
-%operation ordering.
-
 Figure~\ref{fig:httlws} shows how last writer slices help understand the root
-cause of the concurrency bug from Figure~\ref{fig:coreDumpFail}.  The program
-crashes at point {\bf B} when it calls {\tt hts\_mutex\_lock()} with an invalid
-mutex.    The root cause of this bug is that Thread 1 fails to initialize the
-mutex before Thread 2 uses it.   The core dump is of little debugging use, only
-showing that, indeed, the lock is uninitialized.  Even if the programmer
-compared the core dump from the failing run to the memory image from a correct
-execution -- a helpful debugging strategy -- they would find only that in a
-correct execution, the lock contains valid mutex state.
+cause of a concurrency bug from HTTrack-3.43.9.  The program crashes at point
+{\bf B} when it calls {\tt hts\_mutex\_lock()} with an invalid mutex.    The
+root cause of this bug is that Thread 1 fails to initialize the mutex before
+Thread 2 uses it.   The core dump is of little debugging use, only showing
+that, indeed, the lock is uninitialized.  Even if the programmer compared the
+core dump from the failing run to the memory image from a correct execution --
+a helpful debugging strategy -- they would find only that in a correct
+execution, the lock contains valid mutex state.
 
@@
 failing execution, origin tracking shows that the unusable lock is
-uninitialized, but in the non-failing exeuction, origin tracking does not
-monitor the {\em usable} lock value.  By contrast, last writer slicing tracks
-that crucial information, helping debug this complex concurrency bug.
+uninitialized, revealing no more than the core dump.  In the non-failing
+execution, origin tracking does not monitor the {\em usable} lock value, adding
+nothing.  By contrast, comparing the last writer slices for the failing and
+non-failing executions reveals how to fix this bug.
 
@@
 Thus, the last writer slice shows the programmer the need for log rotation to
-be atomic with respect to the insert code to prevent this failure.
+be atomic with respect to the insert code to prevent this failure.  Bad value
+origin tracking does not help debug this bug because {\tt log\_type} only ever
+contains usable values so it would not be tracked.
 

### 2013-11-13 paper.tex
committing with 10 pages, redone related.  Need to add bugs?

@@
 slicing low enough for deployment use.
-
-{\bf TODO: Fix this figure to reflect availability of LWS info if the program crashes/is stopped at *any* point in the exec (like the read X in thread 2).}
 
@@
 
-There are several categories of prior work related to last writer slices and
-\ctraps.
+\paragraph{Debugging with Data Provenance Information}
+The most related prior work looked at using various forms of provenance information for debugging.
+
+We have discussed bad value origin tracking ~\cite{badapples} throughout this
+work and compared directly with that technique in our evaluation.  Bad value
+origin tracking is like our work because they store provenance information to
+be used for debugging -- a similar goal to ours.  Bad value origin tracking
+differs from our work in several ways.  First, their technique only tracks the
+provenance of {\em unusable} values -- a fraction of all values in an
+execution.  Our work, instead, tracks provenance for any value stored in a
+potentially shared memory location, which includes a broader set of provenance
+and targets our technique to concurrency debugging.  Second, their technique
+does not run with overheads tolerable for performance for C and C++ programs.
+They implement their technique for C and C++ programs using Valgrind MemCheck,
+which has high runtime overheads.  Our technique has overheads low enough for
+production in many cases, especially in server applications.  Third, their
+technique does not explicity record thread information with tracked origins.
+Our work focuses on concurrency and we include thread information.  As we show
+in Section~\ref{sec:eval:debugging} thread information is important for
+understanding some bugs.  Fourth, their technique uses bad value piggy-backing
+to store origins, we use a shadow memory.
+
+Like our work, the WhyLine debugger~\cite{whylineicse, whylinechi} allows
+programmers to ask questions about why variables hold particular values during
+debugging.  WhyLine answers by presenting data- and control-flows that
+influenced those variables.  WhyLine differs from our work in that it uses
+dependence slicing on recorded executions to answer provenance questions and
+has overheads too high for production use.  WhyLine also did not explicitly
+focus on concurrency.
 
 \paragraph{Program Slicing}
-Slicing techniques~\cite{tipslicingsurvey} are techniques for identifying
-useful subsequences of a program's instructions.   Slicing can work by
-analyzing program code statically or analyzing execution traces dynamically.
-Many slicing techniques discussed in ~\cite{tipslicingsurvey} monitor control
-and data dependences to identify which parts of a program might be relevant to
-a task ({\em e.g., debugging}).
-
-Thin Slicing~\cite{thinslicing} is especially related to this work. Thin
-Slicing monitors and reports a select subset of dependent operations from an
-execution.  The goal is to eliminate some operations that would be included in
-a traditional slice to make the slice simpler and less costly to collect.
-While this work and Thin Slicing differ in what they collect and how, our
-motivation to monitor a select subset of dependences was similar to theirs.
-
-%is unlike our work in that our work focused on
-%monitoring inter-thread dependences in production code without sampling or
-%hardware support.
-
-\paragraph{Debugging with Data Provenance Information}
-
-Prior work has looked at using data provenance information in
-debugging~\cite{whylineicse, whylinechi}.  The WhyLine debugger allowed
-programmers to ask questions about why particular variables held particular
-values during debugging.  WhyLine answered those questions by presenting data-
-and control-flows that influenced those variables.  WhyLine showed that
-provenance information is intuitive to programmers and makes debugging easier,
-justifying our focus on provenance information in our last writer slicing
-design.
-
-There are several characteristics of this work that distinguish it from prior
-work on debugging with data provenance information.  Unlike prior work, our
-work focuses on isolating a limited subset of dependence information that we
-can collect with run time overheads low enough for production use.
-Furthermore, we focus on {\em concurrent} program debugging.  Prior work on
-provenance debugging that focused mainly on single threaded programs and had
-dependence slicing overheads are probably too high for production.
-
-Bad Apple Tracking~\cite{badapples} is also a form of debugging with data
-provenance information. This work records the {\em origin} of an unusable value
-and can report it to a programmer when it is used in a way that causes a
-failure.
+Program slicing techniques identify useful subsequences of a program's
+instructions.   Slicing can work by analyzing program code statically or
+analyzing execution traces dynamically.  Many slicing techniques discussed in
+the a recent survey~\cite{tipslicingsurvey} monitor control and data
+dependences to identify which parts of a program might be relevant to a task
+({\em e.g., debugging}).
+
+Thin Slicing~\cite{thinslicing}, is a slicing technique that is related to our
+work.  Thin Slicing yields the set of statements that computed on or wrote
+values that influenced a particular variable's value (the ``seed'').  Thin
+Slicing is similar in that its slices provide provenance and it excludes
+control information like our work.  Thin Slicing is different in that it has
+overheads too high for production and focused on minimizing the size 
+of each slice.
 
@@
 
-CCI~\cite{cci}, Bugaboo~\cite{bugaboo}, Recon~\cite{recon}, and
-DefUse~\cite{defuse} all explicitly track inter-thread memory dependences at
-runtime.  Unlike our work, several of these techniques were not designed to run
-in production systems~\cite{recon,defuse}.  A distinguishing characteristic of
-our work is that it is effective and fast enough for production without the
-need to aggregate information from multiple different sampled executions, like
-CCI~\cite{cci}.  Additionally, our system does not require invasive hardware
-changes like Bugaboo~\cite{bugaboo} or LBA~\cite{paralog}.
-
-Other techniques~\cite{threadclustering,threadcriticality} use  hardware debug
-registers to monitor dependences efficiently enough for production use.  These
-techniques are similar to our work in that they track how threads interact.
-These techniques are different from our work in that they track information
-specific to the performance and our system provides general support.
+CCI~\cite{cci}, Bugaboo~\cite{bugaboo}, LBA~\cite{paralog}, Recon~\cite{recon},
+and DefUse~\cite{defuse} all explicitly track some thread interactions.  Unlike
+our work, several of these techniques were not designed to run in production
+systems~\cite{recon,defuse}.  A distinguishing characteristic of our work is
+that it is effective and fast enough for production without the need to
+aggregate information from multiple different sampled executions, like
+CCI~\cite{cci} does.  Additionally, our system does not require invasive
+hardware changes like Bugaboo~\cite{bugaboo} or LBA~\cite{paralog}.
 
 \paragraph{Extensible Program Instrumentation} 
-There are many systems for building analyses using extensible program
-instrumentation.  Binary instrumentation
-systems~\cite{pin,dynamorio,valgrind,roadrunner} provide completely general
-support to instrument arbitrary code -- instrumenting a program to track
-communciation is possible in such systems.
-
-These systems are similar to our work in that they aim to enable dynamic
-analyses and runtime systems.  They are different from our work because they
-rely on dynamic rewriting, yielding overheads often too high for deployment.
-Instead, we trade generality for performance. Last writer slices and \ctraps
-provide extensible support in the restricted domain of tracking communication
-with overheads low enough for production.  Another difference is that our work
-handles some of the delicate, performance sensitive code required to track
-communication.  In general instrumentation systems, programmers write that code
-from scratch at their own risk.
+There are many systems for building analyses using program instrumentation.
+Binary instrumentation~\cite{pin,dynamorio,valgrind,roadrunner} is general and
+tracking provenance and communciation is possible in such systems.
+
+These systems are similar to \ctraps in that they enable dynamic analyses.
+They are different from our work because they rely on dynamic binary
+translation, yielding overheads often too high for deployment.  We trade some
+generality for performance. \ctraps is effective mainly in the restricted
+domain of concurrency analyses, but has overheads low enough for production.
+Another difference is that our work handles some of the delicate, performance
+sensitive code required concurrency analyses, like last writer tracking.  In
+general instrumentation systems, programmers write that code from scratch at
+their own risk.
 

### 2013-11-13 paper.tex
committing before lunch.  need to do a read through.  Then, need to add some more debugging data points.  Is there time?

@@
 Section~\ref{sec:eval:debugging} looking at a set of buggy programs and
-comparing to prior, related work.  We evaluate our performance overheads and
+comparing to prior work.  We evaluate our performance overheads and
 show that last writer slices and \ctraps are feasible for use in deployed
 systems for a majority of workloads we studied.  We then characterize our
-instrumentation and discuss the sources of our system's run time overhead.   We
+instrumentation and discuss sources run time overhead.   We
 show that \ctraps is useful by implementing two existing applications and
@@
 %Section~\ref{sec:eval:debugging:cases}.
-
 Our evaluation shows that bad value origin tracking~\cite{badapples} is not
-helpful in three out of five cases where last writer slices is
-useful.  In one case, a use-after-delete bug in {\tt pbzip2}, bad value origin
-tracking helps.  For this bug, when one thread deletes a variable, it makes
-that variable unusable and bad value origin tracking would track that value.
-When the failure occurs, the programmer would then know why the value was
-deleted.
-
-In the remaining case, {\tt apache}, bad value origin tracking may help.  The
-bug is an atomicity violation that leads to a double free.  Two threads
-manipulate an object's reference count and deallocate the object when the count
-hits zero. Zero-checks and deallocation should be atomic, but the code does not
-ensure they are.  If threads interleave their code, multiple threads might see
-zero reference counts, which is incorrect.  When that happens, both threads try
-to free the object and the second free causes a crash.  Bad value origin
-tracking may help in this case.  At the first free, the freed pointer becomes
-unusable and its origin is tracked.  At the crash, the bad value's reported
-origin is the code point that frees the object.  That information potentially
-helps debug the program by suggesting a double free.  However, in {\tt apache},
-there are many threads, some perform different tasks, and all may free that
-object at that code point.  Bad value origin tracking does not record the
-thread of the origin.  The programmer is left unsure of whether the problem is
-with single-threaded logic or with synchronization.  In contrast, last writer
-slices reports the freed value's origin's thread and code point, making it
-clear that the programmer is dealing with a concurrency bug, and which threads
-are involved.
-
-\subsubsection{Case Studies}
+helpful in three out of five cases where last writer slices is useful.  
+There are two cases where bad value origin tracking may help.  For a
+use-after-delete bug in {\tt pbzip2}, bad value origin tracking helps.  For
+this bug, when one thread deletes a variable, it makes that variable unusable
+and bad value origin tracking would track that value.  When the failure occurs,
+the programmer would then know why the value was deleted.
+
+For {\tt apache}, bad value origin tracking may help.  The bug is an atomicity
+violation that leads to a double free.  Two threads manipulate an object's
+reference count and deallocate the object when the count hits zero. Zero-checks
+and deallocation should be atomic, but the code does not ensure they are.  If
+threads interleave their code, multiple threads might see zero reference
+counts, which is incorrect.  When that happens, both threads try to free the
+object and the second free causes a crash.
+
+Bad value origin tracking may help in this case.  At the first free, the freed
+pointer becomes unusable and its origin is tracked.  At the crash, the bad
+value's reported origin is the code point of the free. That information may
+help debug the program by suggesting a double free.  However, in {\tt apache},
+many threads perform different tasks and all call the same free code.  Bad
+value origin tracking does not record the thread information.  The programmer
+is left unsure whether the problem is with single-threaded logic or
+synchronization.  In contrast, last writer slices reports the code point and
+thread, making it clear the bug is a concurrency bug and identifying the
+involved threads.
+
+\subsubsection{Debugging Case Studies}
 \label{sec:eval:debugging:cases}
@@
 
-\paragraph{JDK-StringBuffer}
+\paragraph{jdk-1.4}
 The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:jdklws},
@@
 
-\paragraph{HTTrack}
+\paragraph{httrack}
 Figure~\ref{fig:httlws} shows how last writer slices help understand the root
@@
 
-\paragraph{MySQL}
+\paragraph{mysql}
 

### 2013-11-13 paper.tex
committing more changes.  Up through page 6 or so.

@@
 challenge to building multi-threaded software because some thread interactions
-have undesirable consequences, like causing a crash or degrading performance.
+have undesirable consequences, like crashing or degrading performance.
 
@@
 value could have originated in any of a large number of threads, making
-reasoning about why a variable contains that value complex and difficult. In
-summary, the challenges in using provenance to debug concurrent programs are:
-(1) collecting the right information to characterize the concurrent execution
-behavior; and (2) making it efficient enough for general programs running in
-production settings.
+reasoning about why a variable contains that value complex and difficult.
+
+%In
+%summary, the challenges in using provenance to debug concurrent programs are:
+%(1) collecting the right information to characterize the concurrent execution
+%behavior; and (2) making it efficient enough for general programs running in
+%production settings.
 
 In this work we address the challenges above and develop a novel form of
-in-production provenance tracking for concurrent programs. We have three
-goals. (1) To collect a general form of provenance information, not limited to
-only certain values. (2) To provide provenance information that helps
-understand how threads interact, which is the key to understanding concurrency
-bugs. (3) To work with overheads low enough for use in production (ideally,
-less than 10\%).
-
-We develop a novel provenance tracking mechanism called {\em last writer
+in-production provenance tracking for concurrent programs. We have three goals.
+(1) To collect a general form of provenance information, not limited to only
+certain values. (2) To provide provenance information that helps understand how
+threads interact, which is the key to understanding concurrency bugs. (3) To
+work with overheads low enough for use in production (ideally, less than 10\%),
+especially for long-running server programs that experience a diversity of production behavior.
+
+We develop a new provenance tracking mechanism called {\em last writer
 slicing} that satisfies these goals.   As the program executes in production,
@@
 
-Last writer slices are primarily useful for concurrency debugging, but we also
-observe that provenance information in the form of last writer slices is a
+Last writer slices are primarily useful for concurrency debugging, but we 
+observe that provenance information is a
 useful foundation for implementing other concurrent program analyses.  Several
-tools from prior work has used {\em ad hoc} last-writer information as part of
+tools from prior work have used {\em ad hoc} last-writer information as part of
 more heavy-weight analyses in domains like bug detection ({\em e.g.},
@@
 
-\item{We describe last writer slicing, an novel data provenance tracking
+\item{We describe last writer slicing, a novel data provenance tracking
 mechanism, that produces information that is useful for debugging concurrent
@@
 
-Data provenance information is directly useful during program debugging.  prior
+Data provenance information is directly useful during program debugging.  Prior
 work~\cite{whylinechi,whylineicse} showed that when programmers debug
@@
 and all of these analyses benefit from a general, high-performance
-communication tracking framework.
-Unfortunately, many of these techniques are implemented using heavy-weight infrastructure, like binary instrumentation~\cite{pin}, making them inappropriate for use in production.
+communication tracking framework.  Unfortunately, many of these techniques are built using heavy-weight infrastructure, like binary instrumentation~\cite{pin}, with overheads too high for production.
 
@@
 side of the figure shows a program execution with three threads.  On the left
-is the \lwt at the point of each memory access.  There are two key things to
+is the \lwt at the point of each memory access.  There are three key things to
 notice.  First, write operations update the \lwt entry of the memory location
-they access.  Second, there is no action necessary for read operations.
-Making read operations cheap is key to making the overhead of last writer
-slicing low enough for deployment use.
+they access.  Second, there is no action necessary for read operations.  Making
+read operations cheap is key to making the overhead of last writer slicing low
+enough for production use.  Third, if the program stops at a point like the
+{\tt Rd Y} illustrated in Thread 1, the \lwt holds provenance information for
+the involved values ({\em i.e.}, Thread 3 last wrote {\tt Y} at the code point
+shown). That information helps understand {\em why} {\tt Y} holds its value.
 
@@
 Instead, a programmer can examine an \lwt entry at any arbitrary point in an
-execution ({\em e.g.}, at a breakpoint), including after a failure, and see a
-value's provenance.  Including the \lwt with core dumps means provenance is
-available when post-mortem debugging production failures.
+execution and see a value's
+provenance ({\em e.g.}, at a breakpoint or after a failure).  Including the \lwt with core dumps means provenance is available
+when post-mortem debugging production failures.
 
@@
 
-The \lwt implemented in the runtime library is a {\em shadow memory} that
-holds one entry for each potentially shared memory address accessed by the
-program.  Each \lwt entry records the thread and program point that last wrote
-to the entry's corresponding memory location.  When a thread performs a write
-to a memory location, it updates the \lwt by recording its thread identifier
-and the program point of its write operation to the memory location's \lwt
-entry.  A program point may be an instruction address, or a call stack.  We use
-instructions by default, but we experimented with 2-address bounded call-stacks
-as well.
+The \lwt implemented in the runtime library is a {\em shadow memory} that holds
+one entry for each potentially shared memory address accessed by the program.
+Each \lwt entry records the thread and program point that last wrote to the
+entry's corresponding memory location.  When a thread performs a write to a
+memory location, our runtime library's \lwt update function updates the \lwt by
+recording the thread's identifier and the program point of the write operation
+in the memory location's \lwt entry.  A program point may be an instruction
+address or a call stack.  We use instructions by default, but we also experimented
+with 2-address call stacks.
 
 Collecting last writer slices requires instrumentation on each write operation.
-We built an instrumenting compiler pass that inserts a call to an \lwt update
+We built an instrumenting compiler pass that inserts a call to the \lwt update
 function in our runtime library just before each write that may access data
-last written by another thread (escaping memory accesses).  Write
-instrumentation calls are passed three pieces of information: the address of
+last written by another thread (escaping memory accesses).  \lwt update 
+calls are passed three pieces of information: the address of
 the memory word involved in the access, the thread identifier of the accessing
@@
 inaccessible to code running in other threads.  Such operations access unshared
-data and they are not instrumented by our compiler.
+data and we do not instrument them.
 
@@
 $P$, our compiler pass transforms it to a new program, $P'$.  Each thread in
-$P'$ executes a modified sequence of instructions: $P'_{t} \in P' = \{
-I(p^{t}_{1}), p^{t}_{1}, \ldots, I(p^{t}_{k_{t}}), p^{t}_{k_{t}} \}$.
+$P'$ executes a modified sequence of instructions: $P'_{t} \in P' = ( 
+I(p^{t}_{1}), p^{t}_{1}, \ldots, I(p^{t}_{k_{t}}), p^{t}_{k_{t}} )$.
 $I(p^{t}_{j})$ sets $LWT[m] = (t,p^{t}_{j})$ if and only if $p^{t}_{j} =
@@
 to the trap hook, as well as the \lwt's record of the program point and thread
-identifier of the last write to the involved memory location.
+identifier of the last write to the involved location.
 
@@
 as a plugin for GCC ({\tt gcc-4.7.0 20111112}).  We used GCC's points-to and
-escape analysis support to prune instrumentation to non-escaping memory
+escape analysis support to prune instrumentation on accesses to non-escaping memory
 locations.  Our compiler instruments calls to {\tt free} and {\tt delete} as
@@
 last writer slices provide information that is useful during debugging in
-Section~\ref{sec:eval:debugging} looking at a set of buggy programs and
+Section~\ref{sec:eval:debugging}, looking at a set of buggy programs and
 comparing to prior work.  We evaluate our performance overheads and
@@
 systems for a majority of workloads we studied.  We then characterize our
-instrumentation and discuss sources run time overhead.   We
+instrumentation and discuss sources of run time overhead.   We
 show that \ctraps is useful by implementing two existing applications and
@@
 
-In Table~\ref{tab:bugs}, we show the bugs we studied describing the program,
+In Table~\ref{tab:bugs}, we show the bugs we studied, describing the program,
 bug type, and bug report identifier where applicable.  For each bug, we
@@
 evaluation criteria - (1) {\bf effectiveness}, (2) {\bf triviality}, and (3)
-{\bf usefulness}.  Our technique is {\bf effective} if the last writer slice
-reports a code point that wrote a value that subsequently caused a failure.  A
-failure is {\bf trivial} to debug if its cause is made obvious by examining the
-core dump alone, {\em i.e.}, a last writer slice is unnecessary.  A last writer
-slice is {\bf useful} if it is effective and the thread and code point in the
-slice help understand why the failure occurred or how to fix it.   We directly
-compare to bad value origin tracking.    The {\bf O.T.} column in
-Table~\ref{tab:bugs} shows  whether their analysis could help debug each bug,
-based on the description of their technique.
+{\bf usefulness}.  Our technique is {\bf effective} if it tracks a correct last
+writer slice for the value or values involved in causing the failure.  A failure is {\bf
+trivial} to debug if its cause is made obvious by examining the core dump
+alone, {\em i.e.}, a last writer slice is unnecessary.  A last writer slice is
+{\bf useful} if it is effective and the thread and code point in the slice help
+understand why the failure occurred or how to fix it.   We directly compare to
+bad value origin tracking.    The {\bf O.T.} column in Table~\ref{tab:bugs}
+shows  whether their analysis could help debug each bug, based on the
+description of their technique.
 

### 2013-11-13 paper.tex
committing after a read through.  I think it's getting there.

@@
 
-\begin{table}
+The data show that escape analysis can prove a substantial fraction of
+operations are local and do not require instrumentation.  Across PARSEC,
+between 8\% and 30\% of accesses were proved local.  In the server programs, a
+larger proportion of accesses were able to be proved local (21\% -- 66\%).  The
+difference is likely due to the fact that server threads perform more
+independent work (as discussed in Section~\ref{sec:eval:parsecserver}).
+Eliminating instrumentation using escape analysis is key to high performance.
+
+\begin{table}[htb]
 \centering
@@
 
-The data show that escape analysis can prove a substantial fraction of
-operations are local and do not require instrumentation.  Across PARSEC,
-between 8\% and 30\% of accesses were proved local.  In the server programs, a
-larger proportion of accesses were able to be proved local (21\% -- 66\%).  The
-difference is likely due to the fact that server threads perform more
-independent work (as discussed in Section~\ref{sec:eval:parsecserver}).
-Eliminating instrumentation using escape analysis is key to high performance.
-
 \section{Related Work}
@@
 
-We have discussed bad value origin tracking ~\cite{badapples} throughout this
+We have discussed bad value origin tracking~\cite{badapples} throughout this
 work and compared directly with that technique in our evaluation.  Bad value
@@
 Slicing is similar in that its slices provide provenance and it excludes
-control information like our work.  Thin Slicing is different in that it has
+most control information like our work.  Thin Slicing is different in that it has
 overheads too high for production and focused on minimizing the size 
-of each slice.
+of each slice compared to prior slicing techniques.
 

### 2013-11-14 paper.tex
committing more in the debugging.stuff file and more text, describing Apache2

@@
 
-\paragraph{Other Applications}
-In addition to those applications, we envision that a variety of applications
-will be enabled by our simple interface.  Anomaly detection and ``likely
-invariant'' techniques similar to AVIO~\cite{avio}, Daikon~\cite{daikon}, and
-others are a natural fit for \ctraps.  \ctraps provides the support necessary
-for implementing specification checkers for specifications related to
-inter-thread communication~\cite{velodrome,oshajava}.  Guided testing
-tools~\cite{cuzz,chess} and bug detection tools~\cite{ctrigger} focused on
-concurrency could also be implemented as \ctraps applications.  \ctraps also
-provides an interesting platform for developing new approaches to
-record/replay~\cite{chimera,fdr}.  \ctraps is an enabling platform for
-developing these or other analyses with low baseline overheads and without the
-need to build an entire concurrency analysis from scratch.
+%\paragraph{Other Applications}
+%In addition to those applications, we envision that a variety of applications
+%will be enabled by our simple interface.  Anomaly detection and ``likely
+%invariant'' techniques similar to AVIO~\cite{avio}, Daikon~\cite{daikon}, and
+%others are a natural fit for \ctraps.  \ctraps provides the support necessary
+%for implementing specification checkers for specifications related to
+%inter-thread communication~\cite{velodrome,oshajava}.  Guided testing
+%tools~\cite{cuzz,chess} and bug detection tools~\cite{ctrigger} focused on
+%concurrency could also be implemented as \ctraps applications.  \ctraps also
+%provides an interesting platform for developing new approaches to
+%record/replay~\cite{chimera,fdr}.  \ctraps is an enabling platform for
+%developing these or other analyses with low baseline overheads and without the
+%need to build an entire concurrency analysis from scratch.
 
@@
 {\bf App} & {\bf Ver. \#} & {\bf Bug \#} &{\bf Type} & {\bf Eff.?} & {\bf Triv.?} & {\bf Usef.?} & {\bf O.T.?}\\ \hline
-mysql  &  4.0.12 &  791     & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
-pbzip2 &  0.9.4  &  N/A     & Order  &Yes  &No    & Yes   & Yes \\
-httrack& 3.43.9  &  N/A     & Order  &Yes  &No    & Yes   & No  \\%Used w/o origin - no help
-jdk1.4 &  1.70   &          & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
-apache&  2.0.48 &  21285   & Atom.  &Yes  &No    & Yes   & Maybe\\%All values usable
-%apache2&  2.2.9  &  45605   & Atom.  &Yes  &No    & Maybe & ??  \\
+jdk1.4 &  1.70   &  N/A     & Atom.  &Yes  &No    &Yes   & No  \\%All values usable
+pbzip2 &  0.9.4  &  N/A     & Order  &Yes  &No    &Yes   & Yes \\
+httrack& 3.43.9  &  N/A     & Order  &Yes  &No    &Yes   & No  \\%Used w/o origin - no help
+transmission & 1.42 & N/A   & Order  &Yes  &No    &Yes   & No  \\
+mysql  &  4.0.12 &  791     & Atom.  &Yes  &No    &Yes   & No  \\%All values usable
+apache1&  2.0.48 &  21285   & Atom.  &Yes  &No    &Yes   & Some \\%All values usable
+apache2&  2.2.9 &  45605    & Atom.  &Yes  &No    &Some  & No \\%All values usable
 \end{tabular}
@@
 
-For {\tt apache}, bad value origin tracking may help.  The bug is an atomicity
+For {\tt apache1}, bad value origin tracking may help.  The bug is an atomicity
 violation that leads to a double free.  Two threads manipulate an object's
@@
 
+\paragraph{apache2}
+We used last writer slices to debug an atomicity violation in Apache-2.2.9 and
+we found that last writer slices help understand the bug, but not how to fix
+it.  The bug involves a group of worker threads that handle web requests and a
+network listener thread that passes requests to the workers.  The queue tracks
+the number of idle workers.  When a worker becomes idle, it increments the
+number of idlers and when the listener passes work to a worker, it decrements
+that number.  The counter's increments and decrements are synchronized using
+condition variables.  If there are no idlers, the listener waits for a signal.
+When a worker becomes idle when no others are, it delivers a signal to the
+listener.
+
+The problem is that the listener does not check the condition governed by the
+condition variable after being signaled on that condition variable.  As a
+result, the listener can incorrectly decrement the idler count twice,
+underflowing the unsigned value, which then causes an assertion to fail.  When
+the failure occurs, the core dump shows the idler count underflowed, revealing
+a primary symptom of the bug, but not providing much help.  We checked the last
+writer slice for the idler count and it reported that the listener thread last
+wrote the idler count at its decrement.  That information shows the programmer
+{\em why} the underflow occurred, which is helpful for understanding the
+failure.  However, we conclude that last writer slices are only somewhat 
+useful in this case.  Solving this complex bug requires understanding not just
+the connection between the underflow and the listener's decrement, but also the
+synchronization protocol that coordinates that decrement and other accesses to
+the variable.
+
 \subsection{Performance Evaluation: Setup and Benchmarks}

### 2013-11-15 paper.tex
added transmission case.  talked about comparative debugging.  seven benchmarks now

@@
 We illustrate the debugging benefits of last writer slices by using our system
-to debug five real-world programs studied previously in the debugging
+to debug several real-world programs studied previously in the debugging
 literature.  Table~\ref{tab:bugs} provides an overview of this evaluation and
@@
 understand why the failure occurred or how to fix it.   We directly compare to
-bad value origin tracking.    The {\bf O.T.} column in Table~\ref{tab:bugs}
+bad value origin tracking~\cite{badapples}.    The {\bf O.T.} column in Table~\ref{tab:bugs}
 shows  whether their analysis could help debug each bug, based on the
@@
 violation bugs, last writer slices are effective.  In all cases, we found that
-the last writer slice revealed the provenance of the data involved in the
-operation that caused the failure, so last writer slicing was effective.  We
-found that none of the failures we looked at were trivially debuggable.  This
-finding is consistent with the difficulty of debugging concurrency bugs -- the
-core dump does not reveal the interaction between threads that led to a
-failure, only the result of the interaction, which leads to the failure.
-
-Last writer slices were obviously useful in all five cases.  In these cases,
-the root cause of the bug becomes clear when last writer slices show the
-inter-thread data flow led to a variable having a value that caused a failure.
-Using that information, we expect it would be straightforward in these cases to
-add synchronization to fix these bugs.  Note that without last writer slices,
-understanding the responsible inter-thread data flow and fixing the bug would
-be difficult.
+the last writer slice revealed the provenance of at least some data involved in
+the failure, so last writer slicing was effective.  We found that none of the
+failures we looked at were trivially debuggable.  This finding is consistent
+with the difficulty of debugging concurrency bugs -- the core dump does not
+reveal the interaction between threads that led to a failure, only the result
+of the interaction, which leads to the failure.
+
+Last writer slices were useful for helping understand how to fix the bug in
+all but one case and useful at least for understanding the bug in the
+remaining case, {\tt apache2}.  The main reason last writer slices are helpful
+is that they reveal the connection between code running in different threads
+that interacts to cause a failure.  Without last writer slices, programmers
+have no easy way of connecting these otherwise disparate parts of a program.
+The case studies in Section~\ref{sec:eval:debugging:cases} show in more detail
+how last writer slices are useful for debugging.
 
@@
 %Section~\ref{sec:eval:debugging:cases}.
-Our evaluation shows that bad value origin tracking~\cite{badapples} is not
-helpful in three out of five cases where last writer slices is useful.  
-There are two cases where bad value origin tracking may help.  For a
-use-after-delete bug in {\tt pbzip2}, bad value origin tracking helps.  For
-this bug, when one thread deletes a variable, it makes that variable unusable
-and bad value origin tracking would track that value.  When the failure occurs,
-the programmer would then know why the value was deleted.
+Our evaluation shows that bad value origin tracking~\cite{badapples} is
+definitely helpful in one case where last writer slices is useful and may help
+in one other case.  For a use-after-delete bug in {\tt pbzip2}, bad value
+origin tracking helps.  For this bug, when one thread deletes a variable, it
+makes that variable unusable and bad value origin tracking would track that
+value.  When the failure occurs, the programmer would then know why the value
+was deleted.
 
@@
 We now provide detailed cases studies showing the debugging role of last writer
-slices in several of the bugs we examined.  Space constraints prevent us from
-providing details for all cases.
+slices in several of the bugs we examined.
+
+%Space constraints prevent us from providing details for all cases.
 
@@
 
+\paragraph{transmission} We used last writer slices to debug a
+use-before-initialization bug in transmission-1.42.  We omit a full case study
+due to space constraints, but we mention a few notable properties of this case.
+First, like {\tt httrack}, we debug {\tt transmission} by comparing the last
+writer slice from failing and non-failing executions.  Second, last writer
+slices reveal all the code and data involved in this bug, making the bug and
+its fix clear.   Third, bad value origin tracking does not work in this case
+because in failing executions, the bad value has no origin and in non-failing
+executions, the involved variable holds only usable values.
+
 %
@@
 results in Table~\ref{tab:char}.  The table shows the percent of
-instrumentation sites pruned and the absolute number of sites in parenthesis.
+calls pruned and the absolute number of sites in parenthesis.
 
@@
 \paragraph{Debugging with Data Provenance Information}
-The most related prior work looked at using various forms of provenance information for debugging.
+Prior work has looked at various forms of data provenance for debugging.
 
@@
 work.  Thin Slicing yields the set of statements that computed on or wrote
-values that influenced a particular variable's value (the ``seed'').  Thin
-Slicing is similar in that its slices provide provenance and it excludes
-most control information like our work.  Thin Slicing is different in that it has
-overheads too high for production and focused on minimizing the size 
-of each slice compared to prior slicing techniques.
+values that influenced a ``seed'' variable's value.  Thin
+Slicing is similar to our work in that slices provide provenance and it excludes most
+control-flow information.  Thin Slicing is different in that it has
+overheads too high for production and focused on minimizing the size of each
+slice compared to prior slicing techniques.
 
@@
 \section{Conclusions}
-This work observes that we can collect a useful form of data provenance
-information with overheads low enough for use in deployed systems.
+Last writer slices are the first system to collect a general form of data provenance information for concurrent programs with overheads low enough for production use.
 

### 2013-11-15 paper.tex
committing a draft that includes the data from the collisions experiment.  back on 10 pages.  fingers crossed.

@@
 variable contain this value?", which is central to debugging. Debugging a
-concurrent program requires understanding how threads interact, and since
+concurrent program requires understanding how threads interact, and because  
 communication happens via reads and writes to shared variables, {\em data
@@
 
-Collecting debugging information, like data provenance, from production
+Efficiently collecting debugging information, like data provenance, from production
 systems is particularly important for concurrent software since bugs
@@
 systems.  We then provide context for \ctraps by discussing some more
-sophisticated concurrency analyses that rely on the kind of dependence
-information we are focusing on in this work.
+sophisticated concurrency analyses that rely on last writer information.
 
@@
 \caption{\label{fig:jdklws}{\bf Using last writer slices to debug an
-ordering violation in the JDK 1.4 StringBuffer library.} The last writer slice reveals the interaction between the erase function and the append function that leads to a failure.}
+ordering violation in the JDK 1.4 StringBuffer library.} The last writer slice reveals the buggy interaction between {\tt erase()} and {\tt append()}.}
 \end{figure}
@@
 real-world inputs.  A primary goal of this work is to show that we can collect
-last writer slices with overheads that are low enough for production use so
-they can be included pervasively in bug reports.
+last writer slices with very low overheads so they can be included pervasively
+in bug reports.
 
@@
 executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
-in $Q$ to write $m$ and $t$ is the thread that executed it.  For a
-never-written location $n$, $LWS[n] = (\emptyset,\emptyset)$ indicating it is
-unwritten.
+in $Q$ to write $m$ and $t$ is the thread that executed it.  For an
+unwritten location $n$, $LWS[n] = (\emptyset,\emptyset)$.
 
@@
 execution.  Correctness in the \lwt requires {\em soundness} and {\em
-completeness}.  A {\em sound} design only updates $LWT[m]$ when $m$ is written.
+completeness}.  A {\em sound} design updates $LWT[m]$ only when $m$ is written.
 A {\em complete} design always updates $LWT[m]$ when $m$ is written.  Our
@@
 Next we discuss two nuances of our design: (1) the interaction of our \lwt
-implementation with code containing data-races; and (2) the correctness of the
-\lwt when a program addresses data that overlap in memory.
+implementation with data-races; and (2) the correctness of the
+\lwt for accesses to overlapping memory regions.
 
@@
 \paragraph{Non-overlapping Memory References} 
-We assume each written memory location corresponds to a non-overlapping memory
-region because our \lwt design is {\em granularity oblivious}.  We discuss this
-issue informally due to space constraints.  Each \lwt entry corresponds to a
-memory location referenced by a program instruction.  Two instructions might
-reference different locations that overlap because the referenced locations
-have different granularities ({\em e.g.}, accessing a byte that is part of a
-previously accessed word).  When that happens, both accesses update different
-\lwt entries, because they reference different locations.  However, the
-overlapping region then has two different entries reflecting the last {\em two}
-updates to its parts rather than just its last write.
+We assume written memory locations do not overlap in memory because our \lwt
+design is {\em granularity oblivious}.  We discuss this issue informally due to
+space constraints.  Each \lwt entry corresponds to a location that was written.
+Two instructions might reference different locations that overlap because the
+referenced locations have different granularities ({\em e.g.}, accessing a byte
+that is part of a previously accessed word).  When that happens, both accesses
+update different \lwt entries, because they reference different locations.
+However, the overlapping region then has two different entries reflecting the
+last {\em two} updates to its parts rather than just its last write.
 
@@
 entry in the \lwt that correctly reflects each memory location's last writer.
-However, the entry for a coarse-grained region may be obsolete because of an
-access to an overlapping sub-region.
+However, the coarse region's entry may be stale because of an
+access to an overlapping region.
 
@@
 recorded edges is the tool's output.  Our \ctraps implementation of
-communication graphs took about 50 lines of code (including output formatting
-code).
+communication graphs took about 50 lines of code.
 
@@
 \section{System Implementation}
-
+\label{sec:implementation}
 We implemented last writer slicing and \ctraps.  We built our compiler support
-as a plugin for GCC ({\tt gcc-4.7.0 20111112}).  We used GCC's points-to and
+as a plugin for GCC ({\tt gcc-4.7}).  We used GCC's points-to and
 escape analysis support to prune instrumentation on accesses to non-escaping memory
@@
 
-We built our runtime system from scratch.  We implemented the LWT as a fixed
-size 2GB array of 64-bit words, fixing our memory overheads.  Each entry packs a program point and a thread
-identifier into the 64-bit word.  By default, program points are single
-instruction addresses, but we also support limited (2-address)
-context-sensitivity by packing the current instruction's address and the
-nearest return address into a 64-bit LWT entry. When an access to a memory
-location occurs, we index into the LWT with the lower bits of the location's
-address.  Our prototype implementation uses a lossy resolution policy for hash
-collisions.  We use this policy in our prototype because it is faster and
-simpler than a chaining concurrent hash table implementation.  Collisions are
-extremely rare, so this simplification is unlikely to be a problem even in a
-more industrial strength implementation.
+We built our runtime system from scratch.  The LWT is a fixed size 2GB array of
+64-bit words.  Each entry packs a program point and a thread identifier into
+the 64-bit word.  By default, program points are single instruction addresses,
+but we also support limited (2-address) context-sensitivity by packing the
+current instruction's address and the nearest return address into a 64-bit LWT
+entry. When an access to a memory location occurs, we index into the LWT with
+the lower bits of the location's address.  Our prototype implementation uses a
+lossy resolution policy for hash collisions.  We use this policy in our
+prototype because it is fast and bounds memory overheads at 2GB.  Collisions
+are rare (see Section~\ref{sec:eval:performance:collisions}), so this
+simplification is unlikely to be a problem.
 
@@
 Since its release, we have seen contributions to our codebase from members of
-the GCC development community and downloads from interested developers.
+the GCC community and downloads from interested developers.
 
@@
 reveal the interaction between threads that led to a failure, only the result
-of the interaction, which leads to the failure.
+of the interaction.
 
@@
 The case studies in Section~\ref{sec:eval:debugging:cases} show in more detail
-how last writer slices are useful for debugging.
+how last writer slices are useful for debugging and illustrate their limitations.
 
@@
 Our evaluation shows that bad value origin tracking~\cite{badapples} is
-definitely helpful in one case where last writer slices is useful and may help
+helpful in one case where last writer slices is useful and provides limited help
 in one other case.  For a use-after-delete bug in {\tt pbzip2}, bad value
@@
 atomic, but the lack of synchronization allows Thread 2 to run its insert
-during log rotation.  In that case, the insert query not logged, manifesting
+during log rotation.  In that case, the insert query is not logged, manifesting
 the failure.  We added an {\tt abort} statement to the non-logging case because
@@
 We used last writer slices to debug an atomicity violation in Apache-2.2.9 and
-we found that last writer slices help understand the bug, but not how to fix
+we found that last writer slices help understand the bug, but not fully how to fix
 it.  The bug involves a group of worker threads that handle web requests and a
-network listener thread that passes requests to the workers.  The queue tracks
+network listener thread that passes requests to the workers.  The server tracks
 the number of idle workers.  When a worker becomes idle, it increments the
@@
 the failure occurs, the core dump shows the idler count underflowed, revealing
-a primary symptom of the bug, but not providing much help.  We checked the last
-writer slice for the idler count and it reported that the listener thread last
-wrote the idler count at its decrement.  That information shows the programmer
+a primary symptom of the bug, but not providing much help.  The last
+writer slice for the idler count reports that the listener thread last
+wrote the idler count at its decrement operation.  That information shows the programmer
 {\em why} the underflow occurred, which is helpful for understanding the
@@
 
-These high-level results support our main claim that the runtime overhead of
-\ctraps is low enough for deployed systems for many applications, especially
+These high-level results support our claim that our overheads 
+are low enough for deployed systems for many applications, especially
 for servers.  We now further analyze our observed overheads, illuminating
@@
 inner-loop's caller, but is never shared.  Eliminating instrumentation on
-accesses to this matrix yielded an overhead of about 46\% for \ctraps, and to
+accesses to this matrix yielded an overhead of 46\% for \ctraps and 
 just 30\% for last writer slicing.
@@
 for these workloads from being unacceptable for any production environment to
-being reasonable for use in production systems.  We expect a more sophisticated
-escape analysis would find these local accesses and achieve these performance
-gains automatically.
+being reasonable for use in production systems. Better escape analysis would
+find these performance gains automatically.
 
@@
 \label{sec:eval:parsecserver}
-In our tests, server programs were better able to tolerate the overhead of
-last writer slices and \ctraps.   We discern three reasons for this difference.
+We discern three reasons that server programs better tolerated the overheads of
+last writer slices and \ctraps in our experiments.
 
@@
 data show that a restricted form of context-sensitive analysis is possible in
-\ctraps with overheads low enough for production.
+\ctraps with overheads low enough for production in some cases.
+
+\subsection{Memory Overheads}
+\label{sec:eval:performance:collisions}
+As we describe in Section~\ref{sec:implementation} our prototype \lwt is a 2GB
+hashtable of 8 byte entries.  We use a lossy hash collision policy: all updates
+always overwrite existing entries.  This strategy fixes memory overheads at
+2GB, but frequent lossy collisions may lead to imprecision.   We instrumented
+our runtime to count collisions and found that for PARSEC applications the
+geometric mean rate of collisions was very low - 89 per 10,000 memory accesses.
+This summary result shows that our prototype implementation is reasonable,
+especially for debugging in production, when low memory use is key and minor
+imprecision is tolerable.  Note that collisions did not introduce noticable
+imprecision in our debugging experiments.  When precision is more important than memory overhead, ({\em
+e.g.}, for sound, offline analysis), a lossless hashtable is a better option.
 
@@
 
-\subsubsection{Performance with \ctraps}
+\subsubsection{Performance with \ctraps Applications}
 
@@
 \section{Conclusions}
-Last writer slices are the first system to collect a general form of data provenance information for concurrent programs with overheads low enough for production use.
+In this work we proposed last writer slicing, the first technique to collect data provenance information
+that is not limited to certain values, is targeted to concurrent programs, and
+has overheads low enough for production use, especially in server programs.
 
 We developed system support that collects last writer slices with overheads
-well within an acceptable range for production.  We showed that, during
-debugging, the provenance information provided by last writer slices reveals
-crucial connections between different threads' otherwise disparate regions of
-code.  Understanding how code in different threads interacts is essential to
-concurrency debugging and this work helps with that understanding.
+well within an acceptable range for always-on use in production.  We showed
+that during debugging, the provenance information provided by last writer
+slices reveals crucial connections between different threads' otherwise
+disparate regions of code.  Understanding how code in different threads
+interacts is essential to concurrency debugging and this work aids that
+understanding.
 

### 2014-03-20 paper.tex
checking in the edits I made to the introduction, related work, and background sections yesterday

@@
-\documentclass[preprint,9pt]{sigplanconf}
+\documentclass[preprint,10pt]{sigplanconf}
 \usepackage{url}
@@
 
-\newcommand{\Caption}[1]{\begin{minipage}{1.0\columnwidth} \caption{#1} \end{minipage}\vspace{-2ex}}
+\newcommand{\Caption}[1]{\begin{minipage}{1.0\columnwidth} \caption{#1} \end{minipage}}
 \newcommand{\CaptionWide}[1]{\caption{#1}\vspace{-2ex}}
@@
 \section{Introduction}
-Multi-threaded programming is challenging.  In contrast to 
-sequential reasoning, multi-threaded programs require complex reasoning about
-many threads of control and their interactions.  In shared memory programs,
-threads interact by reading and writing memory locations.  Threads' reads and
-writes interleave arbitrarily and nondeterministically and each interleaving
-can yield different behavior.  Reasoning about thread interactions is a central
-challenge to building multi-threaded software because some thread interactions
-have undesirable consequences, like crashing or degrading performance.
+Multi-threaded programming is challenging.  In contrast to sequential
+reasoning, multi-threaded programs require complex reasoning about many threads
+of control and their interactions.  In shared memory programs, threads interact
+by reading and writing memory locations.  Threads' reads and writes interleave
+arbitrarily and nondeterministically. Each interleaving can lead to different
+behaviors and memory states, some of which may be undesirable, like crashes or
+data corruption.  Moreover, because complete concurrency testing is infeasiable
+in general, undesirable interleavings can evade testing and lead to production
+failures.
+
+The problem we are addressing in this work is that debugging support for
+understanding thread interactions is inadequate.  Today, programmers use a
+debugger to examine memory contents either during a debugging execution or by
+loading a core dump from a failed execution.  Examining the contents of memory
+reveals {\em what} the state of the program is ({\em e.g.}, the values of
+variables).  Unfortunately, the real question programmers must answer is {\em
+why} the state of the program is what it is.  Understanding why a program
+entered an undesirable state ({\em e.g.}, crashed) is the key to changing the
+code to prevent future executions from entering that state.  Multi-threading
+makes this understanding elusive because implicating particular operations in
+particular threads for a failure is a complex task.  Aiding developers in this
+understanding for failed production executions is important because
+hard-to-find bugs may manifest in production only.
 
@@
 
-A value's provenance is a description of the reason that it is stored in a
-variable. Provenance information helps answer the question: "Why does this
-variable contain this value?", which is central to debugging. Debugging a
-concurrent program requires understanding how threads interact, and because  
-communication happens via reads and writes to shared variables, {\em data
-provenance information is essential}. Provenance can take on many forms, for
-example, a value's provenance could be the instruction that wrote it, the
-function that returned it, and in the context of concurrent programs, the
-thread that produced it.
+Our approach to solving that problem is to provide the programmer with {\em
+provenance} information that can be examined alongside the program's memory
+state ({\em i.e.}, core dump).  A value's provenance describes why a memory
+location contains its value.  In this work we develop a new provenance tracking
+mechanism called {\em last writer slicing} that satisfies these goals.  A
+memory location's last writer slice is a record containing the program point
+and the identifier of the thread that last wrote data to that memory location.
+As the program executes in production, we collect last writer slices for each
+potentially shared memory location and those last writer slices record the
+provenance of the values stored in those locations.  Last writer slices are
+saved with a core dump when the program fails and are available in the debugger
+({\em e.g.}, for breakpoint debugging).
+
+We have three goals for last writer slicing:  (1) To collect a general form of
+provenance information, not limited to certain values or locations only. (2) To
+provide provenance information focused on how threads interact to aid in
+understanding concurrency bugs. (3) To engineer our system to have overheads
+low enough for use in production, aiming for a 10 -- 50\% maximum overhead.
 
@@
 %parts of a program that interact, leading to a failure.
-
-Efficiently collecting debugging information, like data provenance, from production
-systems is particularly important for concurrent software since bugs
-manifest as failures rarely. When a production system fails and a programmer
-receives a bug report, reproducing the reported failure to collect provenance
-information may be difficult or impossible. It is essential to collect as much
-information as performance margins permit from production, to reduce the need
-for time-consuming failure reproduction.
-
-Prior work has previously shown that some limited forms of provenance help
-with debugging~\cite{badapples, whylineicse, tipslicingsurvey}. One effort
-tracked code that wrote unusable values into variables~\cite{badapples}. That
-work showed that revealing the connection between an unusable value's origin
-and a failure it causes is helpful during debugging. Other work showed that
-programmers debug more effectively when they can ask ``why'' questions about
-values during debugging~\cite{tipslicingsurvey,whylineicse}.
-
-These efforts have shown some of the promise of provenance, but are not a
-debugging panacea and are insufficient for dealing with concurrent program
-bugs. First, prior work~\cite{badapples} has focused on providing provenance
-information for only some unusable {\em values}, like {\tt NULL}, which limits its applicability. 
-% Limiting the applicability of their technique to only some failures, like {\tt
-% Null Pointer Exceptions}.
-Second, while prior work has shown that limited provenance tracking for Java
-programs is possible with overheads low enough for production~\cite{badapples},
-a more general form of provenance tracking for C and C++ programs has been
-elusive. Prior approaches that build on slicing~\cite{thinslicing,
-tipslicingsurvey, whylineicse} or heavy-weight memory checkers~\cite{badapples}
-impose overheads too high for production for C and C++ programs. Third, no
-prior technique that we know of has collected provenance information that is
-tailored to debugging concurrent programs with overheads that are low enough
-for use in production systems. Addressing concurrency is important because a
-value could have originated in any of a large number of threads, making
-reasoning about why a variable contains that value complex and difficult.
 
@@
 %production settings.
-
-In this work we address the challenges above and develop a novel form of
-in-production provenance tracking for concurrent programs. We have three goals.
-(1) To collect a general form of provenance information, not limited to only
-certain values. (2) To provide provenance information that helps understand how
-threads interact, which is the key to understanding concurrency bugs. (3) To
-work with overheads low enough for use in production (ideally, less than 10\%),
-especially for long-running server programs that experience a diversity of production behavior.
-
-We develop a new provenance tracking mechanism called {\em last writer
-slicing} that satisfies these goals.   As the program executes in production,
-last writer slicing tracks the thread and program point that last wrote any
-value to each potentially shared memory location --- its provenance.  Last
-writer slices are saved with a core dump when the program fails and are
-available in the debugger ({\em e.g.}, for breakpoint debugging).
 
@@
 called {\em Communication Traps} or {\em \ctraps} on top of last writer slices
-(Section~\ref{sec:ctraps}). \ctraps can be used to implement several dynamic
-analysis tailored to concurrent programs, like the ones discussed above.
-\ctraps works by checking the last writer slice of each potentially shared
-memory location when a thread accesses such a location. If the accessing
-thread is different from the thread that last wrote the location, the access
-signifies {\em communication} between the threads. When communication occurs,
-\ctraps delivers a software {\em trap} to the accessing thread, giving it
-information about how communication occurred. The trap can be handled by an
-analysis-specific trap handler. Trap handlers implement {\em \ctraps
-applications} through an extensible interface that can interpose on the
-execution (Section~\ref{sec:apps}). By default, \ctraps has very low overheads
-and similarly to last writer slicing are appropriate for use in production in
-many cases. 
-%\ctraps applications can perform arbitrary concurrency analyses,
-%but at their own discretion and cost. 
-As we show in our evaluation, the
-overhead of \ctraps applications scales with the complexity of their analysis.
+(Section~\ref{sec:ctraps}). \ctraps is useful for implementing dynamic
+concurrency analyses, like the ones discussed above.  \ctraps runs alongside a
+program and uses last writer slices to identify memory accesses that lead to
+{\em communication} between the threads. Communcation occurs when a thread
+accesses memory last written by a different thread and \ctraps delivers a {\em
+trap} to a thread when it communicates with another thread, giving it
+information about how communication occurred.  Programmer-defined {\em trap
+handlers} can implement arbitrary analysis using that information. By default,
+\ctraps has low overheads that are appropriate production in many cases.  As we
+show in our evaluation, the overhead of \ctraps applications scales with the
+complexity of their analysis.
 
@@
 
-Data provenance information is directly useful during program debugging.  Prior
-work~\cite{whylinechi,whylineicse} showed that when programmers debug
+Prior work shows that some forms of data provenance information is directly
+useful during program debugging.  The WhyLine
+Debugger~\cite{whylinechi,whylineicse} showed that when programmers debug
 programs, they benefit from the ability to directly ask such provenance
-(``why'') questions of their debugger.  Origin tracking~\cite{badapples} and
-program slicing~\cite{tipslicingsurvey} also make debugging easier by reporting
-the provenance of values.  Like these prior efforts, in this work we focus on
-using data provenance information for {\em debugging} -- understanding the root
-cause of a failure -- as opposed to {\em bug detection} -- finding new, unknown
-bugs.
+(``why'') questions of their debugger.  Bad Value Origin
+Tracking~\cite{badapples} is a limited form of provenance analysis that tracks
+intructions that stored certain unusable values.  The authors of Bad Value Origin Tracking showed that even such limited provenance information
+helps understand some bug root causes. Like these
+prior efforts, we focus on using provenance analysis for
+{\em debugging} -- understanding the root cause of a failure -- as opposed to
+{\em bug detection} -- finding new, unknown bugs.
 
@@
 \paragraph{Debugging with Data Provenance Information}
-Prior work has looked at various forms of data provenance for debugging.
-
-We have discussed bad value origin tracking~\cite{badapples} throughout this
-work and compared directly with that technique in our evaluation.  Bad value
-origin tracking is like our work because they store provenance information to
-be used for debugging -- a similar goal to ours.  Bad value origin tracking
-differs from our work in several ways.  First, their technique only tracks the
-provenance of {\em unusable} values -- a fraction of all values in an
-execution.  Our work, instead, tracks provenance for any value stored in a
-potentially shared memory location, which includes a broader set of provenance
-and targets our technique to concurrency debugging.  Second, their technique
-does not run with overheads tolerable for performance for C and C++ programs.
-They implement their technique for C and C++ programs using Valgrind MemCheck,
-which has high runtime overheads.  Our technique has overheads low enough for
-production in many cases, especially in server applications.  Third, their
-technique does not explicity record thread information with tracked origins.
-Our work focuses on concurrency and we include thread information.  As we show
-in Section~\ref{sec:eval:debugging} thread information is important for
-understanding some bugs.  Fourth, their technique uses bad value piggy-backing
-to store origins, we use a shadow memory.
-
-Like our work, the WhyLine debugger~\cite{whylineicse, whylinechi} allows
-programmers to ask questions about why variables hold particular values during
-debugging.  WhyLine answers by presenting data- and control-flows that
-influenced those variables.  WhyLine differs from our work in that it uses
-dependence slicing on recorded executions to answer provenance questions and
-has overheads too high for production use.  WhyLine also did not explicitly
-focus on concurrency.
+Prior work has shown that some limited forms of provenance help
+with debugging~\cite{badapples, whylineicse, tipslicingsurvey}.
+
+Bad Value Origin Tracking~\cite{badapples} tracked code that wrote unusable
+values into variables~\cite{badapples}. That work showed that revealing the
+connection between an unusable value's origin and a failure it causes is
+helpful during debugging.  Bad value origin tracking is the closest prior work
+to this work~\cite{badapples} and we directly compare to it in our evaluation.
+At a high level, bad value origin tracking is like our work because they track
+provenance information for debugging, which is a similar goal to ours.
+
+Bad value origin tracking relates to our work in several other ways.  First,
+their technique tracks the provenance of {\em unusable} values only -- a
+fraction of all values in an execution.  Our work, instead, tracks provenance
+for any value stored in a potentially shared memory location, which includes a
+broader set of provenance and targets our technique to concurrency debugging.
+Second, for Java programs, bad value origin tracking has runtime overheads
+acceptable for production, which was a goal for our system.  Bad value origin
+tracking ``piggybacks'' provenance information on storage allocated by the Java
+runtime for bad values.  Piggybacking is a key to their low overheads.  For
+C/C++, the overhead of bad value origin tracking is much higher.  In C/C++,
+they cannot use piggybacking and, instead, use heavy-weight binary
+instrumentation with overheads unacceptable for production.  Our technique has
+overheads low enough for production for C/C++ programs, especially for server
+applications.  Third, their technique does not explicity record thread
+information with tracked origins.  Our work focuses on concurrency and we
+include thread information.  As we show in Section~\ref{sec:eval:debugging}
+thread information is important for some bugs.
+
+Other work showed that programmers debug more effectively when they can ask
+``why'' questions about values during
+debugging~\cite{tipslicingsurvey,whylineicse}.  Like our work, the WhyLine
+debugger~\cite{whylineicse, whylinechi} allows programmers to ask questions
+about why variables hold particular values during debugging.  WhyLine answers
+by presenting data- and control-flows that influenced those variables.  WhyLine
+differs from our work in that it uses slicing on recorded executions to answer
+provenance questions, it has high run time overheads, and does not focus on
+concurrency.
 
@@
 
-\paragraph{Communication Tracking in Concurrent Programs}
-Several techniques have been proposed do some form of dependence and
-communication tracking to analyze concurrent programs.
+\paragraph{Communication and Sharing Analyses in Concurrent Programs}
+Several techniques have been proposed analyze dependences and
+communication in concurrent programs.
+
+A closely related recent effort in this space is Octet~\cite{octet} that
+captures inter-thread communication soundly and, like \ctraps, can be used to
+build dynamic concurrency analyses.  Octet differs from \ctraps in that it is
+sound and was built to support analyses on Java programs.  Additionally, Octet
+tracks the ``locality state'' of memory locations, rather than the last writer
+slice information.  Octet's average run time overhead (26\%) is comparable to
+the overhead of both last writer slicing and \ctraps and all three techniques
+have low enough overheads in many cases for production use.
 

### 2014-03-20 paper.tex
commit to merge
edits to title and abstract. Brandon, you said you would add that figure I drew on the board to the intro, so I did not touch the intro

@@
 
-\title{Efficient Data Provenance Analysis for Concurrent Programs}
+\title{Data Provenance Tracking for Concurrent Programs}
 
@@
 %dependences typically has overheads too high for production.  
-We propose {\em last writer slices}, an analysis for shared-memory
-multi-threaded programs that tracks {\em data provenance information} in
-production code.  Last writer slices dynamically record the thread and
-operation that last wrote each variable, reflecting the {\em provenance} of
-values.  Provenance helps understand bugs and we provide last writer slices to
-programmers during debugging.  We build {\em communication traps}
-(\ctraps) using last writer slices.  \ctraps efficiently
-monitors and can interpose on operations that share data between
-threads.  \ctraps is extensible, exposing inter-thread dependences
-to {\em \ctraps applications} that implement arbitrary analyses.
-
-We implement last writer slicing and \ctraps.  We conduct debugging studies of
-real, buggy programs showing last writer slices are useful for debugging.  We
-show \ctraps is useful by implementing two concurrency analyses from prior
-work.  We evaluate the performance of our designs using server programs
-and standard benchmarks.  We show that collecting last writer slices
-has low run time overheads (0--15\%) for many applications,including  
-memcached, LevelDB, MySQL, and Apache.  We show that \ctraps imposes low 
-overheads by default and that overheads scale with analysis complexity.
-
+We propose {\em last writer slicing}, a mechanism for tracking {\em data provenance information} in multithreaded code in a production setting.
+Last writer slices dynamically track provenance of values by recording the thread and operation that last wrote each variable.
+We show that this information complements core dumps and greatly improves debugability.
+We also propose {\em communication traps}
+(\ctraps), an extensible framework that uses last writer slices to interpose on operations that lead to inter-thread communication. We show how to use 
+\ctraps implement multiple dynamic analysis tools for mulithreaded programs.
+Our evaluation on server programs and PARSEC benchmarks shows that last writer slices has low run time overheads (0--15\%) for many applications,including
+memcached, LevelDB, MySQL, and Apache. Our debugging case studies of
+real, buggy programs shows that last writer slices are indeed useful for debugging.  
 \end{abstract}

### 2014-03-20 paper.tex
commtting to merge

--- 2014-03-20 paper.tex
@@
 \section{Introduction}
-Multi-threaded programming is challenging.  In contrast to sequential
+Multi-threaded programming is challenging.  In contrast to simple sequential
 reasoning, multi-threaded programs require complex reasoning about many threads
@@
 
-A closely related recent effort in this space is Octet~\cite{octet} that
-captures inter-thread communication soundly and, like \ctraps, can be used to
-build dynamic concurrency analyses.  Octet differs from \ctraps in that it is
-sound and was built to support analyses on Java programs.  Additionally, Octet
-tracks the ``locality state'' of memory locations, rather than the last writer
-slice information.  Octet's average run time overhead (26\%) is comparable to
-the overhead of both last writer slicing and \ctraps and all three techniques
-have low enough overheads in many cases for production use.
+A closely related recent effort on dynamic dependence analysis is
+Octet~\cite{octet}, which captures inter-thread communication soundly and to
+aid in building arbitrary concurrency analyses. Octet tracks the ``locality
+state'' of memory locations to make dependence tracking efficient.   Octet
+differs from \ctraps in that it focuses on soundness, was built primarily to
+optimize dependence tracking, and does not discuss direct use of
+dependence information to help with debugging.  Octet's average run time
+overhead (26\%) is comparable to the overhead of both last writer slicing and
+\ctraps and Octet, last writer slicing, and \ctraps all have low enough
+overheads for production use in many cases.
 

### 2014-03-20 paper.tex
committing the paper with this new figure in the intro

@@
 
-\newcommand{\Caption}[1]{\begin{minipage}{1.0\columnwidth} \caption{#1} \end{minipage}}
+\newcommand{\Caption}[1]{\begin{minipage}{.95\columnwidth} \caption{#1} \end{minipage} \vspace{-1.2ex}}
 \newcommand{\CaptionWide}[1]{\caption{#1}\vspace{-2ex}}
@@
 location contains its value.  In this work we develop a new provenance tracking
-mechanism called {\em last writer slicing} that satisfies these goals.  A
+mechanism called {\em last writer slicing}.  A
 memory location's last writer slice is a record containing the program point
@@
 
-Motivated by these prior {\em ad hoc} efforts, we build a general mechanism
-called {\em Communication Traps} or {\em \ctraps} on top of last writer slices
-(Section~\ref{sec:ctraps}). \ctraps is useful for implementing dynamic
-concurrency analyses, like the ones discussed above.  \ctraps runs alongside a
-program and uses last writer slices to identify memory accesses that lead to
-{\em communication} between the threads. Communcation occurs when a thread
-accesses memory last written by a different thread and \ctraps delivers a {\em
-trap} to a thread when it communicates with another thread, giving it
-information about how communication occurred.  Programmer-defined {\em trap
-handlers} can implement arbitrary analysis using that information. By default,
-\ctraps has low overheads that are appropriate production in many cases.  As we
-show in our evaluation, the overhead of \ctraps applications scales with the
-complexity of their analysis.
+Motivated by these prior {\em ad hoc} efforts, we develop a dynamic analysis
+framework called {\em Communication Traps} or {\em \ctraps}.  \ctraps runs
+alongside a program and uses the provenance information provided by last writer
+slices to identify memory accesses that lead to {\em communication} between the
+threads. Communcation occurs when a thread accesses memory last written by a
+different thread and \ctraps delivers a {\em trap} to a thread when it
+communicates with another thread, giving it information about how communication
+occurred.  Programmer-defined {\em trap handlers} can implement arbitrary
+analysis using that information. By default, \ctraps has low overheads that are
+appropriate production in many cases.  As we show in our evaluation, the
+overhead of \ctraps applications scales with the complexity of their analysis.
+Figure~\ref{fig:overview} illustrates the relationship between last writer
+slices, \ctraps, and \ctraps applications.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=.70\columnwidth]{figs/overview.pdf}
+\Caption{\label{fig:overview}{\bf An overview of last writer slicing and CTraps.} Last writer slicing is a basic mechanism for tracking provenance.
+The CTraps framework uses last writer slicing to identify communication and expose it to CTraps applications.
+CTraps applications implement arbitrary dynamic concurrency analyses, like communication graph collection~\cite{recon} or cooperative debugging~\cite{cci}.}
+\end{figure}
 
@@
 \includegraphics[width=\columnwidth]{figs/JDKStringBufferFail.pdf}
-\caption{\label{fig:coreDumpFail}{\bf Atomicity violation bug in a string buffer implementation.} The core dump does not reveal the connection between {\tt append()} and {\tt erase()}, which is crucial to understand the bug.}
+\Caption{\label{fig:coreDumpFail}{\bf Atomicity violation bug in a string buffer implementation.} The core dump does not reveal the connection between {\tt append()} and {\tt erase()}, which is crucial to understand the bug.}
 \end{figure}
@@
 \includegraphics[scale=.6]{figs/BasicLWT.pdf}
-\caption{\label{fig:basicLWT}Basic operation of the last writer table. Different shapes signify different distinct memory operations. }
+\Caption{\label{fig:basicLWT}Basic operation of the last writer table. Different shapes signify different distinct memory operations. }
 \end{figure}
@@
 \includegraphics[width=\columnwidth]{figs/JDKStringBufferDebug2.pdf}
-\caption{\label{fig:jdklws}{\bf Using last writer slices to debug an
+\Caption{\label{fig:jdklws}{\bf Using last writer slices to debug an
 ordering violation in the JDK 1.4 StringBuffer library.} The last writer slice reveals the buggy interaction between {\tt erase()} and {\tt append()}.}
@@
 \includegraphics[width=.9\columnwidth]{figs/LWS_Overview.pdf}
-\caption{\label{fig:lwsoverview}{\bf Overview of system support for last writer slicing.} Our compiler and runtime library ensure the \lwt remains updated so it can be used during debugging.}
+\Caption{\label{fig:lwsoverview}{\bf Overview of system support for last writer slicing.} Our compiler and runtime library ensure the \lwt remains updated so it can be used during debugging.}
 \end{figure}
@@
 We assume written memory locations do not overlap in memory because our \lwt
-design is {\em granularity oblivious}.  We discuss this issue informally due to
-space constraints.  Each \lwt entry corresponds to a location that was written.
-Two instructions might reference different locations that overlap because the
-referenced locations have different granularities ({\em e.g.}, accessing a byte
-that is part of a previously accessed word).  When that happens, both accesses
-update different \lwt entries, because they reference different locations.
-However, the overlapping region then has two different entries reflecting the
-last {\em two} updates to its parts rather than just its last write.
+design is {\em granularity oblivious}.  Each \lwt entry corresponds to a
+location that was written.  Two instructions might reference different
+locations that overlap because the referenced locations have different
+granularities ({\em e.g.}, accessing a byte that is part of a previously
+accessed word).  When that happens, both accesses update different \lwt
+entries, because they reference different locations.  However, the overlapping
+region then has two different entries reflecting the last {\em two} updates to
+its parts rather than just its last write.
 
@@
 \includegraphics[width=.90\columnwidth]{figs/CTraps_Overview.pdf}
-\caption{\label{fig:systemdiagram}{\bf CTraps system support.}}
+\Caption{\label{fig:systemdiagram}{\bf CTraps system support.}}
 \end{figure}
@@
 \paragraph{Compiler Support}
-Our compiler inserts a call to the \ctraps~{\em trap hook} function immediately
+Our compiler inserts a call to the CTraps {\em trap hook} function immediately
 before all read and write operations that access potentially escaping memory
@@
 %\end{verbatim}
-%\caption{\label{fig:hookapi}The \ctraps trap handling interface.}
+%\Caption{\label{fig:hookapi}The \ctraps trap handling interface.}
 %\end{figure}
@@
 \end{tabular}
-\caption{\label{tab:bugs}{\bf Concurrency debugging with last writer slices.} We describe each bug and summarize the debugging benefit of last writer slices using the evaluation criteria from ~\cite{badapples}.
+\Caption{\label{tab:bugs}{\bf Concurrency debugging with last writer slices.} We describe each bug and summarize the debugging benefit of last writer slices using the evaluation criteria from ~\cite{badapples}.
 }
@@
 cause of a concurrency bug from HTTrack-3.43.9.  The program crashes at point
-{\bf B} when it calls {\tt hts\_mutex\_lock()} with an invalid mutex.    The
+{\bf B} when it calls {\tt hts\_} {\tt mutex\_lock()} with an invalid mutex.    The
 root cause of this bug is that Thread 1 fails to initialize the mutex before
@@
 \includegraphics[width=\columnwidth]{figs/LWSHTTDebug.pdf}
-\caption{\label{fig:httlws}{\bf Using last writer slices to debug an
+\Caption{\label{fig:httlws}{\bf Using last writer slices to debug an
 ordering violation in HTTrack.} The provenance information stored in the last
@@
 \includegraphics[width=\columnwidth]{figs/MySQLDebug.pdf}
-\caption{\label{fig:mysqllws}{\bf Using last writer slices to debug an
+\Caption{\label{fig:mysqllws}{\bf Using last writer slices to debug an
 atomicity violation in MySQL 4.0.12.} The last writer slice reveals the code that closes the log before a database insertion that goes unlogged, which is a security problem.}
@@
 \paragraph{LevelDB-1.5.0}
-LevelDB is a carefully tuned high-performance key-value store written by Google
+LevelDB is a high-performance key-value store written by Google
 developers. To benchmark LevelDB, we used the included {\tt db\_bench} utility,
@@
 \includegraphics[width=.9\columnwidth]{plots/perf.pdf}
-\caption{\label{fig:perfall}Runtime overhead of \ctraps and Last Writer Slices (LWS).}
+\Caption{\label{fig:perfall}Runtime overhead of \ctraps and Last Writer Slices (LWS).}
 \end{figure}
@@
 \includegraphics[width=.9\columnwidth]{plots/appperf.pdf}
-\caption{\label{fig:perfapps}Performance overheads of CCI-Prev and Graph Collection implemented with \ctraps.}
+\Caption{\label{fig:perfapps}Performance overheads of CCI-Prev and Graph Collection implemented with \ctraps.}
 \end{figure}
@@
 \end{tabular}
-\caption{\label{tab:char}Instrumentation pruned by static analysis.}
+\Caption{\label{tab:char}Instrumentation pruned by static analysis.}
 \end{table}

### 2014-08-31 paper.tex
new format for CGO sub
revisions - onto 10.75 pgs

@@
-\documentclass[preprint,10pt]{sigplanconf}
-\usepackage{url}
+\newcount\draft\draft=1 % set to 0 for submission or publication
+\newcount\cameraready\cameraready=0
+
+\documentclass[pageno,nohyperref]{jpaper}
+
+\ifnum\draft=1
+  \input{revision}
+  \usepackage{drafthead}
+\fi
+\usepackage{xxx}
+
+\usepackage{graphicx}
+\DeclareGraphicsExtensions{.pdf,.jpg,.png}
+\graphicspath{{./figs/} {./plots/}}
+
+\usepackage[normalem]{ulem}
+
 \usepackage{listings}
-\usepackage{subfigure}
-\usepackage{times}
-\usepackage{multirow}
-\usepackage{color}
-\usepackage{multicol}
+\lstset{
+    language=C,
+    emphstyle=\bfseries,
+    basicstyle=\ttfamily\small,
+    aboveskip=1mm plus 1mm minus 1mm,
+    belowskip=1mm plus 1mm minus 1mm,
+    mathescape=true,
+    xleftmargin=\parindent,
+}
+\newcommand{\lil}{\lstinline}
+
+\usepackage[labelfont=bf,font=small]{caption}
+
 \usepackage{amsmath}
 \usepackage{amssymb}
-\usepackage{setspace}
-\usepackage{graphicx}
-\usepackage{xspace}
+\usepackage{mathptmx}
+\usepackage{multirow}
+%\usepackage{multicol}
 \usepackage{color}
@@
 
+\usepackage{xspace}
 \newcommand{\addtodo}[1]{\textcolor{red}{[To do: #1]}}
@@
 \newcommand{\ctrapsmm}{CTraps-NRR\xspace}
-\newcommand{\lws}{CTraps-LWS\xspace}
+\newcommand{\lws}{LWS\xspace}
 \newcommand{\lwt}{LWT\xspace}
@@
 %\doublespacing
-\authorinfo{Anonymous for Submission}{}
+%\authorinfo{Anonymous for Submission}{}
+
+%\authorinfo{Brandon Lucia \and Luis Ceze}{University of Washington, Department of Computer Science and Engineering}{\{blucia0a,luisceze\}@cs.washington.edu\\http://sampa.cs.washington.edu\\{\bf In Submission --- Please Do Not Distribute}}
+
+%\author{Anonymous for Submission}
 
 \title{Data Provenance Tracking for Concurrent Programs}
-
-%\authorinfo{Brandon Lucia \and Luis Ceze}{University of Washington, Department of Computer Science and Engineering}{\{blucia0a,luisceze\}@cs.washington.edu\\http://sampa.cs.washington.edu\\{\bf In Submission --- Please Do Not Distribute}}
-
-%\author{Anonymous for Submission}
+\newcommand{\submissionnumber}{65}
+\date{}
+
 \begin{document}
+
+\special{papersize=8.5in,11in}
+\setlength{\pdfpageheight}{\paperheight}
+\setlength{\pdfpagewidth}{\paperwidth}
+
 \maketitle
 
+\thispagestyle{empty}
 \begin{abstract}
@@
 %dependences typically has overheads too high for production.  
-We propose {\em last writer slicing}, a mechanism for tracking {\em data provenance information} in multithreaded code in a production setting.
-Last writer slices dynamically track provenance of values by recording the thread and operation that last wrote each variable.
-We show that this information complements core dumps and greatly improves debugability.
-We also propose {\em communication traps}
-(\ctraps), an extensible framework that uses last writer slices to interpose on operations that lead to inter-thread communication. We show how to use 
-\ctraps implement multiple dynamic analysis tools for mulithreaded programs.
-Our evaluation on server programs and PARSEC benchmarks shows that last writer slices has low run time overheads (0--15\%) for many applications,including
-memcached, LevelDB, MySQL, and Apache. Our debugging case studies of
-real, buggy programs shows that last writer slices are indeed useful for debugging.  
+We propose {\em last writer slicing}, a mechanism for tracking {\em data
+provenance information} in multithreaded code in a production setting.  Last
+writer slices dynamically track provenance of values by recording the thread
+and operation that last wrote each variable. We show that this information
+complements core dumps and greatly improves debugability. We also propose {\em
+communication traps} (\ctraps), an extensible framework that uses last writer
+slices to interpose on operations that lead to inter-thread communication. We
+show how to use \ctraps implement multiple dynamic analysis tools for
+mulithreaded programs. Our evaluation on server programs and PARSEC benchmarks
+shows that last writer slices has low run time overheads (0--15\%) for many
+applications,including  memcached, LevelDB, MySQL, and Apache. Our debugging
+case studies of real, buggy programs shows that last writer slices are indeed
+useful for debugging.  
 \end{abstract}
@@
 reasoning, multi-threaded programs require complex reasoning about many threads
-of control and their interactions.  In shared memory programs, threads interact
-by reading and writing memory locations.  Threads' reads and writes interleave
-arbitrarily and nondeterministically. Each interleaving can lead to different
-behaviors and memory states, some of which may be undesirable, like crashes or
-data corruption.  Moreover, because complete concurrency testing is infeasiable
-in general, undesirable interleavings can evade testing and lead to production
-failures.
-
-The problem we are addressing in this work is that debugging support for
+and their interactions.  Threads primarily interact by reading and writing
+shared memory locations and different threads' reads and writes interleave
+arbitrarily and nondeterministically. Different interleavings can lead to
+different behaviors, some of which may be undesirable, like crashes or data
+corruption. Such undesirable interleavings often evade testing and cause
+production failures because complete concurrency testing is infeasible.
+
+The problem that this work addresses is that debugging support for
 understanding thread interactions is inadequate.  Today, programmers use a
-debugger to examine memory contents either during a debugging execution or by
-loading a core dump from a failed execution.  Examining the contents of memory
-reveals {\em what} the state of the program is ({\em e.g.}, the values of
-variables).  Unfortunately, the real question programmers must answer is {\em
-why} the state of the program is what it is.  Understanding why a program
-entered an undesirable state ({\em e.g.}, crashed) is the key to changing the
-code to prevent future executions from entering that state.  Multi-threading
-makes this understanding elusive because implicating particular operations in
-particular threads for a failure is a complex task.  Aiding developers in this
-understanding for failed production executions is important because
-hard-to-find bugs may manifest in production only.
+debugger to examine memory state during a debugging execution or by loading a
+core dump after a failure.  Examining memory reveals {\em what} the state of
+the program is ({\em e.g.}, ``is the pointer null?'').  Unfortunately, the real
+question programmers must answer is {\em why} the state of the program is what
+it is ({\em e.g.}, ``what thread and instruction made the pointer null?'').
+Understanding why a program entered an undesirable state ({\em e.g.}, crashed)
+is the key to fixing buggy code to prevent that state.  Aiding developers in
+this understanding for failed production executions is especially important
+because hard-to-find bugs may manifest in production only.
 
@@
 
-%During debugging provenance helps a programmer better understand not just where failures occur, but where values came from, connectng disparate
-%parts of a program that interact, leading to a failure.
-
-%In
-%summary, the challenges in using provenance to debug concurrent programs are:
-%(1) collecting the right information to characterize the concurrent execution
-%behavior; and (2) making it efficient enough for general programs running in
-%production settings.
-
 Last writer slices are primarily useful for concurrency debugging, but we 
@@
 threads. Communcation occurs when a thread accesses memory last written by a
-different thread and \ctraps delivers a {\em trap} to a thread when it
+different thread and \ctraps delivers a software {\em trap} to a thread when it
 communicates with another thread, giving it information about how communication
-occurred.  Programmer-defined {\em trap handlers} can implement arbitrary
-analysis using that information. By default, \ctraps has low overheads that are
-appropriate production in many cases.  As we show in our evaluation, the
+occurred.  Programmer-defined {\em \ctraps Applications} can implement arbitrary
+analysis using that information. By default, \ctraps has low overheads, 
+appropriate for production in many cases.  As our evaluation shows, the
 overhead of \ctraps applications scales with the complexity of their analysis.
@@
 \centering
-\includegraphics[width=.70\columnwidth]{figs/overview.pdf}
-\Caption{\label{fig:overview}{\bf An overview of last writer slicing and CTraps.} Last writer slicing is a basic mechanism for tracking provenance.
-The CTraps framework uses last writer slicing to identify communication and expose it to CTraps applications.
-CTraps applications implement arbitrary dynamic concurrency analyses, like communication graph collection~\cite{recon} or cooperative debugging~\cite{cci}.}
+\includegraphics[width=\columnwidth]{figs/systemoverview.pdf}
+\Caption{\label{fig:overview}{\bf System Overview.} Last writer slicing tracks provenance, which augments the core dump and enables provenance debugging.
+CTraps uses provenance to convey communication to CTraps applications, which implement concurrency analyses, like communication graphs~\cite{recon} or CBI~\cite{cci}.}
 \end{figure}
@@
 
-\item{We describe last writer slicing, a novel data provenance tracking
-mechanism, that produces information that is useful for debugging concurrent
-programs.  We design and implement system support for collecting last writer
-slices that is efficient enough for use in production.  }
-
-\item{Building on last writer slicing, we develop \ctraps, an extensible
-framework for implementing analyses of inter-thread communication in
-multi-threaded programs.  }
-
-\item{We evaluate the utility of last writer slices and \ctraps. We first
-present a study of several real concurrent program bugs and show that last
-writer slices are useful for debugging.  We then show how \ctraps is useful for
-building dependence analyses by implementing two analyses from prior
+\item{{\bf Last writer slicing} ({\em LWS}): a provenance tracking mechanism
+useful for concurrency debugging.  We present an \lws implementation
+efficient enough for production use.}
+
+\item{{\bf Communication traps} (\ctraps): an extensible
+framework for implementing concurrency analyses, built on \lws.  }
+
+\item{{\bf Efficacy evaluation}: We show that \lws is useful for debugging with
+a study of several real concurrency bugs.  We show \ctraps is useful for
+building concurrency analyses by implementing two analyses from prior
 work~\cite{cci,defuse,recon}. }
 
-\item{We evaluate our system's performance using a variety of important
-applications ({\em e.g.}, MySQL, memcached and the PARSEC
-benchmarks~\cite{parsec}).  We show that in many cases our designs have
-overheads low enough for use in deployment (0--15\%).}
+\item{{\bf Performance evaluation}: We show that in many cases our designs have
+overheads low enough for use in deployment (0--15\%) with a study of a diverse
+set of real programs ({\em e.g.}, MySQL, memcached and the PARSEC
+benchmarks~\cite{parsec}).  }
 
@@
 example, in order to motivate collecting last writer slices in deployed
-systems.  We then provide context for \ctraps by discussing some more
-sophisticated concurrency analyses that rely on last writer information.
+systems.  We then provide context for \ctraps by discussing more sophisticated
+analyses that use provenance.
 
@@
 
-Prior work shows that some forms of data provenance information is directly
-useful during program debugging.  The WhyLine
-Debugger~\cite{whylinechi,whylineicse} showed that when programmers debug
-programs, they benefit from the ability to directly ask such provenance
+Prior work shows that data provenance information is directly useful during
+program debugging.  The WhyLine Debugger~\cite{whylinechi,whylineicse} showed
+that when programmers debug, they benefit from the ability to ask provenance
 (``why'') questions of their debugger.  Bad Value Origin
 Tracking~\cite{badapples} is a limited form of provenance analysis that tracks
-intructions that stored certain unusable values.  The authors of Bad Value Origin Tracking showed that even such limited provenance information
-helps understand some bug root causes. Like these
-prior efforts, we focus on using provenance analysis for
-{\em debugging} -- understanding the root cause of a failure -- as opposed to
-{\em bug detection} -- finding new, unknown bugs.
+intructions that stored certain unusable values.  The authors of Bad Value
+Origin Tracking showed that even such limited provenance information helps
+understand some bug root causes. Like these prior efforts, we focus on using
+provenance analysis for {\em debugging} -- understanding the root cause of a
+failure -- as opposed to {\em bug detection} -- finding new, unknown bugs.
 
@@
 \label{sec:lwssoundness}
-To discuss the correctness of our \lwt-based last writer slice support, we
-first develop a formal description of last writer slices.  We consider an
-N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_t = (p^{t}_{1},
-p^{t}_{2}, \ldots, p^{t}_{k_{t}})$ is the sequence of $k_{t}$ dynamic
-instructions executed by thread $t$.  Without loss of generality, we assume all
-instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
-$Read$ or $Write$ and $m$ is a uniquely addressed memory location.
-A point in an execution of $P$ after $e$ instructions have
-executed is defined by $Q = (q_{0}, \ldots, q_{e})$ where each $q_{i} \in Q$ is a
-$p^{t}_{j}$ from one of the $k_t$ instructions of thread $t$.  We introduce
-$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
-$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
-q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
-executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
-in $Q$ to write $m$ and $t$ is the thread that executed it.  For an
-unwritten location $n$, $LWS[n] = (\emptyset,\emptyset)$.
-
-For our design to be correct, $LWT[m]$, the entry for a memory location $m$ in
-our \lwt implementation, must be the same as $LWS[m]$ at all points in an
-execution.  Correctness in the \lwt requires {\em soundness} and {\em
-completeness}.  A {\em sound} design updates $LWT[m]$ only when $m$ is written.
-A {\em complete} design always updates $LWT[m]$ when $m$ is written.  Our
-compiler inserts \lwt updates before every write operation (for completeness)
-and only before write operations (for soundness).  Formally, given a program,
-$P$, our compiler pass transforms it to a new program, $P'$.  Each thread in
-$P'$ executes a modified sequence of instructions: $P'_{t} \in P' = ( 
-I(p^{t}_{1}), p^{t}_{1}, \ldots, I(p^{t}_{k_{t}}), p^{t}_{k_{t}} )$.
-$I(p^{t}_{j})$ sets $LWT[m] = (t,p^{t}_{j})$ if and only if $p^{t}_{j} =
-Write(m)$ and is a no-op otherwise.  This reasoning shows our \lwt is both
-sound, complete, and thus correct.
-
-Next we discuss two nuances of our design: (1) the interaction of our \lwt
-implementation with data-races; and (2) the correctness of the
-\lwt for accesses to overlapping memory regions.
-
-\paragraph{Consistency and Atomicity}
-
-Our implementation of the \lwt does not add any synchronization to the program.
-As a result our analysis does not incur the overheads or possible thread
-schedule perturbation synchronization can cause.  However, there are two
-potential issues that stem from this design choice. First, we do not enforce
-memory consistency of meta-data reads and writes in different threads.  Second,
-our implementation does not enforce the atomicity of \lwt updates and their
-corresponding program accesses.  In the absence of data-races our \lwt
-implementation is {\em always correct} and neither of these issues exist.  Both
-atomicity and ordering is enforced by the program synchronization that ensures
-consistency for program data.
-
-We use our \lwt formalism to show that without data-races, \lwt updates and
-their corresponding memory accesses are atomic and well-ordered.  The relation
-$p \prec q$ denotes that $p$ happens before $q$.  In correctly synchronized
-code, any write by thread $t$ to $m$, $p^{t}_{j} = Write(m)$, is ordered by
-program synchronization with $p^{t'}_{j'} = Op(m)$, a later access to $m$ by
-any other thread $t'$ -- {\em i.e.}, $p^{t}_{j} \prec p^{t'}_{j'}$.  Our
-compiler inserts an \lwt update $I(p^{t}_{j})$ into $P'$ before its write
-operation, $p^{t}_{j}$. Thus, $I(p^{t}_{j}) \prec p^{t}_{j}$ due to program
-order.
+
+We developed a formalism of our \lwt-based last writer slices support, but we
+omit it due to space constraints.  We used our formalism to show that the \lwt
+correctly records last writer slices.  There are two important caveats in our
+formal discussion.  The first caveat is that the \lwt implementation introduces
+no synchronization.  We proved that the \lwt is always correct in
+data-race-free programs because \lwt accesses are synchronized by program
+synchronization.  However, if an execution has a data-race, program memory
+operations and \lwt accesses are no longer atomic.  If such a data-race leads
+to an unserializable interleaving of \lwt accesses, the memory location's \lwt
+entry may reflect the wrong last writer.  The second caveat is that the \lwt
+assumes that a memory location accessed at one granularity ({\em e.g.},
+byte-aligned) is never accessed at a different granularity ({\em e.g.}, as part
+of a word-aligned access).  Assuming accesses do not overlap ensures there is
+always only one \lwt entry for each memory location.
+
+%To discuss the correctness of our \lwt-based last writer slice support, we
+%first develop a formal description of last writer slices.  We consider an
+%N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_t = (p^{t}_{1},
+%p^{t}_{2}, \ldots, p^{t}_{k_{t}})$ is the sequence of $k_{t}$ dynamic
+%instructions executed by thread $t$.  Without loss of generality, we assume all
+%instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
+%$Read$ or $Write$ and $m$ is a uniquely addressed memory location.
+%A point in an execution of $P$ after $e$ instructions have
+%executed is defined by $Q = (q_{0}, \ldots, q_{e})$ where each $q_{i} \in Q$ is a
+%$p^{t}_{j}$ from one of the $k_t$ instructions of thread $t$.  We introduce
+%$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
+%$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
+%q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
+%executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
+%in $Q$ to write $m$ and $t$ is the thread that executed it.  For an
+%unwritten location $n$, $LWS[n] = (\emptyset,\emptyset)$.
+%
+%For our design to be correct, $LWT[m]$, the entry for a memory location $m$ in
+%our \lwt implementation, must be the same as $LWS[m]$ at all points in an
+%execution.  Correctness in the \lwt requires {\em soundness} and {\em
+%completeness}.  A {\em sound} design updates $LWT[m]$ only when $m$ is written.
+%A {\em complete} design always updates $LWT[m]$ when $m$ is written.  Our
+%compiler inserts \lwt updates before every write operation (for completeness)
+%and only before write operations (for soundness).  Formally, given a program,
+%$P$, our compiler pass transforms it to a new program, $P'$.  Each thread in
+%$P'$ executes a modified sequence of instructions: $P'_{t} \in P' = ( 
+%I(p^{t}_{1}), p^{t}_{1}, \ldots, I(p^{t}_{k_{t}}), p^{t}_{k_{t}} )$.
+%$I(p^{t}_{j})$ sets $LWT[m] = (t,p^{t}_{j})$ if and only if $p^{t}_{j} =
+%Write(m)$ and is a no-op otherwise.  This reasoning shows our \lwt is both
+%sound, complete, and thus correct.
+%
+%Next we discuss two nuances of our design: (1) the interaction of our \lwt
+%implementation with data-races; and (2) the correctness of the
+%\lwt for accesses to overlapping memory regions. 
+%
+%\paragraph{Consistency and Atomicity}
+%
+%Our implementation of the \lwt does not add any synchronization to the program.
+%As a result our analysis does not incur the overheads or possible thread
+%schedule perturbation synchronization can cause.  However, there are two
+%potential issues that stem from this design choice. First, we do not enforce
+%memory consistency of meta-data reads and writes in different threads.  Second,
+%our implementation does not enforce the atomicity of \lwt updates and their
+%corresponding program accesses.  In the absence of data-races our \lwt
+%implementation is {\em always correct} and neither of these issues exist.  Both
+%atomicity and ordering is enforced by the program synchronization that ensures
+%consistency for program data.   
+%
+%We use our \lwt formalism to show that without data-races, \lwt updates and
+%their corresponding memory accesses are atomic and well-ordered.  The relation
+%$p \prec q$ denotes that $p$ happens before $q$.  In correctly synchronized
+%code, any write by thread $t$ to $m$, $p^{t}_{j} = Write(m)$, is ordered by
+%program synchronization with $p^{t'}_{j'} = Op(m)$, a later access to $m$ by
+%any other thread $t'$ -- {\em i.e.}, $p^{t}_{j} \prec p^{t'}_{j'}$.  Our
+%compiler inserts an \lwt update $I(p^{t}_{j})$ into $P'$ before its write
+%operation, $p^{t}_{j}$. Thus, $I(p^{t}_{j}) \prec p^{t}_{j}$ due to program
+%order.
 $I(p^{t'}_{j'})$ is inserted before $p^{t'}_{j'}$, but after the synchronization that orders $p^{t'}_{j'}$ and $p^{t}_{j}$, so $p^{t}_{j} \prec I(p^{t'}_{j'})$.
 Happens-before is transitive, so without data-races, $I(p^{t}_{j})
-\prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec p^{t'}_{j'}$, meaning $I(p^{t}_{j})$
-happens before all accesses that $p^{t}_{j}$ happens before {\em and before
-their respective \lwt updates}, $I(p^{t'}_{j'})$.  Symmetrically, for any
-access $p^{t''}_{j''}$ in another thread $t''$ that precedes $p^{t}_{j}$,
-$p^{t''}_{j''} \prec p^{t}_{j}$.  Our compiler inserts each $I(p^{t}_{j})$
-immediately before each $p^{t}_{j}$, so $I(p^{t''}_{j''}) \prec p^{t''}_{j''}
-\prec I(p^{t}_{j}) \prec p^{t}_{j}$, happens-before ordering $I(p^{t}_{j})$
-after all preceding accesses to $m$ in other threads, {\em and after their
-repective \lwt updates}.  Thus, at all points $Q$ in a data-race free
-execution, given $q_r = I(p^{t}_{j})$, $q_p = p^{t}_{j}$, and $q_r \prec q_p$,
-$\forall_{ \{q_s | q_r \prec q_s \prec q_p \}}: q_s \ne Op(m) \wedge q_s \ne
-I(Op(m)) $.  In other words, \lwt updates and their accesses are atomic.
-
-In a program with a data-race, there are some accesses, $p^{t}_{j}$ and
-$p^{t'}_{j'}$, that are concurrent.  The \lwt updates for these accesses are
-also concurrent because they are not transitively ordered by the accesses' ordering.  The unordered accesses and \lwt updates can interleave arbitrarily
-because they are concurrent.  Some racy thread schedules may violate the
-atomicity of accesses and their \lwt updates.  Such an execution could reach a
-state $Q_{bad} = (\ldots, I(p^{t}_{j}), I(p^{t'}_{j'}), p^{t'}_{j'}, p^{t}_{j},
-\ldots)$ and in that state, $LWS[m] = (t,p^{t}_{j})$, but $LWT[m] =
-(t',p^{t'}_{j'})$.  The \lwt was effectively not updated when $p^{t}_{j}$
-executed (violating completeness), or was effectively updated to reflect that
-$p^{t'}_{j'}$ was $m$'s last writer, when $p^{t}_{j}$ was (violating
-soundness).
-
-According to the C++11 language specification, in the presence of a data-race,
-a program's behavior is undefined and arbitrary state corruption can occur.  In
-the presence of data-races, {\em any} analysis, including our \lwt analysis,
-running on such a program could experience arbitrary corruption.  In practice
-({\em e.g.}, on x86) these atomicity violations may cause limited \lwt
-corruption, but even this is unlikely.  These atomicity violations only occur
-only when accesses to the same data occur within a very short time interval,
-which is uncommon, {\em and} an unserializable interleaving of accesses and
-\lwt updates also occurs.
-
-\paragraph{Non-overlapping Memory References} 
-We assume written memory locations do not overlap in memory because our \lwt
-design is {\em granularity oblivious}.  Each \lwt entry corresponds to a
-location that was written.  Two instructions might reference different
-locations that overlap because the referenced locations have different
-granularities ({\em e.g.}, accessing a byte that is part of a previously
-accessed word).  When that happens, both accesses update different \lwt
-entries, because they reference different locations.  However, the overlapping
-region then has two different entries reflecting the last {\em two} updates to
-its parts rather than just its last write.
-
-The \lwt remains correct, assuming the program only accesses non-overlapping
-regions.  In programs that access overlapping regions, there is always some
-entry in the \lwt that correctly reflects each memory location's last writer.
-However, the coarse region's entry may be stale because of an
-access to an overlapping region.
-
-One approach to mitigating this issue would be to allocate \lwt entries for
-regions of fixed granularity.  We chose not to do so for two reasons.  First,
-associating \lwt entries with regions of  large granularity only ({\em i.e.},
-cache-line) precludes sound analysis for fine-grained accesses.  Second,
-associating \lwt entries with regions of small granularity ({\em i.e.}, byte)
-requires multiple \lwt updates for larger granularity accesses, increasing
-their overheads proportionately to the granularity of the access.
+%\prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec p^{t'}_{j'}$, meaning $I(p^{t}_{j})$
+%happens before all accesses that $p^{t}_{j}$ happens before {\em and before
+%their respective \lwt updates}, $I(p^{t'}_{j'})$.  Symmetrically, for any
+%access $p^{t''}_{j''}$ in another thread $t''$ that precedes $p^{t}_{j}$,
+%$p^{t''}_{j''} \prec p^{t}_{j}$.  Our compiler inserts each $I(p^{t}_{j})$
+%immediately before each $p^{t}_{j}$, so $I(p^{t''}_{j''}) \prec p^{t''}_{j''}
+%\prec I(p^{t}_{j}) \prec p^{t}_{j}$, happens-before ordering $I(p^{t}_{j})$
+%after all preceding accesses to $m$ in other threads, {\em and after their
+%repective \lwt updates}.  Thus, at all points $Q$ in a data-race free
+%execution, given $q_r = I(p^{t}_{j})$, $q_p = p^{t}_{j}$, and $q_r \prec q_p$,
+%$\forall_{ \{q_s | q_r \prec q_s \prec q_p \}}: q_s \ne Op(m) \wedge q_s \ne
+%I(Op(m)) $.  In other words, \lwt updates and their accesses are atomic.
+%
+%In a program with a data-race, there are some accesses, $p^{t}_{j}$ and
+%$p^{t'}_{j'}$, that are concurrent.  The \lwt updates for these accesses are
+%also concurrent because they are not transitively ordered by the accesses' ordering.  The unordered accesses and \lwt updates can interleave arbitrarily
+%because they are concurrent.  Some racy thread schedules may violate the
+%atomicity of accesses and their \lwt updates.  Such an execution could reach a
+%state $Q_{bad} = (\ldots, I(p^{t}_{j}), I(p^{t'}_{j'}), p^{t'}_{j'}, p^{t}_{j},
+%\ldots)$ and in that state, $LWS[m] = (t,p^{t}_{j})$, but $LWT[m] =
+%(t',p^{t'}_{j'})$.  The \lwt was effectively not updated when $p^{t}_{j}$
+%executed (violating completeness), or was effectively updated to reflect that
+%$p^{t'}_{j'}$ was $m$'s last writer, when $p^{t}_{j}$ was (violating
+%soundness).    
+%
+%According to the C++11 language specification, in the presence of a data-race,
+%a program's behavior is undefined and arbitrary state corruption can occur.  In
+%the presence of data-races, {\em any} analysis, including our \lwt analysis,
+%running on such a program could experience arbitrary corruption.  In practice
+%({\em e.g.}, on x86) these atomicity violations may cause limited \lwt
+%corruption, but even this is unlikely.  These atomicity violations only occur
+%only when accesses to the same data occur within a very short time interval,
+%which is uncommon, {\em and} an unserializable interleaving of accesses and
+%\lwt updates also occurs.
+%
+%\paragraph{Non-overlapping Memory References} 
+%We assume written memory locations do not overlap in memory because our \lwt
+%design is {\em granularity oblivious}.  Each \lwt entry corresponds to a
+%location that was written.  Two instructions might reference different
+%locations that overlap because the referenced locations have different
+%granularities ({\em e.g.}, accessing a byte that is part of a previously
+%accessed word).  When that happens, both accesses update different \lwt
+%entries, because they reference different locations.  However, the overlapping
+%region then has two different entries reflecting the last {\em two} updates to
+%its parts rather than just its last write.
+%
+%The \lwt remains correct, assuming the program only accesses non-overlapping
+%regions.  In programs that access overlapping regions, there is always some
+%entry in the \lwt that correctly reflects each memory location's last writer.
+%However, the coarse region's entry may be stale because of an
+%access to an overlapping region.
+%
+%One approach to mitigating this issue would be to allocate \lwt entries for
+%regions of fixed granularity.  We chose not to do so for two reasons.  First,
+%associating \lwt entries with regions of  large granularity only ({\em i.e.},
+%cache-line) precludes sound analysis for fine-grained accesses.  Second,
+%associating \lwt entries with regions of small granularity ({\em i.e.}, byte)
+%requires multiple \lwt updates for larger granularity accesses, increasing
+%their overheads proportionately to the granularity of the access.
 
@@
 \ctraps is a framework for analyzing and interposing on inter-thread
-communication events during a program's execution.  We build \ctraps on top of
-last writer slicing that we described in Section~\ref{sec:lastwriterslices}.
-Using last writer slice information, \ctraps dynamically tracks 
-inter-thread communication events. \ctraps exposes communication
-events as {\em traps} during an execution that {\em trap handlers} can handle
-to perform analysis and interposition.
+communication.  \ctraps uses \lws (\S~\ref{sec:lastwriterslices}) to track
+inter-thread communication.  \ctraps exposes communication events as {\em
+traps} during an execution that {\em trap handlers} can handle to perform
+analysis and interposition.
 
@@
 
-\subsection{Characterizing \ctraps}
-\label{sec:char}
-We characterized the instrumentation pruned by escape analysis.  We report our
-results in Table~\ref{tab:char}.  The table shows the percent of
-calls pruned and the absolute number of sites in parenthesis.
-
-The data show that escape analysis can prove a substantial fraction of
-operations are local and do not require instrumentation.  Across PARSEC,
-between 8\% and 30\% of accesses were proved local.  In the server programs, a
-larger proportion of accesses were able to be proved local (21\% -- 66\%).  The
-difference is likely due to the fact that server threads perform more
-independent work (as discussed in Section~\ref{sec:eval:parsecserver}).
-Eliminating instrumentation using escape analysis is key to high performance.
-
-\begin{table}[htb]
-\centering
-\small
-\begin{tabular}{l | r r }
-\multirow{3}{*}{\bf Application} & \multicolumn{2}{c}{\bf \em Pruned } \\
-               & \multicolumn{2}{c}{\bf \em Static Local}  \\ 
-               & \multicolumn{2}{c}{\bf \em Operations}    \\ \hline
-
-Apache         &  21.4\%&(12K)                    \\
-MySQL          &  21.3\%&(53K)                    \\
-memcached      &  65.8\%&(7.5K)                   \\
-LevelDB        &  44.0\%&(4.5K)                   \\ \hline
-blackscholes   &   7.5\%&(4)                      \\
-dedup          &   22.0\%&(171)                   \\
-canneal        &   28.6\%&(226)                   \\
-x264           &   29.8\%&(5960)                  \\
-vips           &   24.6\%&(9116)                  \\
-ferret         &   20.9\%&(792)                   \\
-fluidanimate   &   19.6\%&(149)                   \\
-streamcluster  &   11.6\%&(47)                    \\
-swaptions      &   8.3\% &(19)                    \\
-\end{tabular}
-\Caption{\label{tab:char}Instrumentation pruned by static analysis.}
-\end{table}
-
+%\subsection{Characterizing \ctraps}
+%\label{sec:char}
+%We characterized the instrumentation pruned by escape analysis.  We report our
+%results in Table~\ref{tab:char}.  The table shows the percent of
+%calls pruned and the absolute number of sites in parenthesis.
+%
+%The data show that escape analysis can prove a substantial fraction of
+%operations are local and do not require instrumentation.  Across PARSEC,
+%between 8\% and 30\% of accesses were proved local.  In the server programs, a
+%larger proportion of accesses were able to be proved local (21\% -- 66\%).  The
+%difference is likely due to the fact that server threads perform more
+%independent work (as discussed in Section~\ref{sec:eval:parsecserver}).
+%Eliminating instrumentation using escape analysis is key to high performance.  
+%
+%\begin{table}[htb]
+%\centering
+%\small
+%\begin{tabular}{l | r r }
+%\multirow{3}{*}{\bf Application} & \multicolumn{2}{c}{\bf \em Pruned } \\
+%               & \multicolumn{2}{c}{\bf \em Static Local}  \\ 
+%               & \multicolumn{2}{c}{\bf \em Operations}    \\ \hline
+%
+%
+%Apache         &  21.4\%&(12K)                    \\
+%MySQL          &  21.3\%&(53K)                    \\
+%memcached      &  65.8\%&(7.5K)                   \\
+%LevelDB        &  44.0\%&(4.5K)                   \\ \hline
+%blackscholes   &   7.5\%&(4)                      \\
+%dedup          &   22.0\%&(171)                   \\
+%canneal        &   28.6\%&(226)                   \\
+%x264           &   29.8\%&(5960)                  \\
+%vips           &   24.6\%&(9116)                  \\
+%ferret         &   20.9\%&(792)                   \\
+%fluidanimate   &   19.6\%&(149)                   \\
+%streamcluster  &   11.6\%&(47)                    \\
+%swaptions      &   8.3\% &(19)                    \\
+%\end{tabular}
+%\Caption{\label{tab:char}Instrumentation pruned by static analysis.}
+%\end{table}
+%
+%
 \section{Related Work}

### 2014-09-02 paper.tex
inside 10 pages w/ revisions

@@
 %dependences typically has overheads too high for production.  
-We propose {\em last writer slicing}, a mechanism for tracking {\em data
+We propose {\em \lws} (\lws), a mechanism for tracking {\em data
 provenance information} in multithreaded code in a production setting.  Last
@@
 complements core dumps and greatly improves debugability. We also propose {\em
-communication traps} (\ctraps), an extensible framework that uses last writer
-slices to interpose on operations that lead to inter-thread communication. We
-show how to use \ctraps implement multiple dynamic analysis tools for
+communication traps} (\ctraps), an extensible framework that uses \lws to interpose on operations that lead to inter-thread communication. We
+show how to use \ctraps to implement multiple dynamic analysis tools for
 mulithreaded programs. Our evaluation on server programs and PARSEC benchmarks
-shows that last writer slices has low run time overheads (0--15\%) for many
+shows that \lws has low run time overheads (0--15\%) for many
 applications,including  memcached, LevelDB, MySQL, and Apache. Our debugging
-case studies of real, buggy programs shows that last writer slices are indeed
+case studies of real, buggy programs shows that \lws are indeed
 useful for debugging.  
@@
 
-Our approach to solving that problem is to provide the programmer with {\em
-provenance} information that can be examined alongside the program's memory
-state ({\em i.e.}, core dump).  A value's provenance describes why a memory
-location contains its value.  In this work we develop a new provenance tracking
-mechanism called {\em last writer slicing}.  A
-memory location's last writer slice is a record containing the program point
-and the identifier of the thread that last wrote data to that memory location.
-As the program executes in production, we collect last writer slices for each
-potentially shared memory location and those last writer slices record the
-provenance of the values stored in those locations.  Last writer slices are
-saved with a core dump when the program fails and are available in the debugger
-({\em e.g.}, for breakpoint debugging).
-
-We have three goals for last writer slicing:  (1) To collect a general form of
-provenance information, not limited to certain values or locations only. (2) To
-provide provenance information focused on how threads interact to aid in
-understanding concurrency bugs. (3) To engineer our system to have overheads
-low enough for use in production, aiming for a 10 -- 50\% maximum overhead.
-
-Last writer slices are primarily useful for concurrency debugging, but we 
-observe that provenance information is a
-useful foundation for implementing other concurrent program analyses.  Several
-tools from prior work have used {\em ad hoc} last-writer information as part of
-more heavy-weight analyses in domains like bug detection ({\em e.g.},
+Our approach is to provide the programmer with data {\em provenance}
+information that can be examined alongside the program's memory state ({\em
+i.e.}, core dump).  Provenance information describes why a memory location
+contains its value.  In this work we develop a new provenance tracking
+mechanism called {\em \lws} (LWS).  A memory location's last writer slice is a
+record containing the program point and the identifier of the thread that last
+wrote data to that memory location.  As the program executes in production, we
+collect last writer slices for each potentially shared memory location,
+recording the provenance of the data in each location.  Last writer slices are
+saved with a core dump and are available in the debugger.
+
+We have three goals for \lws:  (1) To collect general provenance information,
+not limited to certain values or locations only. (2) To provide provenance
+information focused on how threads interact to aid in understanding concurrency
+bugs. (3) To engineer our system to have overheads low enough for use in
+production, aiming for a 10 -- 50\% maximum overhead.
+
+Last writer slices are primarily useful for concurrency debugging, but
+provenance information is the foundation of many other concurrency analyses.
+Several tools from prior work have used {\em ad hoc} last-writer information as
+part of more heavy-weight analyses in domains like bug detection ({\em e.g.},
 ~\cite{dmtracker,avio}), automatic debugging ({\em e.g.},
-~\cite{recon,conseq,cci}), profiling ({\em
-e.g.},~\cite{threadclustering, schedpredictionmodel}), and program
-understanding~\cite{oshatr}.
-
-Motivated by these prior {\em ad hoc} efforts, we develop a dynamic analysis
-framework called {\em Communication Traps} or {\em \ctraps}.  \ctraps runs
-alongside a program and uses the provenance information provided by last writer
-slices to identify memory accesses that lead to {\em communication} between the
-threads. Communcation occurs when a thread accesses memory last written by a
-different thread and \ctraps delivers a software {\em trap} to a thread when it
-communicates with another thread, giving it information about how communication
-occurred.  Programmer-defined {\em \ctraps Applications} can implement arbitrary
-analysis using that information. By default, \ctraps has low overheads, 
-appropriate for production in many cases.  As our evaluation shows, the
-overhead of \ctraps applications scales with the complexity of their analysis.
+~\cite{recon,conseq,cci}), profiling ({\em e.g.},~\cite{threadclustering,
+schedpredictionmodel}), and program understanding~\cite{oshatr}.
+
+Motivated by these {\em ad hoc} efforts, we develop a dynamic analysis
+framework called {\em Communication Traps} or {\em \ctraps}.  \ctraps uses the
+provenance information provided by last writer slices to identify memory
+accesses that lead to {\em communication} between the threads. Communication
+occurs when a thread accesses memory last written by a different thread and
+\ctraps delivers a software {\em trap} to a thread when it communicates with
+another thread, giving it information about how communication occurred.
+Programmer-defined {\em \ctraps Applications} can implement arbitrary analysis
+using that information. By default, \ctraps has low overheads, appropriate for
+production in many cases.  As our evaluation shows, the overhead of \ctraps
+applications scales with the complexity of their analysis.
 Figure~\ref{fig:overview} illustrates the relationship between last writer
 slices, \ctraps, and \ctraps applications.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=\columnwidth]{figs/systemoverview.pdf}
-\Caption{\label{fig:overview}{\bf System Overview.} Last writer slicing tracks provenance, which augments the core dump and enables provenance debugging.
-CTraps uses provenance to convey communication to CTraps applications, which implement concurrency analyses, like communication graphs~\cite{recon} or CBI~\cite{cci}.}
-\end{figure}
 
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=\columnwidth]{figs/systemoverview.pdf}
+\Caption{\label{fig:overview}{\bf System Overview.} Last writer slicing tracks provenance, which augments the core dump and enables provenance debugging.
+CTraps uses provenance to convey communication to CTraps applications, which implement concurrency analyses, like communication graphs~\cite{recon} or CBI~\cite{cci}.}
+\end{figure}
+
 %In this work, we begin by showing that the cost of monitoring communication has
@@
 
-In this section we discuss debugging with data dependence information, using an
-example, in order to motivate collecting last writer slices in deployed
-systems.  We then provide context for \ctraps by discussing more sophisticated
-analyses that use provenance.
+This section motivates the use of \lws in deployed systems by showing how
+provenance is useful for debugging an example program.  We then provide context
+for \ctraps by discussing other concurrency analyses that use provenance.
 
@@
 understand some bug root causes. Like these prior efforts, we focus on using
-provenance analysis for {\em debugging} -- understanding the root cause of a
-failure -- as opposed to {\em bug detection} -- finding new, unknown bugs.
+provenance analysis for {\em debugging} the root cause of a
+failure, rather than detecting unknown bugs.
 
@@
 \includegraphics[width=\columnwidth]{figs/JDKStringBufferFail.pdf}
-\Caption{\label{fig:coreDumpFail}{\bf Atomicity violation bug in a string buffer implementation.} The core dump does not reveal the connection between {\tt append()} and {\tt erase()}, which is crucial to understand the bug.}
+\Caption{\label{fig:coreDumpFail}{\bf Atomicity violation in string buffer code.} The core dump does not reveal the connection between {\tt append()} and {\tt erase()}, which is crucial to understand the bug.}
 \end{figure}
 
-Figure~\ref{fig:coreDumpFail} illustrates the role of provenance information in
-debugging, using buggy code from the JDK-1.4 stringbuffer library, adapted to
-C++ by other researchers and used variously in prior
+Figure~\ref{fig:coreDumpFail} shows how provenance helps debug code from the
+JDK-1.4 stringbuffer library, adapted to C++ in prior
 work~\cite{concurrencybugs}.  The stringbuffer object pointed to by {\tt sb} in
@@
 
-When the failure manifests, Thread 1 is at the assertion in {\tt append()}, as
-pictured.  However, Thread 2 may have completed its erase operation and moved on
-to execute some unrelated code.  The call stack for Thread 2 may show nothing
-about its involvement in the failure, despite the fact that its interleaving
-with the {\tt append()} code is the root cause.  Examining the core dump, the
-programmer can see that the length and buffer are inconsistent, but is left to
-ask {\em why} they are inconsistent.
+When the failure occurs, Thread 1 is at the assertion in {\tt append()}.
+However, Thread 2 may have completed its erase and moved on to some unrelated
+code.  Thread 2's call stack may lack evidence of its involvement in the
+failure, despite the fact that its interleaving with the {\tt append()} code is
+the root cause.  The core dump shows that length and buffer are inconsistent,
+but does not show {\em why} they are inconsistent.
 
@@
 package it with core dumps sent to programmers with bug reports.  In this work,
-we propose a last writer slicing design that espouses this {\em in situ} data
+we propose a \lws design that espouses this {\em in situ} data
 collection strategy and does so with overheads appropriate for production
@@
 
-Many useful concurrency analyses rely on monitoring inter-thread
+Many concurrency analyses monitor inter-thread
 communication.  The domains of these analyses include 
@@
 avoidance~\cite{aviso,cfix} and concurrent performance
-profiling~\cite{threadcriticality,schedpredictionmodel}.  The breadth of
-applications illustrates the generality of communication tracking.
+profiling~\cite{threadcriticality,schedpredictionmodel}.
 
@@
 \label{sec:lastwriterslices}
-A last writer slice is a dynamic property of a memory location at a point in a
-program's execution.   A memory location's last writer slice is a record
-containing the program point and the identifier of the thread that last wrote
-data to that memory location.
-
-We gather last writer slices at runtime by maintaining a {\em last writer
-table} (\lwt) that maps from a memory location's address to its last writer
-slice.  Just before a thread writes to a memory location, the thread first
-finds the memory location's entry in the \lwt and updates that entry with its
-thread identifier and its current point in the program.  After updating the
-\lwt, the thread performs the original memory access.
+A last writer slice is a dynamic property of a memory location at a point in an
+execution.   A memory location's last writer slice records the program point
+and thread that last wrote data to that memory location.  We gather last writer
+slices at runtime by maintaining a {\em last writer table} (\lwt) that maps
+from a memory location's address to its last writer slice.  Just before a
+thread writes to a memory location, the thread first finds the memory
+location's entry in the \lwt and updates that entry with its thread identifier
+and its current point in the program.  After updating the \lwt, the thread
+performs the original memory access.
 
@@
 they access.  Second, there is no action necessary for read operations.  Making
-read operations cheap is key to making the overhead of last writer slicing low
+read operations cheap is key to making the overhead of \lws low
 enough for production use.  Third, if the program stops at a point like the
@@
 provenance of the data stored in that memory location.   Last writer slicing
-performs no action on read operations and to be useful, it does not need to.
-Instead, a programmer can examine an \lwt entry at any arbitrary point in an
-execution and see a value's
-provenance ({\em e.g.}, at a breakpoint or after a failure).  Including the \lwt with core dumps means provenance is available
-when post-mortem debugging production failures.
+performs {\em no action on a read operation}.  Instead, a programmer can
+examine an \lwt entry at any arbitrary point in an execution and see a value's
+provenance ({\em e.g.}, at a breakpoint or after a failure).  Including the
+\lwt with core dumps means provenance is available when post-mortem debugging
+production failures.
 
@@
 instruments write operations in the program with calls to our runtime that
-update the \lwt.  The \lwt is maintained in the runtime and saved for use
-during debugging, like the core dump.
+update the \lwt.  The \lwt is maintained in the runtime and saved with the core dump.
 
@@
 \includegraphics[width=.9\columnwidth]{figs/LWS_Overview.pdf}
-\Caption{\label{fig:lwsoverview}{\bf Overview of system support for last writer slicing.} Our compiler and runtime library ensure the \lwt remains updated so it can be used during debugging.}
+\Caption{\label{fig:lwsoverview}{\bf Overview of system support for \lws.} Our compiler and runtime library ensure the \lwt remains updated so it can be used during debugging.}
 \end{figure}
@@
 function in our runtime library just before each write that may access data
-last written by another thread (escaping memory accesses).  \lwt update 
-calls are passed three pieces of information: the address of
-the memory word involved in the access, the thread identifier of the accessing
-thread, and the accessing program point.   We use a {\tt GCC}'s local function
-escape analysis to identify accesses to memory locations that are provably
-inaccessible to code running in other threads.  Such operations access unshared
-data and we do not instrument them.
+last written by another thread .  \lwt update calls are passed three pieces of
+information: the address of the memory word involved in the access, the thread
+identifier of the accessing thread, and the accessing program point.   We use
+escape analysis to identify accesses to unshared data and we do not instrument
+them.
 
@@
 system support.  Figure~\ref{fig:systemdiagram} shows an overview of our system
-support.  The programmer writes their program as usual.  They compile it using
+support.  The programmer writes their program as usual and compiles it using
 the \ctraps compiler. The \ctraps compiler includes the compiler support for
-collecting last writer slices.  In addition, the \ctraps compiler also inserts
+collecting last writer slices.  The \ctraps compiler also inserts
 calls to the \ctraps runtime at points where communication may occur.  The
@@
 
-\ctraps delivers software traps before all read and write operations executed
-by one thread that access a memory location that was last written by another
-thread, {\em i.e.}, communicating memory operations.  Traps are simply inserted
-function calls (as opposed to, {\em e.g.}, signals) made before such operations
+\ctraps delivers software traps before each read and write operation executed
+by one thread that accesses memory that was last written by another
+thread, {\em i.e.}, communicating memory operations.  Traps are 
+simply function calls made before such operations
 execution.  \ctraps implements traps using a combination of compiler and
@@
 
-\paragraph{Compiler Support}
-Our compiler inserts a call to the CTraps {\em trap hook} function immediately
-before all read and write operations that access potentially escaping memory
-locations.  Trap hooks are passed the address of the memory location being
-accessed, the thread identifier of the accessing thread, the type of operation
-being performed, and the program point of the access.
+\paragraph{Compiler Support} Our compiler inserts a call to the CTraps {\em
+trap hook} function immediately before all read and write operations to
+potentially shared variables.  Trap hooks are passed the variable's address,
+the accessing thread's ID, the type of access, and the program point of the
+access.
 
@@
 %To discuss the correctness of \ctraps, we first formalize trap delivery by
-%building on our last writer slicing formalism from
+%building on our \lws formalism from
 %Section~\ref{sec:lastwriterslices}.  
@@
 demonstrate that \ctraps can implement real analyses with little work, we
-implemented two applications. The first application is a version of a debugging
+implemented two applications. The first application is a variant of an 
 analysis from CCI~\cite{cci}.  The second application is a communication graph
-collector, similar to DefUse~\cite{defuse}, Recon~\cite{recon}, and
+collector, like DefUse~\cite{defuse}, Recon~\cite{recon}, and
 DMTracker~\cite{dmtracker}.
@@
 
-\paragraph{Communication Graph Implementation}
-Communication graphs are the basis of several prior debugging
-techniques~\cite{recon, bugaboo, defuse}.  We implement communication graph
-collection using \ctraps.  When a trap is delivered, our implementation records
-communication graph edges composed of the code point of the last writer to the
-location being accessed and the code point of the trapping access.  The set of
-recorded edges is the tool's output.  Our \ctraps implementation of
-communication graphs took about 50 lines of code.
+\paragraph{Communication Graph Implementation} Communication graphs are the
+basis of several prior debugging techniques~\cite{recon, bugaboo, defuse}.  We
+implement communication graph collection using \ctraps.  When a trap is
+delivered, our implementation records communication graph edges composed of the
+code point of the last writer to the location being accessed and the code point
+of the trapping access.  The set of recorded edges is the tool's output.  Our
+implementation took about 50 lines of code.
 
@@
 \label{sec:implementation}
-We implemented last writer slicing and \ctraps.  We built our compiler support
+We implemented \lws and \ctraps.  We built our compiler support
 as a plugin for GCC ({\tt gcc-4.7}).  We used GCC's points-to and
@@
 
-\subsection{Release}
+\paragraph{Release}
 We released our implementation, free and open-sourced, on the
@@
 to debug several real-world programs studied previously in the debugging
-literature.  Table~\ref{tab:bugs} provides an overview of this evaluation and
-Section~\ref{sec:eval:debugging:cases} looks in detail at several cases to 
-better illustrate how last writer slices helps with debugging.
-
-In Table~\ref{tab:bugs}, we show the bugs we studied, describing the program,
-bug type, and bug report identifier where applicable.  For each bug, we
-followed the procedure for installation and bug triggering described by the
-author of~\cite{concurrencybugs}.  We study a mixture of concurrency bug types
-in this work, including both single- and multi-variable bugs and both atomicity
-and ordering violations.
-
-The table also summarizes our evaluation of last writer slicing's debugging
+literature.  Table~\ref{tab:bugs} shows the bugs we studied, describing the
+program, bug type, and report information.  We followed the procedure described
+by the author of~\cite{concurrencybugs} for installing and triggering each bug
+.  We study various bug types, including both single- and
+multi-variable bugs and both atomicity and ordering violations.
+
+The table also summarizes our evaluation of \lws's debugging
 benefits.  We mirror the evaluation strategy of bad value origin
-tracking~\cite{badapples}, as it is the only other low-overhead provenance
-tracking mechanism we know of.  We evaluate our system using their three
+tracking~\cite{badapples} and evaluate our system using their three
 evaluation criteria - (1) {\bf effectiveness}, (2) {\bf triviality}, and (3)
@@
 the last writer slice revealed the provenance of at least some data involved in
-the failure, so last writer slicing was effective.  We found that none of the
+the failure, so \lws was effective.  We found that none of the
 failures we looked at were trivially debuggable.  This finding is consistent
@@
 The case studies in Section~\ref{sec:eval:debugging:cases} show in more detail
-how last writer slices are useful for debugging and illustrate their limitations.
+how \lws is useful for debugging and illustrates its limitations.
 
@@
 Our evaluation shows that bad value origin tracking~\cite{badapples} is
-helpful in one case where last writer slices is useful and provides limited help
+helpful in one case where \lws is useful and provides limited help
 in one other case.  For a use-after-delete bug in {\tt pbzip2}, bad value
@@
 threads interleave their code, multiple threads might see zero reference
-counts, which is incorrect.  When that happens, both threads try to free the
+counts, which is incorrect.  When that happens, both threads free the
 object and the second free causes a crash.
@@
 \Caption{\label{fig:httlws}{\bf Using last writer slices to debug an
-ordering violation in HTTrack.} The provenance information stored in the last
-writer slice for the mutex variable helps understand the bug's root cause.}
+ordering violation in HTTrack.} The provenance information in the mutex's last
+writer slice helps understand the bug's root cause.}
 \end{figure}
 
-Looking at last writer slices make this root cause clear: in the failing
+Looking at last writer slices make the root cause clear: in the failing
 execution, the last writer slice indicates the mutex is uninitialized; in the
@@
 We conducted our evaluation on a machine running Linux 2.6.27-7, with a 2.27GHz
-8-core Xeon E5520 processor with 2-way SMT ({\em i.e.}, 16 execution contexts)
-and 10GB of memory.  We evaluated our system using PARSEC, a set of
-architectural benchmark programs~\cite{parsec} and a collection of real server
-and desktop applications.
+8-core Xeon E5520 processor with 2-way SMT and 10GB of memory.  We evaluated
+our system using the PARSEC benchmarks~\cite{parsec} and several 
+real-world applications.
 
@@
 
-\paragraph{MySQL-5.1.65} MySQL is an industrial-strength database
-server. To benchmark MySQL, we used {\tt SysBench OLTP} running with its
-default 8 threaded configuration, measuring MySQL's performance to be the
-throughput reported by {\tt SysBench}.
+\paragraph{MySQL-5.1.65} MySQL is an industrial-strength database server. To
+benchmark MySQL, we used {\tt SysBench OLTP} running 8 threads, measuring
+MySQL's performance as the throughput reported by {\tt SysBench}.
 
@@
 running its ``Read while Writing'' test with 8 threads.  We measure LevelDB's
-performance to be the operation throughput reported by {\tt db\_bench}.
+performance to be the throughput reported by {\tt db\_bench}.
 
@@
 The main performance result of our work is that our designs impose performance
-overheads that are low enough for use in deployed systems for many of our
-experimental workloads.  Figure~\ref{fig:perfall} shows the slowdown suffered
-by our benchmarks due to last writer slicing and \ctraps running with ``no-op''
+overheads that are low enough for use in production for many of our
+workloads.  Figure~\ref{fig:perfall} shows the slowdown suffered
+by our benchmarks due to \lws and \ctraps running with ``no-op''
 trap hooks ({\em i.e.}, no handlers), relative to the natively executing
@@
 \paragraph{Last Writer Slices}
-The data show that last writer slicing has very low overheads with a geometric
+The data show that \lws has very low overheads with a geometric
 mean of less than 10\% across our server programs and less than 50\% across
@@
 
-By comparing \ctraps and last writer slicing, we see four applications ({\tt
-dedup}, {\tt streamcluster}, {\tt fluidanimate}, {\tt ferret}) that have
-\ctraps overhead that is probably too high for production use (averaging
-153\%).  By comparison, those applications have a last writer slicing overhead
-that is likely acceptable for production use (averaging 38\%).  The difference
-indicates that these programs perform relatively more reads than other
-applications, experiencing more overhead due to \ctraps's read instrumentation.
+By comparing \ctraps and \lws, we see four applications ({\tt dedup}, {\tt
+streamcluster}, {\tt fluidanimate}, {\tt ferret}) that have \ctraps overhead
+that is probably too high for production use (averaging 153\%).  By comparison,
+those applications have a \lws overhead that is likely acceptable for
+production use (averaging 38\%).  These programs perform relatively more reads
+than other applications, thus experiencing more overhead due to \ctraps's read
+instrumentation.
 
@@
 are low enough for deployed systems for many applications, especially
-for servers.  We now further analyze our observed overheads, illuminating
-differences between the lower-overhead server programs and the higher-overhead
-PARSEC programs.
+for servers.
 
@@
 \includegraphics[width=.9\columnwidth]{plots/perf.pdf}
-\Caption{\label{fig:perfall}Runtime overhead of \ctraps and Last Writer Slices (LWS).}
+\Caption{\label{fig:perfall}Runtime overhead of \ctraps and \lws.}
 \end{figure}
@@
 instrumentation.  The program's overhead dropped from about 420\% overhead for
-\ctraps to about 300\% overhead; the change for last writer slicing was from
+\ctraps to about 300\% overhead; the change for \lws was from
 300\% to about 130\% overhead.  We also looked at an output matrix passed into
@@
 accesses to this matrix yielded an overhead of 46\% for \ctraps and 
-just 30\% for last writer slicing.
+just 30\% for \lws.
 
@@
 \label{sec:eval:parsecserver}
-We discern three reasons that server programs better tolerated the overheads of
-last writer slices and \ctraps in our experiments.
+We discern several reasons that server programs better tolerated the overheads of
+\lws and \ctraps in our experiments.
 
@@
 In most cases, the servers we evaluated have abundant, completely independent
-parallel work.  Server programs use a pool of threads to handle requests.
-Each thread incurs some instrumentation latency, but that latency is hidden by
-other threads making progress on independent requests.  This analysis also
-applies to the PARSEC applications that had performance similar to the servers.
-For instance, {\tt blackscholes} uses largely independent threads to compute on
+parallel work.  Server programs use a pool of threads to handle requests.  Each
+thread incurs some instrumentation latency, but that latency is hidden by other
+threads making progress on independent requests.  This analysis also applies to
+the PARSEC applications that had performance similar to the servers.  For
+instance, {\tt blackscholes} uses largely independent threads to compute on
 different regions of a matrix.  In addition, threads performing independent
@@
 mostly on non-escaping, local data.  Local data are not instrumented by our
-compiler, sparing the overhead.  We characterize the amount of local data in
-each application in detail in Section~\ref{sec:char}.  We show that the server
-programs have a larger proportion of provably local accesses than PARSEC.  The
-difference supports the fact that the servers have more independent work ({\em
-i.e.}, more local accesses) than PARSEC.
-
-\paragraph{Latency Hiding I/O}
-The servers perform more I/O.  Similarly to the effect of abundant parallelism,
-I/O helps hide instrumentation latency.  While I/O is naturally part of many
-applications, we aimed to minimize the effect of I/O in our experiments.  For
-MySQL and Apache, we used local disks for logs and resources, and connected via
-a local socket, rather than using the network.  In spite of these controls, I/O
-remains, and hides added instrumentation latency in our server applications.
-LevelDB's benchmark is designed to avoid all I/O, creating and manipulating an
-in-memory database in a single process.  The absence of I/O may account for some
-of the difference in overhead between LevelDB and the other servers.  Note that
-the effects of I/O are not merely experimental noise -- in real deployed
-software the latency hiding effects of I/O are likely to be even {\em more}
-pronounced, favoring the use of our system.
+compiler, sparing the overhead.  We analyzed the amount of unshared data in
+each application and found that for our server workloads, 21--66\% of accesses
+were to provably local data.  Far fewer accesses are provably unshared for
+PARSEC: just 8--30\%. The difference supports the fact that the servers have
+more independent work ({\em i.e.}, more local accesses) than PARSEC.
 
 \paragraph{Interference with Optimization}
-The PARSEC programs are heavily hand-optimized. In several cases
-resorting to hand-coded assembly and cache-aware memory access patterns.
-Running instrumentation that accesses the \lwt alongside these computations may
-disrupt their cache performance, or other delicately optimized behavior,
-leading to higher overheads.
-
-Such carefully optimized code is likely to have been written by expert
-programmers.  As with other code written by experts ({\em e.g.}, lock-free
-data-structures), this code may not require as much {\em in situ} analysis
-support as code written by distributed teams of open source developers or
-novice programmers.  Depending on the maturity and level of verification of a
-piece of carefully hand-optimized code, it might make sense to elide 
-last writer slicing and \ctraps code to preserve performance.
+The PARSEC programs are heavily hand-optimized. Several use hand-coded assembly
+and cache-aware algorithms.  The instrumentation that accesses the \lwt may
+disrupt such delicately optimized behavior, leading to higher overheads.  Such
+carefully optimized code is often written by an expert.  As with other
+expert-written code ({\em e.g.}, lock-free data-structures), this code may be
+machine verified or simply not require as much debugging support as code written by an
+average programmer.  Depending on the maturity and level of verification of
+code, a programmer can disable \lws and \ctraps to preserve performance.
 
 \subsection{Context-Sensitivity}
-%{\bf TODO: Was this LWS or CTRAPS-NOOP or what?}
 We briefly evaluated the cost of context-sensitivity using \ctraps with no-op
-trap handlers.  With a bounded 2-address context, the average overhead for
-server programs was 18\% and the average overhead for PARSEC was 145\%.  These
-data show that a restricted form of context-sensitive analysis is possible in
-\ctraps with overheads low enough for production in some cases.
+trap handlers.  With 2-address context, the average overhead for server
+programs was 18\% and for PARSEC was 145\%.  These data show that
+context-sensitive analysis is possible in \ctraps with practically low overheads.
 
@@
 always overwrite existing entries.  This strategy fixes memory overheads at
-2GB, but frequent lossy collisions may lead to imprecision.   We instrumented
-our runtime to count collisions and found that for PARSEC applications the
+2GB, but frequent collisions may lead to imprecision.   We instrumented our
+runtime to count collisions and found that for PARSEC applications the
 geometric mean rate of collisions was very low - 89 per 10,000 memory accesses.
 This summary result shows that our prototype implementation is reasonable,
-especially for debugging in production, when low memory use is key and minor
-imprecision is tolerable.  Note that collisions did not introduce noticable
-imprecision in our debugging experiments.  When precision is more important than memory overhead, ({\em
+especially for debugging in production, when low memory use is key.  Note that
+collisions did not introduce noticable imprecision in our debugging
+experiments.  When precision is more important than memory overhead, ({\em
 e.g.}, for sound, offline analysis), a lossless hashtable is a better option.
@@
 \label{sec:appperf}
-We evaluated our CCI-Prev and communication graph collection CTraps
-applications.  Our results show that these CTraps applications have modest
-overheads that scale with their analysis complexity.  In several cases -- most
-notably, in tests with our server benchmarks -- their overhead is low enough
-for production use.
-
-\subsubsection{Performance with \ctraps Applications}
+We evaluated our CTraps applications and our results show that 
+they have modest overheads that scale with their analysis complexity.
 
 Figure~\ref{fig:perfapps} shows the performance overheads of 
-our implementation as the slowdown with respect to the natively executing
-baseline.  The overheads vary substantially across applications and the average
-run time overheads are higher than for our baseline implementation
-(Section~\ref{sec:eval:perf}).
-
-In six cases (the four servers, {\tt blackscholes}, and {\tt dedup}), the overheads
-for CCI-Prev implementations are tolerable for deployment use (around 50\%).
-This result supports our claim that \ctraps helps build useful analyses that
-have overheads low enough for deployment for an important class of
-applications.
+\ctraps applications as the slowdown relative to native execution.  The overheads
+vary across applications and the average run time overheads are
+higher than our no-op \ctraps application (Section~\ref{sec:eval:perf}).  In
+six cases (the four servers, {\tt blackscholes}, and {\tt dedup}), the overhead
+for CCI-Prev is tolerable for deployment use (around 50\%).  This result
+supports our claim that \ctraps helps build useful analyses that have overheads
+low enough for deployment for an important class of applications.
 
@@
 negative result; instead, they illustrate the ``complexity proportional''
-nature of \ctraps' overheads.  Our communication graph collection
-implementation does more costly data structure manipulations on each trap, so
+nature of \ctraps' overheads.  Our communication graph collector
+does more costly data structure manipulations on each trap, so
 its overheads are higher.  \ctraps provides a low-overhead foundation and
@@
 to this work~\cite{badapples} and we directly compare to it in our evaluation.
-At a high level, bad value origin tracking is like our work because they track
-provenance information for debugging, which is a similar goal to ours.
+At a high level, bad value origin tracking is like our work because they also track
+a form of provenance for debugging.
 
@@
 the a recent survey~\cite{tipslicingsurvey} monitor control and data
-dependences to identify which parts of a program might be relevant to a task
-({\em e.g., debugging}).
+dependences to identify which parts of a program might be relevant to a task.
 
@@
 
-\paragraph{Communication and Sharing Analyses in Concurrent Programs}
-Several techniques have been proposed analyze dependences and
-communication in concurrent programs.
-
-A closely related recent effort on dynamic dependence analysis is
-Octet~\cite{octet}, which captures inter-thread communication soundly and to
-aid in building arbitrary concurrency analyses. Octet tracks the ``locality
-state'' of memory locations to make dependence tracking efficient.   Octet
-differs from \ctraps in that it focuses on soundness, was built primarily to
-optimize dependence tracking, and does not discuss direct use of
-dependence information to help with debugging.  Octet's average run time
-overhead (26\%) is comparable to the overhead of both last writer slicing and
-\ctraps and Octet, last writer slicing, and \ctraps all have low enough
-overheads for production use in many cases.
+\paragraph{Communication and Sharing Analyses}
+Several techniques have been proposed analyze dependences and communication in
+concurrent programs.  A closely related recent effort on dynamic dependence
+analysis is Octet~\cite{octet}, which captures inter-thread communication
+soundly and to aid in building arbitrary concurrency analyses. Octet tracks the
+``locality state'' of memory locations to make dependence tracking efficient.
+Octet differs from \ctraps in that it focuses on soundness, was built primarily
+to optimize dependence tracking, and does not discuss direct use of dependence
+information to help with debugging.  Octet's average run time overhead (26\%)
+is comparable to the overhead of both \lws and \ctraps and Octet, \lws, and
+\ctraps all have low enough overheads for production use in many cases.
 
@@
 \section{Conclusions}
-In this work we proposed last writer slicing, the first technique to collect data provenance information
-that is not limited to certain values, is targeted to concurrent programs, and
-has overheads low enough for production use, especially in server programs.
-
-We developed system support that collects last writer slices with overheads
-well within an acceptable range for always-on use in production.  We showed
-that during debugging, the provenance information provided by last writer
-slices reveals crucial connections between different threads' otherwise
-disparate regions of code.  Understanding how code in different threads
-interacts is essential to concurrency debugging and this work aids that
-understanding.
-
+In this work we proposed \lws, the first technique to collect data provenance
+information that is not limited to certain values, is targeted to concurrent
+programs, and has overheads low enough for production use.  We showed that
+during debugging, the provenance information provided by last writer slices
+reveals crucial connections between different threads' otherwise disparate
+regions of code.  Understanding how code in different threads interacts is
+essential to concurrency debugging and this work aids that understanding.
 Using last writer slices, we then built \ctraps, an extensible framework for
@@
 communication.  Trap handlers implement such analyses in just a few lines of
-code.  We showed that it is possible to build useful analyses using \ctraps
-that are efficient enough for deployment use in an important set of programs
-({\em e.g.}, servers, key-value stores, databases, {\em etc.}).
-
-An interesting direction for future work is to evaluate the role of minimal
-hardware support for \ctraps.  We will also further study the impact of
-provenance information on debugging in both sequential and concurrent
-applications.
+code.
 

### 2014-09-03 paper.tex
added octet discussion.  back on 11

@@
-\newcount\draft\draft=1 % set to 0 for submission or publication
+\newcount\draft\draft=0 % set to 0 for submission or publication
 \newcount\cameraready\cameraready=0
@@
 \paragraph{Communication and Sharing Analyses}
-Several techniques have been proposed analyze dependences and communication in
-concurrent programs.  A closely related recent effort on dynamic dependence
-analysis is Octet~\cite{octet}, which captures inter-thread communication
-soundly and to aid in building arbitrary concurrency analyses. Octet tracks the
-``locality state'' of memory locations to make dependence tracking efficient.
-Octet differs from \ctraps in that it focuses on soundness, was built primarily
-to optimize dependence tracking, and does not discuss direct use of dependence
-information to help with debugging.  Octet's average run time overhead (26\%)
-is comparable to the overhead of both \lws and \ctraps and Octet, \lws, and
-\ctraps all have low enough overheads for production use in many cases.
+Several techniques have been proposed analyze communication in concurrent
+programs and are especially related to \ctraps.
+
+A closely related recent effort on dynamic dependence analysis is
+Octet~\cite{octet}, which exposes inter-thread communication soundly to
+concurrency analyses.   Octet reports an average run time overhead of 26\%,
+which is comparable to the overhead of \lws and \ctraps.
+
+To acheive these low performance overheads, Octet makes fundamental assumptions
+about its target system that \lws and \ctraps do not make.  To track
+communication, a thread in Octet must stop at ``safe points'' in an execution.
+At a safe point, the thread executes extra code to check a message queue that
+reports communication with other threads.  Octet assumes safe points at loop
+back edges, method entries, and blocking function calls.  This assumption is
+reasonable for Java, as many Java VMs use safe points to implement GC.  Octet
+uses existing safe points and their performance baseline assumes a VM with
+existing instrumentation at such safe points.  In contrast, C/C++ programs do
+not have a VM or existing safe points.   Checking a queue on every function
+call and method entry is likely to impose a higher performance penalty than
+Octet saw in Java.  \lws keeps overheads practical by relying on program
+synchronization to keep dependence information consistent --- unlike Octet
+which is always sound, Data-races may leave \lws data inconsistent.  However,
+\lws need not pay the cost of safe points and message queue checks.
+
+Another important difference between Octet and this work is that this work
+illustrates the benefit of directly using \lws's provenance information for
+debugging.   Octet does not explore provenance debugging.
 

### 2014-11-15 paper.tex
add acks section

@@
 
+\section*{Acknowledgements}
+
+We would like to thank our anonymous reviewers for their invaluable comments.
+Our thanks also to Kathryn McKinley and the members of the Sampa group at the University of Washington.
+This work was supported in part by C-FAR, one of six centers of STARnet, a
+Semiconductor Research Corporation program sponsored by MARCO and DARPA.
+It was also supported by a gift from Google.
+
 \bibliography{bib}{}

### 2015-02-04 paper.tex
pre-merge commit
latest latest

@@
 \newcount\draft\draft=0 % set to 0 for submission or publication
-\newcount\cameraready\cameraready=0
-
-\documentclass[pageno,nohyperref]{jpaper}
+\newcount\cameraready\cameraready=1
+
+\documentclass[10pt,nocopyrightspace]{sigplanconf}
 
@@
 \usepackage{xxx}
-
+\usepackage{url}
+\usepackage{flushend}
 \usepackage{graphicx}
@@
 \usepackage[normalem]{ulem}
+\usepackage{fancyhdr}
+
+\let\OLDthebibliography\thebibliography
+\renewcommand\thebibliography[1]{
+  \OLDthebibliography{#1}
+  \setlength{\parskip}{0pt}
+  \setlength{\itemsep}{0pt plus 0.3ex}
+}
 
@@
 \newcommand{\lws}{LWS\xspace}
+\newcommand{\lwsfull}{Last Writer Slicing\xspace}
 \newcommand{\lwt}{LWT\xspace}
+\newcommand{\lwtfull}{Last Writer Table\xspace}
 
@@
 %\doublespacing
-%\authorinfo{Anonymous for Submission}{}
-
-%\authorinfo{Brandon Lucia \and Luis Ceze}{University of Washington, Department of Computer Science and Engineering}{\{blucia0a,luisceze\}@cs.washington.edu\\http://sampa.cs.washington.edu\\{\bf In Submission --- Please Do Not Distribute}}
+\authorinfo{Brandon Lucia}{Carnegie Mellon University, Department of Electrical and Computer Engineering}{blucia@ece.cmu.edu}
+\authorinfo{Luis Ceze}{University of Washington, Department of Computer Science and Engineering}{luisceze@cs.washington.edu}
+
+%\author{Brandon Lucia\thanks{Work done in part as a PhD student at University of Washington.}\\Carnegie Mellon University\\{\em blucia@cmu.edu} \and Luis Ceze\\University of Washington\\{\em luisceze@cs.washington.edu}}
+
+%{blucia@cmu.edu luisceze@cs.washington.edu\\{\bf In Submission --- Please Do Not Distribute}}
 
@@
 
+ \thispagestyle{fancy}
+  %For papers in which all authors are employed by the US government, un-comment the following
+  %\def\ieeenotice{U.S. Government work not protected by U.S. copyright}
+  %For papers in which all authors are employed by a Crown government (UK, Canada, and Australia), uncomment the following
+  %\def\ieeenotice{978-1-4799-8161-8/15/\$31.00 {\copyright}2015 Crown}
+  %For all other papers, uncomment the following:
+  \def\ieeenotice{978-1-4799-8161-8/15/\$31.00 {\copyright}2015 IEEE}
+  \def\confinfo{2015 IEEE/ACM International Symposium on Code Generation and Optimization}
+  \lfoot{\confinfo\\\ieeenotice} \cfoot{}
+  \renewcommand{\headrulewidth}{0pt}
+
 \special{papersize=8.5in,11in}
@@
 
-\thispagestyle{empty}
 \begin{abstract}
@@
 %dependences typically has overheads too high for production.  
-We propose {\em \lws} (\lws), a mechanism for tracking {\em data
+We propose {\em \lwsfull} (\lws), a mechanism for tracking {\em data
 provenance information} in multithreaded code in a production setting.  Last
@@
 contains its value.  In this work we develop a new provenance tracking
-mechanism called {\em \lws} (LWS).  A memory location's last writer slice is a
+mechanism called {\em \lwsfull} (\lws).  A memory location's last writer slice is a
 record containing the program point and the identifier of the thread that last
@@
 applications scales with the complexity of their analysis.
-Figure~\ref{fig:overview} illustrates the relationship between last writer
+Figure~\ref{fig:overview} shows the relationship between last writer
 slices, \ctraps, and \ctraps applications.
@@
 To summarize, this work makes several contributions:
-\begin{itemize}
-
-\item{{\bf Last writer slicing} ({\em LWS}): a provenance tracking mechanism
+
+\noindent {\bf Last writer slicing} ({\em LWS}): a provenance tracking mechanism
 useful for concurrency debugging.  We present an \lws implementation
-efficient enough for production use.}
-
-\item{{\bf Communication traps} (\ctraps): an extensible
-framework for implementing concurrency analyses, built on \lws.  }
-
-\item{{\bf Efficacy evaluation}: We show that \lws is useful for debugging with
+efficient enough for production use.
+
+\vspace{.5ex}
+\noindent {\bf Communication traps} (\ctraps): an extensible
+framework for implementing concurrency analyses, built on \lws.
+
+\vspace{.5ex}
+\noindent {\bf Efficacy evaluation}: We show that \lws is useful for debugging with
 a study of several real concurrency bugs.  We show \ctraps is useful for
 building concurrency analyses by implementing two analyses from prior
-work~\cite{cci,defuse,recon}. }
-
-\item{{\bf Performance evaluation}: We show that in many cases our designs have
+work~\cite{cci,defuse,recon}.
+
+\vspace{.5ex}
+\noindent {\bf Performance evaluation}: We show that in many cases our designs have
 overheads low enough for use in deployment (0--15\%) with a study of a diverse
 set of real programs ({\em e.g.}, MySQL, memcached and the PARSEC
-benchmarks~\cite{parsec}).  }
-
-\end{itemize}
-
-\begin{figure}[h]
+benchmarks~\cite{parsec}).
+
+\begin{figure}[t]
 \centering
